### 2017-10-04 paper.Rnw
fig models paper

\documentclass[12pt]{article}

%\usepackage{endfloat}
%\usepackage{soul}

%\usepackage{type1ec}
\usepackage{subcaption}
\usepackage{fullpage}
\usepackage{enumerate}
%\usepackage{graphicx}
%\usepackage{graphics}
\usepackage{multirow}
\usepackage{comment}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{bbm}
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{natbib}
\usepackage{bm}
\usepackage{pdflscape}
%\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{xr}
%\usepackage{mathabx}
%\usepackage{filecontents}
%\usepackage{bibentry}
%\usepackage{hanging}
%\usepackage{apacite}
%\usepackage{hyperref}
\usepackage{makecell}

%\usepackage{endnotes}

%\let\footnote=\endnote

\title{Mastery Learning in Practice:\\
A Descriptive Analysis of Data from the  Cognitive Tutor Algebra I Effectiveness Trial}

\author{Anita Israni \and Adam C Sales \and John F Pane}

<<include=FALSE>>=
library(knitr)
library(tikzDevice)
opts_chunk$set(
echo=FALSE, results='asis',cache=TRUE,warning=FALSE,error=FALSE,dev='tikz',message=FALSE,autodep = FALSE
    )

options(tikzDefaultEngine = "pdftex")

library(lubridate)
library(scales)
library(reshape2)
library(dplyr)
library(ggplot2)

@

<<data,include=FALSE>>=
source('dataMerge.r')
@

\begin{document}
\maketitle

\section{Introduction}
Mastery learning sits at the foundation of intelligent tutoring systems. The philosophy of mastery learning assumes a well-structured curriculum, and posits that students progress within the curriculum as they master its skills. The Cognitive Tutor Algebra I (CTAI) system, developed by Carnegie Learning, Inc., is one of the best studied and best regarded examples of modern educational software. It is a blended learning system for teaching algebraic concepts and principles, to middle and high school students, including both textbook materials and software. The software component of the curriculum allows students to progress at their own pace and receive individualized feedback on their performance. A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students’ scores on an Algebra-I post-test by about one fifth of a standard deviation (Pane, et al.  2013). CTAI’s success in this experiment would seem to vindicate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.

However, the theory underlying CTAI does not always determine its use. To be sure, the software has a standard set of algebra topics, divided into units and further into sections; and a standard sequence for presenting them. But this precise curriculum is not mandatory. At the request of educators, it can be customized by altering what units or sections are included (including, possibly, material from a different standard curriculum such as geometry), as well as their sequence, to conform to local or state standards. Further, teachers have the option of moving students within the curriculum, regardless of the software’s estimate of their skill mastery.

This article examines teachers’ and schools’ non-adherence to the standard, mastery-based CTAI curriculum, using data from the RAND study, a seven-state randomized controlled trial of CTAI in high schools and middle schools. The study found a significant positive effect of CTAI in high schools, during their second year of implementation but not the first. Students in the treatment group of the study were enrolled in one or more of the standard or customized curriculum during their participation in the study, and the software logged aspects of their usage, including time spent, sections encountered, and whether the software judged the students to have mastered the sections. Both Carnegie Learning and the researchers running the study restricted their software support and oversight to what is typically provided outside of an experimental context. Thus, the data from the study reflect typical software usage. Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects (Sales & Pane, 2015, Sales et al. 2016).

Here we extend the previous findings to investigate patterns of change in the sequence of topics presented to students. As noted earlier, schools could use a standard curriculum, a customized curriculum, or switch from one curriculum to another. Each curriculum consisted of a sequence of units, which was comprised of a sequence of sections. For example, the standard CTAI Algebra I curriculum in Year 1 comprised of 40 units covering approximately 170 sections. The
number of sections within each unit varied. By default, the software operates by automatically moving students from section to section based on the sequence of topics defined by the curriculum they were currently enrolled in. In this software-controlled sequencing, students ideally spend the time necessary to learn the material of a section, are judged by the software to have mastered the material, and then “graduate” to the next section. Students who exhaust a section's material without mastering its skills are “promoted” to the next section. Students left uninterrupted follow the sequence of topics defined by their curriculum, ideally graduating from topics by showing mastery. However, teachers could modify a student’s path within the curriculum. They could reassign students from their current sections to other sections earlier or later in the intended sequence, including sections they had worked on previously.
This study attempts to elucidate teachers' goals in reassigning students. One hypothesis is the need for teachers push ahead students who were falling behind, i.e. to reassign them from sections on which they were struggling, to allow them catch up with the rest of the class. Another hypothesis is that teachers sought to push students past easier sections to begin working on more relevant topics for them. A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exams, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered. This hypothesis may lead to an increase in reassignments as the exam approaches. Overall, understanding patterns of reassignment is the focus of this paper, along with insights into how teachers were using the software component of CTAI.

\section{The RAND Effectiveness Trial}
The study to measure the effectiveness of CTAI included 7 states, 73 high schools, and 74 middle schools with nearly 18,700 high school students and 6,800 middle school students. Schools were enrolled in a total of 52 school districts that were distributed between urban, suburban, and rural areas. Schools were matched on a set of covariates, and then randomly assigned to the treatment or control group. Schools in the control group continued with their current algebra curriculum, and schools in the treatment group used Carnegie Learning’s curriculum which includes textbook materials as well as CTAI. Each school participated for two years, with a different cohort of students taking algebra the second year (with a small fraction of students present in the study both years because they repeated algebra). It should be noted that this study did not include statewide implementations; the study results cannot be generalized to all schools within the state. In some states, one large school district participated, while in other states, a set of smaller school districts participated. The states included Alabama (AL), Connecticut (CT), Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and Texas (TX). Each state participated in both the middle school and high school arms of the study, except AL, which participated only in the middle school arm. The current study focuses on high school students only, due to the large number of students in this cohort and the significant positive treatment effect previously found for this cohort. All discussion that follows refers exclusively to the high school data.

There are some limitations to the available data for this study. In particular, the developer was able to provide the intended sequence of units/sections for only a subset of the customized curricula. Data from some schools, and some students within schools, were missing either because the log files were not retrievable, or because of an imperfect ability to link log data to other study data files. In addition, there was missing completion dates for many customized curricula; making it difficult to analyze the customized curricula when any sort of date was needed, such as time to complete a section.

\section{Standard versus Customized Curricula}

All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year. In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools. The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus. In Year 2, Carnegie Learning also developed many more customized curricula with input from participating schools, for a total of 33 curricula used in 13 schools. This led to fewer students enrolled in the standard curriculum and more enrolled in customized curricula overall in Year 2.

In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204. The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2. It should be noted that students could be enrolled in multiple curricula. For instance, teachers could choose to enroll students in any of the standard curricula provided, such as Algebra I or Algebra II, along with any of the customized curricula that were used by the school.

The use of customized curricula varied substantially by state. For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2. All other states contained schools that used customized curricula in addition to the standard curricula. CT schools maintained similar number of students in Algebra I across both years with 111 in Year 1 and 135 in Year 2. NJ substantially decreased the number of students in the CTAI program from 89 to 32 from Year 1 to Year 2. However, while all other states (KY, LA, MI, and TX) did decrease enrollment in the standard Algebra I curriculum the second year, those same states increased enrollment in the customized curricula. KY added 3 customized curricula with 146 students, LA added 15 customized curricula with 286 students, MI added 5 customized curricula with 370 students, and TX added 10 additional customized curricula and had the largest enrollment in these curricula with 1,531 students. Students enrolled in the standard Algebra I curriculum dropped from 974 in Year 1 to 61 in Year 2 in TX. As noted above, students could be enrolled in more than one curriculum. In Year 2, the average number of curricula per a student ranged from 1.00 (NJ and CT) to 2.53 (TX). The next highest value was LA at 2.10.

Throughout the school year, teachers could have a class of students working on multiple curricula either sequentially, where the students changed curricula in lock step, or simultaneously, where students worked on different curricula at the same time. As an example, two teachers located in KY had their students working on Algebra I throughout most of the year and then reassigned them to Algebra II in the last month of school. Another teacher in KY had students variously enrolled in three different curricula throughout the entire year (Bridge-to-Algebra, Algebra I, and a specialized Geometry curriculum), while another teacher in MI during Year 2 had students enrolled in three curricula sequentially throughout the year covering Algebra I until November, and then starting a customized curriculum and using it until February, and ending the year with another customized curriculum that was used until June. While there are numerous instances of these uses of multiple curricula in Year 2, there are also many occurrences of teachers who had their students enrolled in the standard Algebra I throughout the entire year.  In TX mostly, with some instances in MI, there were teachers who exclusively used customized curricula the whole year.

\section{Student Usage Across Sections and Years}
\begin{figure}
<<usageTime, fig.height=3,fig.width=6>>=
secByStud <- summarize(group_by(data,state,Yr,field_id),numSec=n_distinct(section,na.rm=TRUE),time=sum(total_t1,na.rm=TRUE),mastered=n_distinct(section[status=='graduated'],na.rm=TRUE))

secByStud <- within(secByStud,{
                        numSec[numSec==0] <- NA
                        time[time==0] <- NA
                        mastered[mastered==0] <- NA
                    })

print(timeStateYear <- ggplot(secByStud,aes(Yr,time/3600000))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,100))+labs(x='',y='Time on CTAI Software per Student (Hours)'))
#ggsave('timeStatYear.jpg',width=6,height=3)

@
\caption{Boxplots of hours each student spent on Cognitive Tutor
  software over by year and state. Students spending more than 100
  hours (\Sexpr{sum(secByStud[['time']]>100,na.rm=TRUE)} of \Sexpr{nrow(secByStud)}) and students
  with no timestamp data (\Sexpr{sum(is.na(secByStud[['time']]))})
  were excluded.}
\label{fig:timeByStud}
\end{figure}

\begin{figure}
<<mastSec,dependson='usageTime'>>=
print(mastStateYear <- ggplot(secByStud,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
@
\caption{Boxplots of the number of sections each of Cognitive Tutor
  software each student mastered, by year and state. Students
  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))})
  were excluded.}
\label{fig:mastSec}
\end{figure}

Figure \ref{fig:timeByStud} shows that the number of sections worked varied substantially between students and across states and years. Across years, students in Texas, New Jersey, and Connecticut worked fewer sections each than in Kentucky, Louisiana, and Michigan. From year 1 to 2, the median number of sections a student worked on decreased in every state, decreasing more substantially Texas and New Jersey than in others. The variance of the numbers of sections students worked in Michigan increased from year 1 to 2 as well.

Figure 2 shows the proportion of worked sections that were from a customized curriculum in the three states that used customized curricula substantially: Kentucky, Michigan, and Texas. As students worked fewer sections in year 2, a greater proportion of the sections they did work were from customized curricula. Across states, the average number of sections worked per student enrolled for the standard curricula was 53.83 in Year 1 and was similar at 52.45 in Year 2 across all states. For customized curricula in Year 2, the average number of worked sections per student enrolled was 19.93. However, since students were often enrolled in multiple curricula, these numbers underestimate the number of sections worked by students over the whole year.

Of course, more worked sections does not necessarily mean better worked sections. For instance, students might attempt fewer sections in year two, but graduate a higher percentage. Thus we investigated whether, on average, there were more graduated sections per students the second year, and how this related to whether they were enrolled in the standard versus customized curricula. Overall, the number of graduated sections per student increased in Year 2. The number of graduated sections per student was 17.94 in Year 1 and 25.66 in Year 2 in standard curricula. However, the customized curricula had fewer average graduated sections per student enrolled compared to the standard curricula; these numbers were 5.64 in Year 1 and 11.47 in Year 2. The percentages of average graduated sections to the average total worked sections were: 33.33% in Year 1 and 48.92% in Year 2 in the standard curricula, and 57.55% in Year 2 for the customized curricula. The average number of promoted sections per student also increased from Year 1 to Year 2, though only slightly. In Year 1, the average number of promotions were 1.20 for the customized curricula versus 3.64 for the standard curricula; in Year 2, those numbers were 2.20 and 4.95, respectively.

We also evaluated the time spent per section across different curricula and states.  Some records were omitted from this analysis: sections missing information on date of completion, as well as each student’s last section attempted because there is no completion date recorded for those sections. As presented in Figures 2 and 3 for Year 1 and 2, respectively, students in NJ increased the time spent per section from Year 1 to Year 2; however, there were relatively few students in the study from NJ. From Figure 1, it was shown that CT and TX completed, on average, fewer sections per student. From the information presented in Figures 2 and 3, these states spent the highest average number of days per section. TX, with the largest enrollment of students and the greatest use of customized curricula in Year 2, had students spending slightly more days completing sections in customized curricula relative to the standard curricula.

Figures 4 and 5 present information for only sections that were “graduated” by students, that is, cases in which a student was judged by the software to have mastered the material in the section. Results are similar to the data presented in Figures 2 and 3. However, in Year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time TX students spent mastering sections in the standard curricula. This difference is also seen in KY and LA. If customized curriculum are more aligned with teachers’ instruction, graduating from a section may require fewer days than sections in the standard curricula, possibly because students are working on the same topics in class as they are in the software. Figures 4 and 5 also show that, on average, CT and TX spent more days graduating from sections, regardless whether the section was part of a standard curricula or customized curricula. One explanation for the longer time periods in these states may be due to a teacher’s use of the software simultaneously with the textbook instruction. In this case, the teachers could have switched back-and-forth as they instructed, hence extending the time worked by their students. For the states with shorter time periods, teachers may have followed a more sequential process where they first completed one component, e.g. textbook material, and then allowed students to focus on and complete the second component, e.g. software.

### 2017-10-04 fig1.r

@@ -72,39 +72,6 @@ secByStudCurr <- within(secByStudCurr,{
 
 print(currStateYear <- ggplot(filter(secByStudCurr,!is.na(curriculum)),aes(Yr,numSec,fill=curriculum))+geom_boxplot()+facet_grid(~state))
 
-##########################################################################
-### curricula
-##########################################################################
-data$overall <- factor(ifelse(data$curriculum=='Customized','Customized','Standard'),levels=c('Standard','Customized'))
-data$Curriculum=data$curriculum
-levels(data$Curriculum) <- list(`>Algebra I`=c('Algebra II','Geometry'),`Algebra I`='Algebra I',
-                                `Bridge-to-Algebra`='Bridge-to-Algebra',Customized='Customized')
-
-### apportion customized sections to other curricula:
-custSec <- unique(data$section[data$overall=='Customized'])
-alg1Sec <- unique(data$section[data$Curriculum=='Algebra I'])
-advSec <- unique(data$section[data$Curriculum=='>Algebra'])
-BtASec <- unique(data$section[data$Curriculum=='Bridge-to-Algebra'])
-
-currFunc <- function(sec){
-
-    if(sec%in%alg1Sec) return('Algebra I')
-    if(sec%in%BtASec) return('Bridge-to-Algebra')
-    if(sec%in%advSec) return('>Algebra I')
-    if(grepl('triangle',sec)) return('>Algebra I')
-    if(grepl('geo-',sec)) return('>Algebra I')
-    if(grepl('exponents',sec)) return('Algebra I')
-    if(grepl('cta1',sec)) return('Algebra I')
-    if(grepl('graph',sec)) return('Algebra I')
-    if(grepl('linear',sec)) return('Algebra I')
-    if(grepl('quad',sec)) return('Algebra I')
-    return('na')
-}
-
-standCurr <- vapply(custSec, currFunc,'a')
-standCurr[standCurr=='na'] <- NA
-data$Curriculum[!is.na(data$Curriculum) & data$overall=='Customized' & !is.na(data$section)] <-
-    standCurr[match(data$section[!is.na(data$section) & data$overall=='Customized' & !is.na(data$Curriculum)],custSec)]
 
[…]

### 2017-10-04 models.r

@@ -20,7 +20,7 @@ cpDat$late <- cpDat$month%in%c('Jan','Feb','March','April')
 
 cpDat$cp <- cpDat$status=='changed placement'
 
-cpMod <- glmer(cp~year*(ns(xirt,5)+late)+spec_gifted+spec_speced+spec_esl+state+(ns(xirt,5)+late|classid2)+(1|field_id)+(late|unit),data=cpDat,family=binomial)
+cpMod <- glmer(cp~year*(xirt+late)+spec_gifted+spec_speced+spec_esl+state+(xirt+late|classid2)+(1|field_id)+(late|unit),data=cpDat,family=binomial)
 
 
 ### does failing to master one section increase the odds of failing to master the next?

### 2017-10-04 fig1.r
some merging

@@ -72,9 +72,46 @@ secByStudCurr <- within(secByStudCurr,{
 
 print(currStateYear <- ggplot(filter(secByStudCurr,!is.na(curriculum)),aes(Yr,numSec,fill=curriculum))+geom_boxplot()+facet_grid(~state))
 
+##########################################################################
+### curricula
+##########################################################################
+data$overall <- factor(ifelse(data$curriculum=='Customized','Customized','Standard'),levels=c('Standard','Customized'))
+data$Curriculum=data$curriculum
+levels(data$Curriculum) <- list(`>Algebra I`=c('Algebra II','Geometry'),`Algebra I`='Algebra I',
+                                `Bridge-to-Algebra`='Bridge-to-Algebra',Customized='Customized')
+
+### apportion customized sections to other curricula:
+custSec <- unique(data$section[data$overall=='Customized'])
+alg1Sec <- unique(data$section[data$Curriculum=='Algebra I'])
+advSec <- unique(data$section[data$Curriculum=='>Algebra'])
+BtASec <- unique(data$section[data$Curriculum=='Bridge-to-Algebra'])
+
+currFunc <- function(sec){
+
+    if(sec%in%alg1Sec) return('Algebra I')
+    if(sec%in%BtASec) return('Bridge-to-Algebra')
+    if(sec%in%advSec) return('>Algebra I')
+    if(grepl('triangle',sec)) return('>Algebra I')
+    if(grepl('geo-',sec)) return('>Algebra I')
+    if(grepl('exponents',sec)) return('Algebra I')
+    if(grepl('cta1',sec)) return('Algebra I')
+    if(grepl('graph',sec)) return('Algebra I')
+    if(grepl('linear',sec)) return('Algebra I')
+    if(grepl('quad',sec)) return('Algebra I')
+    return('na')
+}
+
+standCurr <- vapply(custSec, currFunc,'a')
+standCurr[standCurr=='na'] <- NA
+data$Curriculum[!is.na(data$Curriculum) & data$overall=='Customized' & !is.na(data$section)] <-
+    standCurr[match(data$section[!is.na(data$section) & data$overall=='Customized' & !is.na(data$Curriculum)],custSec)]
[…]

### 2017-10-04 paper.Rnw
progress....

@@
 
-\section{Standard versus Customized Curricula}
+\section{Standard and Customized Curricula}
 
-All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year.
-In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools.
+\begin{figure}
+  \centering
+<<curricula,dependson='data',fig.width=6.4,fig.height=3>>=
+
+omar <- par()$oma
+
+par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
+par(mfrow=c(1,9))
+for(st in levels(data$State)){
+#    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
+    cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
+    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
+    for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
+    cwy <- cwy[nrow(cwy):1,]
+    barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main=st,yaxt='n')
+    if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
+    #box()
+}
+cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
+    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
+for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
+cwy <- cwy[nrow(cwy):1,]
+barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main='Overall',yaxt='n')
+
+plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
+legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
+       fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
+       angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,75),xpd = TRUE, bty = "n",inset=c(0,0))
+
+par(oma=omar)
+@
+\caption{Percentage of worked problems coming from various levels
+  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
+  standard and customized variants, denoted by shading.}
+\label{fig:curricula}
+\end{figure}
+
+Students' automatic progress throught the CTAI software is governed by
+curricula---sequences of sections and units embedded in the software.
+Without external meddling, the curriculum a student works on
+determines what section he or she will be directed towards next after
+mastering (or exhausting the problems) from a previous section.
+In the CTAI effectiveness trial, the most common curriculum was,
+naturally, Algebra I.
+This came with three closely related varients, due to software releases.
+Students requiring more remediation were able to work on a less
+advanced curriculum, called ``Bridge to Algebra,'' and more advanced
+students could work on Algebra II or Geometry.
+
+In the second year of the study, some high schools, primarily in
+Texas, Michigan, and Kentucky, requested customized variants of the
+curricula.
+This was typically due to state standards and testing schedules.
+These ``customized curricula'' altered the order of some sections and
+units, and were particular to schools.
+
+Figure \ref{fig:curricula} shows the percentage of worked problems
+from each curriculum, from standard and customized varieties, by state
+and year.
+First, note that the vast majority of worked problems were from the
+Algebra I sequence.
+A small but notable number of more advanced problems were worked in
+Kentucky in Year 2, and some less advanced problems were worked in
+Michigan and Louisiana.
+Secondly, note the rise in ``customized curricula'' in year 2 in
+Texas, Kentucky, and Michigan, the three states with the most students
+in our dataset.
+In particular, Texas shifted almost entirely to customized curriclua
+from years 1 to 2.
+
+% All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year.
+% Figure
+% In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools.
 The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus.
@@
 
-In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204.
+% In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204.
 The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2.
@@
 
-The use of customized curricula varied substantially by state.
+% The use of customized curricula varied substantially by state.
 For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2.
@@
 
-Throughout the school year, teachers could have a class of students working on multiple curricula either sequentially, where the students changed curricula in lock step, or simultaneously, where students worked on different curricula at the same time.
-As an example, two teachers located in KY had their students working on Algebra I throughout most of the year and then reassigned them to Algebra II in the last month of school.
-Another teacher in KY had students variously enrolled in three different curricula throughout the entire year (Bridge-to-Algebra, Algebra I, and a specialized Geometry curriculum), while another teacher in MI during Year 2 had students enrolled in three curricula sequentially throughout the year covering Algebra I until November, and then starting a customized curriculum and using it until February, and ending the year with another customized curriculum that was used until June.
-While there are numerous instances of these uses of multiple curricula in Year 2, there are also many occurrences of teachers who had their students enrolled in the standard Algebra I throughout the entire year.
-In TX mostly, with some instances in MI, there were teachers who exclusively used customized curricula the whole year.
+Throughout the school year, teachers could have a class of students
+working on multiple curricula either sequentially, where the students
+changed curricula in lock step, or simultaneously, where students
+worked on different curricula at the same time. As an example, two
+teachers located in KY had their students working on Algebra I
+throughout most of the year and then reassigned them to Algebra II in
+the last month of school. Another teacher in KY had students variously
+enrolled in three different curricula throughout the entire year
+(Bridge-to-Algebra, Algebra I, and a specialized Geometry curriculum),
+while another teacher in MI during Year 2 had students enrolled in
+three curricula sequentially throughout the year covering Algebra I
+until November, and then starting a customized curriculum and using it
+until February, and ending the year with another customized curriculum
+that was used until June. While there are numerous instances of these
+uses of multiple curricula in Year 2, there are also many occurrences
+of teachers who had their students enrolled in the standard Algebra I
+throughout the entire year.  In TX mostly, with some instances in MI,
+there were teachers who exclusively used customized curricula the
+whole year.
 
@@
 \begin{figure}
-<<usageTime, fig.height=3,fig.width=6>>=
+<<usageTime, fig.height=3,fig.width=6,dependson='data'>>=
 secByStud <- summarize(group_by(data,state,Yr,field_id),numSec=n_distinct(section,na.rm=TRUE),time=sum(total_t1,na.rm=TRUE),mastered=n_distinct(section[status=='graduated'],na.rm=TRUE))
+
+secByStud$time <- secByStud$time/3600000
 
@@
 
-print(timeStateYear <- ggplot(secByStud,aes(Yr,time/3600000))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,100))+labs(x='',y='Time on CTAI Software per Student (Hours)'))
+levels(secByStud$state) <- c(levels(secByStud$state),'Overall')
+secByStud2 <- secByStud
+secByStud2 <- rbind(secByStud,secByStud2)
+secByStud2$state[seq(nrow(secByStud)+1,2*nrow(secByStud))] <- 'Overall'
+
+print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CTAI Software per Student'))
 #ggsave('timeStatYear.jpg',width=6,height=3)
@@
 @
-\caption{Boxplots of hours each student spent on Cognitive Tutor
-  software over by year and state. Students spending more than 100
-  hours (\Sexpr{sum(secByStud[['time']]>100,na.rm=TRUE)} of \Sexpr{nrow(secByStud)}) and students
-  with no timestamp data (\Sexpr{sum(is.na(secByStud[['time']]))})
+\caption{Boxplots (excluding outliers) of hours each student spent on Cognitive Tutor
+  software over by year and state. Students  with no timestamp data
+  (\Sexpr{sum(is.na(secByStud[['time']]))/2})
   were excluded.}
@@
 \begin{figure}
-<<mastSec,dependson='usageTime'>>=
-print(mastStateYear <- ggplot(secByStud,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
+<<mastSec,dependson='usageTime',fig.width=6,fig.height=3,dependson=c('usageTime','data')>>=
+print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
 @
@@
   software each student mastered, by year and state. Students
-  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
-  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))})
+  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
+  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
   were excluded.}
@@
 
-Figure \ref{fig:timeByStud} shows that the number of sections worked varied substantially between students and across states and years.
-Across years, students in Texas, New Jersey, and Connecticut worked fewer sections each than in Kentucky, Louisiana, and Michigan.
-From year 1 to 2, the median number of sections a student worked on decreased in every state, decreasing more substantially Texas and New Jersey than in others.
-The variance of the numbers of sections students worked in Michigan increased from year 1 to 2 as well.
+Figure \ref{fig:timeByStud} shows that the number of hours students
+spent working on the CTAI software
+varied substantially between students and across states and
+years.
+Overall, students spent less time on the software in Year 2 than in
+Year 2.
+Specifically, in the first year the students spent a median of
+\Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours on the software, compared to \Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.
+However, usage varied more in year 2 as well---the interquartile
+range of time spent was \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours in the first year, compared to \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.
 
-Figure 2 shows the proportion of worked sections that were from a customized curriculum in the three states that used customized curricula substantially: Kentucky, Michigan, and Texas.
-As students worked fewer sections in year 2, a greater proportion of the sections they did work were from customized curricula.
-Across states, the average number of sections worked per student enrolled for the standard curricula was 53.83 in Year 1 and was similar at 52.45 in Year 2 across all states.
-For customized curricula in Year 2, the average number of worked sections per student enrolled was 19.93.
-However, since students were often enrolled in multiple curricula, these numbers underestimate the number of sections worked by students over the whole year.
+Time spent varied substantially by state, with students in Texas,
+Connecticut, and New Jersey working far fewer hours than students in
+Kentucky, Louisiana, and Michigan.
+Not every state reduced its usage from years 1 to 2---while students
+in Texas, Kentucky and New Jersey used the software less in the second
+year than in the first, students in Michigan, Louisiana, and
+Connectuct increased their usage.
 
@@
 software.
+
+\end{document}

### 2017-10-04 fig1.r

@@ -75,40 +75,6 @@ print(currStateYear <- ggplot(filter(secByStudCurr,!is.na(curriculum)),aes(Yr,nu
 ##########################################################################
 ### curricula
 ##########################################################################
-data$overall <- factor(ifelse(data$curriculum=='Customized','Customized','Standard'),levels=c('Standard','Customized'))
-data$Curriculum=data$curriculum
-levels(data$Curriculum) <- list(`>Algebra I`=c('Algebra II','Geometry'),`Algebra I`='Algebra I',
-                                `Bridge-to-Algebra`='Bridge-to-Algebra',Customized='Customized')
-
-### apportion customized sections to other curricula:
-custSec <- unique(data$section[data$overall=='Customized'])
-alg1Sec <- unique(data$section[data$Curriculum=='Algebra I'])
-advSec <- unique(data$section[data$Curriculum=='>Algebra'])
-BtASec <- unique(data$section[data$Curriculum=='Bridge-to-Algebra'])
-
-currFunc <- function(sec){
-
-    if(sec%in%alg1Sec) return('Algebra I')
-    if(sec%in%BtASec) return('Bridge-to-Algebra')
-    if(sec%in%advSec) return('>Algebra I')
-    if(grepl('triangle',sec)) return('>Algebra I')
-    if(grepl('geo-',sec)) return('>Algebra I')
-    if(grepl('exponents',sec)) return('Algebra I')
-    if(grepl('cta1',sec)) return('Algebra I')
-    if(grepl('graph',sec)) return('Algebra I')
-    if(grepl('linear',sec)) return('Algebra I')
-    if(grepl('quad',sec)) return('Algebra I')
-    return('na')
-}
-
-standCurr <- vapply(custSec, currFunc,'a')
-standCurr[standCurr=='na'] <- NA
-data$Curriculum[!is.na(data$Curriculum) & data$overall=='Customized' & !is.na(data$section)] <-
-    standCurr[match(data$section[!is.na(data$section) & data$overall=='Customized' & !is.na(data$Curriculum)],custSec)]
-data$Curriculum[grep('algebrai',data$curriculumOrig)] <- 'Algebra I'
-data$Curriculum[grep('algebra1',data$curriculumOrig)] <- 'Algebra I'
-data$Curriculum[grep('geom',data$curriculumOrig)] <- '>Algebra I'
-
 
 #### what proportion of worked sections from each curriculum?
 currWorked <- data%>%distinct(field_id,section,unit,Curriculum,Year,State,overall)

### 2017-10-10 paper.Rnw
more progress

@@
 library(knitr)
-library(tikzDevice)
+#library(tikzDevice)
 opts_chunk$set(
-echo=FALSE, results='asis',cache=TRUE,warning=FALSE,error=FALSE,dev='tikz',message=FALSE,autodep = FALSE
+echo=FALSE, results='asis',cache=TRUE,warning=FALSE,error=FALSE,message=FALSE,autodep = FALSE
     )
 
-options(tikzDefaultEngine = "pdftex")
+#options(tikzDefaultEngine = "pdftex")
+@
+
+<<setup,include=FALSE>>=
 
@@
 library(ggplot2)
+library(xtable)
+
+addOverallState <- function(dat){
+    levels(dat$state) <- c(levels(dat$state),'Overall')
+    dat2 <- dat
+    dat2$state <- 'Overall'
+    dat2 <- rbind(dat,dat2)
+    dat2$state <- factor(dat2$state, levels= levels(dat$state))
+    dat2
+}
 
@@
 <<data,include=FALSE>>=
+if(file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
+    load('cpPaper.RData')
+} else source('dataMerge.r')
+@
+
+<<data1,include=FALSE,cache=FALSE>>=
+dataCache <- grep('data',list.files('cache'),ignore.case=FALSE,value=TRUE)[1]
+if(file.info('dataMerge.r')$mtime>file.info(paste0('cache/',dataCache))$mtime)
     source('dataMerge.r')
@@
 Thus, the data from the study reflect typical software usage.
-Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects (Sales & Pane, 2015, Sales et al.
+Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects (Sales \& Pane, 2015, Sales et al.
 2016).
@@
   \centering
-<<curricula,dependson='data',fig.width=6.4,fig.height=3>>=
+<<curricula,dependson=c('setup','data'),fig.width=6.4,fig.height=3>>=
 
@@
 
-Students' automatic progress throught the CTAI software is governed by
+Students' automatic progress throught the Cognitive Tutor (CT) software is governed by
 curricula---sequences of sections and units embedded in the software.
@@
 
-\section{Student Usage Across Sections and Years}
-\begin{figure}
-<<usageTime, fig.height=3,fig.width=6,dependson='data'>>=
-secByStud <- summarize(group_by(data,state,Yr,field_id),numSec=n_distinct(section,na.rm=TRUE),time=sum(total_t1,na.rm=TRUE),mastered=n_distinct(section[status=='graduated'],na.rm=TRUE))
+\section{Student Usage Across States and Years}
+
+<<usageMedians,dependson=c('data','setup'),results='asis'>>=
+secByStud <- data%>%group_by(state,Yr,field_id) %>% summarize(numSec=n_distinct(section,na.rm=TRUE),
+                                                              numUnit=n_distinct(unit,na.rm=TRUE),
+                                                              time=sum(total_t1,na.rm=TRUE),
+                                                              mastered=n_distinct(section[status=='graduated'],na.rm=TRUE),
+                                                              nprob=n_distinct(unit,section,Prob1,na.rm=TRUE))
 
@@
 secByStud <- within(secByStud,{
+    numUnit[numUnit==0] <- NA
     numSec[numSec==0] <- NA
@@
 
-levels(secByStud$state) <- c(levels(secByStud$state),'Overall')
-secByStud2 <- secByStud
-secByStud2 <- rbind(secByStud,secByStud2)
-secByStud2$state[seq(nrow(secByStud)+1,2*nrow(secByStud))] <- 'Overall'
+secByStud2 <- addOverallState(secByStud)
 
-print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CTAI Software per Student'))
+secByStud2 <-
+  secByStud2 %>%
+  group_by(state,Yr) %>%
+  mutate(outlier.time = time > quantile(time,0.75,na.rm=TRUE) + IQR(time,na.rm=TRUE) * 1.5,
+         outlier.sec = numSec > quantile(numSec,0.75,na.rm=TRUE) + IQR(numSec,na.rm=TRUE) * 1.5,
+         outlier.prob = nprob > quantile(nprob,0.75,na.rm=TRUE) + IQR(nprob,na.rm=TRUE) * 1.5,
+         outlier.unit = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5
+         ) %>%
+  ungroup
+
+tab <- secByStud%>%group_by(Yr)%>%select(time,nprob,numSec,numUnit)%>%summarize_all(median,na.rm=TRUE)
+tab$Yr <- NULL
+tab <- as.data.frame(tab)
+rownames(tab) <- c('Year 1','Year 2')
+names(tab) <- c('Hours','Problems','Sections','Units')
+
+xtable(tab,caption='Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study.
+Students with no usage data were excluded',label='tab:medUsage',digits=c(1,2,0,1,0))
+@
+
+Table \ref{tab:tab:medUsage} shows the median numbers of hours,
+problems, sections, and units worked on by each student in the dataset
+in the two years of the study.
+Apparently, usage decreased markedly in the second year: the median
+student worked about
+\Sexpr{round(tab['Year 1','Hours']-tab['Year 2','Hours'])}
+fewer hours,
+\Sexpr{round(tab['Year 1','Problems']-tab['Year 2','Problems'])}
+fewer problems, and
+\Sexpr{round(tab['Year 1','Sections']-tab['Year 2','Sections'])} fewer
+sections in year 2 than in year 1.
+
+\begin{figure}
+<<usageTime, fig.height=3,fig.width=6,dependson=c('data','setup','usageMedians')>>=
+
+print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
+          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.time), width=0.2)+
+          facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CT Software per Student'))
                                         #ggsave('timeStatYear.jpg',width=6,height=3)
@@
 @
-\caption{Boxplots (excluding outliers) of hours each student spent on Cognitive Tutor
+\caption{Boxplots of hours each student spent on Cognitive Tutor
   software over by year and state. Students  with no timestamp data
-  (\Sexpr{sum(is.na(secByStud[['time']]))/2})
+  (\Sexpr{sum(is.na(secByStud[['time']]))}), with negative time
+  (\Sexpr{sum(secByStud[['time']]<0,na.rm=TRUE)}) or with more than 110 hours (\Sexpr{sum(secByStud[['time']]>110,na.rm=TRUE)})
   were excluded.}
@@
 
-\begin{figure}
-<<mastSec,dependson='usageTime',fig.width=6,fig.height=3,dependson=c('usageTime','data')>>=
-print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
+<<supplementalFigs1,include=FALSE,dependson=c('data','setup','usageMedians')>>=
+secStateYear <- ggplot(secByStud2,aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
+          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.sec), width=0.2)+
+          facet_grid(~state)+labs(x='',y='# of Sections Worked per Student')+coord_cartesian(ylim=c(0,200))
+ggsave('secStateYear.jpg',width=6,height=3)
+
+probStateYear <- ggplot(secByStud2,aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
+          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.prob), width=0.2)+
+          facet_grid(~state)+labs(x='',y='# of Problems Worked per Student')+coord_cartesian(ylim=c(0,2000))
+ggsave('probStateYear.jpg',width=6,height=3)
+
 @
-\caption{Boxplots of the number of sections each of Cognitive Tutor
-  software each student mastered, by year and state. Students
-  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
-  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
-  were excluded.}
-\label{fig:mastSec}
-\end{figure}
+
+% \begin{figure}
+% <<mastSec,dependson=c('usageTime','data','setup'),fig.width=6,fig.height=3>>=
+% print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
+% @
+% \caption{Boxplots of the number of sections each of Cognitive Tutor
+%   software each student mastered, by year and state. Students
+%   mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
+%   with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
+%   were excluded.}
+% \label{fig:mastSec}
+% \end{figure}
 
 Figure \ref{fig:timeByStud} shows that the number of hours students
-spent working on the CTAI software
-varied substantially between students and across states and
+spent working on the CT software in some more detail, via
+state-by-year boxplots.
+Analogous figures for the numbers of problems and sections students
+worked, available in an online supplement, showed similar patterns for
+the numbers of sections and problems students worked.
+Usage time varied substantially between students and across states and
 years.
-Overall, students spent less time on the software in Year 2 than in
-Year 2.
-Specifically, in the first year the students spent a median of
-\Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours on the software, compared to \Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.
-However, usage varied more in year 2 as well---the interquartile
-range of time spent was \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours in the first year, compared to \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.
-
 Time spent varied substantially by state, with students in Texas,
@@
 
+Overall, usage varied a bit more in year 2 than in year
+1---the median absolute deviation of time spent was \Sexpr{round(mad(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE,constant=1),1)} hours in the first year, compared to \Sexpr{round(mad(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE,constant=1),1)} in  the second year.
+The increase in variation seems to be driven both by increasing
+between-state variation, and a between-student increase in Louisiana.
+One intriguing possibility is that CT usage was better tailored to
+teachers and students in the second year than in the first---perhaps
+students who stood to gain more from the software increased their
+usage, and the rest of the students used it less.
+
+In contrast to the numbers of hours, problems, and sections students
+worked, Table \ref{tab:medUsage} shows that the median number of units
+students worked increased by
+\Sexpr{round(tab['Year 2','Units']-tab['Year 1','Units'])}
+from year 1 to year 2.
+This suggests students in year 2 were exposed, on average, to a
+slightly wider range of topics.
+Figure \ref{fig:unitsByStud} shows boxplots of the numbers of units
+worked by state and year.
+The geographic variation in units worked mirrors the pattern in Figure
+\ref{fig:timeByStud}, with more usage in Kentudky, Michigan, and
+Louisiana but less in Texas, Connecticut, and New Jersey.
+However, in every state the median Year 2 student worked at least as
+many different units as the median Year 1 student.
+Variation in the number of units worked also increased slightly
+from years 1 to 2---the interquartile range increased in every state
+except for Kentucky, where a decrease in IQR was accompanied by an
+incrase in the number of outliers.
+
+\begin{figure}
+<<unitsWorked,dependson=c('usageMedians','data','setup'),fig.width=6,fig.height=3>>=
+secByStud2 <-
+  secByStud2 %>%
+  group_by(state,Yr) %>%
+  mutate(outlier = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5) %>%
+  ungroup
+
+print(unitsStateYear <- ggplot(secByStud2,aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
+@
+\caption{Boxplots of the number of units of Cognitive Tutor
+  software each student worked, by year and state. Students
+  working more than 55 units (\Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
+  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
+  were excluded.}
+\label{fig:unitsByStud}
+\end{figure}
+
+All in all, students used CT software less in year 2 than in year
+1.
+On the other hand, students in the second year tended to see a
+slightly wider range of topics, and varied somewhat more in their usage.
+
+\begin{comment}
 Of course, more worked sections does not necessarily mean better worked sections.
@@
 software.
+\end{comment}
+
+\section{Mastery Learning: Working Units in Order---Or Not}
+
+Overally, students used CT less in the second year than in the
+first.
+How was this difference distributed between CTAI units?
+
+\begin{figure}
+<<whichUnits,fig.height=4,fig.width=6,dependson=c('data','setup')>>=
+
+curricula <- read.csv('~/Box Sync/CT/data/sectionLevelUsageData/RAND_study_curricula.csv',stringsAsFactors=FALSE)
+curricula <- subset(curricula,curriculum_name=='algebra i')
+curricula$unit <- tolower(curricula$unit)
+curricula <- subset(curricula,unit%in%intersect(curricula$unit[curricula$ct=='2007'],curricula$unit[curricula$ct=='2008r1']))
+units <- curricula$unit[curricula$ct=='2007']
+sectionStats <- read.csv('~/Box Sync/CT/data/sectionLevelUsageData/section_stats_withAbb.csv',stringsAsFactors=FALSE)
+UnitName <- sectionStats$unit_name_abb[match(units,sectionStats$unit_id)]
+UnitName[units=='inequality-systems-solving'] <- 'Systems of Lin. Ineq.'
+UnitName[units=='intro-pythag-theorem'] <- 'Pythagorean Theorem'
+UnitName[units=='linear-inequality-graphing'] <- 'Graphs of Lin. Ineq.'
+UnitName[units=='linear-systems-solving'] <- 'Systems of Lin. Eq. Solving'
+UnitName[units=='probability'] <- 'Probability'
+UnitName[units=='unit-conversions'] <- 'Unit Conversions'
+
+nstud <- data%>%filter(!is.na(unit))%>%group_by(Year)%>%summarize(nstud=n_distinct(field_id))
+data$Unit <- data$unit
+data$Unit[grep('unit-conversions',data$Unit)] <- 'unit-conversions'
+
+unitLevel <- data%>%filter(Unit%in%units)%>%group_by(Unit,Year)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+
+unitLevel$perWorked <- unitLevel$numWorked/nstud$nstud[match(unitLevel$Year,nstud$Year)]
+
+unitLevel$Unit <- factor(unitLevel$Unit,levels=units)
+levels(unitLevel$Unit) <- UnitName
+
+unitLevel$year <- factor(ifelse(unitLevel$Year=='Year 1',1,2))
+print(unitsWorked <- ggplot(unitLevel,aes(x=Unit,y=perWorked,color=year,group=year))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Year')+scale_y_continuous(labels=percent))
+
+@
+\caption{The percentages of students with usage data who worked at
+  least one problem from each unit in the Algebra I curriculum. The
+  units are arranged in order for the standard curriculum. In the 2008
+version of the software, the ``Unit Conversions'' unit was broken up
+into two smaller units; for the sake of between-year comparisons, we
+re-combined them.}
+\label{fig:unitsWorked}
+\end{figure}
+
+Figure \ref{fig:unitsWorked} shows the percentage of students with
+usage data who worked each section from the standard CTAI curriculum.
+The units are ordered, from left to right, as the are prescribed in
+the standard CTAI curriculum.
+
+In year 1, the curve is almost monontonically decreasing, as one would
+expect if students adhered to the curriculum.
+Students varied in how many units they worked---with the variation due
+to both student ability and the amount of time allocated to CTAI
+within a classroom---but they mostly followed the same curriculum.
+Students who worked fewer units stopped earlier in the curricular
+order, and those who worked more units progressed farther.
+Hence, earlier sections were worked by higher proportions of students
+than later units.
+
+In contrast, in year 2 students were much more likely to buck the
+standard unit order.
+For instance, Figure \ref{fig:unitsWorked} suggests that students
+skipped ``Unit Conversions'' and ``1 step Lin. Eq.'' for ``1st
+Quadrent Linear Graphs'' and ``Ind. Variables in Lin. Mod.''
+Both of the latter sections were worked by a greater proportion of
+students than the former sections that preceded them in the CTAI
+curriculum.
+Most strikingly, ``Lin. Eq. w/ Var's Both Sides'' was worked by a
+greater proportion of students in Year 2 than in Year 1, and by a
+greater proportion of students than any of the previous six sections.
+Presumeably teachers and administrators wanted students to focus on
+that unit, perhaps because they found it to be particularly effective,
+because students tend to struggle with its main topic, or because its
+topic may figure prominantly in an upcoming standardized test.
 

### 2017-10-10 fig1.r

@@ -72,6 +72,18 @@ secByStudCurr <- within(secByStudCurr,{
 
 print(currStateYear <- ggplot(filter(secByStudCurr,!is.na(curriculum)),aes(Yr,numSec,fill=curriculum))+geom_boxplot()+facet_grid(~state))
 
+##### number of skills tried/mastered
+load('~/Box Sync/CT/data/skillsdata/studentSkillTotals.RData')
+studLevel <- within(studLevel,mastPer <- nmast/ntry)
+studLevel <- rename(studLevel,field_id=student)
+secByStud <- left_join(secByStud,studLevel)
+
+print(skillsTry <- ggplot(secByStud,aes(Yr,ntry))+geom_boxplot()+facet_grid(~state)+labs(x='',y='# Skills Attempted'))
+
+print(skillsMast <- ggplot(secByStud,aes(Yr,nmast))+geom_boxplot()+facet_grid(~state)+labs(x='',y='# Skills Mastered'))
+
+
+
 ##########################################################################
 ### curricula
 ##########################################################################
@@ -228,6 +240,26 @@ print(unitsWorked <- ggplot(unitLevel,aes(x=Unit,y=perWorked,color=year,group=ye
 ggsave('unitsWorked.jpg',height=4,width=6)
 
 
+#### by state
+nstudState <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(state)%>%summarize(nstud=n_distinct(field_id))
+unitLevelState <- data%>%filter(Unit%in%units & Year=='Year 2' & state!='NJ' & state!='CT')%>%group_by(Unit,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+
+unitLevelState$perWorked <- unitLevelState$numWorked/nstudState$nstud[match(unitLevelState$state,nstudState$state)]
+
+unitLevelState$Unit <- factor(unitLevelState$Unit,levels=units)
+levels(unitLevelState$Unit) <- UnitName
[…]

### 2017-10-10 fig1.r
which units--by state/school

@@ -250,15 +250,29 @@ unitLevelState$Unit <- factor(unitLevelState$Unit,levels=units)
 levels(unitLevelState$Unit) <- UnitName
 ggplot(unitLevelState,aes(x=Unit,y=perWorked,color=state,group=state))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))#+labs(x='',y='% Worked',color='Year')+scale_y_continuous(labels=percent))
 
-nstudState <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(state)%>%summarize(nstud=n_distinct(field_id))
-unitLevelState <- data%>%filter(Unit%in%units & Year=='Year 2' & state!='NJ' & state!='CT')%>%group_by(Unit,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+### by customized curriculum (at school level)
+cust <- data%>%filter(Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(cust=mean(overall=='Customized',na.rm=T))%>%arrange(cust)
+data$cust <- data$schoolid2%in%cust$schoolid2[cust$cust>0.8]
 
-unitLevelState$perWorked <- unitLevelState$numWorked/nstudState$nstud[match(unitLevelState$state,nstudState$state)]
+nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2' & state=='TX')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
+unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2'&state=='TX' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
 
-unitLevelState$Unit <- factor(unitLevelState$Unit,levels=units)
-levels(unitLevelState$Unit) <- UnitName
[…]

### 2017-10-10 paper.Rnw
more stuff

@@
 number of sections within each unit varied.
-By default, the software operates by automatically moving students from section to section based on the sequence of topics defined by the curriculum they were currently enrolled in.
-In this software-controlled sequencing, students ideally spend the time necessary to learn the material of a section, are judged by the software to have mastered the material, and then “graduate” to the next section.
-Students who exhaust a section's material without mastering its skills are “promoted” to the next section.
-Students left uninterrupted follow the sequence of topics defined by their curriculum, ideally graduating from topics by showing mastery.
-However, teachers could modify a student’s path within the curriculum.
-They could reassign students from their current sections to other sections earlier or later in the intended sequence, including sections they had worked on previously.
 This study attempts to elucidate teachers' goals in reassigning students.
@@
 
-omar <- par()$oma
+bbb <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%
+    group_by(state,Curriculum,overall,Yr)%>%
+    summarize(n=n())
+staten <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%group_by(state,Yr)%>%summarize(n=n())
 
-par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
-par(mfrow=c(1,9))
-for(st in levels(data$State)){
-#    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
-    cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
-    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-    for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-    cwy <- cwy[nrow(cwy):1,]
-    barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main=st,yaxt='n')
-    if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
-    #box()
-}
-cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
-    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-cwy <- cwy[nrow(cwy):1,]
-barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main='Overall',yaxt='n')
+for(i in 1:nrow(bbb)) bbb$n[i] <- bbb$n[i]/staten$n[staten$state==bbb$state[i] & staten$Yr==bbb$Yr[i]]
 
-plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
-legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
-       fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
-       angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,75),xpd = TRUE, bty = "n",inset=c(0,0))
+ggplot(bbb,aes(Yr,n,fill=Curriculum,alpha=overall))+geom_col()+facet_grid(~state)+scale_alpha_manual(values=c(1,0.5))+scale_fill_manual(values=COLS)+scale_y_continuous(labels=percent)+labs(x='',y='% of Problems Worked',alpha='')
 
-par(oma=omar)
+## ggplot(bbb,aes(Yr,n,color=bbb))+geom_col(show.legend=FALSE)+geom_col(aes(fill=Curriculum))+geom_col(aes(fill=overall),alpha=0.5)+facet_grid(~state)
+##     scale_fill_manual(values=c('#8dd3c7','#8dd3c7','#ffffb3','#ffffb3','#bebada','white','black'))
+##     geom_col(aes(Yr,n,fill=overall),alpha=0.5)+
+##     #geom_col(aes(Yr,n,fill=overall))+#scale_fill_manual(values=c('white','black'))+
+##     facet_grid(~state)
+
+## omar <- par()$oma
+
+## par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
+## par(mfrow=c(1,9))
+## for(st in levels(data$State)){
+## #    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
+##     cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
+##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
+##     for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
+##     cwy <- cwy[nrow(cwy):1,]
+##     barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main=st,yaxt='n')
+##     if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
+##     #box()
+## }
+## cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
+##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
+## for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
+## cwy <- cwy[nrow(cwy):1,]
+## barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main='Overall',yaxt='n')
+
+## plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
+## legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
+##        fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
+##        angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,50),xpd = TRUE, bty = "n",inset=c(0,0))
+
+## par(oma=omar)
 @
@@
 \begin{figure}
+\centering
 <<usageTime, fig.height=3,fig.width=6,dependson=c('data','setup','usageMedians')>>=
 
-print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
+print(timeStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
           geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.time), width=0.2)+
@@
 <<supplementalFigs1,include=FALSE,dependson=c('data','setup','usageMedians')>>=
-secStateYear <- ggplot(secByStud2,aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
+secStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
           geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.sec), width=0.2)+
@@
 
-probStateYear <- ggplot(secByStud2,aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
+probStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
           geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.prob), width=0.2)+
@@
 \begin{figure}
+\centering
 <<unitsWorked,dependson=c('usageMedians','data','setup'),fig.width=6,fig.height=3>>=
@@
 
-print(unitsStateYear <- ggplot(secByStud2,aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
+print(unitsStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
 @
@@
 
-\section{Mastery Learning: Working Units in Order---Or Not}
+\section{Working Units in Order---Or Not}
 
@@
 
+\begin{figure}
+  \centering
+<<unitsWorkedCust,dependson=c('data','setup','whichUnits'),fig.height=4,fig.width=6>>=
+### by customized curriculum (at school level)
+cust <- data%>%filter(Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(cust=mean(overall=='Customized',na.rm=T))%>%arrange(cust)
+data$cust <- ifelse(data$schoolid2%in%cust$schoolid2[cust$cust>0.8],'Customized','Standard')
+
+nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
+unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+
+unitLevelCust$perWorked <- unitLevelCust$numWorked/nstudCust$nstud[match(unitLevelCust$cust,nstudCust$cust)]
+
+unitLevelCust$Unit <- factor(unitLevelCust$Unit,levels=units)
+levels(unitLevelCust$Unit) <- UnitName
+ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Curriculum',title='Year 2')+scale_y_continuous(labels=percent)
+@
+\caption{The percentages of year-2 students with usage data who worked at
+  least one problem from each unit in the Algebra I curriculum. The
+  units are arranged in order for the standard curriculum. Students
+  are divided between those attending schools using primarily a customized
+  curriculum and those using primariy the standard Algebra I curriculum.}
+\label{fig:unitsWorkedCust}
+\end{figure}
+
+Most of the variation in unit order was driven by the rise, in year 2,
+of customized curricula.
+Figure \ref{fig:unitsWorkedCust} divides year-2 students into those
+attending schools using primarily a customized curriculum, and those
+attending schools using primarily a standardized
+curriculum.\footnote{At least
+  \Sexpr{trunc(100*min(cust[['cust']][cust[['cust']]>0.8]))}\% of
+  problems at ``Customized'' schools were from a customized curriculum,
+  and at most
+  \Sexpr{ceiling(100*max(cust[['cust']][cust[['cust']]<0.8]))}\% of
+  problems at ``Standard'' schools were from a customized curriculum.}
+Students using a standardized curriculum followed the standard
+sequence---more or less---while students using customized curricula
+did not.
+That said, there were some order violations in the standard group:
+specifically, more students worked problems from units ``4-Quadrant
+Linear Graphs'' and ``Exponents'' than worked the preceding sections;
+this suggests that some teachers used the reassignment tool to
+prioritize particular topics.
+Of course, this may have occurred in schools with customized
+curriculua as well---a possibility we will discuss in the next
+section.
+
+\begin{figure}
+  \centering
+<<unitsBySchool,dependson=c('data','setup','whichUnits'),fig.height=7,fig.width=6>>=
+### by school
+nstudSch <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(nstud=n_distinct(field_id))
+unitLevelSch <- with(filter(data,Unit%in%units & Year=='Year 2'),expand.grid(Unit=unique(Unit),schoolid2=unique(schoolid2)))
+unitLevelSch$state <- data$state[match(unitLevelSch$schoolid2,data$schoolid2)]
+unitLevelSch$cust <- data$cust[match(unitLevelSch$schoolid2,data$schoolid2)]
+unitLevelSch$numWorked <- with(filter(data,Year=='Year 2'),vapply(1:nrow(unitLevelSch),function(i)
+    n_distinct(field_id[schoolid2==unitLevelSch$schoolid2[i] & Unit==unitLevelSch$Unit[i]]),1))
+
+#unitLevelSch <- data%>%filter(Unit%in%units & Year=='Year 2')%>%group_by(Unit,schoolid2,cust,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+
+unitLevelSch$perWorked <- unitLevelSch$numWorked/nstudSch$nstud[match(unitLevelSch$schoolid2,nstudSch$schoolid2)]
+
+unitLevelSch$Unit <- factor(unitLevelSch$Unit,levels=units)
+levels(unitLevelSch$Unit) <- UnitName
+
+ggplot(filter(unitLevelSch,state!='NJ'),aes(x=Unit,y=perWorked,color=schoolid2,group=schoolid2,linetype=cust))+geom_point(size=.5)+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10))+
+                                        #,legend.position=c(.96,.82))+
+    facet_grid(state~.)+scale_y_continuous(labels=percent)+scale_color_discrete(guide=FALSE)+
+    labs(x='',y='% Worked',linetype='Curriculum',title='Year 2')
+@
+\caption{The percentages of year-2 students with usage data in each
+  school who worked at least one problem from each unit in the Algebra I curriculum. The
+  units are arranged in order for the standard curriculum. Schools are
+  classified as either using primarily customized curricula (solid
+  line) or using primarily the standard Algebra I curriculum (dotted).}
+\label{fig:unitsBySchool}
+\end{figure}
+
+Figure \ref{fig:unitsBySchool} further decomposes the year-2 results
+by school and state, showing a large amount of variation between
+states, as well as variation between schools within states.
+In Texas, every school used customized curricula, most of which seem to
+prioritize some of the same units, for instance, ``Linear Patterns,''
+``Ind. Variables in Lin. Mod.,'' and
+``Lin. Eq. w/ Var's Both Side.''
+These topics might be emphasized on Texas standardized tests.
+On the other hand, there was also variance between schools.
+For instance, one school prioritized units ``2 Step Lin. Eq.''
+and ``4−Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''
+
+Between-school variation is evident in the other states, as well.
+In four of the five Kentucky schools, nearly every student worked on
+the first nine units; in the one Kentucky school that used a
+customized curriculum, nearly every student worked on the first 13
+units, omitted the 15th (``Lin. Mod. in General Form''), and worked on
+the 16th and 17th (``Literal Eq.'' and ``Lin. Eq. w/ Var's Both Sides'').
+In the remaining school, nearly every student worked on the first
+section, but usage decreased rapidly from there.
+Curiously, in one Michigan school which used the standard curriculum,
+no students seem to have worked on the ``Lin. Mod. \& Ratios'' section.
+
+If unit order and topic scaffolding are important to CT's mastery
+learning mechanism, the wide variation in students' realized curricula
+would seem to pose a problem.
+The fact that the prescribed order was followed less in the second
+year of the study, when CTAI was effective, than in the first year,
+when it wasn't, suggests that the standard curriculum may play a
+smaller role than one might otherwise imagine.
+
+\section{Mastering the Material---Or Not}
+The central idea behind mastery learning is that students progress
+through the curriculum as they master skills.
+In the context of CT, skills are clustered within sections, which are
+in turn clustered within units.
+Students progress from the current section to the next section after
+mastering all of the current section's skills.
+Ideally, students would master all of the skills in all of the
+sections they work.
+
+By default, the software operates by automatically moving students
+from section to section based on the sequence of topics defined by the
+curriculum they were currently enrolled in. In this
+software-controlled sequencing, students ideally spend the time
+necessary to learn the material of a section, are judged by the
+software to have mastered the material, and then ``graduate'' to the
+next section. Students who exhaust a section's material without
+mastering its skills are “promoted” to the next section.
+Students left uninterrupted follow the sequence of topics defined by
+their curriculum, ideally graduating from topics by showing mastery.
+However, teachers could modify a student’s path within the curriculum.
+They could ``reassign'' students from their current sections to other
+sections earlier or later in the intended sequence, including sections
+they had worked on previously.
+Finally, if the semester ends, or a student stops using CT for some
+other reason, while in the middle of working through a
+section, he or she will leave their ``Final'' section without
+mastering all of its skills.
+All in all, each CT section begun ends in one of four possible ways:
+mastery, promotion, reassignment, or as the student's final section.
+
+\begin{figure}
+  \centering
+<<overallStatus,dependson=c('data','setup'),fig.height=3,fig.width=6>>=
+statusPerSec <- data%>%
+    filter(!is.na(Curriculum) &!(state=='MI'&Curriculum=='Customized'&year==1))%>%
+    group_by(field_id,Yr,state,section,Curriculum,status)%>%
+        summarize(cp=any(status=='changed placement'))%>%group_by(state,Curriculum,Yr,field_id)%>%summarise(pcp=mean(cp,na.rm=TRUE),ncp=sum(cp,na.rm=TRUE))
+
+statusOverall <-  data%>%filter(!is.na(status))%>%
+    group_by(field_id,Yr,state,unit,section,Curriculum,overall)%>%summarize(status=max(status))
+
+statusStateYr <- addOverallState(statusOverall) %>% group_by(state,Yr)%>%
+    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
+              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
+    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
+
+levels(statusStateYr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
+
+ ggplot(filter(statusStateYr,state!='NJ'),aes(Yr,value,fill=variable))+geom_col()+facet_grid(~state)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent)
+
+@
+\caption{The distributions of outcomes of worked sections, by state and across the
+  entire sample, in the two study years.}
+\label{fig:overallStatus}
+\end{figure}
+
+\begin{table}
+ \begin{tabular}{rllllll|llllll}
+
+&   \multicolumn{6}{c}{Year 1}&\multicolumn{6}{c}{Year 2}\\
+<<statusTab,dependson='overallStatus',results='asis'>>=
+
+tab1 <- dcast(subset(statusStateYr,Yr=='Yr 1'&state!='NJ'),variable~state)
+tab2 <- dcast(subset(statusStateYr,Yr=='Yr 2'&state!='NJ'),variable~state)
+tab <- cbind(tab1,tab2[-1])
+
+tab$variable <- as.character(tab$variable)
+cat('&')
+cat(names(tab)[-1],sep='&')
+cat('\\\\')
+for(i in 1:nrow(tab)){
+    cat(tab[i,1],round(unlist(tab[i,-1])*100),sep='&')
+    cat('\\\\')
+}
+
+rownames(tab1) <- rownames(tab2) <- tab1$variable
+tab1 <- as.matrix(tab1[,-1])
+tab2 <- as.matrix(tab2[,-1])
+
+@
+
+\end{tabular}
+\caption{Percentages of worked sections that ended in each of the
+  four possible outcomes, across states and study years.}
+\label{tab:overallStatus}
+\end{table}
+
+Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
+the proportions of worked sections in each state and study year that
+ended with mastery, promotion, or reassignment, or which were the
+student's final section.
+In the first year, about
+\Sexpr{round(mean(tab1['Mastered',c('TX','KY','MI','LA')])*100)}\% of
+worked sections are mastered in every
+state other than Connecticut.
+Other than in Texas, about
+\Sexpr{paste(round(range(tab1['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
+of sections end in promotion.
+About \Sexpr{round(tab1['Reassigned','TX']*100)}
+and \Sexpr{round(tab1['Reassigned','CT']*100)}\% of sections in Texas
+and Connecticut end in reassignment, which is even rarer in the other states.
+
+With the exception of Texas, sections in the second year tend to be
+completed as they were in year 1.
+In Texas, however, the percentage of sections ending in reassignment increased by a factor
+of about about
+\Sexpr{round(tab2['Reassigned','TX']/tab1['Reassigned','TX'])},
+to about \Sexpr{round(tab2['Reassigned','TX']*100)}\%.
+The proportion of Texas sections labeled ``Final'' increased as well,
+most likely due to the decrease in usage.
+
+Across states, just under
+\Sexpr{round(tab1['Reassigned','Overall']*100)}\% of sections in year
+1 eneded in reassignment, as did roughly
+\Sexpr{round(tab2['Reassigned','Overall']*100)}\% in year 2.
+
+\subsection{Section Mastery and Curriculum}
+\begin{figure}
+<<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
+statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curriculum,overall)%>%
+    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
+              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
+    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
+levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
+statusCurr$Curriculum <- factor(statusCurr$Curriculum,levels=rev(levels(statusCurr$Curriculum)))
+print(ggplot(statusCurr,aes(overall,value,fill=variable))+geom_col()+facet_grid(~Curriculum)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))
+
+@
+\caption{The distributions of outcomes of worked sections, by curriculum, in the two study years.}
+\label{fig:statusCur}
+\end{figure}
+
+\section{Digging Deeper into Section Reassignment}
+<<cpDat,dependson=c('data','setup'),include=FALSE>>=
+secLev <- data%>%filter(is.finite(status) & is.finite(timestamp) &is.finite(date))%>%
+    group_by(field_id,unit,section,Year,Yr,classid2,schoolid2)%>%
+    summarize(startDate=min(date),endDate=max(date),startTime=min(timestamp),endTime=max(timestamp),state=state[1],
+              status=max(status),Curriculum=Curriculum[1],overall=overall[1],version=version[1])%>%
+    arrange(endDate)
+
+### cp over time
+secLev <- within(secLev,endMonth <- factor(month(secLev$endDate,TRUE,TRUE),levels=c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr','May','Jun','Jul')))
+
+@
+\subsection{When are Students Reassigned?}
+
+\begin{figure}
+  \centering
+<<byMonth,dependson='cpDat',fig.height=3,fig.width=6>>=
+secLevMonth <- secLev%>%group_by(endMonth,Year)%>%summarize(ncp=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
+secLev$cp <- as.numeric(secLev$status=='changed placement')
+
+ggplot(filter(secLevMonth,ncp>100),aes(endMonth,CPper,group=Year,color=Year,size=ncp))+geom_point()+geom_line(size=1)+
+    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-0.5,cp,group=Year,
+    ##                 color=Year,size=1),
+    ##             method = "glm", formula = y ~ splines::bs(x, 4),data=secLev,method.args=list(family='binomial'),
+                                        #            show.legend=FALSE)+
+scale_y_continuous(breaks=seq(0,0.05,.01),labels=percent)+
+    coord_cartesian(ylim=c(0,0.05))+ labs(x='Month',y='% of Sections Ending in Reassignment')
+
+@
+\end{figure}
+
+\begin{figure}
+  \centering
+<<byMonthState,dependson='cpDat',fig.height=3,fig.width=6>>=
+secLevMonthSt <- secLev%>%filter(state%in%c('TX','KY','MI') & endMonth!='Jul')%>%group_by(endMonth,state,Year)%>%summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
+ggplot(secLevMonthSt,aes(endMonth,CPper,group=state,color=state,size=nsec))+geom_point()+facet_grid(Year~.)+
+    geom_line(size=1)+
+    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-1.5,cp,group=state,
+    ##                 color=state,size=1),
+    ##             method = "glm",
+    ##             formula = y ~ splines::ns(x, knots=seq(1,10,2)),
+    ##             data=filter(secLev,state%in%c('TX','CT','NJ')& endMonth!='Jul'),
+    ##             method.args=list(family='binomial'),
+    ##             show.legend=FALSE)+
+    scale_y_continuous(labels=percent)+
+    labs(x='Month',y='% Sections Ending in Reassignment',size='Worked \n Sections',color='')
+
+@
+\end{figure}
+
+\subsection{Does Reassignment Depend on Classmates?}
+\begin{figure}
+<<classmates,dependson='cpDat',fig.height=6,fig.width=7>>=
+secLev$unitSectionClass <- paste(secLev$unit,secLev$section,secLev$classid2)
+cpDat <- subset(secLev,status=='changed placement' & Curriculum=='Algebra I')
+cpClasses <- unique(cpDat$unitSectionClass)
+secLevCP <- filter(secLev,unitSectionClass%in%cpClasses)
+levels(secLevCP$status) <- list(promGrad=c('promoted','graduated','final_or_incomplete'),cp='changed placement')
+secLevCPsplit <- split(secLevCP[,c('field_id','endDate','status')],secLevCP$unitSectionClass)
+
+ncp <- nrow(cpDat)
+seqDat <- matrix(NA,nrow=ncp,ncol=6)
+for(i in 1:ncp){
+    if(i%%100==0) cat(round(i/ncp*100,2),'% ',sep='')
+    cls <- secLevCPsplit[[cpDat$unitSectionClass[i]]]
+    if(nrow(cls)==1){
+        seqDat[i,] <- 0
+        next
+    }
+    cls <- filter(cls,field_id!=cpDat$field_id[i])
+    date <- cpDat$endDate[i]
+    ord <- factor(ifelse(cls$endDate<date,'bf',
+                  ifelse(cls$endDate==date,'same','after')),levels=c('bf','same','after'))
+    clsSplit <- split(cls,list(ord,cls$status))
+    seqDat[i,] <- vapply(clsSplit,nrow,1)
+}
+cat('\n')
+colnames(seqDat) <- names(clsSplit)
+
+classSize <- data%>%filter(!is.na(section))%>%group_by(classid2)%>%summarize(nstud=n_distinct(field_id))
+cpDat$totClassmates <- classSize$nstud[match(cpDat$classid2,classSize$classid2)]
+
+for(nn in names(clsSplit)) cpDat[[nn]] <- seqDat[,nn]/cpDat$totClassmates
+cpDat$nw <- (cpDat$totClassmates-rowSums(seqDat))/cpDat$totClassmates
+cpDat$total <- rowSums(seqDat)
+cpDat$bf.promGrad <- cpDat$bf.promGrad+cpDat$same.promGrad
+
+### plot them all
+
+mmm <- with(subset(cpDat,totClassmates>15),max(table(state,Year)))
+
+cpDat2 <- cpDat%>%filter(totClassmates>15)%>%arrange(bf.promGrad,bf.cp,same.cp,after.cp,after.promGrad)%>%group_by(Year,state)%>%mutate(cpid=seq(floor((mmm-n())/2)+1,floor((mmm-n())/2)+n()))%>%ungroup()
+
+aaa <- melt(cpDat2,measure.vars=c('bf.promGrad','bf.cp','same.cp','after.cp','after.promGrad','nw'))
+levels(aaa$variable) <- list(
+    `Never Worked\n Section`='nw',
+    `Mastered/Promoted\n Later`='after.promGrad',
+    `Reassigned\n Later`='after.cp',
+    `Reassigned\n Same Day`='same.cp',
+    `Reassigned\n Before`='bf.cp',
+    `Mastered/Promoted\n Before`='bf.promGrad')
+
+#aaa$variable <- factor(aaa$variable,levels=rev(c('pGrad','pProm','pCPbf','pCPsame','pLater','pNW')))
+ggplot(filter(aaa,state%in%c('TX','KY','MI')),aes(as.factor(cpid),value,fill=variable))+geom_col()+facet_grid(Year~state)+theme(axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank()) + scale_y_continuous(label=percent)+labs(fill='% of Classmates')+
+    scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02'))
+
+@
+\end{figure}
+
+\subsection{Where To?}
+\begin{figure}
+<<transition,fig.height=5,fig.width=6,dependson=c('data','setup')>>=
+#### what's the next section each student works?
+secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]))
+
+secOrderCP <- subset(secOrder,status=='changed placement')
+
+secOrderCP$Unit <- secOrderCP$unit
+secOrderCP$Unit[grep('unit-conversions',secOrderCP$Unit)] <- 'unit-conversions'
+
+secOrderCP$unitName <- factor(secOrderCP$Unit,levels=units)
+levels(secOrderCP$unitName) <- UnitName
+
+secOrderCP$nextUnit[grep('unit-conversions',secOrderCP$nextUnit)] <- 'unit-conversions'
+
+secOrderCP$nextUnitName <- factor(secOrderCP$nextUnit,levels=units)
+levels(secOrderCP$nextUnitName) <- UnitName
+
+trans <- with(secOrderCP,table(unitName,nextUnitName))
+
+trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
+trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
+
+transitionPlot(trans)
+
+nsec <- nrow(trans)
+
+plot(1,1,xlim=c(0,6),ylim=c(0,nsec),cex=0)
+text(rep(1,nsec),nsec:1,rownames(trans))
+text(rep(5,nsec),nsec:1,rownames(trans))
+for(i in 1:nsec){
+    arrows(rep(2.5,nsec),rep(nsec-i+1,nsec),rep(3.5,nsec),nsec:1,lwd=5*trans[i,]/rowSums(trans)[i])#,col=rgb(0,0,0,trans[i,]/max(trans)))
+}
+
+@
+\end{figure}
+\section{Notes}
+[[don't forget to include these]]
+
+\begin{itemize}
+  \item \# of students per state?
+  \item NJ excluded from state-by-state comparisons, cuz too
+    small. Included in ``overall,'' etc.
+  \item What schools included?
+  \item Students with no mastery data: maybe they didn't work at all,
+    maybe they worked but we don't have their data. Treating them as unknown.
+\end{itemize}
+
 \end{document}

### 2017-10-10 fig1.r

@@ -254,8 +254,8 @@ ggplot(unitLevelState,aes(x=Unit,y=perWorked,color=state,group=state))+geom_poin
 cust <- data%>%filter(Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(cust=mean(overall=='Customized',na.rm=T))%>%arrange(cust)
 data$cust <- data$schoolid2%in%cust$schoolid2[cust$cust>0.8]
 
-nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2' & state=='TX')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
-unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2'&state=='TX' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
+nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
+unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
 
 unitLevelCust$perWorked <- unitLevelCust$numWorked/nstudCust$nstud[match(unitLevelCust$cust,nstudCust$cust)]
 
@@ -265,14 +265,21 @@ ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()
 
 ### by school
 nstudSch <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(nstud=n_distinct(field_id))
-unitLevelSch <- data%>%filter(Unit%in%units & Year=='Year 2')%>%group_by(Unit,schoolid2,cust,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
[…]

### 2017-10-10 paper.Rnw
fixed bugs

@@
 #### what's the next section each student works?
-secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]))
+secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
+
+secOrder$cp <- secOrder$status=='changed placement'
+secOrder$mast <- secOrder$status=='graduated'
 
@@
 
-transitionPlot(trans)
+#transitionPlot(trans)
 

### 2017-10-16 paper.Rnw
models galore!

@@
 \begin{figure}
-<<transition,fig.height=5,fig.width=6,dependson=c('data','setup')>>=
+<<transition,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
 #### what's the next section each student works?
@@
 
-trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
-trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
 
@@
 nsec <- nrow(trans)
+uuu <- levels(secOrderCP$unitName)
+uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
+nsec <- length(uuu)
 
-plot(1,1,xlim=c(0,6),ylim=c(0,nsec),cex=0)
-text(rep(1,nsec),nsec:1,rownames(trans))
-text(rep(5,nsec),nsec:1,rownames(trans))
-for(i in 1:nsec){
-    arrows(rep(2.5,nsec),rep(nsec-i+1,nsec),rep(3.5,nsec),nsec:1,lwd=5*trans[i,]/rowSums(trans)[i])#,col=rgb(0,0,0,trans[i,]/max(trans)))
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(-.5,4),ylim=c(0,nsec),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+
+for(rr in rownames(trans))
+ for(cc in colnames(trans))
+     if(trans[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>100,1,trans[rr,cc]/100)))
 }
@@
 \end{figure}
+
+\begin{figure}
+<<vcs,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
+load('vcMods.RData')
+library(lme4)
+
+vcFun <- function(nm){
+    mod <- vcMods[[nm]]
+    out <- unlist(summary(mod)$varcor)
+    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
+    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
+    out$state <- strsplit(nm,'_')[[1]][1]
+    out$year <- strsplit(nm,'_')[[1]][2]
+    out$sig2 <- out$sig2/sum(out$sig2)
+    out
+}
+
+vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
+
+yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
+                          state='Overall',
+                  year='1')
+yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
+
+yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
+                          state='Overall',
+                  year='2')
+yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
+
+vcDat <- rbind(vcDat,yr1,yr2)
+
+vcDat$comp <- factor(vcDat$comp)
+levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
+
+vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
+
+ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')
+
+@
+\end{figure}
+
+\begin{figure}
+<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
+load('cpXirtMods.RData')
+
+xirtDat <- data%>%filter(!is.na(status) & !is.na(xirt))%>%
+    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state,unit,section)%>%
+    summarize(status=max(status,na.rm=TRUE),nprob=n())%>%
+    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state)%>%
+    summarize(nprob=sum(nprob),nsec=n_distinct(section,unit),propMast=mean(status=='graduated'),
+              propCP=mean(status=='changed placement'))%>%arrange(state)
+
+xirtDat$fitted <- NA
+for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
+                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
+                                          allow.new.levels=TRUE,re.form=~0,type='response')
+
+xirtDat <- addOverallState(xirtDat)
+
+xirtDat$fitted[xirtDat$state=='Overall'] <-
+    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
+            allow.new.levels=TRUE,re.form=~0,type='response')
+
+ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red')+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))
+
+@
+\end{figure}
+
 \section{Notes}

### 2017-10-16 fig1.r

@@ -673,3 +673,95 @@ text(rep(5,nsec),nsec:1,rownames(trans))
 for(i in 1:nsec){
     arrows(rep(2.5,nsec),rep(nsec-i+1,nsec),rep(3.5,nsec),nsec:1,lwd=5*trans[i,]/rowSums(trans)[i])#,col=rgb(0,0,0,trans[i,]/max(trans)))
 }
+
+
+
+###############################
+#### CP and xirt
+##############################
+
+xirtDat <- data%>%filter(!is.na(status) & !is.na(xirt))%>%
+    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state,unit,section)%>%
+    summarize(status=max(status,na.rm=TRUE),nprob=n())%>%
+    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state)%>%
+    summarize(nprob=sum(nprob),nsec=n_distinct(section,unit),propMast=mean(status=='graduated'),
+              propCP=mean(status=='changed placement'))%>%arrange(state)
+
+
+
+#### prop mastered vs xirt
+ggplot(xirtDat,aes(scale(xirt),propMast,size=nsec))+geom_point()+geom_smooth(size=1)+facet_grid(year~.)
+
+ggplot(xirtDat,aes(scale(xirt),propMast,size=nsec))+geom_point()+geom_smooth(size=1)+facet_grid(year~state)
+
+
+
+### prop cp vs xirt
+ggplot(subset(xirtDat,classid2%in%unique(data$classid2[data$status=='changed placement'])),aes(scale(xirt),propCP,size=nsec))+geom_point()+geom_smooth(size=1)+facet_grid(year~.)
+
+ggplot(subset(xirtDat,classid2%in%unique(data$classid2[data$status=='changed placement'])),aes(scale(xirt),propCP,size=nsec))+geom_point()+geom_smooth(size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.25))
+
+### by classroom
+bbb2 <- data%>%filter(status=='changed placement')%>%group_by(classid2,year,state)%>%summarize(nstud=n_distinct(field_id),nsec=n_distinct(section))
+
+ggplot(subset(xirtDat,classid2%in%bbb2$classid2[bbb2$year==1 & bbb2$nstud>=10]),aes(scale(xirt),propCP,size=nsec,color=state))+
[…]

### 2017-10-16 models.r

@@ -10,7 +10,13 @@ cpDat <- merge(cpDat,stud,all.x=TRUE,all.y=FALSE)
 
 cpDat$month <- months(cpDat$startDate,TRUE)
 
+cpDat$cp <- cpDat$status=='changed placement'
+cpDat$mast <- cpDat$status=='graduated'
+
+
+
 ### take out classrooms, units, with no CPs
+cpDatFull <- cpDat
 cpDat <- subset(cpDat,unit %in% unique(cpDat$unit[cpDat$status=='changed placement']))
 cpDat <- subset(cpDat, classid2 %in% unique(cpDat$classid2[cpDat$status=='changed placement']))
 cpDat <- subset(cpDat,unit %in% unique(cpDat$unit[cpDat$status=='changed placement']))
@@ -18,11 +24,100 @@ cpDat <- subset(cpDat, classid2 %in% unique(cpDat$classid2[cpDat$status=='change
 
 cpDat$late <- cpDat$month%in%c('Jan','Feb','March','April')
 
-cpDat$cp <- cpDat$status=='changed placement'
 
 cpMod <- glmer(cp~year*(xirt+late)+spec_gifted+spec_speced+spec_esl+state+(xirt+late|classid2)+(1|field_id)+(late|unit),data=cpDat,family=binomial)
 
+cpModSimp <- glmer(cp~ year*ns(xirt,5)+(1|field_id),data=cpDat,family=binomial)
+
+cpModSt <- list()
+for(st in c('TX','KY','MI','LA','CT')){
+    cpModSt[[st]] <- glmer(cp~ year*ns(xirt,5)+(1|field_id),data=subset(cpDat,state==st),family=binomial)
+}
+save(cpModSt,cpModSimp,file='cpXirtMods.RData')
+
+plot(cpDat$xirt,predict(cpModSimp,cpDat,allow.new.levels=TRUE,re.form=~0,type='response'),col=cpDat$year)
+legend('topleft',legend=c('year 1','year 2'),col=1:2,pch=1)
+
+### mastMod
+mastMod <- glmer(mast ~ year * (ns(xirt, 5) + late) + spec_gifted + spec_speced +
+    spec_esl + state + (ns(xirt, 5) + late | classid2) + (1 |
+    field_id) + (late | unit),data=cpDat,family=binomial)
+
[…]

### 2017-10-16 paper.Rnw
tweaks

@@
 
-ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')
+ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
 
@@
 
-ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red')+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))
+ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))
 

### 2017-10-17 cpEffect.r
model for effect of cp on post-test
doesn't seem to be working but still...

@@ -1,0 +1,75 @@
+library(dplyr)
+library(lme4)
+library(splines)
+load('cpPaper.RData')
+
+cpDat <- adv%>%filter(field_id%in%data$field_id)
+cpDat <- merge(cpDat,stud,all.x=TRUE,all.y=FALSE)
+
+cpDat$cp <- cpDat$status=='changed placement'
+cpDat$mast <- cpDat$status=='graduated'
+
+
+
+### take out classrooms, sections, with no CPs
+cpDatFull <- cpDat
+cpDat <- subset(cpDat,section %in% unique(cpDat$section[cpDat$status=='changed placement']))
+cpDat <- subset(cpDat, classid2 %in% unique(cpDat$classid2[cpDat$status=='changed placement']))
+
+
+### post-test scores
+### do this one in jags
+library(R2jags)
+
+cpDat <- cpDat[order(cpDat$field_id),]
+cpDat <- cpDat[!is.na(cpDat$section) & !is.na(cpDat$cp),]
+
+cpDat <- droplevels(cpDat)
+
+jagsDat <- list()
+jagsDat$cp <- cpDat$cp
+jagsDat$stud <- as.numeric(as.factor(cpDat$field_id))
+jagsDat$section <- as.numeric(as.factor(cpDat$section))
+
+
+jagsDat$nWS <- nrow(cpDat)
+
+
+### want imputed xirt
+h1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
+h2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')
+
+h2 <- h2[!h2$field_id%in%h1$field_id,]
+
+stud <- rbind(h1[,c(2:56,174:194)],h2[,c(2:56,176:196)])
+stud <- stud[stud$field_id%in%cpDat$field_id,]
+stud <- stud[order(stud$field_id),]
+stud$pretest <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)
+## mode imputation:
+for(vv in c('race','sex','grade','spec_speced','spec_gifted','spec_esl','frl')){
+    stud[[vv]][is.na(stud[[vv]])] <- names(which.max(table(stud[[vv]])))
+}
+stud <- droplevels(stud)
+
+
+jagsDat$nWS <- nrow(cpDat)
[…]

### 2017-10-17 paper.Rnw
movin right along

@@
 <<statusTab,dependson='overallStatus',results='asis'>>=
-
-tab1 <- dcast(subset(statusStateYr,Yr=='Yr 1'&state!='NJ'),variable~state)
-tab2 <- dcast(subset(statusStateYr,Yr=='Yr 2'&state!='NJ'),variable~state)
+statusStateYr2 <- addOverallState(statusOverall)
+statusStateYr2a <- statusStateYr2%>%group_by(state,Yr)%>%
+    summarize(pgrad=sum(status=='graduated'),pfoi=sum(status=='final_or_incomplete'),
+              pcp=sum(status=='changed placement'),pprom=sum(status=='promoted'))%>%
+    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
+statusStateYr2b <- statusStateYr2%>%
+    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
+              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
+    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
+
+levels(statusStateYr2a$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
+levels(statusStateYr2b$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
+
+tab1 <- dcast(subset(statusStateYr2a,Yr=='Yr 1'&state!='NJ'),variable~state)
+tab2 <- dcast(subset(statusStateYr2a,Yr=='Yr 2'&state!='NJ'),variable~state)
 tab <- cbind(tab1,tab2[-1])
+
+tab1b <- dcast(subset(statusStateYr2b,Yr=='Yr 1'&state!='NJ'),variable~state)
+tab2b <- dcast(subset(statusStateYr2b,Yr=='Yr 2'&state!='NJ'),variable~state)
 
@@
 for(i in 1:nrow(tab)){
-    cat(tab[i,1],round(unlist(tab[i,-1])*100),sep='&')
+    cat(tab[i,1],round(unlist(tab[i,-1])),sep='&')
     cat('\\\\')
@@
 tab2 <- as.matrix(tab2[,-1])
+
+rownames(tab1b) <- rownames(tab2b) <- tab1b$variable
+tab1b <- as.matrix(tab1b[,-1])
+tab2b <- as.matrix(tab2b[,-1])
 
@@
 In the first year, about
-\Sexpr{round(mean(tab1['Mastered',c('TX','KY','MI','LA')])*100)}\% of
+\Sexpr{round(mean(tab1b['Mastered',c('TX','KY','MI','LA')])*100)}\% of
 worked sections are mastered in every
@@
 Other than in Texas, about
-\Sexpr{paste(round(range(tab1['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
+\Sexpr{paste(round(range(tab1b['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
 of sections end in promotion.
-About \Sexpr{round(tab1['Reassigned','TX']*100)}
-and \Sexpr{round(tab1['Reassigned','CT']*100)}\% of sections in Texas
+About \Sexpr{round(tab1b['Reassigned','TX']*100)}
+and \Sexpr{round(tab1b['Reassigned','CT']*100)}\% of sections in Texas
 and Connecticut end in reassignment, which is even rarer in the other states.
@@
 of about about
-\Sexpr{round(tab2['Reassigned','TX']/tab1['Reassigned','TX'])},
-to about \Sexpr{round(tab2['Reassigned','TX']*100)}\%.
+\Sexpr{round(tab2b['Reassigned','TX']/tab1b['Reassigned','TX'])},
+to about \Sexpr{round(tab2b['Reassigned','TX']*100)}\%.
 The proportion of Texas sections labeled ``Final'' increased as well,
@@
 Across states, just under
-\Sexpr{round(tab1['Reassigned','Overall']*100)}\% of sections in year
+\Sexpr{round(tab1b['Reassigned','Overall']*100)}\% of sections in year
 1 eneded in reassignment, as did roughly
-\Sexpr{round(tab2['Reassigned','Overall']*100)}\% in year 2.
+\Sexpr{round(tab2b['Reassigned','Overall']*100)}\% in year 2.
 
@@
 <<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
-statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curriculum,overall)%>%
+statusOverall$Curr2 <- with(statusOverall,
+                        ifelse(Curriculum=='Algebra I',
+                                     ifelse(overall=='Standard','Algebra I','Algebra I (Cust.)'),
+                                     as.character(Curriculum)))
+statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curr2,Yr)%>%
     summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
@@
 levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
-statusCurr$Curriculum <- factor(statusCurr$Curriculum,levels=rev(levels(statusCurr$Curriculum)))
-print(ggplot(statusCurr,aes(overall,value,fill=variable))+geom_col()+facet_grid(~Curriculum)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))
-
-@
-\caption{The distributions of outcomes of worked sections, by curriculum, in the two study years.}
+statusCurr$Curr2 <- factor(statusCurr$Curr2,levels=c('Bridge-to-Algebra','Algebra I','Algebra I (Cust.)','>Algebra I'))
+print(ggplot(statusCurr,aes(Yr,value,fill=variable))+geom_col()+facet_grid(~Curr2)+
+      labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))
+
+@
+\caption{The distributions of outcomes of worked sections, by
+  curriculum, in the two study years. (There were no Bridge-to-Algebra
+sections in customized curricula in our dataset.)}
 \label{fig:statusCur}
@@
 
+A well-designed curriculum, can, in theory, play an important role in
+students' attainment of mastery.
+Students who work on appropriate problems that build on their current
+set of skills should be more likely to master new skills than students
+working on problems above their level.
+What role did curriculum play in mastery during the CTAI effectiveness trial?
+
+Figure \ref{fig:statusCur} shows the proportions of worked sections
+that were mastered or ended in promotion, reassignment, or finality, in
+standard and customized versions of each CT curriculum.
+Mastery proportions do, indeed, depend on curriculum.
+Specifically, students mastered sections from more advanced curriculua
+less frequently.
+Sections from the most basic curriculum, Bridge to Algebra, were
+mastered
+\Sexpr{round(statusCurr[['value']][statusCurr[['Curriculum']]=='Bridge-to-Algebra'
+  & statusCurr[['variable']]=='Graduated']*100)}\% of the time;
+those from Algebra I were mastered
+\Sexpr{round(
+  mean(statusOverall$status[statusOverall$Curriculum=='Algebra I']=='graduated',na.rm=T)*100)}\%
+of the time, and those from more advanced curriculua were mastered
+at a rate of
+\Sexpr{round(
+  mean(statusOverall$status[statusOverall$Curriculum=='>Algebra I']=='graduated',na.rm=T)*100)}\%.
+This is unsurprising, since more advanced curriculua may be expected
+to be more challenging.
+However, it may suggest that some students studying advanced topics
+would fair better in more standard curricula.
+
+Algebra I sections from customized curricula tended to end in
+reassignment more often than sections from the standard Algebra I
+curriculum (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &
+  variable=='Reassigned')[['value']])}\% vs.
+\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Reassigned')[['value']])}\%).
+This may an overall skepticism towards the Carnegie Learning standards
+among certain schools and teachers.
+
+% A somewhat lower proportion of Algebra I sections from customized
+% curricula were mastered
+% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])}\%)
+% than from the standard Algebra I curriculum
+% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])}\%).
+% This difference must be interpreted with caution, since a greater
+% proportion of sections of the customized curriculum ended in
+% reassignment
+
+% or were final
+% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])}\% vs.
+% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])}\%)
+% , as well.
+% The proportion of sections ending in promotion was almost identical:
+% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])}\% vs.
+% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])}\%).
+
 \section{Digging Deeper into Section Reassignment}
+The proportion of worked sections in our dataset ending in
+reassignment was small.
+Nevertheless, since reassignement represents the only mechanism by
+which individual teachers can affect their students' progress through
+the Cognitive Tutor, exploring patterns of reassignment can provide
+insight into how CT was used.
+
 <<cpDat,dependson=c('data','setup'),include=FALSE>>=
@@
 @
+
+\subsection{How Do Reassignment Patterns Vary?}
+Teachers alone control reassignment.
+Nevertheless, the factors influencing student reassignment vary at a
+number of levels.
+For instance, state and district standards may prod teachers into
+reassigning students to particular units.
+Some prinicpals may encourage teachers to adhere to the official
+curriculum and avoid reassignment.
+Some students may be more prone to reassignment than others.
+Certain units in the CTAI curriculum may be harder than others,
+causing students to tarry and teachers to reassign.
+Finally, a host of other factors, at these levels and others, may spur reassignment.
+
+Figure \ref{fig:vc}
+
+\begin{figure}
+<<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
+load('vcMods.RData')
+library(lme4)
+
+vcFun <- function(nm){
+    mod <- vcMods[[nm]]
+    out <- unlist(summary(mod)$varcor)
+    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
+    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
+    out$state <- strsplit(nm,'_')[[1]][1]
+    out$year <- strsplit(nm,'_')[[1]][2]
+    out$sig2 <- out$sig2/sum(out$sig2)
+    out
+}
+
+vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
+
+yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
+                          state='Overall',
+                  year='1')
+yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
+
+yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
+                          state='Overall',
+                  year='2')
+yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
+
+vcDat <- rbind(vcDat,yr1,yr2)
+
+vcDat$comp <- factor(vcDat$comp)
+levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
+
+vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
+
+ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
+
+@
+\caption{Results from a set of eight multilevel logistic regressions
+  predicting section reassignment. For each year, in the entire sample (``Overall'')
+  and in the three states with the largest numbers of reassignments (Texas,
+  Kentucky, and Michigan), we regressed a binary variable indicating
+  whether a section ended in reassignment on random intercepts for
+  school, class, student, and unit, and in the overall case, for state
+  as well, and recorded their variance. The residual variance was set
+  as the variance of the standard logistic distribution,
+  $\pi/3$. These bar charts give the proportion of the total variance
+  attributable to each random effect.}
+\label{fig:vc}
+\end{figure}
+
 \subsection{When are Students Reassigned?}
@@
 \subsection{Where To?}
-\begin{figure}
-<<transition,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
+
+<<secOrder,include=FALSE,dependson=c('data','setup')>>=
 #### what's the next section each student works?
@@
 
-trans <- with(secOrderCP,table(unitName,nextUnitName))
+@
+
+\begin{figure}
+<<transition1,fig.height=3,fig.width=6.5,dependson='secOrder'>>=
+
+trans <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
 
@@
 
+propTotCP <- sum(trans)/sum(secOrderCP$year==1)
 #transitionPlot(trans)
@@
 \begin{figure}
-<<vcs,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
-load('vcMods.RData')
-library(lme4)
-
-vcFun <- function(nm){
-    mod <- vcMods[[nm]]
-    out <- unlist(summary(mod)$varcor)
-    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
-    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
-    out$state <- strsplit(nm,'_')[[1]][1]
-    out$year <- strsplit(nm,'_')[[1]][2]
-    out$sig2 <- out$sig2/sum(out$sig2)
-    out
+<<transition2,fig.height=7,fig.width=6.5,dependson='secOrder'>>=
+
+trans <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
+
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+
+propTotCP <- sum(trans)/sum(secOrderCP$year==2)
+
+#transitionPlot(trans)
+
+nsec <- nrow(trans)
+uuu <- levels(secOrderCP$unitName)
+uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
+nsec <- length(uuu)
+
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(-.5,4),ylim=c(0,nsec),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+
+for(rr in rownames(trans))
+ for(cc in colnames(trans))
+     if(trans[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>100,1,trans[rr,cc]/100)))
 }
 
-vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
-
-yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
-                          state='Overall',
-                  year='1')
-yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
-
-yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
-                          state='Overall',
-                  year='2')
-yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
-
-vcDat <- rbind(vcDat,yr1,yr2)
-
-vcDat$comp <- factor(vcDat$comp)
-levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
-
-vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
-
-ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
-
-@
-\end{figure}
-
-\begin{figure}
-<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
-load('cpXirtMods.RData')
-
-xirtDat <- data%>%filter(!is.na(status) & !is.na(xirt))%>%
-    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state,unit,section)%>%
-    summarize(status=max(status,na.rm=TRUE),nprob=n())%>%
-    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state)%>%
-    summarize(nprob=sum(nprob),nsec=n_distinct(section,unit),propMast=mean(status=='graduated'),
-              propCP=mean(status=='changed placement'))%>%arrange(state)
-
-xirtDat$fitted <- NA
+@
+\end{figure}
+
+\subsection{Does Reassignment Depend on Student Ability?}
+
+<<xirtMods,include=FALSE,dependson=c('data','setup')>>=
+## don't worry about nesting, etc--we aren't reporting standard errors anyway
+cpDat <- data%>%filter(!is.na(status))%>%group_by(field_id,unit,section,state,year,classid2,schoolid2)%>%
+    summarize(cp=max(status)=='changed placement')
+
+### get imputed xirt
+##
+h1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
+h2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')
+
+h2 <- h2[!h2$field_id%in%h1$field_id,]
+
+stud <- rbind(h1[,c(174:194)],h2[,c(176:196)])
+stud <- stud[stud$field_id%in%cpDat$field_id,]
+stud <- stud[order(stud$field_id),]
+stud$xirt <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)
+
+cpDat$xirt <- stud$xirt[match(cpDat$field_id,stud$field_id)]
+
+cpModSimp <- glm(cp~splines::ns(xirt,5)*year,data=cpDat)
+
+cpModSt <- list()
+for(st in c('TX','KY','MI','LA','CT')){
+    cpModSt[[st]] <- update(cpModSimp,data=subset(cpDat,state==st))
+}
+
+@
+
+\begin{figure}
+<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup','xirtMods')>>=
+#load('cpXirtMods.RData')
+
+xirtDat <- cpDat%>%group_by(field_id,state,year,classid2,schoolid2,xirt)%>%
+    summarize(nsec=n_distinct(section,unit),propCP=mean(cp,na.rm=TRUE))
+
 for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
                                   predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
-                                          allow.new.levels=TRUE,re.form=~0,type='response')
+                                         type='response')
 
@@
     predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
-            allow.new.levels=TRUE,re.form=~0,type='response')
-
-ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))
+            type='response')
+
+ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.5))
+
+@
+\end{figure}
+
+\begin{figure}
+<<cp-xirtClassroom1,dependson='xirtCP',fig.height=7,fig.width=6.5>>=
+bbb2 <- data%>%filter(status=='changed placement')%>%group_by(classid2,year,state)%>%summarize(ncp=n_distinct(field_id,section,unit),nstud=n_distinct(field_id),nsec=n_distinct(section))
+
+cpcls1 <- xirtDat%>%
+    filter(classid2%in%bbb2$classid2[bbb2$year==1 & bbb2$ncp>=20] & state!='Overall')%>%
+    arrange(state)
+cpcls1$classid2 <- factor(cpcls1$classid2,levels=unique(cpcls1$classid2))
+
+cpcls1 <- cpcls1%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()
+
+ggplot(cpcls1,aes(xirt,propCP,size=nsec,color=state))+
+    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
+    ylim(0,1)
+
+@
+\end{figure}
+
+\begin{figure}
+<<cp-xirtClassroom2,dependson='cp-xirtClassroom1',fig.height=7,fig.width=6.5>>=
+
+cpcls2 <- xirtDat%>%
+    filter(classid2%in%bbb2$classid2[bbb2$year==2 & bbb2$ncp>=20] & state!='Overall')%>%
+    arrange(state)
+cpcls2$classid2 <- factor(cpcls2$classid2,levels=unique(cpcls2$classid2))
+
+cpcls2 <- cpcls2%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()
+
+ggplot(cpcls2,aes(xirt,propCP,size=nsec,color=state))+
+    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
+    ylim(0,1)
 

### 2017-10-17 cpEffectSimp.r

@@ -1,0 +1,47 @@
+library(dplyr)
+library(lme4)
+library(splines)
+load('cpPaper.RData')
+
+cpDat <- data%>%filter(!is.na(status)& status!='final_or_incomplete')%>%
+ group_by(field_id,year,state,section,unit,race,sex,grade,spec_speced,xirt,spec_gifted,
+  spec_esl,frl,pretest,y_yirt,classid2,schoolid2)%>%
+ summarize(status=max(status,na.rm=TRUE),date=max(date),totalTime=sum(total_t1,na.rm=TRUE),
+  ttNA=sum(is.na(total_t1)),nprob=n(),probNA=sum(is.na(Prob1)))%>%ungroup()%>%
+ group_by(field_id,year,state,race,sex,grade,spec_speced,xirt,spec_gifted,spec_esl,frl,pretest,y_yirt,
+  classid2,schoolid2)%>%
+ summarize(nsec=n(),nprob=sum(nprob,na.rm=TRUE),totalTime=sum(totalTime,na.rm=TRUE),
+  timeNA=sum(ttNA>0),probNA=sum(probNA>0),ncp=sum(status=='changed placement'),
+  nprom=sum(status=='promoted'))%>%mutate(gainScore=y_yirt-pretest)%>%droplevels()
+
+
+
+## mode imputation:
+for(vv in c('race','sex','grade','spec_speced','spec_gifted','spec_esl','frl')){
+    cpDat[[paste0(vv,'Imp')]] <- cpDat[[vv]]
+    cpDat[[paste0(vv,'Imp')]][is.na(cpDat[[vv]])] <- names(which.max(table(cpDat[[vv]])))
+}
+
+
+
+cpDat <- mutate(cpDat,totalTime=totalTime/3600000,totalTime=ifelse(totalTime<0,NA,totalTime),
+                totalTime=ifelse(totalTime>360,NA,totalTime))
+
+cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=3,'3+',cpDat$ncp))
+cpDat$everCP <- cpDat$ncp>0
+
+#### models!!
+mod1 <- lmer(gainScore~ncp+nsec+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
+mod1.1 <- update(mod1,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
[…]

### 2017-10-17 dataMerge.r

@@ -9,13 +9,16 @@ usageIDs <- unique(c(unique(x$field_id),unique(advance$field_id)))
 hs1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
 hs2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')
 
-stopifnot(all.equal(names(hs1)[1:100],names(hs2)[1:100]))
+hs2 <- hs2[!hs2$field_id%in%hs1$field_id,]
+stopifnot(all.equal(names(hs1)[c(2:100,174:194)],names(hs2)[c(2:100,176:196)]))
+stud <- rbind(hs1[,c(2:100,174:194)],hs2[,c(2:100,176:196)])
 
-stud <- rbind(hs1[,1:100],hs2[!hs2$field_id%in%hs1$field_id,1:100])
-stud$field_id <- c(hs1$field_id,hs2$field_id[!hs2$field_id%in%hs1$field_id])
 stud <- stud[stud$treatment==1,]
 
 stud$obsUsage <- stud$field_id%in%usageIDs
+
+stud$pretest <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)
+
 
 schoolMiss <- NULL
 for(scl in unique(stud$schoolid2)){

### 2017-10-17 fig1.r

@@ -629,18 +629,24 @@ secLev$cp <- secLev$status=='changed placement'
 library(lme4)
 cpMod <- glmer(cp~Year*(xirt+grade+spec_speced+spec_gifted+pBefore+totClassmates+endDate+state)+(Year*(xirt+grade+spec_speced+spec_gifted+pBefore+totClassmates+endDate+state)|unit)+(xirt+grade+spec_speced+spec_gifted+pBefore+totClassmates+endDate+state|classid2)+(1|schoolid2),data=secLev)
 save(cpMod,file='cpMod.RData')
-=======
-#### what's the next section each student works?
-secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]))
 
 
 #### what's the next section each student works?
-secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
+secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(date=max(date,na.rm=TRUE),time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(status=as.character(status),prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
 
 secOrder$cp <- secOrder$status=='changed placement'
[…]

### 2017-10-18 paper.Rnw
more

@@
 dataCache <- grep('data',list.files('cache'),ignore.case=FALSE,value=TRUE)[1]
-if(file.info('dataMerge.r')$mtime>file.info(paste0('cache/',dataCache))$mtime)
-    source('dataMerge.r')
+
+if(file.info('dataMerge.r')$mtime>file.info(paste0('cache/',dataCache))$mtime){
+    if(file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
+        load('cpPaper.RData')
+    } else source('dataMerge.r')
+}
 @
@@
 for(i in 1:nrow(bbb)) bbb$n[i] <- bbb$n[i]/staten$n[staten$state==bbb$state[i] & staten$Yr==bbb$Yr[i]]
+
+COLS <- c('#1b9e77','#d95f02','#7570b3')
 
@@
     melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
-statusStateYr2b <- statusStateYr2%>%
+statusStateYr2b <- statusStateYr2%>%group_by(state,Yr)%>%
     summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
@@
 
-Figure \ref{fig:vc}
+To better understand the source of the variation in
+reassignment---what drives some, but not other, sections to end in
+reassignment---we fit a set of multilevel models.
+We fit separate models to data from each the three states with the highest numbers of reassignments, Texas,
+Kentucky, and Michigan, and in the sample as a whole, in each of the
+two study years, yielding a total of eight models.
+Each model was a logistic regression: a binary indicator for section
+reassignment was regressed on a random intercept for unit, as well as
+nested random intercepts for student, classroom, and school.
+Models fit to data from all six states included an additional random
+intercept for state.
+
+Logistic regression can be represented in terms of an underlying
+latent variable $Z^*$: student $i$ working section $sec$ is reassigned
+when $Z_{sec,i}^*>0$.
+The model for $Z^*$ is:
+\begin{equation*}
+ Z^*_{sec,i}=\alpha_0+\beta_{u[sec]}+\gamma_i+\delta_{c[i]}+\epsilon_{s[i]}+e_{sec,i}
+\end{equation*}
+Where $\alpha_0$ is an overall intercept,
+and $\beta_{u[sec]}$, $\gamma_i$, $\delta_{c[i]}$, and $\epsilon_{s[i]}$
+are random intercepts for the unit in which $sec$ appears, for student
+$i$, for $i$'s classroom, and for $i$'s school, respectively.
+Again, the model fit to all six states includes an additional random
+intercept for state.
+The random intercepts are modeled as independent and normally
+distributed, each with its own variance.
+The regression error $e_{sec,i}$ is given the standard logistic
+distribution, with ``residual'' variance $\pi/3$.
+It is convenient to represent variance in reassignment probabilities
+in terms of the variance of $Z^*$.
+
+Figure \ref{fig:vc} gives the variance components estimated from these
+logistic regressions: variances of the random intercept terms, as a
+percentage of the total variance of $Z^*$.
+Overall, in both years of the study, the largest determinant of
+reassignment was school, accounting for
+\Sexpr{vcs[['Overall_1']]['schoolid','sig2']}\% of the variation in year 1,
+and \Sexpr{vcs[['Overall_2']]['schoolid',1]}\% in year 2.
+After school, state was the most important, accounting for
+\Sexpr{vcs[['Overall_1']]['state',1]} and
+\Sexpr{vcs[['Overall_2']]['state',1]}\% in the two years, and unit,
+accounting for \Sexpr{vcs[['Overall_1']]['unit',1]} and \Sexpr{vcs[['Overall_2']]['unit',1]}\%.
+Surprisingly, classroom and student-level factors only accounted for
+\Sexpr{vcs[['Overall_1']]['classid',1]} and
+\Sexpr{vcs[['Overall_1']]['field_id',1]}\% in year 1, respectively,
+and \Sexpr{vcs[['Overall_2']]['classid',1]} and
+\Sexpr{vcs[['Overall_2']]['field_id',1]}\% in year 2.
+The pattern was similar in Texas, where school accounted for over half
+the variation in reassignment in both years, and in Michigan.
+In Kentucky, unit played the largest role
+(\Sexpr{vcs[['KY_1']]['unit',1]}\%) in year 1, and classroom played
+the largest role in year 2 (\Sexpr{vcs[['KY_2']]['classid',1]}\%).
+Across states and years, student level factors never accounted for
+more than \Sexpr{max(sapply(vcs,function(x) x['field_id',1]))}\% of
+the variation in reassignment.
+Other than in Kentucky in year 2, classroom never accounted for more
+than \Sexpr{sort(sapply(vcs,function(x) x['classid',1]),dec=T)}\%  of
+the variation.
+
+Although teachers control reassignment, their decisions are largely
+dictated by broader policies, ocurring at the state or school level.
 
@@
 vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
+
+vcs <- list()
+for(s in unique(vcDat$state)) for(y in 1:2){
+ ddd <- round(subset(vcDat,state==s&year==y,select=c(sig2))*100)
+ rownames(ddd) <- gsub('\\d','',rownames(ddd))
+ rownames(ddd)[nrow(ddd)] <- 'resid'
+ vcs[[paste0(s,'_',y)]] <- ddd
+}
 
@@
 
+The timing of reassignments can provide a window into what drives
+teachers' decisions to reassign students.
+Figure \ref{fig:byMonth} shows the proportion of worked sections in
+each month that end in reassignment.
+In both years, reassignments were much more common in the second half
+of the school year than in the first.
+This may be the result of teachers learning how to use the software as
+the year progresses, or responding the pressure of upcoming
+standardized tests by accelerating students' progress and reassigning
+students to relevent sections.
+
+As we've seen, reassignment was more common in year 2 than in year 1.
+In fact, reassignment increases fairly steadily over the entire length
+of the study.
+Through December of the first year, reassignment was rare, and from
+January through May of year 1, between one and two percent of sections
+ended in reassignment.
+Year 2 begin where year 1 left off, with one to two percent of
+sections reassigned.
+Finally, from February through May of the second year, the rate of
+reassignment incrased again.
+
 \begin{figure}
@@
 <<byMonth,dependson='cpDat',fig.height=3,fig.width=6>>=
-secLevMonth <- secLev%>%group_by(endMonth,Year)%>%summarize(ncp=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
+secLevMonth <- secLev%>%group_by(endMonth,Year)%>%
+ summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
 secLev$cp <- as.numeric(secLev$status=='changed placement')
 
-ggplot(filter(secLevMonth,ncp>100),aes(endMonth,CPper,group=Year,color=Year,size=ncp))+geom_point()+geom_line(size=1)+
+ggplot(filter(secLevMonth,nsec>100),aes(endMonth,CPper,group=Year,color=Year,size=nsec))+geom_point()+geom_line(size=1)+
     ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-0.5,cp,group=Year,
@@
 scale_y_continuous(breaks=seq(0,0.05,.01),labels=percent)+
-    coord_cartesian(ylim=c(0,0.05))+ labs(x='Month',y='% of Sections Ending in Reassignment')
-
-@
-\end{figure}
+    coord_cartesian(ylim=c(0,0.05))+
+    labs(x='Month',y='% of Sections Ending in Reassignment',size='# Worked \\Sections')
+
+@
+\caption{The proportion of worked sections ending in reassignment, by
+  month and year.}
+\label{fig:byMonth}
+\end{figure}
+
+\begin{table}
+  \begin{tabular}{rll|ll}
+    &\multicolumn{2}{c}{Year 1}&\multicolumn{2}{c}{Year 2}\\
+    &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
+<<byMonthTab,results='asis'>>=
+
+tab <- NULL
+for(st in c('TX','KY','MI','LA','CT','NJ')){
+ tab <- rbind(tab,
+  c(sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 1'&
+     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
+    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 1'&
+     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement'),
+    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 2'&
+     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
+    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 2'&
+     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement')))
+}
+tab <- rbind(tab,
+    c(sum(secLev$status[secLev$Yr=='Yr 1'&
+     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
+    sum(secLev$status[ secLev$Yr=='Yr 1'&
+     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement'),
+    sum(secLev$status[ secLev$Yr=='Yr 2'&
+     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
+    sum(secLev$status[ secLev$Yr=='Yr 2'&
+     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement')))
+rownames(tab) <- c('TX','KY','MI','LA','CT','NJ','Overall')
+
+print(xtable(tab),only.contents=TRUE,include.colnames=FALSE,hline.after=6)
+@
+\hline
+\end{tabular}
+\caption{The number of reassignments in each state and overall in the
+  first and second halves of the year}
+\label{tab:byMonth}
+\end{table}
+
+Figure \ref{fig:byMonthState} and Table \ref{tab:byMonth} decompose
+these trends by state, Figure \ref{fig:byMonthState} for
+the three states with the highest number of reassignments, Texas,
+Kentucky, and Michigan, and Table \ref{tab:byMonth} for all six states
+and overall.
+For the two study years, Figure \ref{fig:byMonthState} shows the proportion of all of each
+state's reassignments the occurred in each month of the school year.
+(Note that while both Figure \ref{fig:byMonth} and \ref{fig:byMonth}
+show proportions for each month, the denominators are not the same:
+Figure \ref{fig:byMonth} gives the proportion of each month's worked
+sections that ended in reassignment, and Figure \ref{fig:byMonthState}
+gives the proportion of all of each state's reassignmens over the
+course of the year, that occurred in each month.)
+
+The pattern in Figure \ref{fig:byMonth} seems largely driven by Texas.
+In contrast, the bulk of Kentucky's reassignments took place in May.
+A fifth of Michigan's reassignments in year 1 occurred in October;
+for the rest of the year, Michigan roughly followed the same rough pattern as Texas.
+In year 2, the vast majority of Michigan's reassignments occurred at
+the beginning of the year, in September and October.
 
@@
 <<byMonthState,dependson='cpDat',fig.height=3,fig.width=6>>=
-secLevMonthSt <- secLev%>%filter(state%in%c('TX','KY','MI') & endMonth!='Jul')%>%group_by(endMonth,state,Year)%>%summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
-ggplot(secLevMonthSt,aes(endMonth,CPper,group=state,color=state,size=nsec))+geom_point()+facet_grid(Year~.)+
+
+secLevMonthSt <- secLev%>%filter(status=='changed placement'&state%in%c('TX','KY','MI') & endMonth!='Jul')%>%
+ group_by(endMonth,Year,state)%>%summarize(cpPer=n())
+secLevNum <- secLev%>%filter(status=='changed placement'&state%in%c('TX','KY','MI'))%>%
+ group_by(Year,state)%>%summarize(ncp=n())
+
+for(st in c('TX','KY','MI')) for(yr in c('Year 1','Year 2'))
+ secLevMonthSt$cpPer[secLevMonthSt$state==st & secLevMonthSt$Year==yr] <-
+  secLevMonthSt$cpPer[secLevMonthSt$state==st & secLevMonthSt$Year==yr]/
+   secLevNum$ncp[secLevNum$state==st & secLevNum$Year==yr]
+
+## secLevMonthSt <- secLev%>%filter(state%in%c('TX','KY','MI') & endMonth!='Jul')%>%group_by(endMonth,state,Year)%>%summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
+ggplot(secLevMonthSt,aes(endMonth,cpPer,group=state,color=state))+geom_point()+facet_grid(Year~.)+
     geom_line(size=1)+
-    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-1.5,cp,group=state,
-    ##                 color=state,size=1),
-    ##             method = "glm",
-    ##             formula = y ~ splines::ns(x, knots=seq(1,10,2)),
-    ##             data=filter(secLev,state%in%c('TX','CT','NJ')& endMonth!='Jul'),
-    ##             method.args=list(family='binomial'),
-    ##             show.legend=FALSE)+
     scale_y_continuous(labels=percent)+
-    labs(x='Month',y='% Sections Ending in Reassignment',size='Worked \n Sections',color='')
-
-@
+    labs(x='Month',y='% of State\'s Total Reassignments')
+
+@
+\caption{The proportion of a state's reassignments that occurred in
+  each month of the year.}
+\ref{fig:byMonthState}
 \end{figure}
@@
 \subsection{Does Reassignment Depend on Classmates?}
+
+Student individuality and independence might be the most important
+motivating factors behind mastery learning---each student learns at
+his or her own pace, and struggles on a unique set of skills.
+Students are supposed to move through the CT curriculum independently
+of each other.
+However, reassignments give teachers the ability to override this
+feature, and coordinate students' progress.
+Teachers can identify students who are behind their classmates and
+reassign them to later sections.
+They may also move an entire class together to a particular section or
+unit of interest.
+To what extent did these and similar considerations drive reassignment
+in the CTAI study?
+
 \begin{figure}
@@
 cpDat$bf.promGrad <- cpDat$bf.promGrad+cpDat$same.promGrad
+cpDat$same.promGrad <- NULL
 
@@
 
+behind <- with(cpDat2,mean(bf.promGrad+bf.cp+same.cp>.75))
+behind1 <- with(subset(cpDat2,Yr=='Yr 1'),mean(bf.promGrad+bf.cp+same.cp>.75))
+behind2 <- with(subset(cpDat2,Yr=='Yr 2'),mean(bf.promGrad+bf.cp+same.cp>.75))
+
+behind1tx <- with(subset(cpDat2,Yr=='Yr 1'&state=='TX'),mean(bf.promGrad+bf.cp+same.cp>.75))
+behind1ky <- with(subset(cpDat2,Yr=='Yr 1'&state=='KY'),mean(bf.promGrad+bf.cp+same.cp>.75))
+behind1mi <- with(subset(cpDat2,Yr=='Yr 1'&state=='MI'),mean(bf.promGrad+bf.cp+same.cp>.75))
+
+behind2tx <- with(subset(cpDat2,Yr=='Yr 2'&state=='TX'),mean(bf.promGrad+bf.cp+same.cp>.75))
+behind2ky <- with(subset(cpDat2,Yr=='Yr 2'&state=='KY'),mean(bf.promGrad+bf.cp+same.cp>.75))
+behind2mi <- with(subset(cpDat2,Yr=='Yr 2'&state=='MI'),mean(bf.promGrad+bf.cp+same.cp>.75))
+
+mi2cp <- round(with(subset(cpDat2,Yr=='Yr 2'&state=='MI'),mean(bf.cp+same.cp+after.cp+nw>.75))*100)
+ky1cp <- round(with(subset(cpDat2,Yr=='Yr 1'&state=='KY'),mean(bf.cp+same.cp+after.cp+nw>.75))*100)
+
 #aaa$variable <- factor(aaa$variable,levels=rev(c('pGrad','pProm','pCPbf','pCPsame','pLater','pNW')))
@@
         axis.ticks.x=element_blank()) + scale_y_continuous(label=percent)+labs(fill='% of Classmates')+
-    scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02'))
-
-@
-\end{figure}
+    scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02'))+
+    labs(y='')
+
+@
+\caption{Distributions of classmates' statuses at each reassignment in
+  Texas, Kentucky, and Michigan. Each reassignment that took place in
+  these three states is represented by a bar showing the proportions of
+  the reassigned student's classmates who had exited the section on
+  the same day or earlier via promotion or mastery, who had been
+  reassigned from that section on an earlier date, on the same date,
+  or on a later date, who mastered the section or were promoted on a
+  later date, or who never worked the section at all. The bars are
+  ordered according to those proportions, in the order listed.}
+\label{fig:classmates}
+\end{figure}
+
+Figure \ref{fig:classmates} addresses this question by plotting a
+students' classmates' statuses at the time he or she is reassigned.
+For each reassignment in Texas, Kentucky, and Michigan, Figure \ref{fig:classmates}
+plots a vertical bar colored to show the proportions of the reassigned
+students' classmates (represented in the usage data) who  had exited the section on
+the same day or earlier via promotion or mastery, who had been
+reassigned from that section on an earlier date, on the same date,
+or on a later date, who mastered the section or were promoted on a
+later date, or who never worked the section at all.
+The bars are ordered according to those same proportions: first by the
+proportion of classmates who were promoted or mastered the section on
+the same day as the reassignment in question, or earlier, next by the
+proportion who were reassigned from the same section on an earlier
+date, next by the proportion who were reassigned from the same section
+on the same date, and finally by the proportion who were reassigned
+from the same section on a later date.
+
+Do teachers reassign students in order to help them catch up with classmates?
+According to Figure \ref{fig:classmates}, that might be part of the
+story, but isn't all of it.
+Across years and states, in about \Sexpr{round(behind*100)}\% of
+reassignments, at least 75\% of the rest of the class had exited the
+section on the same date or earlier.
+In fact, smaller proportions of students had graduated or been
+promoted from the same section on the same day or earlier in year 2
+than in year 1, especially in Texas and Michigan.
+
+In Texas, year 2 saw a dramatic increase in the proportions of
+students reassigned from the same section on the same day, suggesting
+that some teachers may have been moving the class together through the
+curriculum.
+In Michigan in year 2, teachers reassigned almost all students who
+worked certain sections.
+That is, \Sexpr{mi2cp}\% of the instances in which year-two Michigan
+students were reassigned from a section, at least 75\% of their
+classmates were, at some point, reassigned from the same section, or
+never worked it.
+Accross all three states and both years, it is exceedingly rare for
+students to master or be promoted from a section after someone in
+their class has been reassigned from the same section.
+
+All in all, three different patterns emerge from Figure
+\ref{fig:classmates}: teachers reassigning students who have fallen
+behind their classmates, teachers reassigning an entire class from the
+same section on the same day, and teachers reassigning almost all
+students who begin to work on particular sections.
+Each of these patterns takes place in different states and
+years.
 
@@
 
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+## top senders and receivers
+cols <- which(colSums(trans)>30)
+rows <- which(rowSums(trans)>30)
+
+cols2 <- cols
+rows2 <- rows
+
+## top 2 receivers for each sender, vice versa
+for(r in rows)
+    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
+    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
+}
+
+for(cc in cols){
+ #   print('new col')
+  #  print(cc)
+    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
+        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
+    }
+
+}
+
+cols <- sort(unique(c(cols2,rows2)))
+rows <- sort(unique(rows2))
+
+trans <- trans[rows,cols]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
 
@@
 par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(-.5,4),ylim=c(0,nsec),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
+text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
+
 for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
@@
 for(rr in rownames(trans))
- for(cc in colnames(trans))
-     if(trans[rr,cc]>20){
+ for(cc in colnames(trans)){
+#     if(trans[rr,cc]>20){
       arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>100,1,trans[rr,cc]/100)))
-}
-
-@
-\end{figure}
-
-\begin{figure}
-<<transition2,fig.height=7,fig.width=6.5,dependson='secOrder'>>=
+       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
+}
+
+@
+\end{figure}
+
+\begin{figure}
+<<transition2,fig.height=5,fig.width=6.5,dependson='secOrder'>>=
 
@@
 
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-
-propTotCP <- sum(trans)/sum(secOrderCP$year==2)
-
+## top senders and receivers
+cols <- which(colSums(trans)>30)
+rows <- which(rowSums(trans)>30)
+
+cols2 <- cols
+rows2 <- rows
+
+## top 2 receivers for each sender, vice versa
+for(r in rows)
+    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
+    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
+}
+
+for(cc in cols){
+ #   print('new col')
+  #  print(cc)
+    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
+        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
+    }
+
+}
+
+cols <- sort(unique(c(cols2,rows2)))
+rows <- sort(unique(rows2))
+
+trans <- trans[rows,cols]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
+
+propTotCP <- sum(trans)/sum(secOrderCP$year==1)
 #transitionPlot(trans)
@@
 for(rr in rownames(trans))
- for(cc in colnames(trans))
-     if(trans[rr,cc]>20){
+ for(cc in colnames(trans)){
+#     if(trans[rr,cc]>20){
       arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>100,1,trans[rr,cc]/100)))
+       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
 }

### 2017-10-18 cpEffectSimp.r

@@ -45,3 +45,5 @@ mod4.1 <- update(mod4,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
 save(list=grep('mod',ls()),file='outcomeMods.RData')
 
 
+library(optmatch)
+propScore <- glm(everCP~(race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)*schoolid2,family=binomial,da

### 2017-10-19 paper.Rnw
sss

@@
 
+Teachers who reassign students may simply move them to the next
+section within the same unit.
+Say that a teacher believes that a particular student had already mastered
+the skills in one of the CTAI sections, or is wheel-spinning---working
+problems without learning.
+Still, the teacher wants the student to learn as much as possible from
+the student's current unit.
+Then moving the student to the next section within the same unit would
+make sense.
+Alternatively, teachers who believe that some of their students are
+not progressing quickly enough may reassign them out of their current
+units entirely, and into the next unit in the sequence---whatever that
+may be.
+Finally, a teacher who wanted his or her students to focus on a
+particular topic might reassign them all to the appropriate section
+when the time is appropriate.
+
+Which of these patterns is most prevelant?
+More generally, when teachers reassign their students, where in the
+curriculum do they send them?
+
+To address these questions, we focus on the units in the standard
+Algebra I sequence.
+Figures \ref{fig:trans1} and \ref{fig:trans2} give transition plots
+for reassignment in the two study years.
+On the left of each figure, marked ``From,'' are the top ``sending''
+units: the units that students were reassigned \emph{from} at least 30
+times.
+They are listed in the order they appear in the standard Algebra I curriculum.
+On the right, under ``To,'' are the top ``receiving'' units: the units
+that students were reassigned \emph{to} at least 30 times.
+For completeness, each of the top two receivers for each sending unit,
+and each of the top two senders for each receiving unit were also included.
+Finally, all of the sending units were also listed on the left, in the
+receiving column.
+
+The arrows in the plot represent reassignments.
+An arrow from a sending unit to a receiving unit indicates
+reassignment from the former to the latter.
+The thickness of the arrows represents the proportion of reassignments
+originating in the sending unit whose destination was the receiving unit.
+The darkness of the arrows represents the number of such reassignments.
+For instance, in Figure \ref{fig:trans1}, the thick line from
+``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
+Graphs'' on the right is fairly thick, since
+
 <<secOrder,include=FALSE,dependson=c('data','setup')>>=
@@
 \begin{figure}
-<<transition1,fig.height=3,fig.width=6.5,dependson='secOrder'>>=
-
-trans <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
+  \begin{subfigure}{.7\textwidth}
+<<transition1,fig.height=3,fig.width=4.5,dependson='secOrder'>>=
+
+trans1 <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
 
 ## top senders and receivers
-cols <- which(colSums(trans)>30)
-rows <- which(rowSums(trans)>30)
+cols <- which(colSums(trans1)>30)
+rows <- which(rowSums(trans1)>30)
 
@@
 for(r in rows)
-    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
-    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
+    while(sum(trans1[r,cols2])/sum(trans1[r,])<0.75){
+    nextBest <- which.max(trans1[r,seq(ncol(trans1))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans1)==names(nextBest)))
 }
@@
 for(cc in cols){
- #   print('new col')
-  #  print(cc)
-    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
-        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
+    while(sum(trans1[rows2,cc])/sum(trans1[,cc])<0.75){
+        nextBest <- which.max(trans1[seq(nrow(trans1))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans1)==names(nextBest)))
     }
@@
 
-trans <- trans[rows,cols]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-
-propTotCP <- sum(trans)/sum(secOrderCP$year==1)
-#transitionPlot(trans)
-
-nsec <- nrow(trans)
+trans1 <- trans1[rows,cols]
+
+propTotCP <- sum(trans1)/sum(secOrderCP$year==1)
+#trans1itionPlot(trans1)
+
+nsec <- nrow(trans1)
 uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
+uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
 nsec <- length(uuu)
@@
 
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
-
-for(rr in rownames(trans))
- for(cc in colnames(trans)){
-#     if(trans[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+
+for(rr in rownames(trans1))
+ for(cc in colnames(trans1)){
+#     if(trans1[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans1[rr,cc]/sum(trans1[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans1[rr,cc]>50,1,trans1[rr,cc]/50)))
 }
@@
 @
+\end{subfigure}
+\begin{subfigure}{.25\textwidth}
+<<trans1legend,fig.width=1.6,fig.height=3>>=
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(0,2),ylim=c(0,11),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+text(1,10,expression(underline(bold(paste(Legend)))))
+text(1,8,'Thickness: Percent')
+text(.5,7,'25%')
+text(.5,6,'50%')
+text(.5,5,'75%')
+arrows(c(1.5,1.5,1.5),c(7,6,5),c(2,2,2),lwd=5*c(.25,.5,.75))
+text(1,4,'Darkness: Number')
+text(.5,3,'10')
+text(.5,2,'25')
+text(.5,1,'50+')
+arrows(c(1.5,1.5,1.5),c(3,2,1),c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
+
+@
+\caption{Reassignment transition plot for study year 1. The units on
+  the left of the plot are the Algebra I units that ended in reassignment at
+  least 30 times in year 1. The units on the right are those that were
+  the destination of at least 30 reassignments. Also included on the
+  left are the top two sending units for each of the units on the
+  right, and also included on the right are all the units on the left,
+  along with their top two receiving units. The units are listed
+  following their order in the standard Algebra I curriculum. There is an arrow from a
+  unit on the left to a unit on the right if a student was assigned
+  from the unit on the left to the one on the right. The thickness of
+  the arrows is proportional to the percentage of all reassignments from
+  the sending unit that ended in that receiving unit. The darkness of
+  the arrows is proportional to the number of reassignments from the
+  sending unit to the receiving unit.}
+\label{fig:trans1}
 \end{figure}
@@
 
-trans <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
+trans2 <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
 
 ## top senders and receivers
-cols <- which(colSums(trans)>30)
-rows <- which(rowSums(trans)>30)
+cols <- which(colSums(trans2)>30)
+rows <- which(rowSums(trans2)>30)
 
@@
 for(r in rows)
-    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
-    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
+    while(sum(trans2[r,cols2])/sum(trans2[r,])<0.75){
+    nextBest <- which.max(trans2[r,seq(ncol(trans2))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans2)==names(nextBest)))
 }
@@
   #  print(cc)
-    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
-        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
+    while(sum(trans2[rows2,cc])/sum(trans2[,cc])<0.75){
+        nextBest <- which.max(trans2[seq(nrow(trans2))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans2)==names(nextBest)))
     }
@@
 
-trans <- trans[rows,cols]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
-
-propTotCP <- sum(trans)/sum(secOrderCP$year==1)
-#transitionPlot(trans)
-
-nsec <- nrow(trans)
+trans2 <- trans2[rows,cols]
+
+propTotCP <- sum(trans2)/sum(secOrderCP$year==1)
+
+nsec <- nrow(trans2)
 uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
+uuu <- uuu[uuu%in%c(colnames(trans2),rownames(trans2))]
 nsec <- length(uuu)
@@
 
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
-
-for(rr in rownames(trans))
- for(cc in colnames(trans)){
-#     if(trans[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+
+for(rr in rownames(trans2))
+ for(cc in colnames(trans2)){
+#     if(trans2[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
 }
-
-@
+@
+\caption{Reassignment transition plot for study year 2.}
+\label{fig:trans2}
 \end{figure}

### 2017-10-19 cpEffectSimp.r

@@ -30,20 +30,66 @@ cpDat <- mutate(cpDat,totalTime=totalTime/3600000,totalTime=ifelse(totalTime<0,N
 cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=3,'3+',cpDat$ncp))
 cpDat$everCP <- cpDat$ncp>0
 
+exirt <- stud[match(cpDat$field_id,stud$field_id),grep('Exirt2',names(stud))]
+names(exirt) <- gsub('_0','_',names(exirt))
+for(n in names(exirt)) cpDat[[n]] <- exirt[[n]]
+
+
 #### models!!
-mod1 <- lmer(gainScore~ncp+nsec+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
+mod1 <- lmer(gainScore~ncp+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
 mod1.1 <- update(mod1,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
-mod2 <- lmer(gainScore~year*(ncp+nsec)+state+(1|classid2)+(1|schoolid2),data=cpDat)
+mod2 <- lmer(gainScore~year*(ncp)+state+(1|classid2)+(1|schoolid2),data=cpDat)
 mod2.1 <- update(mod2,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
 mod3 <- lmer(gainScore~everCP+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
-mod3.1 <- update(mod3,.~.+nsec)
-mod3.1.1 <- update(mod3.1,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
-mod3.2 <- update(mod3,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
-mod4 <- lmer(gainScore~ncpCat+nsec+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
+mod3.1 <- update(mod3,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
+mod4 <- lmer(gainScore~ncpCat+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
 mod4.1 <- update(mod4,.~.+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)
+cpDat$Year=as.factor(cpDat$year)
[…]

### 2017-10-19 fig1.r

@@ -99,7 +99,8 @@ currWorkedYr <- within(currWorkedYr,p <- n/ifelse(Year=='Year 1',nByYr$n[nByYr$Y
 ggplot(currWorkedYr,aes(Year,p,fill=Curriculum,shape=overall))+geom_col()
 
 
-cwy <- currWorkedYr%>%group_by(Curriculum,overall)%>%summarize(`Year 1`=p[Year=='Year 1'],`Year 2`=p[Year=='Year 2'])%>%arrange(Curriculum,overall)
+##cwy <- currWorkedYr%>%group_by(Curriculum,overall)%>%summarize(`Year 1`=p[Year=='Year 1'],`Year 2`=p[Year=='Year 2'])%>%arrange(Curriculum,overall)
+cwy <- dcast(currWorkedYr,Curriculum+overall~Year,value.var='p')
 COLS <- c('#1b9e77','#d95f02','#7570b3')
 
 cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%

### 2017-10-19 mastNext.r
subsequent mastery models

@@ -1,0 +1,33 @@
+library(lme4)
+library(dplyr)
+
+load('cpPaper.RData')
+
+#### what's the next section each student works?
+secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp) )%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(date=max(date,na.rm=TRUE),time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(status=as.character(status),prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
+
+
+secOrder$cp <- secOrder$status=='changed placement'
+secOrder$mast <- secOrder$status=='graduated'
+
+cpdStuds <- unique(data$field_id[!is.na(data$status) & data$status=='changed placement'])
+
+mnDat <- filter(secOrder,prevStatus!='final_or_incomplete' &  field_id%in%cpdStuds)
+
+mnDat$prevStatus <- factor(mnDat$prevStatus,levels=c('changed placement','promoted','graduated'))
+
+### try with glm:
+mnDat$secUnit <- paste(mnDat$section,mnDat$unit)
+
+mnDat <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>100])
+
+mn1 <- glm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat,family=binomial,subset=year==1)
+mn2 <- glm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat,family=binomial,subset=year==2)
+
+mastNext <- glmer(mast~(1|field_id/classid2)+(1|section/unit)+as.factor(prevStatus),data=mnDat,family=binomial)
+
+mastNext2 <- update(mastNext,data=subset(mnDat,status!='changed placement' & status!='final_or_incomplete'))
+
+mnDat$month <- months(mnDat$date)
+
+mastNext3 <- update(mastNext,.~.+Year+month+schoolid2)

### 2017-10-19 paper.Rnw
more modeling...

@@
 xtable(tab,caption='Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study.
-Students with no usage data were excluded',label='tab:medUsage',digits=c(1,2,0,1,0))
+Students with no usage data were excluded.',label='tab:medUsage',digits=c(1,2,0,1,0))
 @
@@
   software each student worked, by year and state. Students
-  working more than 55 units (\Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
+  working more than 55 units (
+  \Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)} of
+  \Sexpr{nrow(secByStud)}) and students
   with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
@@
 \subsection{How Do Reassignment Patterns Vary?}
+
+\begin{figure}
+<<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
+load('vcMods.RData')
+library(lme4)
+
+vcFun <- function(nm){
+    mod <- vcMods[[nm]]
+    out <- unlist(summary(mod)$varcor)
+    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
+    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
+    out$state <- strsplit(nm,'_')[[1]][1]
+    out$year <- strsplit(nm,'_')[[1]][2]
+    out$sig2 <- out$sig2/sum(out$sig2)
+    out
+}
+
+vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
+
+yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
+                          state='Overall',
+                  year='1')
+yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
+
+yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
+                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
+                          state='Overall',
+                  year='2')
+yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
+
+vcDat <- rbind(vcDat,yr1,yr2)
+
+vcDat$comp <- factor(vcDat$comp)
+levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
+
+vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
+
+vcs <- list()
+for(s in unique(vcDat$state)) for(y in 1:2){
+ ddd <- round(subset(vcDat,state==s&year==y,select=c(sig2))*100)
+ rownames(ddd) <- gsub('\\d','',rownames(ddd))
+ rownames(ddd)[nrow(ddd)] <- 'resid'
+ vcs[[paste0(s,'_',y)]] <- ddd
+}
+
+ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
+
+@
+\caption{Results from a set of eight multilevel logistic regressions
+  predicting section reassignment. For each year, in the entire sample (``Overall'')
+  and in the three states with the largest numbers of reassignments (Texas,
+  Kentucky, and Michigan), we regressed a binary variable indicating
+  whether a section ended in reassignment on random intercepts for
+  school, class, student, and unit, and in the overall case, for state
+  as well, and recorded their variance. The residual variance was set
+  as the variance of the standard logistic distribution,
+  $\pi/3$. These bar charts give the proportion of the total variance
+  attributable to each random effect.}
+\label{fig:vc}
+\end{figure}
 Teachers alone control reassignment.
@@
 
-\begin{figure}
-<<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
-load('vcMods.RData')
-library(lme4)
-
-vcFun <- function(nm){
-    mod <- vcMods[[nm]]
-    out <- unlist(summary(mod)$varcor)
-    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
-    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
-    out$state <- strsplit(nm,'_')[[1]][1]
-    out$year <- strsplit(nm,'_')[[1]][2]
-    out$sig2 <- out$sig2/sum(out$sig2)
-    out
-}
-
-vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
-
-yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
-                          state='Overall',
-                  year='1')
-yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
-
-yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
-                          state='Overall',
-                  year='2')
-yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
-
-vcDat <- rbind(vcDat,yr1,yr2)
-
-vcDat$comp <- factor(vcDat$comp)
-levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
-
-vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
-
-vcs <- list()
-for(s in unique(vcDat$state)) for(y in 1:2){
- ddd <- round(subset(vcDat,state==s&year==y,select=c(sig2))*100)
- rownames(ddd) <- gsub('\\d','',rownames(ddd))
- rownames(ddd)[nrow(ddd)] <- 'resid'
- vcs[[paste0(s,'_',y)]] <- ddd
-}
-
-ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
-
-@
-\caption{Results from a set of eight multilevel logistic regressions
-  predicting section reassignment. For each year, in the entire sample (``Overall'')
-  and in the three states with the largest numbers of reassignments (Texas,
-  Kentucky, and Michigan), we regressed a binary variable indicating
-  whether a section ended in reassignment on random intercepts for
-  school, class, student, and unit, and in the overall case, for state
-  as well, and recorded their variance. The residual variance was set
-  as the variance of the standard logistic distribution,
-  $\pi/3$. These bar charts give the proportion of the total variance
-  attributable to each random effect.}
-\label{fig:vc}
-\end{figure}
-
 \subsection{When are Students Reassigned?}
@@
 receiving column.
+All in all, \Sexpr{round(propTotCP1*100)}\% of the first-year reassignments and
+\Sexpr{round(propTotCP1*100)}\% of the second-year reassignments are
+captured in the figures.
 
@@
 The darkness of the arrows represents the number of such reassignments.
-For instance, in Figure \ref{fig:trans1}, the thick line from
+For instance, in Figure \ref{fig:trans1}, the arrow from
 ``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
 Graphs'' on the right is fairly thick, since
+\Sexpr{round(trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']/
+  sum(trans1['4-Quadrant Linear Graphs',])*100)}\%
+of the reassignments begun in ``4-Quadrante Linear Graphs'' end in the
+same unit.
+Yet it is also fairly faint, since it only represents
+\Sexpr{trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']}
+reassignments.
 
@@
 \begin{figure}
-  \begin{subfigure}{.7\textwidth}
-<<transition1,fig.height=3,fig.width=4.5,dependson='secOrder'>>=
+  \begin{subfigure}{5in}
+<<transition1,fig.height=3,fig.width=5,dependson='secOrder'>>=
 
@@
 
-propTotCP <- sum(trans1)/sum(secOrderCP$year==1)
-#trans1itionPlot(trans1)
-
-nsec <- nrow(trans1)
+propTotCP1 <- sum(trans1)/sum(secOrderCP$year==1)
+
 uuu <- levels(secOrderCP$unitName)
@@
 
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)
 
@@
 \end{subfigure}
-\begin{subfigure}{.25\textwidth}
-<<trans1legend,fig.width=1.6,fig.height=3>>=
+\begin{subfigure}{1in}
+<<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
 par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(0,2),ylim=c(0,11),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-text(1,10,expression(underline(bold(paste(Legend)))))
-text(1,8,'Thickness: Percent')
-text(.5,7,'25%')
-text(.5,6,'50%')
-text(.5,5,'75%')
-arrows(c(1.5,1.5,1.5),c(7,6,5),c(2,2,2),lwd=5*c(.25,.5,.75))
-text(1,4,'Darkness: Number')
-text(.5,3,'10')
-text(.5,2,'25')
-text(.5,1,'50+')
-arrows(c(1.5,1.5,1.5),c(3,2,1),c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
-
-@
+plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+conv <- (nsec+2)/11
+text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
+text(1,9*conv,'Thickness:\n Percent',cex=0.8)
+text(.5,7.5*conv,'25%')
+text(.5,6.5*conv,'50%')
+text(.5,5.5*conv,'75%')
+arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
+text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
+text(.5,3*conv,'10')
+text(.5,2*conv,'25')
+text(.5,1*conv,'50+')
+arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
+
+@
+\end{subfigure}
 \caption{Reassignment transition plot for study year 1. The units on
@@
 
+Inspecting the figures shows that the most common pattern is for
+students to be reassigned to the next unit in the curriculum.
+On the other hand, it is rare for students to be reassigned
+
 \begin{figure}
-<<transition2,fig.height=5,fig.width=6.5,dependson='secOrder'>>=
+  \begin{subfigure}{5in}
+<<transition2,fig.height=5,fig.width=5,dependson='secOrder'>>=
 
@@
 
-propTotCP <- sum(trans2)/sum(secOrderCP$year==1)
+propTotCP2 <- sum(trans2)/sum(secOrderCP$year==1)
 
@@
 
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)
 
@@
 @
+\end{subfigure}
+\begin{subfigure}{1in}
+<<trans2legend,fig.width=1,fig.height=3,dependson='transition2'>>=
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+conv <- (nsec+2)/11
+text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
+text(1,9*conv,'Thickness:\n Percent',cex=0.8)
+text(.5,7.5*conv,'25%')
+text(.5,6.5*conv,'50%')
+text(.5,5.5*conv,'75%')
+arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
+text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
+text(.5,3*conv,'10')
+text(.5,2*conv,'25')
+text(.5,1*conv,'50+')
+arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
+
+@
+\end{subfigure}
 \caption{Reassignment transition plot for study year 2.}
@@
 
-\subsection{Does Reassignment Depend on Student Ability?}
-
-<<xirtMods,include=FALSE,dependson=c('data','setup')>>=
-## don't worry about nesting, etc--we aren't reporting standard errors anyway
-cpDat <- data%>%filter(!is.na(status))%>%group_by(field_id,unit,section,state,year,classid2,schoolid2)%>%
-    summarize(cp=max(status)=='changed placement')
-
-### get imputed xirt
-##
-h1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
-h2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')
-
-h2 <- h2[!h2$field_id%in%h1$field_id,]
-
-stud <- rbind(h1[,c(174:194)],h2[,c(176:196)])
-stud <- stud[stud$field_id%in%cpDat$field_id,]
-stud <- stud[order(stud$field_id),]
-stud$xirt <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)
-
-cpDat$xirt <- stud$xirt[match(cpDat$field_id,stud$field_id)]
-
-cpModSimp <- glm(cp~splines::ns(xirt,5)*year,data=cpDat)
-
-cpModSt <- list()
-for(st in c('TX','KY','MI','LA','CT')){
-    cpModSt[[st]] <- update(cpModSimp,data=subset(cpDat,state==st))
-}
-
-@
-
-\begin{figure}
-<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup','xirtMods')>>=
-#load('cpXirtMods.RData')
-
-xirtDat <- cpDat%>%group_by(field_id,state,year,classid2,schoolid2,xirt)%>%
-    summarize(nsec=n_distinct(section,unit),propCP=mean(cp,na.rm=TRUE))
-
-for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
-                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
-                                         type='response')
-
-xirtDat <- addOverallState(xirtDat)
-
-xirtDat$fitted[xirtDat$state=='Overall'] <-
-    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
-            type='response')
-
-ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.5))
-
-@
-\end{figure}
-
-\begin{figure}
-<<cp-xirtClassroom1,dependson='xirtCP',fig.height=7,fig.width=6.5>>=
-bbb2 <- data%>%filter(status=='changed placement')%>%group_by(classid2,year,state)%>%summarize(ncp=n_distinct(field_id,section,unit),nstud=n_distinct(field_id),nsec=n_distinct(section))
-
-cpcls1 <- xirtDat%>%
-    filter(classid2%in%bbb2$classid2[bbb2$year==1 & bbb2$ncp>=20] & state!='Overall')%>%
-    arrange(state)
-cpcls1$classid2 <- factor(cpcls1$classid2,levels=unique(cpcls1$classid2))
-
-cpcls1 <- cpcls1%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()
-
-ggplot(cpcls1,aes(xirt,propCP,size=nsec,color=state))+
-    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
-    ylim(0,1)
-
-@
-\end{figure}
-
-\begin{figure}
-<<cp-xirtClassroom2,dependson='cp-xirtClassroom1',fig.height=7,fig.width=6.5>>=
-
-cpcls2 <- xirtDat%>%
-    filter(classid2%in%bbb2$classid2[bbb2$year==2 & bbb2$ncp>=20] & state!='Overall')%>%
-    arrange(state)
-cpcls2$classid2 <- factor(cpcls2$classid2,levels=unique(cpcls2$classid2))
-
-cpcls2 <- cpcls2%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()
-
-ggplot(cpcls2,aes(xirt,propCP,size=nsec,color=state))+
-    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
-    ylim(0,1)
-
-@
-\end{figure}
+\section{Effects of Reassignment}
+The goal of CTAI is to help students learn Algebra, so
+the most important questions about reassignment are about its effect on learning.
+Although the data from this study came from a randomized trial, it was
+CTAI as a whole that was randomized, not individual behaviors within CTAI.
+Specifically, student reassignment was not randomized.
+Therefore, claims regarding the causal effects of reassignment on
+learning require strong untestable assumptions that are unlikely to be
+true.
+As in all observational studies, this includes the assumption that all
+counfounding variables---variables that predict both reassignment and
+learning---have been measured well and modeled correctly.
+Further complicating matters, although reassignment itself is a
+well-defined process, in practice it can take many forms, as we have
+endevoured to show.
+There is no reason to expect the effects of reassignment to be
+the same regardless of wheather the teacher used it to help lagging
+students catch up, to allocate time to important topics, or for some
+other reason.
+
+All that said, observational estimates of reassignment's average
+effects can be valuable, if interpreted cautiously.
+They may help guide research
+
+\subsection{Effects on Subsequent Mastery}
+<<>>=
+@
+
+\subsection{Effects on Post-Test Scores}
 
@@
 \end{document}
+<<eval=FALSE>>=
+
+ nextDat <- secOrder%>%filter(!is.na(status))%>%group_by(field_id,classid2,year,state)%>%
+  summarize(naf=sum(is.na(prevStatus) & status=='final_or_incomplete')/sum(is.na(prevStatus)),
+   nacp=sum(is.na(prevStatus) & status=='changed placement')/sum(is.na(prevStatus)),
+   naprom=sum(is.na(prevStatus) & status=='promoted')/sum(is.na(prevStatus)),
+   nagrad=sum(is.na(prevStatus) & status=='graduated')/sum(is.na(prevStatus)),
+   ff=sum(prevStatus==1 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fcp=sum(prevStatus==1 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fprom=sum(prevStatus==1 & status=='promoted',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fgrad=sum(prevStatus==1 & status=='graduated',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   cpf=sum(prevStatus==2 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpcp=sum(prevStatus==2 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpprom=sum(prevStatus==2 & status=='promoted',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpgrad=sum(prevStatus==2 & status=='graduated',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   promf=sum(prevStatus==3 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
+   promcp=sum(prevStatus==3 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
+   promprom=sum(prevStatus==3 & status=='promoted',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
+   promgrad=sum(prevStatus==3 & status=='graduated',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
+   gradf=sum(prevStatus==4 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
+   gradcp=sum(prevStatus==4 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
+   gradprom=sum(prevStatus==4 & status=='promoted',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
+   gradgrad=sum(prevStatus==4 & status=='graduated',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
+   npna=sum(is.na(prevStatus)),npf=sum(prevStatus==1,na.rm=TRUE),npcp=sum(prevStatus==2,na.rm=TRUE),
+    npprom=sum(prevStatus==3,na.rm=TRUE),npgrad=sum(prevStatus==4,na.rm=TRUE))
+
+@

### 2017-10-19 cpEffectSimp.r

@@ -86,7 +86,7 @@ mi <- function(mod){
     miPool(miMod(mod))
 }
 
-ggplot(filter(cpDat,state%in%c('TX','KY','MI')),aes(ncp,gainScore))+geom_jitter()+geom_boxplot(aes(group=ncp),alpha=0.5)+geom_smooth(method='loess')+facet_grid(year~.)
+ggplot(filter(cpDat,state%in%c('TX')),aes(ncpCat,gainScore))+geom_jitter()+geom_boxplot(aes(group=ncp),alpha=0.5)+geom_smooth(method='loess')+facet_grid(year~.)
 
 library(optmatch)
 propScore <- glm(everCP~(race+sex+grade+spec_speced+spec_gifted+spec_esl+frl)*schoolid2,family=binomial,da

### 2017-10-19 mastNext.r

@@ -16,12 +16,48 @@ mnDat <- filter(secOrder,prevStatus!='final_or_incomplete' &  field_id%in%cpdStu
 
 mnDat$prevStatus <- factor(mnDat$prevStatus,levels=c('changed placement','promoted','graduated'))
 
-### try with glm:
+### try with lm:
 mnDat$secUnit <- paste(mnDat$section,mnDat$unit)
 
-mnDat <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>100])
+mnDat25 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>25])
+mnDat50 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>50])
+mnDat100 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>100])
+
+mn1ols25 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat25,subset=year==1)
+mn1ols50 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat50,subset=year==1)
+mn1ols100 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat100,subset=year==1)
+
+print(head(coef(mn1ols25)))
+print(head(coef(mn1ols50)))
+print(head(coef(mn1ols100)))
+
+mn2ols25 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat25,subset=year==2)
+
+
+library(sandwich)
+mn1vcov <- vcovHC(mn1ols25,'HC')
+mn2vcov <- vcovHC(mn2ols25,'HC')
+
+
+### what if remove cp?
+mn1olsNoCP <- update(mn1ols25,subset=status!='changed placement')
+mn2olsNoCP <- update(mn2ols25,subset=status!='changed placement')
+
+mn1vcovNoCP <- vcovHC(mn1olsNoCP,'HC')
+mn2vcovNoCP <- vcovHC(mn2olsNoCP,'HC')
+
[…]

### 2017-10-23 paper.Rnw
effects

@@
 Say that a teacher believes that a particular student had already mastered
-the skills in one of the CTAI sections, or is wheel-spinning---working
-problems without learning.
+the skills in one of the CTAI sections or is wheel-spinning---working
+problems without learning---or dislikes one of the sections in a CT unit.
 Still, the teacher wants the student to learn as much as possible from
@@
 
+<<secOrder,include=FALSE,dependson=c('data','setup')>>=
+#### what's the next section each student works?
+secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
+
+secOrder$cp <- secOrder$status=='changed placement'
+secOrder$mast <- secOrder$status=='graduated'
+
+secOrderCP <- subset(secOrder,status=='changed placement')
+
+secOrderCP$Unit <- secOrderCP$unit
+secOrderCP$Unit[grep('unit-conversions',secOrderCP$Unit)] <- 'unit-conversions'
+
+secOrderCP$unitName <- factor(secOrderCP$Unit,levels=units)
+levels(secOrderCP$unitName) <- UnitName
+
+secOrderCP$nextUnit[grep('unit-conversions',secOrderCP$nextUnit)] <- 'unit-conversions'
+
+secOrderCP$nextUnitName <- factor(secOrderCP$nextUnit,levels=units)
+levels(secOrderCP$nextUnitName) <- UnitName
+
+@
+
+\begin{figure}
+  \begin{subfigure}{5in}
+<<transition1,fig.height=3,fig.width=5,dependson='secOrder'>>=
+
+trans1 <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
+trans1o <- trans1
+
+## top senders and receivers
+cols <- which(colSums(trans1)>30)
+rows <- which(rowSums(trans1)>30)
+
+cols2 <- cols
+rows2 <- rows
+
+## top 2 receivers for each sender, vice versa
+for(r in rows)
+    while(sum(trans1[r,cols2])/sum(trans1[r,])<0.75){
+    nextBest <- which.max(trans1[r,seq(ncol(trans1))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans1)==names(nextBest)))
+}
+
+for(cc in cols){
+    while(sum(trans1[rows2,cc])/sum(trans1[,cc])<0.75){
+        nextBest <- which.max(trans1[seq(nrow(trans1))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans1)==names(nextBest)))
+    }
+
+}
+
+cols <- sort(unique(c(cols2,rows2)))
+rows <- sort(unique(rows2))
+
+trans1 <- trans1[rows,cols]
+
+propTotCP1 <- sum(trans1)/sum(secOrderCP$year==1)
+
+uuu <- levels(secOrderCP$unitName)
+uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
+nsec <- length(uuu)
+
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
+text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
+
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans1))
+  text(1,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(1,NA),cex=0.75)
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans1))
+  text(2,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(0,NA),cex=0.75)
+
+for(rr in rownames(trans1))
+ for(cc in colnames(trans1)){
+#     if(trans1[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans1[rr,cc]/sum(trans1o[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans1[rr,cc]>50,1,trans1[rr,cc]/50)))
+}
+
+@
+\end{subfigure}
+\begin{subfigure}{1in}
+<<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+conv <- (nsec+2)/11
+text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
+text(1,9*conv,'Thickness:\n Percent',cex=0.75)
+text(.5,7.5*conv,'25%')
+text(.5,6.5*conv,'50%')
+text(.5,5.5*conv,'75%')
+arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
+text(1,4.2*conv,'Darkness:\n Number',cex=0.75)
+text(.5,3*conv,'10')
+text(.5,2*conv,'25')
+text(.5,1*conv,'50+')
+arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
+
+@
+\end{subfigure}
+\caption{Reassignment transition plot for study year 1. The units on
+  the left of the plot are the Algebra I units that ended in reassignment at
+  least 30 times in year 1. The units on the right are those that were
+  the destination of at least 30 reassignments. Also included on the
+  left are the top two sending units for each of the units on the
+  right, and also included on the right are all the units on the left,
+  along with their top two receiving units. The units are numbered
+  according their order in the standard Algebra I curriculum. There is an arrow from a
+  unit on the left to a unit on the right if a student was assigned
+  from the unit on the left to the one on the right. The thickness of
+  the arrows is proportional to the percentage of all reassignments from
+  the sending unit that ended in that receiving unit. The darkness of
+  the arrows is proportional to the number of reassignments from the
+  sending unit to the receiving unit.}
+\label{fig:trans1}
+\end{figure}
+
+\begin{figure}
+  \begin{subfigure}{5in}
+<<transition2,fig.height=5,fig.width=5,dependson='secOrder'>>=
+
+trans2 <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
+trans2o <- trans2
+
+## top senders and receivers
+cols <- which(colSums(trans2)>30)
+rows <- which(rowSums(trans2)>30)
+
+cols2 <- cols
+rows2 <- rows
+
+## top 2 receivers for each sender, vice versa
+for(r in rows)
+    while(sum(trans2[r,cols2])/sum(trans2[r,])<0.75){
+    nextBest <- which.max(trans2[r,seq(ncol(trans2))[-cols2]])
+    cols2 <- c(cols2,which(colnames(trans2)==names(nextBest)))
+}
+
+for(cc in cols){
+ #   print('new col')
+  #  print(cc)
+    while(sum(trans2[rows2,cc])/sum(trans2[,cc])<0.75){
+        nextBest <- which.max(trans2[seq(nrow(trans2))[-rows2],cc])
+        rows2 <- c(rows2,which(rownames(trans2)==names(nextBest)))
+    }
+
+}
+
+cols <- sort(unique(c(cols2,rows2)))
+rows <- sort(unique(rows2))
+
+trans2 <- trans2[rows,cols]
+
+propTotCP2 <- sum(trans2)/sum(secOrderCP$year==1)
+
+nsec <- nrow(trans2)
+uuu <- levels(secOrderCP$unitName)
+uuu <- uuu[uuu%in%c(colnames(trans2),rownames(trans2))]
+nsec <- length(uuu)
+
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
+text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
+
+for(i in 1:nsec) if(uuu[i]%in%rownames(trans2))
+  text(1,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(1,NA),cex=0.75)
+for(i in 1:nsec) if(uuu[i]%in%colnames(trans2))
+  text(2,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(0,NA),cex=0.75)
+
+for(rr in rownames(trans2))
+ for(cc in colnames(trans2)){
+#     if(trans2[rr,cc]>20){
+      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2o[rr,]),
+       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
+}
+@
+\end{subfigure}
+\begin{subfigure}{1in}
+<<trans2legend,fig.width=1,fig.height=3,dependson='transition2'>>=
+par(mar=c(.1,.1,.1,.1))
+plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
+conv <- (nsec+2)/11
+text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
+text(1,9*conv,'Thickness:\n Percent',cex=0.75)
+text(.5,7.5*conv,'25%')
+text(.5,6.5*conv,'50%')
+text(.5,5.5*conv,'75%')
+arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
+text(1,4.2*conv,'Darkness:\n Number',cex=0.75)
+text(.5,3*conv,'10')
+text(.5,2*conv,'25')
+text(.5,1*conv,'50+')
+arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)),lwd=2.5)
+
+@
+\end{subfigure}
+\caption{Reassignment transition plot for study year 2.}
+\label{fig:trans2}
+\end{figure}
+
 To address these questions, we focus on the units in the standard
@@
 times.
-They are listed in the order they appear in the standard Algebra I curriculum.
+They are numbered according to the order they appear in the standard Algebra I curriculum.
 On the right, under ``To,'' are the top ``receiving'' units: the units
@@
 For instance, in Figure \ref{fig:trans1}, the arrow from
-``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
+``4-Quadrant Linear Graphs'' on the left to ``4-Quadrant Linear
 Graphs'' on the right is fairly thick, since
-\Sexpr{round(trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']/
-  sum(trans1['4-Quadrant Linear Graphs',])*100)}\%
-of the reassignments begun in ``4-Quadrante Linear Graphs'' end in the
+\Sexpr{round(trans1o['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']/
+  sum(trans1o['4-Quadrant Linear Graphs',])*100)}\%
+of the reassignments begun in ``4-Quadrant Linear Graphs'' end in the
 same unit.
 Yet it is also fairly faint, since it only represents
-\Sexpr{trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']}
+\Sexpr{trans1o['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']}
 reassignments.
 
-<<secOrder,include=FALSE,dependson=c('data','setup')>>=
-#### what's the next section each student works?
-secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
-
-secOrder$cp <- secOrder$status=='changed placement'
-secOrder$mast <- secOrder$status=='graduated'
-
-secOrderCP <- subset(secOrder,status=='changed placement')
-
-secOrderCP$Unit <- secOrderCP$unit
-secOrderCP$Unit[grep('unit-conversions',secOrderCP$Unit)] <- 'unit-conversions'
-
-secOrderCP$unitName <- factor(secOrderCP$Unit,levels=units)
-levels(secOrderCP$unitName) <- UnitName
-
-secOrderCP$nextUnit[grep('unit-conversions',secOrderCP$nextUnit)] <- 'unit-conversions'
-
-secOrderCP$nextUnitName <- factor(secOrderCP$nextUnit,levels=units)
-levels(secOrderCP$nextUnitName) <- UnitName
-
-@
-
-\begin{figure}
-  \begin{subfigure}{5in}
-<<transition1,fig.height=3,fig.width=5,dependson='secOrder'>>=
-
-trans1 <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
-
-## top senders and receivers
-cols <- which(colSums(trans1)>30)
-rows <- which(rowSums(trans1)>30)
-
-cols2 <- cols
-rows2 <- rows
-
-## top 2 receivers for each sender, vice versa
-for(r in rows)
-    while(sum(trans1[r,cols2])/sum(trans1[r,])<0.75){
-    nextBest <- which.max(trans1[r,seq(ncol(trans1))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans1)==names(nextBest)))
-}
-
-for(cc in cols){
-    while(sum(trans1[rows2,cc])/sum(trans1[,cc])<0.75){
-        nextBest <- which.max(trans1[seq(nrow(trans1))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans1)==names(nextBest)))
-    }
-
-}
-
-cols <- sort(unique(c(cols2,rows2)))
-rows <- sort(unique(rows2))
-
-trans1 <- trans1[rows,cols]
-
-propTotCP1 <- sum(trans1)/sum(secOrderCP$year==1)
-
-uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
-nsec <- length(uuu)
-
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
-text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
-
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)
-
-for(rr in rownames(trans1))
- for(cc in colnames(trans1)){
-#     if(trans1[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans1[rr,cc]/sum(trans1[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans1[rr,cc]>50,1,trans1[rr,cc]/50)))
-}
-
-@
-\end{subfigure}
-\begin{subfigure}{1in}
-<<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-conv <- (nsec+2)/11
-text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
-text(1,9*conv,'Thickness:\n Percent',cex=0.8)
-text(.5,7.5*conv,'25%')
-text(.5,6.5*conv,'50%')
-text(.5,5.5*conv,'75%')
-arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
-text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
-text(.5,3*conv,'10')
-text(.5,2*conv,'25')
-text(.5,1*conv,'50+')
-arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
-
-@
-\end{subfigure}
-\caption{Reassignment transition plot for study year 1. The units on
-  the left of the plot are the Algebra I units that ended in reassignment at
-  least 30 times in year 1. The units on the right are those that were
-  the destination of at least 30 reassignments. Also included on the
-  left are the top two sending units for each of the units on the
-  right, and also included on the right are all the units on the left,
-  along with their top two receiving units. The units are listed
-  following their order in the standard Algebra I curriculum. There is an arrow from a
-  unit on the left to a unit on the right if a student was assigned
-  from the unit on the left to the one on the right. The thickness of
-  the arrows is proportional to the percentage of all reassignments from
-  the sending unit that ended in that receiving unit. The darkness of
-  the arrows is proportional to the number of reassignments from the
-  sending unit to the receiving unit.}
-\label{fig:trans1}
-\end{figure}
-
 Inspecting the figures shows that the most common pattern is for
-students to be reassigned to the next unit in the curriculum.
-On the other hand, it is rare for students to be reassigned
-
-\begin{figure}
-  \begin{subfigure}{5in}
-<<transition2,fig.height=5,fig.width=5,dependson='secOrder'>>=
-
-trans2 <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
-
-## top senders and receivers
-cols <- which(colSums(trans2)>30)
-rows <- which(rowSums(trans2)>30)
-
-cols2 <- cols
-rows2 <- rows
-
-## top 2 receivers for each sender, vice versa
-for(r in rows)
-    while(sum(trans2[r,cols2])/sum(trans2[r,])<0.75){
-    nextBest <- which.max(trans2[r,seq(ncol(trans2))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans2)==names(nextBest)))
-}
-
-for(cc in cols){
- #   print('new col')
-  #  print(cc)
-    while(sum(trans2[rows2,cc])/sum(trans2[,cc])<0.75){
-        nextBest <- which.max(trans2[seq(nrow(trans2))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans2)==names(nextBest)))
-    }
-
-}
-
-cols <- sort(unique(c(cols2,rows2)))
-rows <- sort(unique(rows2))
-
-trans2 <- trans2[rows,cols]
-
-propTotCP2 <- sum(trans2)/sum(secOrderCP$year==1)
-
-nsec <- nrow(trans2)
-uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans2),rownames(trans2))]
-nsec <- length(uuu)
-
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
-text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
-
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)
-
-for(rr in rownames(trans2))
- for(cc in colnames(trans2)){
-#     if(trans2[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
-}
-@
-\end{subfigure}
-\begin{subfigure}{1in}
-<<trans2legend,fig.width=1,fig.height=3,dependson='transition2'>>=
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-conv <- (nsec+2)/11
-text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
-text(1,9*conv,'Thickness:\n Percent',cex=0.8)
-text(.5,7.5*conv,'25%')
-text(.5,6.5*conv,'50%')
-text(.5,5.5*conv,'75%')
-arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
-text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
-text(.5,3*conv,'10')
-text(.5,2*conv,'25')
-text(.5,1*conv,'50+')
-arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
-
-@
-\end{subfigure}
-\caption{Reassignment transition plot for study year 2.}
-\label{fig:trans2}
-\end{figure}
+students to be reassigned to the next unit in the curriculum---this pattern comprises 
+\Sexpr{round(sum(diag(trans1o[,-1]))/sum(trans1o)*100)}\% of the reassignments in year 1 and 
+\Sexpr{round(sum(diag(trans2o[,-1]))/sum(trans2o)*100)}\% of those in year2.
+On the other hand, it is rare for students to be reassigned within the same unit 
+(these reassignments account for \Sexpr{round(sum(diag(trans1o))/sum(trans1o)*100)}\% and \Sexpr{round(sum(diag(trans2o))/sum(trans2o)*100)}\% in the two years).
+It is also rare for students to be reassigned to units earlier in the curriculum (comprising 
+\Sexpr{round(sum(trans1o[lower.tri(trans1o)])/sum(trans1o)*100)}\% and
+\Sexpr{round(sum(trans2o[lower.tri(trans2o)])/sum(trans2o)*100)}\%).
+
+The transition plots also reveal some interesting cases worth highlighting.
+In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions.'' on the part of the reassigning teachers.
+Similarly, 
+\Sexpr{trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']} students 
+(\Sexpr{round(trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']/sum(trans1o['Ind. Variables in Lin. Mod.',])*100)}\%)
+of the students reassigned from the section ``Ind. Variables in Lin. Mod.'' were placed 13 units later in ``Systems of Lin. Eq.''
+and all \Sexpr{trans1o['Quad. Mod. & Area','Exponents']} of students reassigned from ``Quad. Mod. \& Area'' were placed in ``Exponents.''
+This suggests that some teachers may have considered these units, ``Systems of Lin. Eq.'' and ``Exponents'' to contain particularly important material.
+
+Since the total number of reassignments was higher in year 2, the corresponding plot (Figure \ref{fig:trans2}) is larger and more complex. 
+It is also more common for students to be reassigned to units other than the next unit in the sequences.
+This may be partly due to the proliferation of customized curricula. 
+Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Eq. w/ Var's Both Sides,'' were common destinations from a wide variety of earlier units, suggesting a strong interest in those units. 
+The majority (\Sexpr{round(trans2o['Probability','Probability']/sum(trans2o['Probability',])*100)}\%)
+of the students reassigned from ``Probability'' were placed in another section of the same unit, perhaps suggesting problems with some of that unit's sections; in fact, all of the students reassigned within the ``Probability'' unit were reassigned from one its first three sections (of seven).
+Finally, \Sexpr{trans2['Pythagorean Theorem','Product Rule for Exponents']} 
+(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignements from sections in the ``Pythagorean Theorem'' unit ended in the earliear ``Product Rule for Exponents'' sections.
+
+In sum, the most prevalent pattern was for students to be reassigned to the following unit, suggesting that teachers are mostly interested in helping lagging students progress.
+On the other hand, a number of examples of other patterns---students moving within the same unit, or to units out of sequence---appear as well, suggesting that some teachers may be finely manipulating their students' curricula.
 
@@
 effects can be valuable, if interpreted cautiously.
-They may help guide research
-
-\subsection{Effects on Subsequent Mastery}
-<<>>=
-@
-
-\subsection{Effects on Post-Test Scores}
+In the absence of evidence from randomized trials, observational
+studies can help guide intuition, future research, and even---when
+combined with other relevent information and theory---practice.
+
+<<effects,include=FALSE,dependson=c('data','setup')>>=
+cpDat <- data%>%filter(!is.na(status)& status!='final_or_incomplete')%>%
+ group_by(field_id,year,state,section,unit,race,sex,grade,spec_speced,xirt,spec_gifted,
+  spec_esl,frl,pretest,y_yirt,classid2,schoolid2)%>%
+ summarize(status=max(status,na.rm=TRUE),date=max(date),totalTime=sum(total_t1,na.rm=TRUE),
+  ttNA=sum(is.na(total_t1)),nprob=n(),probNA=sum(is.na(Prob1)))%>%ungroup()%>%
+ group_by(field_id,year,state,race,sex,grade,spec_speced,xirt,spec_gifted,spec_esl,frl,pretest,y_yirt,
+  classid2,schoolid2)%>%
+ summarize(nsec=n(),nprob=sum(nprob,na.rm=TRUE),totalTime=sum(totalTime,na.rm=TRUE),
+  timeNA=sum(ttNA>0),probNA=sum(probNA>0),ncp=sum(status=='changed placement'),
+  nprom=sum(status=='promoted'))%>%mutate(gainScore=y_yirt-pretest)%>%droplevels()
+
+## mode imputation:
+for(vv in c('race','sex','grade','spec_speced','spec_gifted','spec_esl','frl')){
+    num <- is.numeric(cpDat[[vv]])
+    cpDat[[paste0(vv,'MIS')]] <- is.na(cpDat[[vv]])
+    cpDat[[vv]][is.na(cpDat[[vv]])] <- names(which.max(table(cpDat[[vv]])))
+    if(num) cpDat[[vv]] <- as.numeric(cpDat[[vv]])
+
+}
+
+cpDat$grade <- factor(ifelse(cpDat$grade==9,'9','10+'))
+levels(cpDat$race) <- list(White=c('WHITE NON-HISPANIC','ASIAN / PACIFIC ISLANDER'),Black=c('BLACK NON-HISPANIC','OTHER RACE / MULTI-RACIAL'),Hispanic=c('HISPANIC','AMERICAN INDIAN / ALASKAN NATIVE'))
+
+cpDat <- mutate(cpDat,totalTime=totalTime/3600000,totalTime=ifelse(totalTime<0,NA,totalTime),
+                totalTime=ifelse(totalTime>360,NA,totalTime))
+
+cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=4,'4+',cpDat$ncp))
+cpDat$everCP <- cpDat$ncp>0
+
+exirt <- data[match(cpDat$field_id,data$field_id),grep('Exirt2',names(data))]
+names(exirt) <- gsub('_0','_',names(exirt))
+for(n in names(exirt)) cpDat[[n]] <- exirt[[n]]
+
+@
+<<models,include=FALSE,dependson='effects'>>=
+#### models!!
+## ind var: ncp
+addCovs <- .~.+pretest+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl+frlMIS
+mod1 <- lm(y_yirt~ncp+classid2,data=cpDat)
+mod1.1 <- update(mod1,addCovs)
+
+## ind var: ncpCat
+mod2 <- lm(y_yirt~ncpCat+classid2,data=cpDat)
+mod2.1 <- update(mod2,addCovs)
+
+## ind var: everCP
+mod3 <- lm(y_yirt~everCP+classid2,data=cpDat)
+mod3.1 <- update(mod3,addCovs)
+
+miMod <- function(mod){
+  if(inherits(mod,'lm')) CCC <- coef
+  if(inherits(mod,'lmerMod')) CCC <- fixef
+    ests <- NULL
+    covs <- list()
+    for(i in 1:20){
+        formNew <- as.formula(paste0('.~.-pretest+Exirt2_',i))
+        modNew <- update(mod,formNew)
+        ests <- cbind(ests,CCC(modNew))
+        covs[[i]] <- vcov(modNew)
+    }
+    return(list(ests=ests,covs=covs))
+}
+miPool <- function(fitMods){
+    ests <- fitMods$ests
+    covs <- fitMods$covs
+    interest <- grep('CP',rownames(ests),ignore.case=TRUE)
+    SEs <- sqrt(rowMeans(do.call('cbind',lapply(covs,function(cc) diag(as.matrix(cc[interest,interest])))))+
+                1.05*apply(rbind(ests[interest,]),1,var))
+    ests <- rowMeans(rbind(ests[interest,]))
+    cbind(ests,SEs)
+}
+
+mi <- function(mod){
+    miPool(miMod(mod))
+}
+
+adjustedEsts1 <- list()
+for(i in 1:3){
+  mm <- paste0('mod',i)
+  mm1 <- paste0('mod',i,'.1')
+  adjustedEsts1[[mm]] <- mi(get(mm))
+  adjustedEsts1[[mm1]] <- mi(get(mm1))
+}
+@
+<<heterogeneityMods,include=FALSE,dependson='models'>>=
+### heterogeneity
+## year
+cpDat$Year=as.factor(cpDat$year)
+mod4 <- lmer(y_yirt~everCP*Year+state+(1|classid2)+(1|schoolid2),data=cpDat)
+mod4.1 <- update(mod4,addCovs)
+
+## classroom
+mod5 <- lmer(y_yirt~everCP+year+state+(everCP|classid2)+(1|schoolid2),data=cpDat)
+mod5.1 <- update(mod5,addCovs)
+@
+
+\begin{figure}
+<<cpYyear,fig.width=6,fig.height=4,dependson='effects'>>=
+levels(cpDat$Year) <- list(`Year 1`='1',`Year 2`='2')
+ggplot(cpDat,aes(ncp,gainScore))+geom_jitter(height = 0,width=.3)+geom_boxplot(aes(group=ncp),alpha=0.5)+geom_smooth(method='loess')+facet_grid(Year~.)+labs(x='# of Reassignments',y='Gain Score (posttest - pretest)') + scale_x_continuous(breaks=seq(0,max(cpDat$ncp)))
+@
+\caption{Students' Gain scores (post-test minus pre-test) versus the number of times the were reassigned over the course of the year (jittered), with a Loess smoother added.}
+\label{fig:cpYyear}
+\end{figure}
+
+\subsection{Average Effects}
+Figure \ref{fig:cpYyear} shows students gain scores---the difference between their post-test and pre-test scores---as a function of the number of times they were reassigned.
+(The number of reassignments was jittered---random noise was added---to avoid overplotting.)
+Overall, the relationship between the two variables is positive.
+Nevertheless, some non-linearity seems to be present, especially in year 2.
+Further, the distribution of the number of reassignments is right-skewed---again, especially in year 2. 
+Care in modeling the number of reassignments, then, is especially important---observations from students reassigned an unusually large number of times can exert undo influence on a regression model and generate misleading results, particularly in the presence of non-monotonic relationships.
+We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for stundents who were. 
+Next, $R^{cat}$ defines a categorical variable taking the values $R^{cat}=0,1,2,3,4+$ for students who were reassigned 0, 1, 2, 3 or four or more times, respectively. 
+Finally, $R^{num}$ is the raw number of reassignments, which we include for completeness.
+
+We used linear models to estimate the effect of reassignment, regressing post-test scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
+Since this is unlikely to be the case (even approximately), we ran a second set of models including student level covariates: 
+pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
+\Sexpr{round(mean(is.na(cpDat$xirt))*100)}\%).
+To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2013effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligability.
+
+\begin{table}
+\begin{tabular}{r|c|c|c|}
+&\multicolumn{3}{c}{Parametrization}\\
+&$R^{bin}$&$R^{cat}$&$R^{num}$\\
+&\makecell[c]{Effect of\\ $\ge 1$ Reassignment}&\makecell[l]{Effect of\\  \# Reassignments:}& \makecell[c]{Effect per\\ Reassignment:}\\
+\hline
+<<effectResults,results='asis',dependson='models'>>=
+## model 3
+cat('\\makecell[r]{No\\\\Covariates}&')
+cat(round(adjustedEsts1[['mod3']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod3']][1,'SEs'],2),'&')  
+## model 2
+res <- adjustedEsts1[['mod2']]
+cat('\\makecell[l]{')
+for(rr in 1:(nrow(res)-1)) 
+  cat(rr,':  ',round(res[rr,1],2),'$\\pm$',round(2*res[rr,'SEs'],2),'\\\\')
+cat(nrow(res),'+: ',round(res[nrow(res),1],2),'$\\pm$',round(2*res[nrow(res),'SEs'],2),'}&')
+## model 1
+    cat(round(adjustedEsts1[['mod1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod1']][1,'SEs'],2),'\\\\')
+cat('\n')
+cat('\\hline')
+###################################
+cat('\\makecell[r]{Covariate\\\\ Adjusted}&')
+cat(    round(adjustedEsts1[['mod3.1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod3.1']][1,'SEs'],2),'&')
+## model 2
+cat('\\makecell[l]{')
+res <- adjustedEsts1[['mod2.1']]
+for(rr in 1:(nrow(res)-1)) cat(rr,': ',round(res[rr,1],2),'$\\pm$',round(2*res[rr,'SEs'],2),'\\\\')
+cat(nrow(res),'+: ',round(res[nrow(res),1],2),'$\\pm$',round(2*res[nrow(res),'SEs'],2),'}&')
+## model 1
+cat(
+    round(adjustedEsts1[['mod1.1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod1.1']][1,'SEs'],2),'\\\\')
+
+@
+\hline
+\end{tabular}
+\caption{Estimates of effects of reassignment on posttests, with 95\% margins of error.}
+\label{effectResults}
+\end{table}
+
+The results are reported in Table \ref{effectResults}.
+All the effect estimates are negative, indicating that reassigning students may hurt their algebra learning.
+The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times. 
+The magnitudes of the effects are rather large: \citet{pane2013effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests']*100)}\% of that.
+
+But what of unmeasured confounding?
+\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on benchmarking from observed confounders.
+In order to confound the causal relationship between reassignment and post-tests, a confounder would have to predict both.
+Roughly speaking, the idea is to widen the confidence interval from a causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
+These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omssion of a confounder.
+As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of post-test scores.
+The sensitivity interval for the effect of being reassigned at least once on post-test scores, 
+\begin{table}
+\begin{tabular}{rrl}
+Benchmark&\multicolumn{2}{c}{Sensitivity Interval}\\
+<<sens,dependson='models'>>=
+source('hhh.r')
+X <- as.data.frame(model.matrix(mod3.1)[,-1])
+Ts <- Tz(X = X,treatment = 'everCPTRUE')
+Rhos <- vapply(names(X)[-grep('classid2',names(X))],
+               function(nn) partialCor(cpDat$y_yirt,X,nn),1)[-1]^2
+se <- adjustedEsts1[['mod3.1']][1,2]
+cat(names(Rhos)[1],'&',round(adjustedEsts1[['mod3.1']][1,'ests'],2),'&$\\pm$',
+    round(se*MEmult(abs(Ts[names(Rhos)[1]]),R = Rhos[1],df = mod3.1$df.residual,1.96),2),'\\\\')
+for(i in 2:length(Rhos))
+  cat(gsub('_','',names(Rhos)[i]),'&&$\\pm$',round(se*MEmult(T=abs(Ts[names(Rhos)[i]]),R = Rhos[i],df = mod3.1$df.residual,1.96),2),'\\\\')
+@
+\end{tabular}
+\caption{Sensitivity intervals for the $R^{bin}$ model.}
+\label{tab:sens}
+\end{table}
 

### 2017-10-23 cpEffectSimp.r

@@ -1,6 +1,7 @@
 library(dplyr)
 library(lme4)
 library(splines)
+library(ggplot2)
 load('cpPaper.RData')
 
 cpDat <- data%>%filter(!is.na(status)& status!='final_or_incomplete')%>%
@@ -18,16 +19,20 @@ cpDat <- data%>%filter(!is.na(status)& status!='final_or_incomplete')%>%
 
 ## mode imputation:
 for(vv in c('race','sex','grade','spec_speced','spec_gifted','spec_esl','frl')){
-    cpDat[[paste0(vv,'Imp')]] <- cpDat[[vv]]
-    cpDat[[paste0(vv,'Imp')]][is.na(cpDat[[vv]])] <- names(which.max(table(cpDat[[vv]])))
+    num <- is.numeric(cpDat[[vv]])
+    cpDat[[paste0(vv,'MIS')]] <- is.na(cpDat[[vv]])
+    cpDat[[vv]][is.na(cpDat[[vv]])] <- names(which.max(table(cpDat[[vv]])))
+    if(num) cpDat[[vv]] <- as.numeric(cpDat[[vv]])
+
 }
 
-
+cpDat$grade <- factor(ifelse(cpDat$grade==9,'9','10+'))
+levels(cpDat$race) <- list(White=c('WHITE NON-HISPANIC','ASIAN / PACIFIC ISLANDER'),Black=c('BLACK NON-HISPANIC','OTHER RACE / MULTI-RACIAL'),Hispanic=c('HISPANIC','AMERICAN INDIAN / ALASKAN NATIVE'))
 
 cpDat <- mutate(cpDat,totalTime=totalTime/3600000,totalTime=ifelse(totalTime<0,NA,totalTime),
                 totalTime=ifelse(totalTime>360,NA,totalTime))
 
-cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=3,'3+',cpDat$ncp))
+cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=4,'4+',cpDat$ncp))
 cpDat$everCP <- cpDat$ncp>0
 
 exirt <- stud[match(cpDat$field_id,stud$field_id),grep('Exirt2',names(stud))]
@@ -36,29 +41,50 @@ for(n in names(exirt)) cpDat[[n]] <- exirt[[n]]
 
 
 #### models!!
-mod1 <- lmer(gainScore~ncp+state+year+(1|classid2)+(1|schoolid2),data=cpDat)
[…]

### 2017-10-23 hhh.r

@@ -1,0 +1,164 @@
+library(arm)
+library(optmatch)
+
+between <- function(x,a,b) x>=a & x<=b
+## calibrate rho and T
+
+partialCor <- function(y,X,varb){
+    ex <- extract(X,varb,TRUE)
+    yres <- residuals(lm(y~.,data=ex$Xdot,na.action=na.exclude))
+    Xres <- residuals(lm(ex$x~.,data=ex$Xdot,na.action=na.exclude))
+    cor(yres,Xres,use='comp')
+}
+
+## takes logistic regression model and spits out
+##treatment and covariates vectors
+mod2dat <- function(PSmod){
+ Zname <- as.character(PSmod$formula[2])
+ Covnames <- names(coef(PSmod))[-1]
+ Covnames <- intersect(Covnames,names(PSmod$data))
+ X <- data.frame(PSmod$data[,Covnames])
+ names(X)<-Covnames
+ X$Z <- PSmod$data[[Zname]]
+ X
+}
+
+rhos <- function(y,PSmod,X,treatment){
+
+    if(!missing(treatment)){
+        if(is.integer(treatment)) names(X)[treatment] <- 'Z'
+        if(is.character(treatment)) names(X)[which(names(X)==treatment)] <- 'Z'
+    }
+
+    if(!missing(PSmod))
+        X <- mod2dat(PSmod)
+
+    rs <- NULL
+    if(ncol(X)==2) rs=cor(X[names(X)!='Z'],y)
+    else{
+       for(i in 1:ncol(X)){
+           if(names(X)[i]!='Z')
+       	      rs <- c(rs,partialCor(y,X,i))
+    }}
+    names(rs) <- names(X)[names(X)!='Z']
+    rs^2
+}
+
+newPS <- function(ex){
+    psMod <- bayesglm(ex$Z~as.matrix(ex$Xdot),family=binomial(logit))
+    print(summary(psMod))
+    psMod$linear
+}
+
+newMatch <- function(X,mBayes,varb){
+    X <- X[!is.na(mBayes),]
+    ex <- extract(X,varb,FALSE)
+    PS <- newPS(ex)
+    match(ex$Z,PS)
+}
+
+extract <- function(X,varb,includeZ=T){
+  varbCol <- ifelse(is.character(varb),which(names(X)==varb),varb)
+  Xdot <- X[,-varbCol]
+  x <- X[,varbCol]
+
+  Z <- X$Z
+  if(!includeZ)
[…]

### 2017-10-23 mastNext.r

@@ -1,5 +1,7 @@
 library(lme4)
 library(dplyr)
+library(lubridate)
+library(sandwich)
 
 load('cpPaper.RData')
 
@@ -9,6 +11,9 @@ secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp) )%>%group_by(
 
 secOrder$cp <- secOrder$status=='changed placement'
 secOrder$mast <- secOrder$status=='graduated'
+secOrder$month <- months(secOrder$date)
+
+secOrder$prevCP <- secOrder$prevStatus=='changed placement'
 
 cpdStuds <- unique(data$field_id[!is.na(data$status) & data$status=='changed placement'])
 
@@ -23,47 +28,36 @@ mnDat25 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat
 mnDat50 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>50])
 mnDat100 <- subset(mnDat,mnDat$secUnit%in%names(table(mnDat$secUnit))[table(mnDat$secUnit)>100])
 
-mn1ols25 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat25,subset=year==1)
-mn1ols50 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat50,subset=year==1)
-mn1ols100 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat100,subset=year==1)
+mn1ols25 <- lm(mast~prevCP+as.factor(field_id)+as.factor(secUnit),data=mnDat25,subset=year==1)
+mn1ols50 <- lm(mast~prevCP+as.factor(field_id)+as.factor(secUnit),data=mnDat50,subset=year==1)
+mn1ols100 <- lm(mast~prevCP+as.factor(field_id)+as.factor(secUnit),data=mnDat100,subset=year==1)
 
 print(head(coef(mn1ols25)))
 print(head(coef(mn1ols50)))
 print(head(coef(mn1ols100)))
 
-mn2ols25 <- lm(mast~prevStatus+as.factor(field_id)+as.factor(secUnit),data=mnDat25,subset=year==2)
[…]

### 2017-10-23 paper.Rnw
finishing (-conclusion)

@@
 \title{Mastery Learning in Practice:\\
-A Descriptive Analysis of Data from the  Cognitive Tutor Algebra I Effectiveness Trial}
+A (Mostly) Descriptive Analysis Based on the Cognitive Tutor Algebra I Effectiveness Trial}
 
@@
 <<data,include=FALSE>>=
-if(file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
+if('cpPaper.RData'%in%list.files() & file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
     load('cpPaper.RData')
@@
 \section{Introduction}
-Mastery learning sits at the foundation of intelligent tutoring systems.
-The philosophy of mastery learning assumes a well-structured curriculum, and posits that students progress within the curriculum as they master its skills.
-The Cognitive Tutor Algebra I (CTAI) system, developed by Carnegie Learning, Inc., is one of the best studied and best regarded examples of modern educational software.
+Mastery learning sits at the foundation of intelligent tutoring
+systems \citep[e.g.][]{corbett2001cognitive,wenger2014artificial}.
+The philosophy of mastery learning assumes a well-structured
+curriculum, and posits that students progress within the curriculum as
+they master its skills
+\citep{bloom1968learning,kulik1990effectiveness}. The Cognitive Tutor
+Algebra I (CTAI) system, developed by Carnegie Learning, Inc., is one of the best studied and best regarded examples of modern educational software.
 It is a blended learning system for teaching algebraic concepts and principles, to middle and high school students, including both textbook materials and software.
 The software component of the curriculum allows students to progress at their own pace and receive individualized feedback on their performance.
-A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students’ scores on an Algebra-I post-test by about one fifth of a standard deviation (Pane, et al.
-2013).
-CTAI’s success in this experiment would seem to vindicate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
+A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students' scores on an Algebra-I posttest by about one fifth of a standard deviation \citep{pane2014effectiveness}.
+CTAI's success in this experiment would seem to vindicate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
 
@@
 At the request of educators, it can be customized by altering what units or sections are included (including, possibly, material from a different standard curriculum such as geometry), as well as their sequence, to conform to local or state standards.
-Further, teachers have the option of moving students within the curriculum, regardless of the software’s estimate of their skill mastery.
-
-This article examines teachers’ and schools’ non-adherence to the standard, mastery-based CTAI curriculum, using data from the RAND study, a seven-state randomized controlled trial of CTAI in high schools and middle schools.
+Further, teachers have the option of moving students within the curriculum, regardless of the software's estimate of their skill mastery.
+
+This article examines teachers' and schools' non-adherence to the standard, mastery-based CTAI curriculum, using data from the RAND study, a seven-state randomized controlled trial of CTAI in high schools and middle schools.
 The study found a significant positive effect of CTAI in high schools, during their second year of implementation but not the first.
@@
 Thus, the data from the study reflect typical software usage.
-Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects (Sales \& Pane, 2015, Sales et al.
-2016).
-
-Here we extend the previous findings to investigate patterns of change in the sequence of topics presented to students.
-As noted earlier, schools could use a standard curriculum, a customized curriculum, or switch from one curriculum to another.
-Each curriculum consisted of a sequence of units, which was comprised of a sequence of sections.
-For example, the standard CTAI Algebra I curriculum in Year 1 comprised of 40 units covering approximately 170 sections.
-The
-number of sections within each unit varied.
-This study attempts to elucidate teachers' goals in reassigning students.
+Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects \citep{sales2015exploring,sales2016student}.
+\citet{sales2017role} found that students more likely to master worked
+sections of the CTAI software may experience \emph{smaller} effects
+than those less likely to achieve mastery, casting doubt on the role
+of mastery learning as a mechanism for the treatment effect.
+
+Here we extend the previous findings to describe, in detail, the ways
+in which schools, teachers, and students violate mastery learning.
+We will begin with short discussions of the RAND effectiveness trial and the usage
+data it produced.
+Sections \ref{sec:curricula}---\ref{sec:order} will describe overall
+patterns of usage.
+First, we will discuss standard and customized CTAI curricula used in the study (\ref{sec:curricula}).
+Next, in Section\ref{sec:usage}, we will describe patterns in the amount CTAI usage, showing that
+it varied widely between states, between years, and between students.
+In particular, the amount of usage decreased from years 1 to
+2---though more in some states than others.
+Then in section \ref{sec:order} we will describe how the amount of usage
+changed---which units of the CTAI curriculum were worked more and less
+from years 1 to 2---and find that order in which students worked
+CTAI's units varied across years. We show that this change is due
+mostly, but not completely, to the presence of customized curricula in
+year 2.
+
+In Section \ref{sec:mastery} we will describe patterns of mastery in general,
+and in Section \ref{sec:cp} we will delve deeply into ``reassignment'': the process in which a
+teacher moves a student out of a section he or she has not (yet)
+mastered into a new section.
+In particular, we will attempt to elucidate teachers' goals in reassigning students.
 One hypothesis is the need for teachers push ahead students who were falling behind, i.e.
@@
 This hypothesis may lead to an increase in reassignments as the exam approaches.
-Overall, understanding patterns of reassignment is the focus of this paper, along with insights into how teachers were using the software component of CTAI.
-
-\section{The RAND Effectiveness Trial}
-The study to measure the effectiveness of CTAI included 7 states, 73 high schools, and 74 middle schools with nearly 18,700 high school students and 6,800 middle school students.
-Schools were enrolled in a total of 52 school districts that were distributed between urban, suburban, and rural areas.
-Schools were matched on a set of covariates, and then randomly assigned to the treatment or control group.
-Schools in the control group continued with their current algebra curriculum, and schools in the treatment group used Carnegie Learning’s curriculum which includes textbook materials as well as CTAI.
-Each school participated for two years, with a different cohort of students taking algebra the second year (with a small fraction of students present in the study both years because they repeated algebra).
-It should be noted that this study did not include statewide implementations; the study results cannot be generalized to all schools within the state.
-In some states, one large school district participated, while in other states, a set of smaller school districts participated.
-The states included Alabama (AL), Connecticut (CT), Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and Texas (TX).
-Each state participated in both the middle school and high school arms of the study, except AL, which participated only in the middle school arm.
-The current study focuses on high school students only, due to the large number of students in this cohort and the significant positive treatment effect previously found for this cohort.
-All discussion that follows refers exclusively to the high school data.
+We find evidence of all three motivations for reassignment, varying
+across classrooms, schools, states, and study  years.
+
+Finally, in Section \ref{sec:effects} will will give
+quasi-experimental estimates of the effects of reassignment on
+students' posttest scores.
+We find that reassignment probably decreases student learning,
+though the effects vary widely across classrooms.
+Section \ref{sec:discussion} will conclude.
+
+\section{The RAND Effectiveness Trial}\label{sec:RANDtrial}
+The study to measure the effectiveness of CTAI included 7 states, 73
+high schools, and 74 middle schools with nearly 18,700 high school
+students and 6,800 middle school students. Schools were enrolled in a
+total of 52 school districts that were distributed between urban,
+suburban, and rural areas. Schools were matched on a set of
+covariates, and then randomly assigned to the treatment or control
+group. Schools in the control group continued with their current
+algebra curriculum, and schools in the treatment group used Carnegie
+Learning's curriculum which includes textbook materials as well as
+CTAI. Each school participated for two years, with a different cohort
+of students taking algebra the second year (with a small fraction of
+students present in the study both years because they repeated
+algebra). It should be noted that this study did not include statewide
+implementations; the study results cannot be generalized to all
+schools within the state. In some states, one large school district
+participated, while in other states, a set of smaller school districts
+participated. The states included Alabama (AL), Connecticut (CT),
+Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and
+Texas (TX). Each state participated in both the middle school and high
+school arms of the study, except AL, which participated only in the
+middle school arm. The current study focuses on high school students
+only. %, due to the large number of students in this cohort and the
+% significant positive treatment effect previously found for this
+% cohort. All discussion that follows refers exclusively to the high
+% school data.
 
 There are some limitations to the available data for this study.
-In particular, the developer was able to provide the intended sequence of units/sections for only a subset of the customized curricula.
-Data from some schools, and some students within schools, were missing either because the log files were not retrievable, or because of an imperfect ability to link log data to other study data files.
-In addition, there was missing completion dates for many customized curricula; making it difficult to analyze the customized curricula when any sort of date was needed, such as time to complete a section.
-
-\section{Standard and Customized Curricula}
+Log data from some schools, and some students within schools, were
+missing either because the log files were not retrievable, or because
+of an imperfect ability to link log data to other study data files.
+For this reason, this study uses only data from on the
+\Sexpr{n_distinct(data[['schoolid2']])} schools for which at least 80\%
+students in both study years appear in the log data file.
+This sample includes \Sexpr{n_distinct(data[['field_id']])} students, around
+\Sexpr{round(n_distinct(data[['field_id']])/n_distinct(stud[['field_id']]))*100}\%
+of the total high-school sample.
+Table \ref{tab:nByState} gives the number of students in each state of
+the sample in each year.
+The states in the table are ordered by the total number of students
+they represent in the sample; they will appear in this order in
+all of the forthcoming tables and figures.
+Some figures will only show data from a subset of states; since so few
+students were in New Jersey, it will be excluded from almost all
+state-by-state comparisons (but included analyses that pool across states).
+<<nByState,dependson='data'>>=
+xtable(data%>%group_by(state,year)%>%summarize(n=n_distinct(field_id))%>%dcast(year~state),
+       caption='Numbers of students in the sample by state and study year',label='tab:nByState')
+@
+
+In this smaller sample, \Sexpr{sum(data[['obsUsage']]==FALSE)}
+students do not appear in the log data; they may have not used the CTAI
+software at all, or may have been excluded from the log data for other
+reasons.
+Since we don't know which is true, we exclude these students from most
+analyses.
+
+It is likely that some usage data were missing, even for students who
+appear in the usage dataset.
+However, it is impossible to know in which cases these data were
+missing or why; for the most part, we ignore this problem, but it
+should be kept in mind nonetheless.
+
+\section{Standard and Customized Curricula}\label{sec:curricula}
 
@@
 
-Students' automatic progress throught the Cognitive Tutor (CT) software is governed by
+Students' automatic progress through the Cognitive Tutor (CT) software is governed by
 curricula---sequences of sections and units embedded in the software.
@@
 naturally, Algebra I.
-This came with three closely related varients, due to software releases.
+This came with three closely related variants, due to software releases.
 Students requiring more remediation were able to work on a less
@@
 in our dataset.
-In particular, Texas shifted almost entirely to customized curriclua
+In particular, Texas shifted almost entirely to customized curricula
 from years 1 to 2.
@@
 
-\section{Student Usage Across States and Years}
+\section{Student Usage Across States and Years}\label{sec:usage}
 
@@
 
-tab <- secByStud%>%group_by(Yr)%>%select(time,nprob,numSec,numUnit)%>%summarize_all(median,na.rm=TRUE)
+tab <- ungroup(secByStud)%>%group_by(Yr)%>%dplyr::select(time,nprob,numSec,numUnit)%>%summarize_all(median,na.rm=TRUE)
 tab$Yr <- NULL
@@
 
-Table \ref{tab:tab:medUsage} shows the median numbers of hours,
+Table \ref{tab:medUsage} shows the median numbers of hours,
 problems, sections, and units worked on by each student in the dataset
@@
 year than in the first, students in Michigan, Louisiana, and
-Connectuct increased their usage.
+Connecticut increased their usage.
 
@@
 The geographic variation in units worked mirrors the pattern in Figure
-\ref{fig:timeByStud}, with more usage in Kentudky, Michigan, and
+\ref{fig:timeByStud}, with more usage in Kentucky, Michigan, and
 Louisiana but less in Texas, Connecticut, and New Jersey.
@@
 except for Kentucky, where a decrease in IQR was accompanied by an
-incrase in the number of outliers.
+increase in the number of outliers.
 
@@
 We also evaluated the time spent per section across different curricula and states.
-Some records were omitted from this analysis: sections missing information on date of completion, as well as each student’s last section attempted because there is no completion date recorded for those sections.
+Some records were omitted from this analysis: sections missing information on date of completion, as well as each student's last section attempted because there is no completion date recorded for those sections.
 As presented in Figures 2 and 3 for Year 1 and 2, respectively, students in NJ increased the time spent per section from Year 1 to Year 2; however, there were relatively few students in the study from NJ.
@@
 
-Figures 4 and 5 present information for only sections that were “graduated” by students, that is, cases in which a student was judged by the software to have mastered the material in the section.
+Figures 4 and 5 present information for only sections that were mastered by students, that is, cases in which a student was judged by the software to have mastered the material in the section.
 Results are similar to the data presented in Figures 2 and 3.
@@
 This difference is also seen in KY and LA.
-If customized curriculum are more aligned with teachers’ instruction, graduating from a section may require fewer days than sections in the standard curricula, possibly because students are working on the same topics in class as they are in the software.
+If customized curriculum are more aligned with teachers' instruction, graduating from a section may require fewer days than sections in the standard curricula, possibly because students are working on the same topics in class as they are in the software.
 Figures 4 and 5 also show that, on average, CT and TX spent more days graduating from sections, regardless whether the section was part of a standard curricula or customized curricula.
-One explanation for the longer time periods in these states may be due to a teacher’s use of the software simultaneously with the textbook instruction.
+One explanation for the longer time periods in these states may be due to a teacher's use of the software simultaneously with the textbook instruction.
 In this case, the teachers could have switched back-and-forth as they instructed, hence extending the time worked by their students.
@@
 
-\section{Working Units in Order---Or Not}
-
-Overally, students used CT less in the second year than in the
+\section{Working Units in Order---Or Not}\label{sec:order}
+
+Overall, students used CT less in the second year than in the
 first.
@@
 \begin{figure}
+  \centering
 <<whichUnits,fig.height=4,fig.width=6,dependson=c('data','setup')>>=
@@
 
-In year 1, the curve is almost monontonically decreasing, as one would
+In year 1, the curve is almost monotonically decreasing, as one would
 expect if students adhered to the curriculum.
@@
 skipped ``Unit Conversions'' and ``1 step Lin. Eq.'' for ``1st
-Quadrent Linear Graphs'' and ``Ind. Variables in Lin. Mod.''
+Quadrant Linear Graphs'' and ``Ind. Variables in Lin. Mod.''
 Both of the latter sections were worked by a greater proportion of
@@
 greater proportion of students than any of the previous six sections.
-Presumeably teachers and administrators wanted students to focus on
+Presumably teachers and administrators wanted students to focus on
 that unit, perhaps because they found it to be particularly effective,
 because students tend to struggle with its main topic, or because its
-topic may figure prominantly in an upcoming standardized test.
+topic may figure prominently in an upcoming standardized test.
 
@@
   are divided between those attending schools using primarily a customized
-  curriculum and those using primariy the standard Algebra I curriculum.}
+  curriculum and those using primarily the standard Algebra I curriculum.}
 \label{fig:unitsWorkedCust}
@@
 Of course, this may have occurred in schools with customized
-curriculua as well---a possibility we will discuss in the next
+curricula as well---a possibility we will discuss in the next
 section.
@@
 For instance, one school prioritized units ``2 Step Lin. Eq.''
-and ``4−Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''
+and ``4-Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''
 
@@
 
-\section{Mastering the Material---Or Not}
+\section{Mastering the Material---Or Not}\label{sec:mastery}
 The central idea behind mastery learning is that students progress
@@
 next section. Students who exhaust a section's material without
-mastering its skills are “promoted” to the next section.
+mastering its skills are ``promoted'' to the next section.
 Students left uninterrupted follow the sequence of topics defined by
 their curriculum, ideally graduating from topics by showing mastery.
-However, teachers could modify a student’s path within the curriculum.
+However, teachers could modify a student's path within the curriculum.
 They could ``reassign'' students from their current sections to other
@@
 \begin{table}
- \begin{tabular}{rllllll|llllll}
-
-&   \multicolumn{6}{c}{Year 1}&\multicolumn{6}{c}{Year 2}\\
+  \centering
+ \begin{tabular}{rllllll}%|llllll}
+
+&   \multicolumn{6}{c}{Year 1}\\%&\multicolumn{6}{c}{Year 2}\\
 <<statusTab,dependson='overallStatus',results='asis'>>=
@@
 tab2 <- dcast(subset(statusStateYr2a,Yr=='Yr 2'&state!='NJ'),variable~state)
-tab <- cbind(tab1,tab2[-1])
+#tab <- cbind(tab1,tab2[-1])
 
@@
 
-tab$variable <- as.character(tab$variable)
+tab1$variable <- as.character(tab1$variable)
+tab2$variable <- as.character(tab2$variable)
 cat('&')
-cat(names(tab)[-1],sep='&')
+cat(names(tab1)[-1],sep='&')
 cat('\\\\')
-for(i in 1:nrow(tab)){
-    cat(tab[i,1],round(unlist(tab[i,-1])),sep='&')
+for(i in 1:nrow(tab1)){
+    cat(tab1[i,1],round(unlist(tab1[i,-1])),sep='&')
     cat('\\\\')
 }
-
+cat('&\\multicolumn{6}{c}{Year 2}\\\\ \n')
+for(i in 1:nrow(tab2)){
+    cat(tab2[i,1],round(unlist(tab2[i,-1])),sep='&')
+    cat('\\\\')
+}
 rownames(tab1) <- rownames(tab2) <- tab1$variable
@@
 \end{tabular}
-\caption{Percentages of worked sections that ended in each of the
+\caption{Numbers of worked sections that ended in each of the
   four possible outcomes, across states and study years.}
@@
 \Sexpr{round(tab1b['Reassigned','Overall']*100)}\% of sections in year
-1 eneded in reassignment, as did roughly
+1 ended in reassignment, as did roughly
 \Sexpr{round(tab2b['Reassigned','Overall']*100)}\% in year 2.
@@
 \begin{figure}
+  \centering
 <<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
@@
 levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
-statusCurr$Curr2 <- factor(statusCurr$Curr2,levels=c('Bridge-to-Algebra','Algebra I','Algebra I (Cust.)','>Algebra I'))
+statusCurr$Curr2 <-
+    factor(statusCurr$Curr2,levels=c('Bridge-to-Algebra','Algebra I','Algebra I (Cust.)','>Algebra I'))
+levels(statusCurr$Curr2)[1] <- 'Bridge-to-Alg.'
 print(ggplot(statusCurr,aes(Yr,value,fill=variable))+geom_col()+facet_grid(~Curr2)+
@@
 Mastery proportions do, indeed, depend on curriculum.
-Specifically, students mastered sections from more advanced curriculua
+Specifically, students mastered sections from more advanced curricula
 less frequently.
@@
   mean(statusOverall$status[statusOverall$Curriculum=='Algebra I']=='graduated',na.rm=T)*100)}\%
-of the time, and those from more advanced curriculua were mastered
+of the time, and those from more advanced curricula were mastered
 at a rate of
@@
   mean(statusOverall$status[statusOverall$Curriculum=='>Algebra I']=='graduated',na.rm=T)*100)}\%.
-This is unsurprising, since more advanced curriculua may be expected
+This is unsurprising, since more advanced curricula may be expected
 to be more challenging.
@@
 
-\section{Digging Deeper into Section Reassignment}
+\section{Digging Deeper into Section Reassignment}\label{sec:cp}
 The proportion of worked sections in our dataset ending in
 reassignment was small.
-Nevertheless, since reassignement represents the only mechanism by
+Nevertheless, since reassignment represents the only mechanism by
 which individual teachers can affect their students' progress through
@@
 \begin{figure}
+  \centering
 <<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
-load('vcMods.RData')
+#load('vcMods.RData')
 library(lme4)
+
+vcDat <- data%>%filter(unit%in%unique(data$unit[data$curriculum=='Algebra I']))%>%group_by(field_id,unit,section,classid2,state)%>%summarize(status=max(status,na.rm=TRUE),startDate=min(date,na.rm=TRUE))
+vcDat <- merge(vcDat,stud,all.x=TRUE,all.y=FALSE)
+
+vcDat$month <- months(vcDat$startDate,TRUE)
+
+vcDat$cp <- vcDat$status=='changed placement'
+vcDat$mast <- vcDat$status=='graduated'
+
+vcModYr <- list(yr1=glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|state)+(1|unit),family=binomial,data=subset(vcDat,year==1)),yr2=glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|state)+(1|unit),family=binomial,data=subset(vcDat,year==2)))
+
+vcMods <- list()
+for(st in c('TX','KY','MI'))
+ for(yr in c(1,2))
+     vcMods[[paste0(st,'_',yr)]] <- glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|unit),family=binomial,data=subset(vcDat,state==st & year==yr))
 
@@
 reassigning students to particular units.
-Some prinicpals may encourage teachers to adhere to the official
+Some principals may encourage teachers to adhere to the official
 curriculum and avoid reassignment.
@@
 Although teachers control reassignment, their decisions are largely
-dictated by broader policies, ocurring at the state or school level.
+dictated by broader policies, occurring at the state or school level.
 
@@
 standardized tests by accelerating students' progress and reassigning
-students to relevent sections.
+students to relevant sections.
 
@@
 Finally, from February through May of the second year, the rate of
-reassignment incrased again.
+reassignment increased again.
 
@@
     coord_cartesian(ylim=c(0,0.05))+
-    labs(x='Month',y='% of Sections Ending in Reassignment',size='# Worked \\Sections')
+    labs(x='Month',y='% of Sections Ending in Reassignment',size='# Worked\nSections',color='')
 
@@
 \begin{table}
+  \centering
   \begin{tabular}{rll|ll}
@@
 sections that ended in reassignment, and Figure \ref{fig:byMonthState}
-gives the proportion of all of each state's reassignmens over the
+gives the proportion of all of each state's reassignments over the
 course of the year, that occurred in each month.)
@@
     scale_y_continuous(labels=percent)+
-    labs(x='Month',y='% of State\'s Total Reassignments')
+    labs(x='Month',y='% of State\'s Total Reassignments',color='')
 
@@
   each month of the year.}
-\ref{fig:byMonthState}
+\label{fig:byMonthState}
 \end{figure}
@@
 \begin{figure}
+  \centering
 <<classmates,dependson='cpDat',fig.height=6,fig.width=7>>=
@@
 for(i in 1:ncp){
-    if(i%%100==0) cat(round(i/ncp*100,2),'% ',sep='')
+    #if(i%%100==0) cat(round(i/ncp*100,2),'% ',sep='')
     cls <- secLevCPsplit[[cpDat$unitSectionClass[i]]]
@@
 never worked it.
-Accross all three states and both years, it is exceedingly rare for
+Across all three states and both years, it is exceedingly rare for
 students to master or be promoted from a section after someone in
@@
 
-Which of these patterns is most prevelant?
+Which of these patterns is most prevalent?
 More generally, when teachers reassign their students, where in the
@@
 \begin{figure}
+  \centering
   \begin{subfigure}{5in}
@@
 \begin{subfigure}{1in}
+  \centering
 <<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
@@
 Finally, \Sexpr{trans2['Pythagorean Theorem','Product Rule for Exponents']}
-(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignements from sections in the ``Pythagorean Theorem'' unit ended in the earliear ``Product Rule for Exponents'' sections.
+(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignments from sections in the ``Pythagorean Theorem'' unit ended in the earlier ``Product Rule for Exponents'' sections.
 
@@
 
-\section{Effects of Reassignment}
+\section{Effects of Reassignment}\label{sec:effects}
 The goal of CTAI is to help students learn Algebra, so
@@
 As in all observational studies, this includes the assumption that all
-counfounding variables---variables that predict both reassignment and
+confounding variables---variables that predict both reassignment and
 learning---have been measured well and modeled correctly.
@@
 well-defined process, in practice it can take many forms, as we have
-endevoured to show.
+endeavored to show.
 There is no reason to expect the effects of reassignment to be
-the same regardless of wheather the teacher used it to help lagging
+the same regardless of whether the teacher used it to help lagging
 students catch up, to allocate time to important topics, or for some
@@
 studies can help guide intuition, future research, and even---when
-combined with other relevent information and theory---practice.
+combined with other relevant information and theory---practice.
 
@@
 @
+<<sensitivity,include=FALSE,dependson='models'>>=
+source('hhh.r')
+### we're only computing sensitivity for omitting pretest
+X <- as.data.frame(model.matrix(mod3.1)[,-1])
+Ts <- Tz(X = X,treatment = 'everCPTRUE')
+X$everCPTRUE <- NULL
+R <-  partialCor(cpDat$y_yirt,X,'pretest')^2
+multPre <- MEmult(T=Ts['pretest'],R=R,df=mod3.1$df,q=1.96)
+
+Rhos <- vapply(names(X)[-grep('classid2|pretest',names(X))],
+               function(nn) partialCor(cpDat$y_yirt,X,nn),1)^2
+multOth <- MEmult(T=max(abs(Ts[names(Rhos)])),max(Rhos),df=mod3.1$df,q=1.96)
+
+@
+
 <<heterogeneityMods,include=FALSE,dependson='models'>>=
@@
 \begin{figure}
+  \centering
 <<cpYyear,fig.width=6,fig.height=4,dependson='effects'>>=
@@
 @
-\caption{Students' Gain scores (post-test minus pre-test) versus the number of times the were reassigned over the course of the year (jittered), with a Loess smoother added.}
+\caption{Students' Gain scores (posttest minus pre-test) versus the number of times the were reassigned over the course of the year (jittered), with a Loess smoother added.}
 \label{fig:cpYyear}
@@
 
-\subsection{Average Effects}
-Figure \ref{fig:cpYyear} shows students gain scores---the difference between their post-test and pre-test scores---as a function of the number of times they were reassigned.
+%\subsection{Average Effects}
+Figure \ref{fig:cpYyear} shows students gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
 (The number of reassignments was jittered---random noise was added---to avoid overplotting.)
@@
 Care in modeling the number of reassignments, then, is especially important---observations from students reassigned an unusually large number of times can exert undo influence on a regression model and generate misleading results, particularly in the presence of non-monotonic relationships.
-We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for stundents who were. 
+We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for students who were.
 Next, $R^{cat}$ defines a categorical variable taking the values $R^{cat}=0,1,2,3,4+$ for students who were reassigned 0, 1, 2, 3 or four or more times, respectively.
@@
 
-We used linear models to estimate the effect of reassignment, regressing post-test scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
-Since this is unlikely to be the case (even approximately), we ran a second set of models including student level covariates: 
+We used linear models to estimate the effect of reassignment, regressing posttest scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
+Since this is unlikely to be the case (even approximately), we ran a
+second set of models including student level covariates as well:
 pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
-\Sexpr{round(mean(is.na(cpDat$xirt))*100)}\%).
-To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2013effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligability.
+\Sexpr{round(mean(is.na(cpDat[['xirt']]))*100)}\%).
+To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2014effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligibility.
 
 \begin{table}
+  \centering
 \begin{tabular}{r|c|c|c|}
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2013effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests']*100)}\% of that.
+The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests']*100)}\% of that.
 
 But what of unmeasured confounding?
-\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on benchmarking from observed confounders.
-In order to confound the causal relationship between reassignment and post-tests, a confounder would have to predict both.
+\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
+In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
 Roughly speaking, the idea is to widen the confidence interval from a causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
-These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omssion of a confounder.
-As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of post-test scores.
-The sensitivity interval for the effect of being reassigned at least once on post-test scores, 
-\begin{table}
-\begin{tabular}{rrl}
-Benchmark&\multicolumn{2}{c}{Sensitivity Interval}\\
-<<sens,dependson='models'>>=
-source('hhh.r')
-X <- as.data.frame(model.matrix(mod3.1)[,-1])
-Ts <- Tz(X = X,treatment = 'everCPTRUE')
-Rhos <- vapply(names(X)[-grep('classid2',names(X))],
-               function(nn) partialCor(cpDat$y_yirt,X,nn),1)[-1]^2
-se <- adjustedEsts1[['mod3.1']][1,2]
-cat(names(Rhos)[1],'&',round(adjustedEsts1[['mod3.1']][1,'ests'],2),'&$\\pm$',
-    round(se*MEmult(abs(Ts[names(Rhos)[1]]),R = Rhos[1],df = mod3.1$df.residual,1.96),2),'\\\\')
-for(i in 2:length(Rhos))
-  cat(gsub('_','',names(Rhos)[i]),'&&$\\pm$',round(se*MEmult(T=abs(Ts[names(Rhos)[i]]),R = Rhos[i],df = mod3.1$df.residual,1.96),2),'\\\\')
-@
-\end{tabular}
-\caption{Sensitivity intervals for the $R^{bin}$ model.}
-\label{tab:sens}
-\end{table}
-
-\section{Notes}
-[[don't forget to include these]]
-
-\begin{itemize}
-  \item \# of students per state?
-  \item NJ excluded from state-by-state comparisons, cuz too
-    small. Included in ``overall,'' etc.
-  \item What schools included?
-  \item Students with no mastery data: maybe they didn't work at all,
-    maybe they worked but we don't have their data. Treating them as unknown.
-\end{itemize}
+These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.
+As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of posttest scores.
+The sensitivity interval for the effect of being reassigned at least
+once on posttest scores, allowing for the possible omission of a
+hypothetical confounder at most as important as pretest, is
+\Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests'],2)} $\pm$
+\Sexpr{round(adjustedEsts1[['mod3.1']][1,'SEs']*multPre,2)}.
+This interval is quite wide, implying that such a covariate could
+explain much of the observed relationship between reassignment and
+posttest scores (or that the relationship may be much stronger).
+On the other hand, the sensitivity interval allowing for the possible
+omission of a less important hypothetical confounder---one that
+predicts posttests and reassignment as well as the next best observed
+covariates---is
+\Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests'],2)} $\pm$
+\Sexpr{round(adjustedEsts1[['mod3.1']][1,'SEs']*multOth,2)}.
+This interval is substantially tighter.
+All in all, unmeasured confounding may play an important role here,
+but there is good reason to believe that it does not explain all of
+the observed relationship.
+
+The wide variety in the use of reassignment that we have documented
+here might suggest that reassignment's treatment effect might vary as
+well.
+Figure \ref{fig:trtHet} shows estimated classroom-specific treatment
+effects of $R^{bin}$, being reassigned at least once.
+The estimates came from a multilevel model in which posttest scores
+were regressed on $R^{bin}$ and student-level covariates, along with
+random effects for classroom and random slopes for $R^{bin}$ varying
+by classroom.
+Unlike fixed effects models, multilevel models ``partially pool'' data
+across classrooms to estimate classroom specific effects more precisely \citep{gelmanHill}.
+This is especially important given the small sample sizes within
+classrooms.
+Figure \ref{fig:trtHet} shows a wide variation in the effect of
+reassignment across classrooms---the estimated standard deviation of
+these effects was
+\Sexpr{round(sqrt(summary(mod5.1)$varcor$classid2[2,2]),2)},
+larger than the average effect itself.
+While the effect was negative in most classrooms, it was positive in some.
+This variation could be due to a number of factors, including
+differences in the composition of classrooms, but supports the
+hypothesis that differences in when and how reassignment is used
+lead to differences in its effect.
+
+\begin{figure}
+  \centering
+<<hetroPlot,fig.width=6,fig.height=3,dependson='heterogeneityMods'>>=
+re <- ranef(mod5.1,condVar=TRUE)
+reVar <- attr(re$classid2,'postVar')[2,2,]
+feVar <- vcov(mod5.1)['everCPTRUE','everCPTRUE']
+
+hetPlotDat <- data.frame(eff=re$classid2$everCPTRUE+fixef(mod5.1)['everCPTRUE'],
+                         se=sqrt(reVar+feVar),
+                         year=as.factor(data$year[match(rownames(re$classid2),data$classid2)]),
+                         state=data$state[match(rownames(re$classid2),data$classid2)])
+hetPlotDat$ord <- rank(hetPlotDat$eff)
+
+ggplot(hetPlotDat,aes(ord,eff))+geom_point()+
+#    geom_errorbar(aes(ymin=eff-2*se,ymax=eff+2*se),lwd=.2)+
+    geom_hline(yintercept=0,lty=2)+
+#    geom_hline(yintercept=fixef(mod5.1)['everCPTRUE'])
+    scale_x_continuous(labels=NULL,breaks=NULL)+
+    labs(y='Effect of Any Reassignment',x='Classroom')
+@
+\caption{The effect of being reassigned at least once, in each
+  classroom, as estimated by a multilevel model. The classrooms are
+  sorted by these estimated effects. The dotted line indicated an
+  effect of zero.}
+\label{fig:trtHet}
+\end{figure}
+
+% \section{Notes}
+% [[don't forget to include these]]
+
+% \begin{itemize}
+%   \item \# of students per state?
+%   \item NJ excluded from state-by-state comparisons, cuz too
+%     small. Included in ``overall,'' etc.
+%   \item What schools included?
+%   \item Students with no mastery data: maybe they didn't work at all,
+%     maybe they worked but we don't have their data. Treating them as unknown.
+% \end{itemize}
+
+\section{Discussion}\ref{sec:discussion}
+
+\bibliographystyle{plainnat}
+\bibliography{ct,cp}
 
 \end{document}
-<<eval=FALSE>>=
-
- nextDat <- secOrder%>%filter(!is.na(status))%>%group_by(field_id,classid2,year,state)%>%
-  summarize(naf=sum(is.na(prevStatus) & status=='final_or_incomplete')/sum(is.na(prevStatus)),
-   nacp=sum(is.na(prevStatus) & status=='changed placement')/sum(is.na(prevStatus)),
-   naprom=sum(is.na(prevStatus) & status=='promoted')/sum(is.na(prevStatus)),
-   nagrad=sum(is.na(prevStatus) & status=='graduated')/sum(is.na(prevStatus)),
-   ff=sum(prevStatus==1 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
-   fcp=sum(prevStatus==1 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
-   fprom=sum(prevStatus==1 & status=='promoted',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
-   fgrad=sum(prevStatus==1 & status=='graduated',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
-   cpf=sum(prevStatus==2 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
-   cpcp=sum(prevStatus==2 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
-   cpprom=sum(prevStatus==2 & status=='promoted',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
-   cpgrad=sum(prevStatus==2 & status=='graduated',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
-   promf=sum(prevStatus==3 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
-   promcp=sum(prevStatus==3 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
-   promprom=sum(prevStatus==3 & status=='promoted',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
-   promgrad=sum(prevStatus==3 & status=='graduated',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
-   gradf=sum(prevStatus==4 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
-   gradcp=sum(prevStatus==4 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
-   gradprom=sum(prevStatus==4 & status=='promoted',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
-   gradgrad=sum(prevStatus==4 & status=='graduated',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
-   npna=sum(is.na(prevStatus)),npf=sum(prevStatus==1,na.rm=TRUE),npcp=sum(prevStatus==2,na.rm=TRUE),
-    npprom=sum(prevStatus==3,na.rm=TRUE),npgrad=sum(prevStatus==4,na.rm=TRUE))
-
-@

### 2017-10-23 cpEffectSimp.r

@@ -67,6 +67,13 @@ mod4.1 <- update(mod6,addCovs)
 mod5 <- lmer(y_yirt~everCP+year+state+pretest+(everCP|classid2)+(1|schoolid2),data=cpDat)
 mod5.1 <- update(mod7,addCovs)
 
+
+## classes with some cp
+cpClass <- unique(cpDat$classid2[cpDat$everCP])
+
+print('mod3.1 with only CP classes included')
+print(summary(mod3.1.2 <- update(mod3.1,subset=classid2%in%cpClass))$coef['everCPTRUE',])
+
 ## for(mm in grep('mod',ls(),value=TRUE))
 ##     print(summary(assign(paste0(mm,'Y'),update(get(mm),y_yirt~.+pretest))))
 
@@ -74,8 +81,6 @@ save(list=grep('mod',ls(),value=TRUE),file='outcomeMods.RData')
 
 
 
-## classes with some cp
-cpClass <- unique(cpDat$classid2[cpDat$everCP])
 ols5.1 <- lm(y_yirt~everCP+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl+pretest+classid2,
            data=cpDat,subset=classid2%in%cpClass)#)$coef['everCPTRUE',]
 ols5.1.1 <- lm(y_yirt~everCP+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl+ns(pretest,5)+classid2,
@@ -128,18 +133,25 @@ byState2 <- ggplot(filter(cpDat,year==2 & state %in% c('TX','KY','MI')),aes(ncp,
 classLevel <- aggregate(cpDat[,sapply(cpDat,class)%in%c('numeric','integer','logical')],
                         by=list(classid2=cpDat$classid2),FUN=mean,na.rm=TRUE)
 classLevel$state <- cpDat$state[match(classLevel$classid2,cpDat$classid2)]
-classLevel$schoolid2 <- cpDat$schoolid2[match(classLevel$classid2,cpDat$classid2)]
+classLevel$schoolid2 <- cpDat$schoolid2[match(classLevel$classid2,cpDat$classid2)
+]classLevel$grade <- aggregate(cpDat$grade=='9',
+                               by=list(classid2=cpDat$classid2),FUN=mean,na.rm=TRUE)$x
+classLevel$race <- aggregate(cpDat$race=='White',
[…]

### 2017-10-23 mastNext.r

@@ -61,3 +61,28 @@ m2Splitvcov.1 <- vcovHC(m2ols25Split.1,'HC')
 
 save(list=ls(),
      file='mastNextols.RData')
+
+
+ nextDat <- secOrder%>%filter(!is.na(status))%>%group_by(field_id,classid2,year,state)%>%
+  summarize(naf=sum(is.na(prevStatus) & status=='final_or_incomplete')/sum(is.na(prevStatus)),
+   nacp=sum(is.na(prevStatus) & status=='changed placement')/sum(is.na(prevStatus)),
+   naprom=sum(is.na(prevStatus) & status=='promoted')/sum(is.na(prevStatus)),
+   nagrad=sum(is.na(prevStatus) & status=='graduated')/sum(is.na(prevStatus)),
+   ff=sum(prevStatus==1 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fcp=sum(prevStatus==1 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fprom=sum(prevStatus==1 & status=='promoted',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   fgrad=sum(prevStatus==1 & status=='graduated',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
+   cpf=sum(prevStatus==2 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpcp=sum(prevStatus==2 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpprom=sum(prevStatus==2 & status=='promoted',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   cpgrad=sum(prevStatus==2 & status=='graduated',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
+   promf=sum(prevStatus==3 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
+   promcp=sum(prevStatus==3 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
[…]

### 2017-10-24 paper.tex
upload figures

--- 2017-10-23 paper.Rnw
@@
-\documentclass[12pt]{article}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 
-<<include=FALSE>>=
-library(knitr)
-#library(tikzDevice)
-opts_chunk$set(
-echo=FALSE, results='asis',cache=TRUE,warning=FALSE,error=FALSE,message=FALSE,autodep = FALSE
-    )
-
-#options(tikzDefaultEngine = "pdftex")
-@
-
-<<setup,include=FALSE>>=
-
-library(lubridate)
-library(scales)
-library(reshape2)
-library(dplyr)
-library(ggplot2)
-library(xtable)
-
-addOverallState <- function(dat){
-    levels(dat$state) <- c(levels(dat$state),'Overall')
-    dat2 <- dat
-    dat2$state <- 'Overall'
-    dat2 <- rbind(dat,dat2)
-    dat2$state <- factor(dat2$state, levels= levels(dat$state))
-    dat2
-}
-
-@
-
-<<data,include=FALSE>>=
-if('cpPaper.RData'%in%list.files() & file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
-    load('cpPaper.RData')
-} else source('dataMerge.r')
-@
-
-<<data1,include=FALSE,cache=FALSE>>=
-dataCache <- grep('data',list.files('cache'),ignore.case=FALSE,value=TRUE)[1]
-
-if(file.info('dataMerge.r')$mtime>file.info(paste0('cache/',dataCache))$mtime){
-    if(file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
-        load('cpPaper.RData')
-    } else source('dataMerge.r')
-}
-@
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 For this reason, this study uses only data from on the
-\Sexpr{n_distinct(data[['schoolid2']])} schools for which at least 80\%
+18 schools for which at least 80\%
 students in both study years appear in the log data file.
-This sample includes \Sexpr{n_distinct(data[['field_id']])} students, around
-\Sexpr{round(n_distinct(data[['field_id']])/n_distinct(stud[['field_id']]))*100}\%
+This sample includes 4460 students, around
+100\%
 of the total high-school sample.
@@
 state-by-state comparisons (but included analyses that pool across states).
-<<nByState,dependson='data'>>=
-xtable(data%>%group_by(state,year)%>%summarize(n=n_distinct(field_id))%>%dcast(year~state),
-       caption='Numbers of students in the sample by state and study year',label='tab:nByState')
-@
-
-In this smaller sample, \Sexpr{sum(data[['obsUsage']]==FALSE)}
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Mon Oct 23 15:09:52 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrrrr}
+  \hline
+ & year & TX & KY & MI & LA & CT & NJ \\ 
+  \hline
+1 &   1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
+  2 &   2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
+   \hline
+\end{tabular}
+\caption{Numbers of students in the sample by state and study year} 
+\label{tab:nByState}
+\end{table}
+
+In this smaller sample, 164
 students do not appear in the log data; they may have not used the CTAI
@@
   \centering
-<<curricula,dependson=c('setup','data'),fig.width=6.4,fig.height=3>>=
-
-bbb <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%
-    group_by(state,Curriculum,overall,Yr)%>%
-    summarize(n=n())
-staten <- addOverallState(data)%>%filter(!is.na(Curriculum) & State!='NJ')%>%group_by(state,Yr)%>%summarize(n=n())
-
-for(i in 1:nrow(bbb)) bbb$n[i] <- bbb$n[i]/staten$n[staten$state==bbb$state[i] & staten$Yr==bbb$Yr[i]]
-
-COLS <- c('#1b9e77','#d95f02','#7570b3')
-
-ggplot(bbb,aes(Yr,n,fill=Curriculum,alpha=overall))+geom_col()+facet_grid(~state)+scale_alpha_manual(values=c(1,0.5))+scale_fill_manual(values=COLS)+scale_y_continuous(labels=percent)+labs(x='',y='% of Problems Worked',alpha='')
-
-## ggplot(bbb,aes(Yr,n,color=bbb))+geom_col(show.legend=FALSE)+geom_col(aes(fill=Curriculum))+geom_col(aes(fill=overall),alpha=0.5)+facet_grid(~state)
-##     scale_fill_manual(values=c('#8dd3c7','#8dd3c7','#ffffb3','#ffffb3','#bebada','white','black'))
-##     geom_col(aes(Yr,n,fill=overall),alpha=0.5)+
-##     #geom_col(aes(Yr,n,fill=overall))+#scale_fill_manual(values=c('white','black'))+
-##     facet_grid(~state)
-
-## omar <- par()$oma
-
-## par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
-## par(mfrow=c(1,9))
-## for(st in levels(data$State)){
-## #    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
-##     cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
-##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-##     for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-##     cwy <- cwy[nrow(cwy):1,]
-##     barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main=st,yaxt='n')
-##     if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
-##     #box()
-## }
-## cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
-##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-## for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-## cwy <- cwy[nrow(cwy):1,]
-## barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main='Overall',yaxt='n')
-
-## plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
-## legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
-##        fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
-##        angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,50),xpd = TRUE, bty = "n",inset=c(0,0))
-
-## par(oma=omar)
-@
+
+\includegraphics[width=\maxwidth]{figure/curricula-1}
+
 \caption{Percentage of worked problems coming from various levels
@@
 
-<<usageMedians,dependson=c('data','setup'),results='asis'>>=
-secByStud <- data%>%group_by(state,Yr,field_id) %>% summarize(numSec=n_distinct(section,na.rm=TRUE),
-                                                              numUnit=n_distinct(unit,na.rm=TRUE),
-                                                              time=sum(total_t1,na.rm=TRUE),
-                                                              mastered=n_distinct(section[status=='graduated'],na.rm=TRUE),
-                                                              nprob=n_distinct(unit,section,Prob1,na.rm=TRUE))
-
-secByStud$time <- secByStud$time/3600000
-
-secByStud <- within(secByStud,{
-    numUnit[numUnit==0] <- NA
-    numSec[numSec==0] <- NA
-    time[time==0] <- NA
-    mastered[mastered==0] <- NA
-})
-
-secByStud2 <- addOverallState(secByStud)
-
-secByStud2 <-
-  secByStud2 %>%
-  group_by(state,Yr) %>%
-  mutate(outlier.time = time > quantile(time,0.75,na.rm=TRUE) + IQR(time,na.rm=TRUE) * 1.5,
-         outlier.sec = numSec > quantile(numSec,0.75,na.rm=TRUE) + IQR(numSec,na.rm=TRUE) * 1.5,
-         outlier.prob = nprob > quantile(nprob,0.75,na.rm=TRUE) + IQR(nprob,na.rm=TRUE) * 1.5,
-         outlier.unit = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5
-         ) %>%
-  ungroup
-
-tab <- ungroup(secByStud)%>%group_by(Yr)%>%dplyr::select(time,nprob,numSec,numUnit)%>%summarize_all(median,na.rm=TRUE)
-tab$Yr <- NULL
-tab <- as.data.frame(tab)
-rownames(tab) <- c('Year 1','Year 2')
-names(tab) <- c('Hours','Problems','Sections','Units')
-
-xtable(tab,caption='Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study.
-Students with no usage data were excluded.',label='tab:medUsage',digits=c(1,2,0,1,0))
-@
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Mon Oct 23 15:23:21 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrr}
+  \hline
+ & Hours & Problems & Sections & Units \\ 
+  \hline
+Year 1 & 33.47 & 309 & 43.0 & 9 \\ 
+  Year 2 & 23.83 & 228 & 36.5 & 10 \\ 
+   \hline
+\end{tabular}
+\caption{Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study. Students with no usage data were excluded.} 
+\label{tab:medUsage}
+\end{table}
 
@@
 student worked about
-\Sexpr{round(tab['Year 1','Hours']-tab['Year 2','Hours'])}
+10
 fewer hours,
-\Sexpr{round(tab['Year 1','Problems']-tab['Year 2','Problems'])}
+81
 fewer problems, and
-\Sexpr{round(tab['Year 1','Sections']-tab['Year 2','Sections'])} fewer
+6 fewer
 sections in year 2 than in year 1.
@@
 \centering
-<<usageTime, fig.height=3,fig.width=6,dependson=c('data','setup','usageMedians')>>=
-
-print(timeStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,time))+geom_boxplot(outlier.shape=NA)+
-          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.time), width=0.2)+
-          facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CT Software per Student'))
-                                        #ggsave('timeStatYear.jpg',width=6,height=3)
-
-@
+
+\includegraphics[width=\maxwidth]{figure/usageTime-1}
+
 \caption{Boxplots of hours each student spent on Cognitive Tutor
   software over by year and state. Students  with no timestamp data
-  (\Sexpr{sum(is.na(secByStud[['time']]))}), with negative time
-  (\Sexpr{sum(secByStud[['time']]<0,na.rm=TRUE)}) or with more than 110 hours (\Sexpr{sum(secByStud[['time']]>110,na.rm=TRUE)})
+  (193), with negative time
+  (3) or with more than 110 hours (91)
   were excluded.}
@@
 \end{figure}
-
-<<supplementalFigs1,include=FALSE,dependson=c('data','setup','usageMedians')>>=
-secStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numSec))+geom_boxplot(outlier.shape=NA)+
-          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.sec), width=0.2)+
-          facet_grid(~state)+labs(x='',y='# of Sections Worked per Student')+coord_cartesian(ylim=c(0,200))
-ggsave('secStateYear.jpg',width=6,height=3)
-
-probStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,nprob))+geom_boxplot(outlier.shape=NA)+
-          geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier.prob), width=0.2)+
-          facet_grid(~state)+labs(x='',y='# of Problems Worked per Student')+coord_cartesian(ylim=c(0,2000))
-ggsave('probStateYear.jpg',width=6,height=3)
-
-@
 
@@
 %   software each student mastered, by year and state. Students
-%   mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
-%   with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
+%   mastering more than 250 sections (sum(secByStud[['mastered']]>250,na.rm=TRUE)/2) of nrow(secByStud)/2) and students
+%   with no mastery data (sum(is.na(secByStud[['mastered']]))/2)
 %   were excluded.}
@@
 Overall, usage varied a bit more in year 2 than in year
-1---the median absolute deviation of time spent was \Sexpr{round(mad(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE,constant=1),1)} hours in the first year, compared to \Sexpr{round(mad(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE,constant=1),1)} in  the second year.
+1---the median absolute deviation of time spent was 15.9 hours in the first year, compared to 16.9 in  the second year.
 The increase in variation seems to be driven both by increasing
@@
 students worked increased by
-\Sexpr{round(tab['Year 2','Units']-tab['Year 1','Units'])}
+1
 from year 1 to year 2.
@@
 \centering
-<<unitsWorked,dependson=c('usageMedians','data','setup'),fig.width=6,fig.height=3>>=
-secByStud2 <-
-  secByStud2 %>%
-  group_by(state,Yr) %>%
-  mutate(outlier = numUnit > quantile(numUnit,0.75,na.rm=TRUE) + IQR(numUnit,na.rm=TRUE) * 1.5) %>%
-  ungroup
-
-print(unitsStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
-@
+
+\includegraphics[width=\maxwidth]{figure/unitsWorked-1}
+
 \caption{Boxplots of the number of units of Cognitive Tutor
@@
   working more than 55 units (
-  \Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)} of
-  \Sexpr{nrow(secByStud)}) and students
-  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
+  7 of
+  4460) and students
+  with no usage data (189)
   were excluded.}
@@
   \centering
-<<whichUnits,fig.height=4,fig.width=6,dependson=c('data','setup')>>=
-
-curricula <- read.csv('~/Box Sync/CT/data/sectionLevelUsageData/RAND_study_curricula.csv',stringsAsFactors=FALSE)
-curricula <- subset(curricula,curriculum_name=='algebra i')
-curricula$unit <- tolower(curricula$unit)
-curricula <- subset(curricula,unit%in%intersect(curricula$unit[curricula$ct=='2007'],curricula$unit[curricula$ct=='2008r1']))
-units <- curricula$unit[curricula$ct=='2007']
-sectionStats <- read.csv('~/Box Sync/CT/data/sectionLevelUsageData/section_stats_withAbb.csv',stringsAsFactors=FALSE)
-UnitName <- sectionStats$unit_name_abb[match(units,sectionStats$unit_id)]
-UnitName[units=='inequality-systems-solving'] <- 'Systems of Lin. Ineq.'
-UnitName[units=='intro-pythag-theorem'] <- 'Pythagorean Theorem'
-UnitName[units=='linear-inequality-graphing'] <- 'Graphs of Lin. Ineq.'
-UnitName[units=='linear-systems-solving'] <- 'Systems of Lin. Eq. Solving'
-UnitName[units=='probability'] <- 'Probability'
-UnitName[units=='unit-conversions'] <- 'Unit Conversions'
-
-nstud <- data%>%filter(!is.na(unit))%>%group_by(Year)%>%summarize(nstud=n_distinct(field_id))
-data$Unit <- data$unit
-data$Unit[grep('unit-conversions',data$Unit)] <- 'unit-conversions'
-
-unitLevel <- data%>%filter(Unit%in%units)%>%group_by(Unit,Year)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
-
-unitLevel$perWorked <- unitLevel$numWorked/nstud$nstud[match(unitLevel$Year,nstud$Year)]
-
-unitLevel$Unit <- factor(unitLevel$Unit,levels=units)
-levels(unitLevel$Unit) <- UnitName
-
-unitLevel$year <- factor(ifelse(unitLevel$Year=='Year 1',1,2))
-print(unitsWorked <- ggplot(unitLevel,aes(x=Unit,y=perWorked,color=year,group=year))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Year')+scale_y_continuous(labels=percent))
-
-@
+
+\includegraphics[width=\maxwidth]{figure/whichUnits-1}
+
 \caption{The percentages of students with usage data who worked at
@@
   \centering
-<<unitsWorkedCust,dependson=c('data','setup','whichUnits'),fig.height=4,fig.width=6>>=
-### by customized curriculum (at school level)
-cust <- data%>%filter(Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(cust=mean(overall=='Customized',na.rm=T))%>%arrange(cust)
-data$cust <- ifelse(data$schoolid2%in%cust$schoolid2[cust$cust>0.8],'Customized','Standard')
-
-nstudCust <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(cust)%>%summarize(nstud=n_distinct(field_id))
-unitLevelCust <- data%>%filter(Unit%in%units & Year=='Year 2' )%>%group_by(Unit,cust)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
-
-unitLevelCust$perWorked <- unitLevelCust$numWorked/nstudCust$nstud[match(unitLevelCust$cust,nstudCust$cust)]
-
-unitLevelCust$Unit <- factor(unitLevelCust$Unit,levels=units)
-levels(unitLevelCust$Unit) <- UnitName
-ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Curriculum',title='Year 2')+scale_y_continuous(labels=percent)
-@
+
+\includegraphics[width=\maxwidth]{figure/unitsWorkedCust-1}
+
 \caption{The percentages of year-2 students with usage data who worked at
@@
 curriculum.\footnote{At least
-  \Sexpr{trunc(100*min(cust[['cust']][cust[['cust']]>0.8]))}\% of
+  90\% of
   problems at ``Customized'' schools were from a customized curriculum,
   and at most
-  \Sexpr{ceiling(100*max(cust[['cust']][cust[['cust']]<0.8]))}\% of
+  1\% of
   problems at ``Standard'' schools were from a customized curriculum.}
@@
   \centering
-<<unitsBySchool,dependson=c('data','setup','whichUnits'),fig.height=7,fig.width=6>>=
-### by school
-nstudSch <- data%>%filter(!is.na(unit) & Year=='Year 2')%>%group_by(schoolid2,state)%>%summarize(nstud=n_distinct(field_id))
-unitLevelSch <- with(filter(data,Unit%in%units & Year=='Year 2'),expand.grid(Unit=unique(Unit),schoolid2=unique(schoolid2)))
-unitLevelSch$state <- data$state[match(unitLevelSch$schoolid2,data$schoolid2)]
-unitLevelSch$cust <- data$cust[match(unitLevelSch$schoolid2,data$schoolid2)]
-unitLevelSch$numWorked <- with(filter(data,Year=='Year 2'),vapply(1:nrow(unitLevelSch),function(i)
-    n_distinct(field_id[schoolid2==unitLevelSch$schoolid2[i] & Unit==unitLevelSch$Unit[i]]),1))
-
-#unitLevelSch <- data%>%filter(Unit%in%units & Year=='Year 2')%>%group_by(Unit,schoolid2,cust,state)%>%summarize(numWorked= n_distinct(field_id,na.rm=TRUE),numCP=sum(status=='changed placement',na.rm=TRUE),meanCP=mean(status=='changed placement',na.rm=TRUE))
-
-unitLevelSch$perWorked <- unitLevelSch$numWorked/nstudSch$nstud[match(unitLevelSch$schoolid2,nstudSch$schoolid2)]
-
-unitLevelSch$Unit <- factor(unitLevelSch$Unit,levels=units)
-levels(unitLevelSch$Unit) <- UnitName
-
-ggplot(filter(unitLevelSch,state!='NJ'),aes(x=Unit,y=perWorked,color=schoolid2,group=schoolid2,linetype=cust))+geom_point(size=.5)+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10))+
-                                        #,legend.position=c(.96,.82))+
-    facet_grid(state~.)+scale_y_continuous(labels=percent)+scale_color_discrete(guide=FALSE)+
-    labs(x='',y='% Worked',linetype='Curriculum',title='Year 2')
-@
+
+\includegraphics[width=\maxwidth]{figure/unitsBySchool-1}
+
 \caption{The percentages of year-2 students with usage data in each
@@
   \centering
-<<overallStatus,dependson=c('data','setup'),fig.height=3,fig.width=6>>=
-statusPerSec <- data%>%
-    filter(!is.na(Curriculum) &!(state=='MI'&Curriculum=='Customized'&year==1))%>%
-    group_by(field_id,Yr,state,section,Curriculum,status)%>%
-        summarize(cp=any(status=='changed placement'))%>%group_by(state,Curriculum,Yr,field_id)%>%summarise(pcp=mean(cp,na.rm=TRUE),ncp=sum(cp,na.rm=TRUE))
-
-statusOverall <-  data%>%filter(!is.na(status))%>%
-    group_by(field_id,Yr,state,unit,section,Curriculum,overall)%>%summarize(status=max(status))
-
-statusStateYr <- addOverallState(statusOverall) %>% group_by(state,Yr)%>%
-    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
-              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
-    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
-
-levels(statusStateYr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
-
- ggplot(filter(statusStateYr,state!='NJ'),aes(Yr,value,fill=variable))+geom_col()+facet_grid(~state)+labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent)
-
-@
+
+\includegraphics[width=\maxwidth]{figure/overallStatus-1}
+
 \caption{The distributions of outcomes of worked sections, by state and across the
@@
 &   \multicolumn{6}{c}{Year 1}\\%&\multicolumn{6}{c}{Year 2}\\
-<<statusTab,dependson='overallStatus',results='asis'>>=
-statusStateYr2 <- addOverallState(statusOverall)
-statusStateYr2a <- statusStateYr2%>%group_by(state,Yr)%>%
-    summarize(pgrad=sum(status=='graduated'),pfoi=sum(status=='final_or_incomplete'),
-              pcp=sum(status=='changed placement'),pprom=sum(status=='promoted'))%>%
-    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
-statusStateYr2b <- statusStateYr2%>%group_by(state,Yr)%>%
-    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
-              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
-    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
-
-levels(statusStateYr2a$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
-levels(statusStateYr2b$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Mastered='pgrad')
-
-tab1 <- dcast(subset(statusStateYr2a,Yr=='Yr 1'&state!='NJ'),variable~state)
-tab2 <- dcast(subset(statusStateYr2a,Yr=='Yr 2'&state!='NJ'),variable~state)
-#tab <- cbind(tab1,tab2[-1])
-
-tab1b <- dcast(subset(statusStateYr2b,Yr=='Yr 1'&state!='NJ'),variable~state)
-tab2b <- dcast(subset(statusStateYr2b,Yr=='Yr 2'&state!='NJ'),variable~state)
-
-tab1$variable <- as.character(tab1$variable)
-tab2$variable <- as.character(tab2$variable)
-cat('&')
-cat(names(tab1)[-1],sep='&')
-cat('\\\\')
-for(i in 1:nrow(tab1)){
-    cat(tab1[i,1],round(unlist(tab1[i,-1])),sep='&')
-    cat('\\\\')
-}
-cat('&\\multicolumn{6}{c}{Year 2}\\\\ \n')
-for(i in 1:nrow(tab2)){
-    cat(tab2[i,1],round(unlist(tab2[i,-1])),sep='&')
-    cat('\\\\')
-}
-rownames(tab1) <- rownames(tab2) <- tab1$variable
-tab1 <- as.matrix(tab1[,-1])
-tab2 <- as.matrix(tab2[,-1])
-
-rownames(tab1b) <- rownames(tab2b) <- tab1b$variable
-tab1b <- as.matrix(tab1b[,-1])
-tab2b <- as.matrix(tab2b[,-1])
-
-@
+&TX&KY&MI&LA&CT&Overall\\Final&699&625&242&129&91&1796\\Reassigned&600&140&220&5&92&1082\\Promoted&1619&9628&4582&1459&261&17561\\Mastered&15815&56199&31200&8648&1303&113374\\&\multicolumn{6}{c}{Year 2}\\
+Final&1252&660&236&137&114&2412\\Reassigned&1215&253&275&9&144&2042\\Promoted&1000&6296&4717&1998&450&14477\\Mastered&4929&36676&26509&10108&1513&79874\\
 
@@
 In the first year, about
-\Sexpr{round(mean(tab1b['Mastered',c('TX','KY','MI','LA')])*100)}\% of
+85\% of
 worked sections are mastered in every
@@
 Other than in Texas, about
-\Sexpr{paste(round(range(tab1b['Promoted',c('CT','KY','MI','LA')])*100),collapse='--')}\%
+13--15\%
 of sections end in promotion.
-About \Sexpr{round(tab1b['Reassigned','TX']*100)}
-and \Sexpr{round(tab1b['Reassigned','CT']*100)}\% of sections in Texas
+About 3
+and 5\% of sections in Texas
 and Connecticut end in reassignment, which is even rarer in the other states.
@@
 of about about
-\Sexpr{round(tab2b['Reassigned','TX']/tab1b['Reassigned','TX'])},
-to about \Sexpr{round(tab2b['Reassigned','TX']*100)}\%.
+5,
+to about 14\%.
 The proportion of Texas sections labeled ``Final'' increased as well,
@@
 Across states, just under
-\Sexpr{round(tab1b['Reassigned','Overall']*100)}\% of sections in year
+1\% of sections in year
 1 ended in reassignment, as did roughly
-\Sexpr{round(tab2b['Reassigned','Overall']*100)}\% in year 2.
+2\% in year 2.
 
@@
   \centering
-<<statusCur,dependson='overallStatus',fig.height=3,fig.width=6>>=
-statusOverall$Curr2 <- with(statusOverall,
-                        ifelse(Curriculum=='Algebra I',
-                                     ifelse(overall=='Standard','Algebra I','Algebra I (Cust.)'),
-                                     as.character(Curriculum)))
-statusCurr <- statusOverall%>%filter(!is.na(Curriculum))%>% group_by(Curr2,Yr)%>%
-    summarize(pgrad=mean(status=='graduated'),pfoi=mean(status=='final_or_incomplete'),
-              pcp=mean(status=='changed placement'),pprom=mean(status=='promoted'))%>%
-    melt(measure.vars=c('pgrad','pfoi','pcp','pprom'))
-levels(statusCurr$variable) <- list(Final='pfoi',Reassigned='pcp',Promoted='pprom',Graduated='pgrad')
-statusCurr$Curr2 <-
-    factor(statusCurr$Curr2,levels=c('Bridge-to-Algebra','Algebra I','Algebra I (Cust.)','>Algebra I'))
-levels(statusCurr$Curr2)[1] <- 'Bridge-to-Alg.'
-print(ggplot(statusCurr,aes(Yr,value,fill=variable))+geom_col()+facet_grid(~Curr2)+
-      labs(y='% of Sections Worked',x='',fill='Exit Status')+scale_y_continuous(labels=percent))
-
-@
+
+\includegraphics[width=\maxwidth]{figure/statusCur-1}
+
 \caption{The distributions of outcomes of worked sections, by
@@
 mastered
-\Sexpr{round(statusCurr[['value']][statusCurr[['Curriculum']]=='Bridge-to-Algebra'
-  & statusCurr[['variable']]=='Graduated']*100)}\% of the time;
+\% of the time;
 those from Algebra I were mastered
-\Sexpr{round(
-  mean(statusOverall$status[statusOverall$Curriculum=='Algebra I']=='graduated',na.rm=T)*100)}\%
+83\%
 of the time, and those from more advanced curricula were mastered
 at a rate of
-\Sexpr{round(
-  mean(statusOverall$status[statusOverall$Curriculum=='>Algebra I']=='graduated',na.rm=T)*100)}\%.
+75\%.
 This is unsurprising, since more advanced curricula may be expected
@@
 reassignment more often than sections from the standard Algebra I
-curriculum (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &
-  variable=='Reassigned')[['value']])}\% vs.
-\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Reassigned')[['value']])}\%).
+curriculum (3\% vs.
+1, 1\%).
 This may an overall skepticism towards the Carnegie Learning standards
@@
 % curricula were mastered
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])}\%)
+% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])\%)
 % than from the standard Algebra I curriculum
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])}\%).
+% (round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])\%).
 % This difference must be interpreted with caution, since a greater
@@
 % or were final
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])}\% vs.
-% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])}\%)
+% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])\% vs.
+% round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])\%)
 % , as well.
 % The proportion of sections ending in promotion was almost identical:
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])}\% vs.
-% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])}\%).
+% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])\% vs.
+% round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])\%).
 
@@
 
-<<cpDat,dependson=c('data','setup'),include=FALSE>>=
-secLev <- data%>%filter(is.finite(status) & is.finite(timestamp) &is.finite(date))%>%
-    group_by(field_id,unit,section,Year,Yr,classid2,schoolid2)%>%
-    summarize(startDate=min(date),endDate=max(date),startTime=min(timestamp),endTime=max(timestamp),state=state[1],
-              status=max(status),Curriculum=Curriculum[1],overall=overall[1],version=version[1])%>%
-    arrange(endDate)
-
-### cp over time
-secLev <- within(secLev,endMonth <- factor(month(secLev$endDate,TRUE,TRUE),levels=c('Aug','Sep','Oct','Nov','Dec','Jan','Feb','Mar','Apr','May','Jun','Jul')))
-
-@
-
 \subsection{How Do Reassignment Patterns Vary?}
@@
   \centering
-<<vcs,fig.height=3,fig.width=6.5,dependson=c('data','setup')>>=
-#load('vcMods.RData')
-library(lme4)
-
-vcDat <- data%>%filter(unit%in%unique(data$unit[data$curriculum=='Algebra I']))%>%group_by(field_id,unit,section,classid2,state)%>%summarize(status=max(status,na.rm=TRUE),startDate=min(date,na.rm=TRUE))
-vcDat <- merge(vcDat,stud,all.x=TRUE,all.y=FALSE)
-
-vcDat$month <- months(vcDat$startDate,TRUE)
-
-vcDat$cp <- vcDat$status=='changed placement'
-vcDat$mast <- vcDat$status=='graduated'
-
-vcModYr <- list(yr1=glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|state)+(1|unit),family=binomial,data=subset(vcDat,year==1)),yr2=glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|state)+(1|unit),family=binomial,data=subset(vcDat,year==2)))
-
-vcMods <- list()
-for(st in c('TX','KY','MI'))
- for(yr in c(1,2))
-     vcMods[[paste0(st,'_',yr)]] <- glmer(cp~(1|field_id)+(1|classid2)+(1|schoolid2)+(1|unit),family=binomial,data=subset(vcDat,state==st & year==yr))
-
-vcFun <- function(nm){
-    mod <- vcMods[[nm]]
-    out <- unlist(summary(mod)$varcor)
-    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
-    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
-    out$state <- strsplit(nm,'_')[[1]][1]
-    out$year <- strsplit(nm,'_')[[1]][2]
-    out$sig2 <- out$sig2/sum(out$sig2)
-    out
-}
-
-vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))
-
-yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
-                          state='Overall',
-                  year='1')
-yr1$sig2 <- yr1$sig2/sum(yr1$sig2)
-
-yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
-                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
-                          state='Overall',
-                  year='2')
-yr2$sig2 <- yr2$sig2/sum(yr2$sig2)
-
-vcDat <- rbind(vcDat,yr1,yr2)
-
-vcDat$comp <- factor(vcDat$comp)
-levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')
-
-vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))
-
-vcs <- list()
-for(s in unique(vcDat$state)) for(y in 1:2){
- ddd <- round(subset(vcDat,state==s&year==y,select=c(sig2))*100)
- rownames(ddd) <- gsub('\\d','',rownames(ddd))
- rownames(ddd)[nrow(ddd)] <- 'resid'
- vcs[[paste0(s,'_',y)]] <- ddd
-}
-
-ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')+scale_y_continuous(labels=percent)
-
-@
+
+\includegraphics[width=\maxwidth]{figure/vcs-1}
+
 \caption{Results from a set of eight multilevel logistic regressions
@@
 reassignment was school, accounting for
-\Sexpr{vcs[['Overall_1']]['schoolid','sig2']}\% of the variation in year 1,
-and \Sexpr{vcs[['Overall_2']]['schoolid',1]}\% in year 2.
+30\% of the variation in year 1,
+and 43\% in year 2.
 After school, state was the most important, accounting for
-\Sexpr{vcs[['Overall_1']]['state',1]} and
-\Sexpr{vcs[['Overall_2']]['state',1]}\% in the two years, and unit,
-accounting for \Sexpr{vcs[['Overall_1']]['unit',1]} and \Sexpr{vcs[['Overall_2']]['unit',1]}\%.
+20 and
+21\% in the two years, and unit,
+accounting for 16 and 14\%.
 Surprisingly, classroom and student-level factors only accounted for
-\Sexpr{vcs[['Overall_1']]['classid',1]} and
-\Sexpr{vcs[['Overall_1']]['field_id',1]}\% in year 1, respectively,
-and \Sexpr{vcs[['Overall_2']]['classid',1]} and
-\Sexpr{vcs[['Overall_2']]['field_id',1]}\% in year 2.
+7 and
+3\% in year 1, respectively,
+and 6 and
+1\% in year 2.
 The pattern was similar in Texas, where school accounted for over half
@@
 In Kentucky, unit played the largest role
-(\Sexpr{vcs[['KY_1']]['unit',1]}\%) in year 1, and classroom played
-the largest role in year 2 (\Sexpr{vcs[['KY_2']]['classid',1]}\%).
+(52\%) in year 1, and classroom played
+the largest role in year 2 (50\%).
 Across states and years, student level factors never accounted for
-more than \Sexpr{max(sapply(vcs,function(x) x['field_id',1]))}\% of
+more than 5\% of
 the variation in reassignment.
 Other than in Kentucky in year 2, classroom never accounted for more
-than \Sexpr{sort(sapply(vcs,function(x) x['classid',1]),dec=T)}\%  of
+than 50, 12, 12, 10, 7, 6, 5, 2\%  of
 the variation.
@@
   \centering
-<<byMonth,dependson='cpDat',fig.height=3,fig.width=6>>=
-secLevMonth <- secLev%>%group_by(endMonth,Year)%>%
- summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
-secLev$cp <- as.numeric(secLev$status=='changed placement')
-
-ggplot(filter(secLevMonth,nsec>100),aes(endMonth,CPper,group=Year,color=Year,size=nsec))+geom_point()+geom_line(size=1)+
-    ## geom_smooth(aes(as.numeric(endMonth)+day(endDate)/31-0.5,cp,group=Year,
-    ##                 color=Year,size=1),
-    ##             method = "glm", formula = y ~ splines::bs(x, 4),data=secLev,method.args=list(family='binomial'),
-                                        #            show.legend=FALSE)+
-scale_y_continuous(breaks=seq(0,0.05,.01),labels=percent)+
-    coord_cartesian(ylim=c(0,0.05))+
-    labs(x='Month',y='% of Sections Ending in Reassignment',size='# Worked\nSections',color='')
-
-@
+
+\includegraphics[width=\maxwidth]{figure/byMonth-1}
+
 \caption{The proportion of worked sections ending in reassignment, by
@@
     &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
-<<byMonthTab,results='asis'>>=
-
-tab <- NULL
-for(st in c('TX','KY','MI','LA','CT','NJ')){
- tab <- rbind(tab,
-  c(sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 1'&
-     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
-    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 1'&
-     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement'),
-    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 2'&
-     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
-    sum(secLev$status[secLev$state==st & secLev$Yr=='Yr 2'&
-     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement')))
-}
-tab <- rbind(tab,
-    c(sum(secLev$status[secLev$Yr=='Yr 1'&
-     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
-    sum(secLev$status[ secLev$Yr=='Yr 1'&
-     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement'),
-    sum(secLev$status[ secLev$Yr=='Yr 2'&
-     secLev$endMonth%in%c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')]=='changed placement'),
-    sum(secLev$status[ secLev$Yr=='Yr 2'&
-     secLev$endMonth%in%c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun')]=='changed placement')))
-rownames(tab) <- c('TX','KY','MI','LA','CT','NJ','Overall')
-
-print(xtable(tab),only.contents=TRUE,include.colnames=FALSE,hline.after=6)
-@
+% latex table generated in R 3.3.2 by xtable 1.8-2 package
+% Mon Oct 23 10:33:17 2017
+ TX &  18 & 582 & 281 & 932 \\ 
+  KY &   9 & 131 &  52 & 201 \\ 
+  MI &  47 & 173 & 245 &  30 \\ 
+  LA &   0 &   5 &   9 &   0 \\ 
+  CT &  14 &  78 &  84 &  60 \\ 
+  NJ &   3 &  22 &  17 & 129 \\ 
+   \hline
+Overall &  91 & 991 & 688 & 1352 \\
+
 \hline
@@
   \centering
-<<byMonthState,dependson='cpDat',fig.height=3,fig.width=6>>=
-
-secLevMonthSt <- secLev%>%filter(status=='changed placement'&state%in%c('TX','KY','MI') & endMonth!='Jul')%>%
- group_by(endMonth,Year,state)%>%summarize(cpPer=n())
-secLevNum <- secLev%>%filter(status=='changed placement'&state%in%c('TX','KY','MI'))%>%
- group_by(Year,state)%>%summarize(ncp=n())
-
-for(st in c('TX','KY','MI')) for(yr in c('Year 1','Year 2'))
- secLevMonthSt$cpPer[secLevMonthSt$state==st & secLevMonthSt$Year==yr] <-
-  secLevMonthSt$cpPer[secLevMonthSt$state==st & secLevMonthSt$Year==yr]/
-   secLevNum$ncp[secLevNum$state==st & secLevNum$Year==yr]
-
-## secLevMonthSt <- secLev%>%filter(state%in%c('TX','KY','MI') & endMonth!='Jul')%>%group_by(endMonth,state,Year)%>%summarize(nsec=n(),CPper=mean(status=='changed placement',na.rm=TRUE))
-ggplot(secLevMonthSt,aes(endMonth,cpPer,group=state,color=state))+geom_point()+facet_grid(Year~.)+
-    geom_line(size=1)+
-    scale_y_continuous(labels=percent)+
-    labs(x='Month',y='% of State\'s Total Reassignments',color='')
-
-@
+
+\includegraphics[width=\maxwidth]{figure/byMonthState-1}
+
 \caption{The proportion of a state's reassignments that occurred in
@@
   \centering
-<<classmates,dependson='cpDat',fig.height=6,fig.width=7>>=
-secLev$unitSectionClass <- paste(secLev$unit,secLev$section,secLev$classid2)
-cpDat <- subset(secLev,status=='changed placement' & Curriculum=='Algebra I')
-cpClasses <- unique(cpDat$unitSectionClass)
-secLevCP <- filter(secLev,unitSectionClass%in%cpClasses)
-levels(secLevCP$status) <- list(promGrad=c('promoted','graduated','final_or_incomplete'),cp='changed placement')
-secLevCPsplit <- split(secLevCP[,c('field_id','endDate','status')],secLevCP$unitSectionClass)
-
-ncp <- nrow(cpDat)
-seqDat <- matrix(NA,nrow=ncp,ncol=6)
-for(i in 1:ncp){
-    #if(i%%100==0) cat(round(i/ncp*100,2),'% ',sep='')
-    cls <- secLevCPsplit[[cpDat$unitSectionClass[i]]]
-    if(nrow(cls)==1){
-        seqDat[i,] <- 0
-        next
-    }
-    cls <- filter(cls,field_id!=cpDat$field_id[i])
-    date <- cpDat$endDate[i]
-    ord <- factor(ifelse(cls$endDate<date,'bf',
-                  ifelse(cls$endDate==date,'same','after')),levels=c('bf','same','after'))
-    clsSplit <- split(cls,list(ord,cls$status))
-    seqDat[i,] <- vapply(clsSplit,nrow,1)
-}
-cat('\n')
-colnames(seqDat) <- names(clsSplit)
-
-classSize <- data%>%filter(!is.na(section))%>%group_by(classid2)%>%summarize(nstud=n_distinct(field_id))
-cpDat$totClassmates <- classSize$nstud[match(cpDat$classid2,classSize$classid2)]
-
-for(nn in names(clsSplit)) cpDat[[nn]] <- seqDat[,nn]/cpDat$totClassmates
-cpDat$nw <- (cpDat$totClassmates-rowSums(seqDat))/cpDat$totClassmates
-cpDat$total <- rowSums(seqDat)
-cpDat$bf.promGrad <- cpDat$bf.promGrad+cpDat$same.promGrad
-cpDat$same.promGrad <- NULL
-
-### plot them all
-
-mmm <- with(subset(cpDat,totClassmates>15),max(table(state,Year)))
-
-cpDat2 <- cpDat%>%filter(totClassmates>15)%>%arrange(bf.promGrad,bf.cp,same.cp,after.cp,after.promGrad)%>%group_by(Year,state)%>%mutate(cpid=seq(floor((mmm-n())/2)+1,floor((mmm-n())/2)+n()))%>%ungroup()
-
-aaa <- melt(cpDat2,measure.vars=c('bf.promGrad','bf.cp','same.cp','after.cp','after.promGrad','nw'))
-levels(aaa$variable) <- list(
-    `Never Worked\n Section`='nw',
-    `Mastered/Promoted\n Later`='after.promGrad',
-    `Reassigned\n Later`='after.cp',
-    `Reassigned\n Same Day`='same.cp',
-    `Reassigned\n Before`='bf.cp',
-    `Mastered/Promoted\n Before`='bf.promGrad')
-
-behind <- with(cpDat2,mean(bf.promGrad+bf.cp+same.cp>.75))
-behind1 <- with(subset(cpDat2,Yr=='Yr 1'),mean(bf.promGrad+bf.cp+same.cp>.75))
-behind2 <- with(subset(cpDat2,Yr=='Yr 2'),mean(bf.promGrad+bf.cp+same.cp>.75))
-
-behind1tx <- with(subset(cpDat2,Yr=='Yr 1'&state=='TX'),mean(bf.promGrad+bf.cp+same.cp>.75))
-behind1ky <- with(subset(cpDat2,Yr=='Yr 1'&state=='KY'),mean(bf.promGrad+bf.cp+same.cp>.75))
-behind1mi <- with(subset(cpDat2,Yr=='Yr 1'&state=='MI'),mean(bf.promGrad+bf.cp+same.cp>.75))
-
-behind2tx <- with(subset(cpDat2,Yr=='Yr 2'&state=='TX'),mean(bf.promGrad+bf.cp+same.cp>.75))
-behind2ky <- with(subset(cpDat2,Yr=='Yr 2'&state=='KY'),mean(bf.promGrad+bf.cp+same.cp>.75))
-behind2mi <- with(subset(cpDat2,Yr=='Yr 2'&state=='MI'),mean(bf.promGrad+bf.cp+same.cp>.75))
-
-mi2cp <- round(with(subset(cpDat2,Yr=='Yr 2'&state=='MI'),mean(bf.cp+same.cp+after.cp+nw>.75))*100)
-ky1cp <- round(with(subset(cpDat2,Yr=='Yr 1'&state=='KY'),mean(bf.cp+same.cp+after.cp+nw>.75))*100)
-
-#aaa$variable <- factor(aaa$variable,levels=rev(c('pGrad','pProm','pCPbf','pCPsame','pLater','pNW')))
-ggplot(filter(aaa,state%in%c('TX','KY','MI')),aes(as.factor(cpid),value,fill=variable))+geom_col()+facet_grid(Year~state)+theme(axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank()) + scale_y_continuous(label=percent)+labs(fill='% of Classmates')+
-    scale_fill_manual(values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02'))+
-    labs(y='')
-
-@
+
+\includegraphics[width=\maxwidth]{figure/classmates-1}
+
 \caption{Distributions of classmates' statuses at each reassignment in
@@
 story, but isn't all of it.
-Across years and states, in about \Sexpr{round(behind*100)}\% of
+Across years and states, in about 40\% of
 reassignments, at least 75\% of the rest of the class had exited the
@@
 worked certain sections.
-That is, \Sexpr{mi2cp}\% of the instances in which year-two Michigan
+That is, 65\% of the instances in which year-two Michigan
 students were reassigned from a section, at least 75\% of their
@@
 
-<<secOrder,include=FALSE,dependson=c('data','setup')>>=
-#### what's the next section each student works?
-secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))
-
-secOrder$cp <- secOrder$status=='changed placement'
-secOrder$mast <- secOrder$status=='graduated'
-
-secOrderCP <- subset(secOrder,status=='changed placement')
-
-secOrderCP$Unit <- secOrderCP$unit
-secOrderCP$Unit[grep('unit-conversions',secOrderCP$Unit)] <- 'unit-conversions'
-
-secOrderCP$unitName <- factor(secOrderCP$Unit,levels=units)
-levels(secOrderCP$unitName) <- UnitName
-
-secOrderCP$nextUnit[grep('unit-conversions',secOrderCP$nextUnit)] <- 'unit-conversions'
-
-secOrderCP$nextUnitName <- factor(secOrderCP$nextUnit,levels=units)
-levels(secOrderCP$nextUnitName) <- UnitName
-
-@
-
 \begin{figure}
@@
   \begin{subfigure}{5in}
-<<transition1,fig.height=3,fig.width=5,dependson='secOrder'>>=
-
-trans1 <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))
-trans1o <- trans1
-
-## top senders and receivers
-cols <- which(colSums(trans1)>30)
-rows <- which(rowSums(trans1)>30)
-
-cols2 <- cols
-rows2 <- rows
-
-## top 2 receivers for each sender, vice versa
-for(r in rows)
-    while(sum(trans1[r,cols2])/sum(trans1[r,])<0.75){
-    nextBest <- which.max(trans1[r,seq(ncol(trans1))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans1)==names(nextBest)))
-}
-
-for(cc in cols){
-    while(sum(trans1[rows2,cc])/sum(trans1[,cc])<0.75){
-        nextBest <- which.max(trans1[seq(nrow(trans1))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans1)==names(nextBest)))
-    }
-
-}
-
-cols <- sort(unique(c(cols2,rows2)))
-rows <- sort(unique(rows2))
-
-trans1 <- trans1[rows,cols]
-
-propTotCP1 <- sum(trans1)/sum(secOrderCP$year==1)
-
-uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
-nsec <- length(uuu)
-
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
-text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
-
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans1))
-  text(1,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(1,NA),cex=0.75)
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans1))
-  text(2,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(0,NA),cex=0.75)
-
-for(rr in rownames(trans1))
- for(cc in colnames(trans1)){
-#     if(trans1[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans1[rr,cc]/sum(trans1o[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans1[rr,cc]>50,1,trans1[rr,cc]/50)))
-}
-
-@
+
+\includegraphics[width=\maxwidth]{figure/transition1-1}
+
 \end{subfigure}
@@
   \centering
-<<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-conv <- (nsec+2)/11
-text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
-text(1,9*conv,'Thickness:\n Percent',cex=0.75)
-text(.5,7.5*conv,'25%')
-text(.5,6.5*conv,'50%')
-text(.5,5.5*conv,'75%')
-arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
-text(1,4.2*conv,'Darkness:\n Number',cex=0.75)
-text(.5,3*conv,'10')
-text(.5,2*conv,'25')
-text(.5,1*conv,'50+')
-arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))
-
-@
+
+\includegraphics[width=\maxwidth]{figure/trans1legend-1}
+
 \end{subfigure}
@@
   \begin{subfigure}{5in}
-<<transition2,fig.height=5,fig.width=5,dependson='secOrder'>>=
-
-trans2 <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))
-trans2o <- trans2
-
-## top senders and receivers
-cols <- which(colSums(trans2)>30)
-rows <- which(rowSums(trans2)>30)
-
-cols2 <- cols
-rows2 <- rows
-
-## top 2 receivers for each sender, vice versa
-for(r in rows)
-    while(sum(trans2[r,cols2])/sum(trans2[r,])<0.75){
-    nextBest <- which.max(trans2[r,seq(ncol(trans2))[-cols2]])
-    cols2 <- c(cols2,which(colnames(trans2)==names(nextBest)))
-}
-
-for(cc in cols){
- #   print('new col')
-  #  print(cc)
-    while(sum(trans2[rows2,cc])/sum(trans2[,cc])<0.75){
-        nextBest <- which.max(trans2[seq(nrow(trans2))[-rows2],cc])
-        rows2 <- c(rows2,which(rownames(trans2)==names(nextBest)))
-    }
-
-}
-
-cols <- sort(unique(c(cols2,rows2)))
-rows <- sort(unique(rows2))
-
-trans2 <- trans2[rows,cols]
-
-propTotCP2 <- sum(trans2)/sum(secOrderCP$year==1)
-
-nsec <- nrow(trans2)
-uuu <- levels(secOrderCP$unitName)
-uuu <- uuu[uuu%in%c(colnames(trans2),rownames(trans2))]
-nsec <- length(uuu)
-
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(-.5,3.5),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-text(1,nsec+1.5,expression(underline(bold(paste(From)))),adj=c(1,NA))
-text(2,nsec+1.5,expression(underline(bold(paste(To)))),adj=c(0,NA))
-
-for(i in 1:nsec) if(uuu[i]%in%rownames(trans2))
-  text(1,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(1,NA),cex=0.75)
-for(i in 1:nsec) if(uuu[i]%in%colnames(trans2))
-  text(2,nsec-i+1,paste0(uuu[i],' (',which(levels(secOrderCP$unitName)==uuu[i]),')'),adj=c(0,NA),cex=0.75)
-
-for(rr in rownames(trans2))
- for(cc in colnames(trans2)){
-#     if(trans2[rr,cc]>20){
-      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2o[rr,]),
-       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
-}
-@
+
+\includegraphics[width=\maxwidth]{figure/transition2-1}
+
 \end{subfigure}
 \begin{subfigure}{1in}
-<<trans2legend,fig.width=1,fig.height=3,dependson='transition2'>>=
-par(mar=c(.1,.1,.1,.1))
-plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
-conv <- (nsec+2)/11
-text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
-text(1,9*conv,'Thickness:\n Percent',cex=0.75)
-text(.5,7.5*conv,'25%')
-text(.5,6.5*conv,'50%')
-text(.5,5.5*conv,'75%')
-arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
-text(1,4.2*conv,'Darkness:\n Number',cex=0.75)
-text(.5,3*conv,'10')
-text(.5,2*conv,'25')
-text(.5,1*conv,'50+')
-arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)),lwd=2.5)
-
-@
+
+\includegraphics[width=\maxwidth]{figure/trans2legend-1}
+
 \end{subfigure}
@@
 receiving column.
-All in all, \Sexpr{round(propTotCP1*100)}\% of the first-year reassignments and
-\Sexpr{round(propTotCP1*100)}\% of the second-year reassignments are
+All in all, 75\% of the first-year reassignments and
+75\% of the second-year reassignments are
 captured in the figures.
@@
 Graphs'' on the right is fairly thick, since
-\Sexpr{round(trans1o['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']/
-  sum(trans1o['4-Quadrant Linear Graphs',])*100)}\%
+44\%
 of the reassignments begun in ``4-Quadrant Linear Graphs'' end in the
@@
 Yet it is also fairly faint, since it only represents
-\Sexpr{trans1o['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']}
+12
 reassignments.
@@
 students to be reassigned to the next unit in the curriculum---this pattern comprises
-\Sexpr{round(sum(diag(trans1o[,-1]))/sum(trans1o)*100)}\% of the reassignments in year 1 and
-\Sexpr{round(sum(diag(trans2o[,-1]))/sum(trans2o)*100)}\% of those in year2.
+52\% of the reassignments in year 1 and
+24\% of those in year2.
 On the other hand, it is rare for students to be reassigned within the same unit
-(these reassignments account for \Sexpr{round(sum(diag(trans1o))/sum(trans1o)*100)}\% and \Sexpr{round(sum(diag(trans2o))/sum(trans2o)*100)}\% in the two years).
+(these reassignments account for 15\% and 18\% in the two years).
 It is also rare for students to be reassigned to units earlier in the curriculum (comprising
-\Sexpr{round(sum(trans1o[lower.tri(trans1o)])/sum(trans1o)*100)}\% and
-\Sexpr{round(sum(trans2o[lower.tri(trans2o)])/sum(trans2o)*100)}\%).
+2\% and
+8\%).
 
@@
 Similarly,
-\Sexpr{trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']} students
-(\Sexpr{round(trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']/sum(trans1o['Ind. Variables in Lin. Mod.',])*100)}\%)
+9 students
+(27\%)
 of the students reassigned from the section ``Ind. Variables in Lin. Mod.'' were placed 13 units later in ``Systems of Lin. Eq.''
-and all \Sexpr{trans1o['Quad. Mod. & Area','Exponents']} of students reassigned from ``Quad. Mod. \& Area'' were placed in ``Exponents.''
+and all 62 of students reassigned from ``Quad. Mod. \& Area'' were placed in ``Exponents.''
 This suggests that some teachers may have considered these units, ``Systems of Lin. Eq.'' and ``Exponents'' to contain particularly important material.
@@
 Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Eq. w/ Var's Both Sides,'' were common destinations from a wide variety of earlier units, suggesting a strong interest in those units.
-The majority (\Sexpr{round(trans2o['Probability','Probability']/sum(trans2o['Probability',])*100)}\%)
+The majority (76\%)
 of the students reassigned from ``Probability'' were placed in another section of the same unit, perhaps suggesting problems with some of that unit's sections; in fact, all of the students reassigned within the ``Probability'' unit were reassigned from one its first three sections (of seven).
-Finally, \Sexpr{trans2['Pythagorean Theorem','Product Rule for Exponents']}
-(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignments from sections in the ``Pythagorean Theorem'' unit ended in the earlier ``Product Rule for Exponents'' sections.
+Finally, 44
+(81\%) of the reassignments from sections in the ``Pythagorean Theorem'' unit ended in the earlier ``Product Rule for Exponents'' sections.
 
@@
 
-<<effects,include=FALSE,dependson=c('data','setup')>>=
-cpDat <- data%>%filter(!is.na(status)& status!='final_or_incomplete')%>%
- group_by(field_id,year,state,section,unit,race,sex,grade,spec_speced,xirt,spec_gifted,
-  spec_esl,frl,pretest,y_yirt,classid2,schoolid2)%>%
- summarize(status=max(status,na.rm=TRUE),date=max(date),totalTime=sum(total_t1,na.rm=TRUE),
-  ttNA=sum(is.na(total_t1)),nprob=n(),probNA=sum(is.na(Prob1)))%>%ungroup()%>%
- group_by(field_id,year,state,race,sex,grade,spec_speced,xirt,spec_gifted,spec_esl,frl,pretest,y_yirt,
-  classid2,schoolid2)%>%
- summarize(nsec=n(),nprob=sum(nprob,na.rm=TRUE),totalTime=sum(totalTime,na.rm=TRUE),
-  timeNA=sum(ttNA>0),probNA=sum(probNA>0),ncp=sum(status=='changed placement'),
-  nprom=sum(status=='promoted'))%>%mutate(gainScore=y_yirt-pretest)%>%droplevels()
-
-## mode imputation:
-for(vv in c('race','sex','grade','spec_speced','spec_gifted','spec_esl','frl')){
-    num <- is.numeric(cpDat[[vv]])
-    cpDat[[paste0(vv,'MIS')]] <- is.na(cpDat[[vv]])
-    cpDat[[vv]][is.na(cpDat[[vv]])] <- names(which.max(table(cpDat[[vv]])))
-    if(num) cpDat[[vv]] <- as.numeric(cpDat[[vv]])
-
-}
-
-cpDat$grade <- factor(ifelse(cpDat$grade==9,'9','10+'))
-levels(cpDat$race) <- list(White=c('WHITE NON-HISPANIC','ASIAN / PACIFIC ISLANDER'),Black=c('BLACK NON-HISPANIC','OTHER RACE / MULTI-RACIAL'),Hispanic=c('HISPANIC','AMERICAN INDIAN / ALASKAN NATIVE'))
-
-cpDat <- mutate(cpDat,totalTime=totalTime/3600000,totalTime=ifelse(totalTime<0,NA,totalTime),
-                totalTime=ifelse(totalTime>360,NA,totalTime))
-
-cpDat$ncpCat <- factor(ifelse(cpDat$ncp>=4,'4+',cpDat$ncp))
-cpDat$everCP <- cpDat$ncp>0
-
-exirt <- data[match(cpDat$field_id,data$field_id),grep('Exirt2',names(data))]
-names(exirt) <- gsub('_0','_',names(exirt))
-for(n in names(exirt)) cpDat[[n]] <- exirt[[n]]
-
-@
-<<models,include=FALSE,dependson='effects'>>=
-#### models!!
-## ind var: ncp
-addCovs <- .~.+pretest+race+sex+grade+spec_speced+spec_gifted+spec_esl+frl+frlMIS
-mod1 <- lm(y_yirt~ncp+classid2,data=cpDat)
-mod1.1 <- update(mod1,addCovs)
-
-## ind var: ncpCat
-mod2 <- lm(y_yirt~ncpCat+classid2,data=cpDat)
-mod2.1 <- update(mod2,addCovs)
-
-## ind var: everCP
-mod3 <- lm(y_yirt~everCP+classid2,data=cpDat)
-mod3.1 <- update(mod3,addCovs)
-
-miMod <- function(mod){
-  if(inherits(mod,'lm')) CCC <- coef
-  if(inherits(mod,'lmerMod')) CCC <- fixef
-    ests <- NULL
-    covs <- list()
-    for(i in 1:20){
-        formNew <- as.formula(paste0('.~.-pretest+Exirt2_',i))
-        modNew <- update(mod,formNew)
-        ests <- cbind(ests,CCC(modNew))
-        covs[[i]] <- vcov(modNew)
-    }
-    return(list(ests=ests,covs=covs))
-}
-miPool <- function(fitMods){
-    ests <- fitMods$ests
-    covs <- fitMods$covs
-    interest <- grep('CP',rownames(ests),ignore.case=TRUE)
-    SEs <- sqrt(rowMeans(do.call('cbind',lapply(covs,function(cc) diag(as.matrix(cc[interest,interest])))))+
-                1.05*apply(rbind(ests[interest,]),1,var))
-    ests <- rowMeans(rbind(ests[interest,]))
-    cbind(ests,SEs)
-}
-
-mi <- function(mod){
-    miPool(miMod(mod))
-}
-
-adjustedEsts1 <- list()
-for(i in 1:3){
-  mm <- paste0('mod',i)
-  mm1 <- paste0('mod',i,'.1')
-  adjustedEsts1[[mm]] <- mi(get(mm))
-  adjustedEsts1[[mm1]] <- mi(get(mm1))
-}
-@
-<<sensitivity,include=FALSE,dependson='models'>>=
-source('hhh.r')
-### we're only computing sensitivity for omitting pretest
-X <- as.data.frame(model.matrix(mod3.1)[,-1])
-Ts <- Tz(X = X,treatment = 'everCPTRUE')
-X$everCPTRUE <- NULL
-R <-  partialCor(cpDat$y_yirt,X,'pretest')^2
-multPre <- MEmult(T=Ts['pretest'],R=R,df=mod3.1$df,q=1.96)
-
-Rhos <- vapply(names(X)[-grep('classid2|pretest',names(X))],
-               function(nn) partialCor(cpDat$y_yirt,X,nn),1)^2
-multOth <- MEmult(T=max(abs(Ts[names(Rhos)])),max(Rhos),df=mod3.1$df,q=1.96)
-
-@
-
-<<heterogeneityMods,include=FALSE,dependson='models'>>=
-### heterogeneity
-## year
-cpDat$Year=as.factor(cpDat$year)
-mod4 <- lmer(y_yirt~everCP*Year+state+(1|classid2)+(1|schoolid2),data=cpDat)
-mod4.1 <- update(mod4,addCovs)
-
-## classroom
-mod5 <- lmer(y_yirt~everCP+year+state+(everCP|classid2)+(1|schoolid2),data=cpDat)
-mod5.1 <- update(mod5,addCovs)
-@
-
-\begin{figure}
-  \centering
-<<cpYyear,fig.width=6,fig.height=4,dependson='effects'>>=
-levels(cpDat$Year) <- list(`Year 1`='1',`Year 2`='2')
-ggplot(cpDat,aes(ncp,gainScore))+geom_jitter(height = 0,width=.3)+geom_boxplot(aes(group=ncp),alpha=0.5)+geom_smooth(method='loess')+facet_grid(Year~.)+labs(x='# of Reassignments',y='Gain Score (posttest - pretest)') + scale_x_continuous(breaks=seq(0,max(cpDat$ncp)))
-@
+\begin{figure}
+  \centering
+
+\includegraphics[width=\maxwidth]{figure/cpYyear-1}
+
 \caption{Students' Gain scores (posttest minus pre-test) versus the number of times the were reassigned over the course of the year (jittered), with a Loess smoother added.}
@@
 pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
-\Sexpr{round(mean(is.na(cpDat[['xirt']]))*100)}\%).
+15\%).
 To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2014effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligibility.
@@
 \hline
-<<effectResults,results='asis',dependson='models'>>=
-## model 3
-cat('\\makecell[r]{No\\\\Covariates}&')
-cat(round(adjustedEsts1[['mod3']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod3']][1,'SEs'],2),'&')
-## model 2
-res <- adjustedEsts1[['mod2']]
-cat('\\makecell[l]{')
-for(rr in 1:(nrow(res)-1))
-  cat(rr,':  ',round(res[rr,1],2),'$\\pm$',round(2*res[rr,'SEs'],2),'\\\\')
-cat(nrow(res),'+: ',round(res[nrow(res),1],2),'$\\pm$',round(2*res[nrow(res),'SEs'],2),'}&')
-## model 1
-    cat(round(adjustedEsts1[['mod1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod1']][1,'SEs'],2),'\\\\')
-cat('\n')
-cat('\\hline')
-###################################
-cat('\\makecell[r]{Covariate\\\\ Adjusted}&')
-cat(    round(adjustedEsts1[['mod3.1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod3.1']][1,'SEs'],2),'&')
-## model 2
-cat('\\makecell[l]{')
-res <- adjustedEsts1[['mod2.1']]
-for(rr in 1:(nrow(res)-1)) cat(rr,': ',round(res[rr,1],2),'$\\pm$',round(2*res[rr,'SEs'],2),'\\\\')
-cat(nrow(res),'+: ',round(res[nrow(res),1],2),'$\\pm$',round(2*res[nrow(res),'SEs'],2),'}&')
-## model 1
-cat(
-    round(adjustedEsts1[['mod1.1']][1,'ests'],2),'$\\pm$',round(2*adjustedEsts1[['mod1.1']][1,'SEs'],2),'\\\\')
-
-@
+\makecell[r]{No\\Covariates}&-0.14 $\pm$ 0.06 &\makecell[l]{1 :   -0.15 $\pm$ 0.07 \\2 :   -0.15 $\pm$ 0.1 \\3 :   -0.07 $\pm$ 0.15 \\4 +:  -0.12 $\pm$ 0.17 }&-0.04 $\pm$ 0.03 \\
+\hline\makecell[r]{Covariate\\ Adjusted}&-0.15 $\pm$ 0.06 &\makecell[l]{1 :  -0.15 $\pm$ 0.07 \\2 :  -0.16 $\pm$ 0.1 \\3 :  -0.08 $\pm$ 0.15 \\4 +:  -0.14 $\pm$ 0.17 }&-0.04 $\pm$ 0.03 \\
 \hline
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests']*100)}\% of that.
+The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is -15\% of that.
 
@@
 hypothetical confounder at most as important as pretest, is
-\Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests'],2)} $\pm$
-\Sexpr{round(adjustedEsts1[['mod3.1']][1,'SEs']*multPre,2)}.
+-0.15 $\pm$
+0.14.
 This interval is quite wide, implying that such a covariate could
@@
 covariates---is
-\Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests'],2)} $\pm$
-\Sexpr{round(adjustedEsts1[['mod3.1']][1,'SEs']*multOth,2)}.
+-0.15 $\pm$
+0.07.
 This interval is substantially tighter.
@@
 these effects was
-\Sexpr{round(sqrt(summary(mod5.1)$varcor$classid2[2,2]),2)},
+0.18,
 larger than the average effect itself.
@@
   \centering
-<<hetroPlot,fig.width=6,fig.height=3,dependson='heterogeneityMods'>>=
-re <- ranef(mod5.1,condVar=TRUE)
-reVar <- attr(re$classid2,'postVar')[2,2,]
-feVar <- vcov(mod5.1)['everCPTRUE','everCPTRUE']
-
-hetPlotDat <- data.frame(eff=re$classid2$everCPTRUE+fixef(mod5.1)['everCPTRUE'],
-                         se=sqrt(reVar+feVar),
-                         year=as.factor(data$year[match(rownames(re$classid2),data$classid2)]),
-                         state=data$state[match(rownames(re$classid2),data$classid2)])
-hetPlotDat$ord <- rank(hetPlotDat$eff)
-
-ggplot(hetPlotDat,aes(ord,eff))+geom_point()+
-#    geom_errorbar(aes(ymin=eff-2*se,ymax=eff+2*se),lwd=.2)+
-    geom_hline(yintercept=0,lty=2)+
-#    geom_hline(yintercept=fixef(mod5.1)['everCPTRUE'])
-    scale_x_continuous(labels=NULL,breaks=NULL)+
-    labs(y='Effect of Any Reassignment',x='Classroom')
-@
+
+\includegraphics[width=\maxwidth]{figure/hetroPlot-1}
+
 \caption{The effect of being reassigned at least once, in each

### 2017-11-06 paper.Rnw
conclusion+summaries+edits

@@
 
-<<data1,include=FALSE,cache=FALSE>>=
-dataCache <- grep('data',list.files('cache'),ignore.case=FALSE,value=TRUE)[1]
-
-if(file.info('dataMerge.r')$mtime>file.info(paste0('cache/',dataCache))$mtime){
-    if(file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
-        load('cpPaper.RData')
-    } else source('dataMerge.r')
-}
-@
-
 \begin{document}
@@
 A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students' scores on an Algebra-I posttest by about one fifth of a standard deviation \citep{pane2014effectiveness}.
-CTAI's success in this experiment would seem to vindicate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
-
-However, the theory underlying CTAI does not always determine its use.
-To be sure, the software has a standard set of algebra topics, divided into units and further into sections; and a standard sequence for presenting them.
-But this precise curriculum is not mandatory.
-At the request of educators, it can be customized by altering what units or sections are included (including, possibly, material from a different standard curriculum such as geometry), as well as their sequence, to conform to local or state standards.
-Further, teachers have the option of moving students within the curriculum, regardless of the software's estimate of their skill mastery.
-
-This article examines teachers' and schools' non-adherence to the standard, mastery-based CTAI curriculum, using data from the RAND study, a seven-state randomized controlled trial of CTAI in high schools and middle schools.
-The study found a significant positive effect of CTAI in high schools, during their second year of implementation but not the first.
-Students in the treatment group of the study were enrolled in one or more of the standard or customized curriculum during their participation in the study, and the software logged aspects of their usage, including time spent, sections encountered, and whether the software judged the students to have mastered the sections.
-Both Carnegie Learning and the researchers running the study restricted their software support and oversight to what is typically provided outside of an experimental context.
-Thus, the data from the study reflect typical software usage.
+CTAI's success in this experiment would seem to validate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
+
+However, the theory underlying CTAI does not always determine its
+use. To be sure, the software has a standard set of algebra topics,
+divided into units and further into sections; and a standard sequence
+for presenting them. But this precise curriculum is not mandatory. At
+the request of educators, it can be customized by altering what units
+or sections are included (including, possibly, material from a
+different standard curriculum such as geometry), as well as their
+sequence, to conform to local or state standards or scope and sequence
+guides. Further, teachers have the option of moving students within
+the curriculum, regardless of the software's estimate of their skill
+mastery.
+
+This article examines teachers' and schools' adherence and
+non-adherence to the standard, mastery-based CTAI curriculum, using
+data from the RAND study, a seven-state randomized controlled trial of
+CTAI in high schools and middle schools. The study found a significant
+positive effect of CTAI in high schools, during their second year of
+implementation but not the first. Students in the treatment group of
+the study were enrolled in one or more of the standard or customized
+curricula during their participation in the study, and the software
+logged aspects of their usage, including time spent, sections
+encountered, and whether the software judged the students to have
+mastered the sections. Adopting standard procedures of an
+effectiveness trial, both Carnegie Learning and the researchers
+running the study restricted their support and oversight to what is
+typically provided outside of an experimental context. Thus, the
+software data from the study reflect typical usage.
 Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects \citep{sales2015exploring,sales2016student}.
@@
 
-Here we extend the previous findings to describe, in detail, the ways
+Here we contextualize the previous findings to describe, in detail, the ways
 in which schools, teachers, and students violate mastery learning.
@@
 mastered into a new section.
-In particular, we will attempt to elucidate teachers' goals in reassigning students.
-One hypothesis is the need for teachers push ahead students who were falling behind, i.e.
-to reassign them from sections on which they were struggling, to allow them catch up with the rest of the class.
-Another hypothesis is that teachers sought to push students past easier sections to begin working on more relevant topics for them.
-A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exams, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered.
+In particular, we will attempt to elucidate teachers' goals in
+reassigning students. One hypothesis is the need for teachers to push
+ahead students who were falling behind, i.e. to reassign them from
+sections on which they were struggling, to allow them catch up with
+the rest of the class. Another hypothesis is that teachers sought to
+push students past easier sections to begin working on more relevant
+or challenging topics for them.
+A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exam, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered.
 This hypothesis may lead to an increase in reassignments as the exam approaches.
@@
 high schools, and 74 middle schools with nearly 18,700 high school
-students and 6,800 middle school students. Schools were enrolled in a
-total of 52 school districts that were distributed between urban,
+students and 6,800 middle school students participating. Schools were enrolled in a
+total of 52 school districts that were distributed among urban,
 suburban, and rural areas. Schools were matched on a set of
@@
 CTAI. Each school participated for two years, with a different cohort
-of students taking algebra the second year (with a small fraction of
+of students participating the second year (with a small fraction of
 students present in the study both years because they repeated
@@
 middle school arm. The current study focuses on high school students
-only. %, due to the large number of students in this cohort and the
+only. %\marginpar{I think it reads better to use the spelled-out names,
+%  but I'm happy to switch if you feel strongly about this.}%, due to the large number of students in this cohort and the
 % significant positive treatment effect previously found for this
@@
 
-Students' automatic progress through the Cognitive Tutor (CT) software is governed by
-curricula---sequences of sections and units embedded in the software.
+Students' automatic progress through the Cognitive Tutor (CT) software
+is normally governed by
+the sequences of sections and units embedded in the software.
 Without external meddling, the curriculum a student works on
-determines what section he or she will be directed towards next after
+determines the sequence, and thus what section he or she will be directed towards next after
 mastering (or exhausting the problems) from a previous section.
@@
 naturally, Algebra I.
-This came with three closely related variants, due to software releases.
+This came with three closely related variants, due to new software releases.
 Students requiring more remediation were able to work on a less
@@
 curricula.
-This was typically due to state standards and testing schedules.
+This was typically due to state standards, testing schedules, or local
+scope and sequence guidelines.
 These ``customized curricula'' altered the order of some sections and
-units, and were particular to schools.
+units, and were usually particular to schools.
 
@@
 worked on different curricula at the same time. As an example, two
-teachers located in KY had their students working on Algebra I
+teachers located in Kentucky had their students working on Algebra I
 throughout most of the year and then reassigned them to Algebra II in
-the last month of school. Another teacher in KY had students variously
+the last month of school. In contrast, a different teacher in Kentucky had students variously
 enrolled in three different curricula throughout the entire year
 (Bridge-to-Algebra, Algebra I, and a specialized Geometry curriculum),
-while another teacher in MI during Year 2 had students enrolled in
-three curricula sequentially throughout the year covering Algebra I
-until November, and then starting a customized curriculum and using it
-until February, and ending the year with another customized curriculum
-that was used until June. While there are numerous instances of these
+while a year 2 teacher in Michigan enrolled students in
+three curricula sequentially throughout the year: Algebra I
+until November, followed by a customized curriculum util February, and
+ending with a different customized curriculum until June. While there are numerous instances of these
 uses of multiple curricula in Year 2, there are also many occurrences
 of teachers who had their students enrolled in the standard Algebra I
-throughout the entire year.  In TX mostly, with some instances in MI,
+throughout the entire year.  In TX mostly, with some instances in Michigan,
 there were teachers who exclusively used customized curricula the
@@
 Apparently, usage decreased markedly in the second year: the median
-student worked about
+of hours worked decreased by
 \Sexpr{round(tab['Year 1','Hours']-tab['Year 2','Hours'])}
-fewer hours,
-\Sexpr{round(tab['Year 1','Problems']-tab['Year 2','Problems'])}
-fewer problems, and
-\Sexpr{round(tab['Year 1','Sections']-tab['Year 2','Sections'])} fewer
-sections in year 2 than in year 1.
+, the median number of problems decreased by
+\Sexpr{round(tab['Year 1','Problems']-tab['Year 2','Problems'])}, and
+the median number of sections decreased by
+\Sexpr{round(tab['Year 1','Sections']-tab['Year 2','Sections'])} from
+years 1 to 2.
 
@@
   software over by year and state. Students  with no timestamp data
-  (\Sexpr{sum(is.na(secByStud[['time']]))}), with negative time
-  (\Sexpr{sum(secByStud[['time']]<0,na.rm=TRUE)}) or with more than 110 hours (\Sexpr{sum(secByStud[['time']]>110,na.rm=TRUE)})
+  ($n=$\Sexpr{sum(is.na(secByStud[['time']]))}), with anomalous negative time
+  ($n=$\Sexpr{sum(secByStud[['time']]<0,na.rm=TRUE)}) or with more than 110 hours ($n=$\Sexpr{sum(secByStud[['time']]>110,na.rm=TRUE)})
   were excluded.}
@@
 years.
-Time spent varied substantially by state, with students in Texas,
-Connecticut, and New Jersey working far fewer hours than students in
+Students in Texas,
+Connecticut, and New Jersey worked far fewer hours than students in
 Kentucky, Louisiana, and Michigan.
@@
 between-state variation, and a between-student increase in Louisiana.
-One intriguing possibility is that CT usage was better tailored to
+One intriguing possibility is that CT usage may have been better tailored to
 teachers and students in the second year than in the first---perhaps
@@
 Variation in the number of units worked also increased slightly
-from years 1 to 2---the interquartile range increased in every state
+from years 1 to 2---the interquartile range (IQR) increased in every state
 except for Kentucky, where a decrease in IQR was accompanied by an
@@
 Results are similar to the data presented in Figures 2 and 3.
-However, in Year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time TX students spent mastering sections in the standard curricula.
-This difference is also seen in KY and LA.
+However, in Year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time Texas students spent mastering sections in the standard curricula.
+This difference is also seen in Kentucky and Louisiana.
 If customized curriculum are more aligned with teachers' instruction, graduating from a section may require fewer days than sections in the standard curricula, possibly because students are working on the same topics in class as they are in the software.
-Figures 4 and 5 also show that, on average, CT and TX spent more days graduating from sections, regardless whether the section was part of a standard curricula or customized curricula.
+Figures 4 and 5 also show that, on average, Connecticut and Texas spent more days graduating from sections, regardless whether the section was part of a standard curricula or customized curricula.
 One explanation for the longer time periods in these states may be due to a teacher's use of the software simultaneously with the textbook instruction.
@@
 first.
-How was this difference distributed between CTAI units?
+How was this difference distributed across CTAI units?
 
@@
 
-Figure \ref{fig:unitsWorked} shows the percentage of students with
-usage data who worked each section from the standard CTAI curriculum.
-The units are ordered, from left to right, as the are prescribed in
-the standard CTAI curriculum.
+Figure \ref{fig:unitsWorked} shows the units of algebra along the
+horizontal axis, according to their order in the standard CTAI
+curriculum.
+The vertical axis shows the percentage of students with usage data who
+worked each unit.
 
@@
 to both student ability and the amount of time allocated to CTAI
-within a classroom---but they mostly followed the same curriculum.
-Students who worked fewer units stopped earlier in the curricular
-order, and those who worked more units progressed farther.
+within a classroom---but they mostly followed the standard curriculum.
+Students who worked fewer units stopped earlier in the sequence, and those who worked more units progressed farther.
 Hence, earlier sections were worked by higher proportions of students
@@
 
-In contrast, in year 2 students were much more likely to buck the
+In contrast, in year 2 students were much more likely to depart from the
 standard unit order.
-For instance, Figure \ref{fig:unitsWorked} suggests that students
-skipped ``Unit Conversions'' and ``1 step Lin. Eq.'' for ``1st
-Quadrant Linear Graphs'' and ``Ind. Variables in Lin. Mod.''
-Both of the latter sections were worked by a greater proportion of
-students than the former sections that preceded them in the CTAI
-curriculum.
-Most strikingly, ``Lin. Eq. w/ Var's Both Sides'' was worked by a
+For instance, Figure \ref{fig:unitsWorked} suggests that some students
+skipped ``Unit Conversions'' to work on  ``1st
+Quadrant Linear Graphs'' or skipped ``1 step Linear Equations'' to work on ``Independent Variables in Linear Models''
+In both these cases, the subsequent unit was worked on by a greater
+proportion of students than the immediately prior unit.
+
+Most strikingly, ``Linear Equations with Variables on Both Sides'' was worked by a
 greater proportion of students in Year 2 than in Year 1, and by a
@@
 levels(unitLevelCust$Unit) <- UnitName
-ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.96,.82))+labs(x='',y='% Worked',color='Curriculum',title='Year 2')+scale_y_continuous(labels=percent)
+ggplot(unitLevelCust,aes(x=Unit,y=perWorked,color=cust,group=cust))+geom_point()+geom_line()+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5,size=10),legend.text=element_text(size=10),legend.position=c(.90,.82))+labs(x='',y='% Worked',color='Curriculum',title='Year 2')+scale_y_continuous(labels=percent)
 @
@@
   \Sexpr{trunc(100*min(cust[['cust']][cust[['cust']]>0.8]))}\% of
-  problems at ``Customized'' schools were from a customized curriculum,
+  problems worked by students at ``Customized'' schools were from a customized curriculum,
   and at most
@@
 That said, there were some order violations in the standard group:
-specifically, more students worked problems from units ``4-Quadrant
-Linear Graphs'' and ``Exponents'' than worked the preceding sections;
+specifically, more students worked problems from units ``2-step Linear
+Equations'' and ``Exponents'' than worked the preceding sections;
 this suggests that some teachers used the reassignment tool to
 prioritize particular topics.
-Of course, this may have occurred in schools with customized
+Of course, teacher reassignment may have occurred in schools with customized
 curricula as well---a possibility we will discuss in the next
@@
 prioritize some of the same units, for instance, ``Linear Patterns,''
-``Ind. Variables in Lin. Mod.,'' and
-``Lin. Eq. w/ Var's Both Side.''
-These topics might be emphasized on Texas standardized tests.
+``Independent Variables in Linear Models,'' and
+``Linear Equations with Variables on Both Sides''
 On the other hand, there was also variance between schools.
-For instance, one school prioritized units ``2 Step Lin. Eq.''
+For instance, one school prioritized units ``2 Step Linear Equations''
 and ``4-Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''
@@
 units, omitted the 15th (``Lin. Mod. in General Form''), and worked on
-the 16th and 17th (``Literal Eq.'' and ``Lin. Eq. w/ Var's Both Sides'').
+the 16th and 17th (``Literal Equations'' and ``Linear Equations with Variables on Both Sides'').
 In the remaining school, nearly every student worked on the first
 section, but usage decreased rapidly from there.
-Curiously, in one Michigan school which used the standard curriculum,
-no students seem to have worked on the ``Lin. Mod. \& Ratios'' section.
+In one Michigan school which used the standard curriculum,
+no students seem to have worked on the ``Linear Models \& Ratios'' section.
 
@@
 software to have mastered the material, and then ``graduate'' to the
-next section. Students who exhaust a section's material without
-mastering its skills are ``promoted'' to the next section.
-Students left uninterrupted follow the sequence of topics defined by
-their curriculum, ideally graduating from topics by showing mastery.
-However, teachers could modify a student's path within the curriculum.
-They could ``reassign'' students from their current sections to other
+next section.
+However, the software will also ``promote'' a student to the next
+section if the student exhausts a section's material without mastering
+its skills.
+Additionally, teachers are able to modify a student's path within the curriculum.
+They can ``reassign'' students from their current sections to other
 sections earlier or later in the intended sequence, including sections
-they had worked on previously.
+they worked on previously.
 Finally, if the semester ends, or a student stops using CT for some
 other reason, while in the middle of working through a
-section, he or she will leave their ``Final'' section without
-mastering all of its skills.
-All in all, each CT section begun ends in one of four possible ways:
+section, the section is designated ``final.''
+All in all, each CT section a student encounters ends in one of four possible ways:
 mastery, promotion, reassignment, or as the student's final section.
@@
 \Sexpr{round(mean(tab1b['Mastered',c('TX','KY','MI','LA')])*100)}\% of
-worked sections are mastered in every
-state other than Connecticut.
+worked sections are mastered, except in Connecticut.
 Other than in Texas, about
@@
 of sections end in promotion.
-About \Sexpr{round(tab1b['Reassigned','TX']*100)}
-and \Sexpr{round(tab1b['Reassigned','CT']*100)}\% of sections in Texas
-and Connecticut end in reassignment, which is even rarer in the other states.
-
-With the exception of Texas, sections in the second year tend to be
-completed as they were in year 1.
+About \Sexpr{round(tab1b['Reassigned','TX']*100)}\% of sections in Texas
+and \Sexpr{round(tab1b['Reassigned','CT']*100)}\%
+in Connecticut end in reassignment, which is even rarer in the other states.
+
+With the exception of Texas, sections tended to be completed similarly
+in both years.
 In Texas, however, the percentage of sections ending in reassignment increased by a factor
@@
 The proportion of Texas sections labeled ``Final'' increased as well,
-most likely due to the decrease in usage.
-
-Across states, just under
-\Sexpr{round(tab1b['Reassigned','Overall']*100)}\% of sections in year
-1 ended in reassignment, as did roughly
+most likely due to the decrease in usage, and thus the total number of
+sections.
+
+Across states, sections ended in reassignment at a rate of about
+\Sexpr{round(tab1b['Reassigned','Overall']*100)}\% in year 1 and
 \Sexpr{round(tab2b['Reassigned','Overall']*100)}\% in year 2.
@@
 working on problems above their level.
-What role did curriculum play in mastery during the CTAI effectiveness trial?
+What role did variations in the CT curriculum play in mastery during the effectiveness trial?
 
@@
 \Sexpr{round(statusCurr[['value']][statusCurr[['Curriculum']]=='Bridge-to-Algebra'
-  & statusCurr[['variable']]=='Graduated']*100)}\% of the time;
+  & statusCurr[['variable']]=='graduated']*100)}\% of the time;
 those from Algebra I were mastered
@@
 This may an overall skepticism towards the Carnegie Learning standards
-among certain schools and teachers.
+among certain schools and teachers, manifested in both adoption of
+alternative curricula and reassignment.%\marginpar{does that make more sense?}
 
@@
 To better understand the source of the variation in
-reassignment---what drives some, but not other, sections to end in
+reassignment---what drives some, but not other, sections worked by
+students to end in
 reassignment---we fit a set of multilevel models.
@@
 Surprisingly, classroom and student-level factors only accounted for
-\Sexpr{vcs[['Overall_1']]['classid',1]} and
+\Sexpr{vcs[['Overall_1']]['classid',1]}\% and
 \Sexpr{vcs[['Overall_1']]['field_id',1]}\% in year 1, respectively,
-and \Sexpr{vcs[['Overall_2']]['classid',1]} and
+and \Sexpr{vcs[['Overall_2']]['classid',1]}\% and
 \Sexpr{vcs[['Overall_2']]['field_id',1]}\% in year 2.
-The pattern was similar in Texas, where school accounted for over half
-the variation in reassignment in both years, and in Michigan.
+The pattern was similar in Texas---where school accounted for over half
+the variation in reassignment in both years---and in Michigan to a
+lesser extent.\footnote{Percentages in state specific models, in which
+  there is no between-state variance, cannot be
+  directly compared to those from the overall model.}
 In Kentucky, unit played the largest role
@@
 Other than in Kentucky in year 2, classroom never accounted for more
-than \Sexpr{sort(sapply(vcs,function(x) x['classid',1]),dec=T)}\%  of
+than \Sexpr{sort(sapply(vcs,function(x) x['classid',1]),dec=T)[2]}\%  of
 the variation.
 
-Although teachers control reassignment, their decisions are largely
-dictated by broader policies, occurring at the state or school level.
+\textbf{Summary.} Although teachers control reassignment, their decisions appear to be largely
+determined by broader policies, occurring at the state or school level.
 
@@
 
-The timing of reassignments can provide a window into what drives
+The timing of reassignments can also provide a window into what drives
 teachers' decisions to reassign students.
@@
 of the study.
-Through December of the first year, reassignment was rare, and from
+Through December of the first year, reassignment was rare. From
 January through May of year 1, between one and two percent of sections
@@
 Figure \ref{fig:byMonthState} and Table \ref{tab:byMonth} decompose
-these trends by state, Figure \ref{fig:byMonthState} for
+these trends by state, Figure \ref{fig:byMonthState} includes just
 the three states with the highest number of reassignments, Texas,
-Kentucky, and Michigan, and Table \ref{tab:byMonth} for all six states
-and overall.
+Kentucky, and Michigan. Table \ref{tab:byMonth} shows data for each of
+the six states and overall.
 For the two study years, Figure \ref{fig:byMonthState} shows the proportion of all of each
@@
 sections that ended in reassignment, and Figure \ref{fig:byMonthState}
-gives the proportion of all of each state's reassignments over the
+gives the proportion of all of the state's reassignments over the
 course of the year, that occurred in each month.)
 
-The pattern in Figure \ref{fig:byMonth} seems largely driven by Texas.
-In contrast, the bulk of Kentucky's reassignments took place in May.
+Figure \ref{fig:byMonthState} reveals patterns that are not apparent
+in Figure 10.
+For example, the bulk of Kentucky's reassignments took place in May.
 A fifth of Michigan's reassignments in year 1 occurred in October;
-for the rest of the year, Michigan roughly followed the same rough pattern as Texas.
-In year 2, the vast majority of Michigan's reassignments occurred at
+for the rest of the year, Michigan roughly followed the same pattern as Texas.
+In year 2, the vast majority of Michigan's reassignments also occurred near
 the beginning of the year, in September and October.
@@
 \end{figure}
+
+\textbf{Summary.} In Texas and overall, reassignments were more common
+in the second half of the school year than in the first half. In
+Kentucky and Michigan, they were clustered in a few specific months.
 
@@
 feature, and coordinate students' progress.
-Teachers can identify students who are behind their classmates and
+Teachers might identify students who are behind their classmates and
 reassign them to later sections.
@@
   later date, or who never worked the section at all. The bars are
-  ordered according to those proportions, in the order listed.}
+  rank ordered according to those proportions, in the order listed.}
 \label{fig:classmates}
@@
 later date, or who never worked the section at all.
-The bars are ordered according to those same proportions: first by the
+The bars are rank ordered according to those same proportions: first by the
 proportion of classmates who were promoted or mastered the section on
@@
 section on the same date or earlier.
-In fact, smaller proportions of students had graduated or been
-promoted from the same section on the same day or earlier in year 2
+The figure also reveals that the proportions of students who had graduated or been
+promoted from the same section on the same day or earlier was smaller in year 2
 than in year 1, especially in Texas and Michigan.
@@
 worked certain sections.
-That is, \Sexpr{mi2cp}\% of the instances in which year-two Michigan
+That is, in \Sexpr{mi2cp}\% of the instances in which year-two Michigan
 students were reassigned from a section, at least 75\% of their
@@
 
-All in all, three different patterns emerge from Figure
+\textbf{Summary.} Three different patterns emerge from Figure
 \ref{fig:classmates}: teachers reassigning students who have fallen
@@
 students who begin to work on particular sections.
-Each of these patterns takes place in different states and
+Each of these patterns mostly takes place in different states and
 years.
@@
 the skills in one of the CTAI sections or is wheel-spinning---working
-problems without learning---or dislikes one of the sections in a CT unit.
+problems without learning---or a teacher dislikes one of the sections in a CT unit.
 Still, the teacher wants the student to learn as much as possible from
-the student's current unit.
-Then moving the student to the next section within the same unit would
+the current unit.
+Then moving the student to the next section within the same unit might
 make sense.
@@
 Finally, a teacher who wanted his or her students to focus on a
-particular topic might reassign them all to the appropriate section
+particular topic might reassign them all to the appropriate unit
 when the time is appropriate.
@@
   right, and also included on the right are all the units on the left,
-  along with their top two receiving units. The units are numbered
+  along with those units' top two receiving units. The units are numbered
   according their order in the standard Algebra I curriculum. There is an arrow from a
@@
 \end{subfigure}
-\caption{Reassignment transition plot for study year 2.}
+\caption{Reassignment transition plot for study year 2. See caption of
+Figure \ref{fig:trans1} for details.}
 \label{fig:trans2}
@@
 and each of the top two senders for each receiving unit were also included.
-Finally, all of the sending units were also listed on the left, in the
+Finally, all of the sending units listed on the left were also included in the
 receiving column.
@@
 \Sexpr{round(sum(diag(trans1o[,-1]))/sum(trans1o)*100)}\% of the reassignments in year 1 and
-\Sexpr{round(sum(diag(trans2o[,-1]))/sum(trans2o)*100)}\% of those in year2.
+\Sexpr{round(sum(diag(trans2o[,-1]))/sum(trans2o)*100)}\% of those in
+year 2.
 On the other hand, it is rare for students to be reassigned within the same unit
@@
 The transition plots also reveal some interesting cases worth highlighting.
-In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions.'' on the part of the reassigning teachers.
+In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions'' on the part of the reassigning teachers.
 Similarly,
-\Sexpr{trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']} students
+\Sexpr{trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']}
+of the students
 (\Sexpr{round(trans1o['Ind. Variables in Lin. Mod.','Systems of Lin. Eq.']/sum(trans1o['Ind. Variables in Lin. Mod.',])*100)}\%)
-of the students reassigned from the section ``Ind. Variables in Lin. Mod.'' were placed 13 units later in ``Systems of Lin. Eq.''
-and all \Sexpr{trans1o['Quad. Mod. & Area','Exponents']} of students reassigned from ``Quad. Mod. \& Area'' were placed in ``Exponents.''
-This suggests that some teachers may have considered these units, ``Systems of Lin. Eq.'' and ``Exponents'' to contain particularly important material.
+reassigned from the section ``Independent Variables in Linear Models'' were placed 13 units later in ``Systems of Linear Equations''
+and all \Sexpr{trans1o['Quad. Mod. & Area','Exponents']} of the students reassigned from ``Quadratic Models \& Area'' were placed in ``Exponents.''
+This suggests that some teachers may have considered the units ``Systems of Linear Equations'' and ``Exponents'' to contain particularly important material.
 
 Since the total number of reassignments was higher in year 2, the corresponding plot (Figure \ref{fig:trans2}) is larger and more complex.
-It is also more common for students to be reassigned to units other than the next unit in the sequences.
+It is also more common in year 2 for students to be reassigned to units other than the next unit in the sequences.
 This may be partly due to the proliferation of customized curricula.
-Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Eq. w/ Var's Both Sides,'' were common destinations from a wide variety of earlier units, suggesting a strong interest in those units.
+Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Equations with
+Variables on Both Sides,'' were common destinations from a wide
+variety of earlier units, suggesting strong teacher interest in those units.
 The majority (\Sexpr{round(trans2o['Probability','Probability']/sum(trans2o['Probability',])*100)}\%)
@@
 Finally, \Sexpr{trans2['Pythagorean Theorem','Product Rule for Exponents']}
-(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignments from sections in the ``Pythagorean Theorem'' unit ended in the earlier ``Product Rule for Exponents'' sections.
-
-In sum, the most prevalent pattern was for students to be reassigned to the following unit, suggesting that teachers are mostly interested in helping lagging students progress.
+(\Sexpr{round(trans2['Pythagorean Theorem','Product Rule for Exponents']/sum(trans2o['Pythagorean Theorem',])*100)}\%) of the reassignments from ``Pythagorean Theorem'' ended in the earlier ``Product Rule for Exponents'' unit.
+
+\textbf{Summary.} The most prevalent pattern was for students to be reassigned
+to the following unit, suggesting that teachers may be mostly
+interested in advancing students who are behind.
 On the other hand, a number of examples of other patterns---students moving within the same unit, or to units out of sequence---appear as well, suggesting that some teachers may be finely manipulating their students' curricula.
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests']*100)}\% of that.
+The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(abs(adjustedEsts1[['mod3.1']][1,'ests'])/0.2*100)}\% of that.
 
 But what of unmeasured confounding?
+For instance, the negative effect may be due to baseline differences
+in ability, beyond what is captured in pretest scores.
 \citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
@@
 
-\section{Discussion}\ref{sec:discussion}
+\section{Summary and Discussion}\label{sec:discussion}
+
+The effectiveness of the Cognitive Tutor software presumably depends
+on how much, and how, it is used.
+This paper exploited available log data from the high-school arm of the CTAI
+effectiveness trial---in the second year, an impressive result for the
+software---to describe variation and patterns in the software's
+usage, paying particular attention to issues of mastery learning.
+We found that:
+\begin{itemize}
+ \item The amount the software was used varied widely between states
+   and decreased, overall, from years 1 to 2.
+ \item Year 2 saw the proliferation of ``customized'' curricula in
+   three states, altering which units students worked, and in what
+   order.
+ \item Year 2 saw frequent departures from the standard CTAI unit
+   sequence, driven mainly, but not entirely, by customized
+   curricula.
+ \item Section mastery:
+ \begin{itemize}
+  \item About \Sexpr{round(tab1b['Mastered','Overall']*100)}\% of
+    worked sections in year 1 and
+    \Sexpr{round(tab2b['Mastered','Overall']*100)}\% in  year 2 ended
+    with the student having mastered all of the included skills.
+  \item There were three ways students worked a section with
+    mastering its contents: exhausting its problems and being
+    promoted, being reassigned to a new section by the teacher, and
+    ending CT use altogether.
+  \item Reassignment was rare, though it was more prevalent in Texas
+    and Connecticut than in other states, and more common in year 2
+    than 1.
+  \item Mastery rates were lower for more advanced curricula.
+ \end{itemize}
+ \item Reassignment Patterns
+  \begin{itemize}
+   \item Reassignment rates were determined more by factors varying by state and school
+     than by classroom.
+   \item Reassignment was more common in the second half of the year
+     than in the first---particularly in Texas.
+   \item Depending on state and year, reassignment typically takes
+     place in one of three scenarios: teachers reassigning students
+     who have fallen behind, teachers reassigning (almost) the entire
+     class together, and teachers assigning (almost) all students
+     \emph{out} of a particular section.
+   \item Students are most commonly reassigned to the next unit in the
+     curriculum---suggesting that teachers may be advancing lagging
+     students---but in some cases they might be finely manipulating
+     their students' curricula.
+   \end{itemize}
+ \item Reassignment appears to lower students' performance on the
+   post-test relative to that of their peers; however, the effect
+   appears to vary widely between classrooms (and, presumably, how
+   reassignment is used).
+ \end{itemize}
+
+In practice, mastery learning and topic scaffolding often unfolded
+quite differently from the vision of CT's designers.
+In some cases, the departures were a matter of a student's inability
+to achieve mastery quickly enough, and in other cases departures were
+driven by educators' preferences that ran counter to CTAI's design.
+
+Notably, both the amount of usage and fidelity to CTAI's design
+decreased from years 1 to 2---just as the estimated \emph{effect} of
+CTAI increased.
+In year 2, when the effect was substantial, students spent less time,
+and followed the CTAI curriculum and guidelines less closely than in
+year 1, when the estimated effect was negative but statistically
+insignificant.
+This raises questions as to the roles of structure and mastery in
+CTAI's effectiveness.
+Does flexibility lead to higher effects? Is mastery learning an
+important mechanism for CTAI?
+Answers to these questions could prove crucial for optimizing the
+realized effectiveness of CTAI and other intelligent tutors.
+
+On the other hand, reassignment appears to hurt students' performance
+on the posttest, relative to their classmates (though based on data
+from a randomized experiment, this analysis was observational, so
+confounding can't be ruled out).
+Does this, in contrast, suggest that achieving mastery is an important
+component of effective intelligent tutoring?
+
+As educational technology spreads, attention to the details of
+implementation may yield important insights---or important
+questions---about effectiveness, and the science of when
+intelligent tutors work, and when they don't.
 

### 2017-11-06 paper.tex
conclusion+summaries+edits

@@
 state-by-state comparisons (but included analyses that pool across states).
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Mon Oct 23 15:09:52 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Nov  6 13:55:49 2017
 \begin{table}[ht]
@@
 
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Mon Oct 23 15:23:21 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Nov  6 13:58:09 2017
 \begin{table}[ht]
@@
 Apparently, usage decreased markedly in the second year: the median
-student worked about
+of hours worked decreased by
 10
-fewer hours,
-81
-fewer problems, and
-6 fewer
-sections in year 2 than in year 1.
+, the median number of problems decreased by
+81, and
+the median number of sections decreased by
+6 from
+years 1 to 2.
 
@@
   software over by year and state. Students  with no timestamp data
-  (193), with negative time
-  (3) or with more than 110 hours (91)
+  ($n=$193), with anomalous negative time
+  ($n=$3) or with more than 110 hours ($n=$91)
   were excluded.}
@@
   90\% of
-  problems at ``Customized'' schools were from a customized curriculum,
+  problems worked by students at ``Customized'' schools were from a customized curriculum,
   and at most
@@
 85\% of
-worked sections are mastered in every
-state other than Connecticut.
+worked sections are mastered, except in Connecticut.
 Other than in Texas, about
@@
 of sections end in promotion.
-About 3
-and 5\% of sections in Texas
-and Connecticut end in reassignment, which is even rarer in the other states.
-
-With the exception of Texas, sections in the second year tend to be
-completed as they were in year 1.
+About 3\% of sections in Texas
+and 5\%
+in Connecticut end in reassignment, which is even rarer in the other states.
+
+With the exception of Texas, sections tended to be completed similarly
+in both years.
 In Texas, however, the percentage of sections ending in reassignment increased by a factor
@@
 The proportion of Texas sections labeled ``Final'' increased as well,
-most likely due to the decrease in usage.
-
-Across states, just under
-1\% of sections in year
-1 ended in reassignment, as did roughly
+most likely due to the decrease in usage, and thus the total number of
+sections.
+
+Across states, sections ended in reassignment at a rate of about
+1\% in year 1 and
 2\% in year 2.
@@
 Surprisingly, classroom and student-level factors only accounted for
-7 and
+7\% and
 3\% in year 1, respectively,
-and 6 and
+and 6\% and
 1\% in year 2.
-The pattern was similar in Texas, where school accounted for over half
-the variation in reassignment in both years, and in Michigan.
+The pattern was similar in Texas---where school accounted for over half
+the variation in reassignment in both years---and in Michigan to a
+lesser extent.\footnote{Percentages in state specific models, in which
+  there is no between-state variance, cannot be
+  directly compared to those from the overall model.}
 In Kentucky, unit played the largest role
@@
 Other than in Kentucky in year 2, classroom never accounted for more
-than 50, 12, 12, 10, 7, 6, 5, 2\%  of
+than 12\%  of
 the variation.
 
-Although teachers control reassignment, their decisions are largely
-dictated by broader policies, occurring at the state or school level.
+\textbf{Summary.} Although teachers control reassignment, their decisions appear to be largely
+determined by broader policies, occurring at the state or school level.
 
@@
     &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
-% latex table generated in R 3.3.2 by xtable 1.8-2 package
-% Mon Oct 23 10:33:17 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Nov  6 14:19:52 2017
  TX &  18 & 582 & 281 & 932 \\ 
@@
 worked certain sections.
-That is, 65\% of the instances in which year-two Michigan
+That is, in 65\% of the instances in which year-two Michigan
 students were reassigned from a section, at least 75\% of their
@@
 52\% of the reassignments in year 1 and
-24\% of those in year2.
+24\% of those in
+year 2.
 On the other hand, it is rare for students to be reassigned within the same unit
@@
 The transition plots also reveal some interesting cases worth highlighting.
-In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions.'' on the part of the reassigning teachers.
+In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions'' on the part of the reassigning teachers.
 Similarly,
-9 students
+9
+of the students
 (27\%)
-of the students reassigned from the section ``Ind. Variables in Lin. Mod.'' were placed 13 units later in ``Systems of Lin. Eq.''
-and all 62 of students reassigned from ``Quad. Mod. \& Area'' were placed in ``Exponents.''
-This suggests that some teachers may have considered these units, ``Systems of Lin. Eq.'' and ``Exponents'' to contain particularly important material.
+reassigned from the section ``Independent Variables in Linear Models'' were placed 13 units later in ``Systems of Linear Equations''
+and all 62 of the students reassigned from ``Quadratic Models \& Area'' were placed in ``Exponents.''
+This suggests that some teachers may have considered the units ``Systems of Linear Equations'' and ``Exponents'' to contain particularly important material.
 
 Since the total number of reassignments was higher in year 2, the corresponding plot (Figure \ref{fig:trans2}) is larger and more complex.
-It is also more common for students to be reassigned to units other than the next unit in the sequences.
+It is also more common in year 2 for students to be reassigned to units other than the next unit in the sequences.
 This may be partly due to the proliferation of customized curricula.
-Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Eq. w/ Var's Both Sides,'' were common destinations from a wide variety of earlier units, suggesting a strong interest in those units.
+Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Equations with
+Variables on Both Sides,'' were common destinations from a wide
+variety of earlier units, suggesting strong teacher interest in those units.
 The majority (76\%)
@@
 Finally, 44
-(81\%) of the reassignments from sections in the ``Pythagorean Theorem'' unit ended in the earlier ``Product Rule for Exponents'' sections.
-
-In sum, the most prevalent pattern was for students to be reassigned to the following unit, suggesting that teachers are mostly interested in helping lagging students progress.
+(81\%) of the reassignments from ``Pythagorean Theorem'' ended in the earlier ``Product Rule for Exponents'' unit.
+
+\textbf{Summary.} The most prevalent pattern was for students to be reassigned
+to the following unit, suggesting that teachers may be mostly
+interested in advancing students who are behind.
 On the other hand, a number of examples of other patterns---students moving within the same unit, or to units out of sequence---appear as well, suggesting that some teachers may be finely manipulating their students' curricula.
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is -15\% of that.
+The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is 73\% of that.
 
 But what of unmeasured confounding?
+For instance, the negative effect may be due to baseline differences
+in ability, beyond what is captured in pretest scores.
 \citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
@@
 
-\section{Discussion}\ref{sec:discussion}
+\section{Summary and Discussion}\label{sec:discussion}
+
+The effectiveness of the Cognitive Tutor software presumably depends
+on how much, and how, it is used.
+This paper exploited available log data from the high-school arm of the CTAI
+effectiveness trial---in the second year, an impressive result for the
+software---to describe variation and patterns in the software's
+usage, paying particular attention to issues of mastery learning.
+We found that:
+\begin{itemize}
+ \item The amount the software was used varied widely between states
+   and decreased, overall, from years 1 to 2.
+ \item Year 2 saw the proliferation of ``customized'' curricula in
+   three states, altering which units students worked, and in what
+   order.
+ \item Year 2 saw frequent departures from the standard CTAI unit
+   sequence, driven mainly, but not entirely, by customized
+   curricula.
+ \item Section mastery:
+ \begin{itemize}
+  \item About 85\% of
+    worked sections in year 1 and
+    81\% in  year 2 ended
+    with the student having mastered all of the included skills.
+  \item There were three ways students worked a section with
+    mastering its contents: exhausting its problems and being
+    promoted, being reassigned to a new section by the teacher, and
+    ending CT use altogether.
+  \item Reassignment was rare, though it was more prevalent in Texas
+    and Connecticut than in other states, and more common in year 2
+    than 1.
+  \item Mastery rates were lower for more advanced curricula.
+ \end{itemize}
+ \item Reassignment Patterns
+  \begin{itemize}
+   \item Reassignment rates were determined more by factors varying by state and school
+     than by classroom.
+   \item Reassignment was more common in the second half of the year
+     than in the first---particularly in Texas.
+   \item Depending on state and year, reassignment typically takes
+     place in one of three scenarios: teachers reassigning students
+     who have fallen behind, teachers reassigning (almost) the entire
+     class together, and teachers assigning (almost) all students
+     \emph{out} of a particular section.
+   \item Students are most commonly reassigned to the next unit in the
+     curriculum---suggesting that teachers may be advancing lagging
+     students---but in some cases they might be finely manipulating
+     their students' curricula.
+   \end{itemize}
+ \item Reassignment appears to lower students' performance on the
+   post-test relative to that of their peers; however, the effect
+   appears to vary widely between classrooms (and, presumably, how
+   reassignment is used).
+ \end{itemize}
+
+In practice, mastery learning and topic scaffolding often unfolded
+quite differently from the vision of CT's designers.
+In some cases, the departures were a matter of a student's inability
+to achieve mastery quickly enough, and in other cases departures were
+driven by educators' preferences that ran counter to CTAI's design.
+
+Notably, both the amount of usage and fidelity to CTAI's design
+decreased from years 1 to 2---just as the estimated \emph{effect} of
+CTAI increased.
+In year 2, when the effect was substantial, students spent less time,
+and followed the CTAI curriculum and guidelines less closely than in
+year 1, when the estimated effect was negative but statistically
+insignificant.
+This raises questions as to the roles of structure and mastery in
+CTAI's effectiveness.
+Does flexibility lead to higher effects? Is mastery learning an
+important mechanism for CTAI?
+Answers to these questions could prove crucial for optimizing the
+realized effectiveness of CTAI and other intelligent tutors.
+
+On the other hand, reassignment appears to hurt students' performance
+on the posttest, relative to their classmates (though based on data
+from a randomized experiment, this analysis was observational, so
+confounding can't be ruled out).
+Does this, in contrast, suggest that achieving mastery is an important
+component of effective intelligent tutoring?
+
+As educational technology spreads, attention to the details of
+implementation may yield important insights---or important
+questions---about effectiveness, and the science of when
+intelligent tutors work, and when they don't.
 

### 2017-12-05 paper.Rnw
couple tweaks
added a "total" line to one of the tables and fixed the formatting on
table 1

@@
 <<nByState,dependson='data'>>=
-xtable(data%>%group_by(state,year)%>%summarize(n=n_distinct(field_id))%>%dcast(year~state),
-       caption='Numbers of students in the sample by state and study year',label='tab:nByState')
+sampleSizeTable <- data%>%group_by(state,year)%>%summarize(n=n_distinct(field_id))%>%dcast(year~state)
+sampleSizeTable <- sampleSizeTable[,-1]
+rownames(sampleSizeTable) <- c('Year 1','Year 2')
+
+print(xtable(sampleSizeTable,caption='Numbers of students in the sample by state and study year',label='tab:nByState'),include.rownames=TRUE)
 @
@@
 }
+cat('Total',round(colSums(tab1[,-1])),sep='&')
+cat('\\\\')
 cat('&\\multicolumn{6}{c}{Year 2}\\\\ \n')
@@
 }
+cat('Total',round(colSums(tab2[,-1])),sep='&')
+cat('\\\\')
+
 rownames(tab1) <- rownames(tab2) <- tab1$variable

### 2017-12-05 paper.tex
couple tweaks
added a "total" line to one of the tables and fixed the formatting on
table 1

@@
 state-by-state comparisons (but included analyses that pool across states).
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Nov  6 13:55:49 2017
+% latex table generated in R 3.4.2 by xtable 1.8-2 package
+% Tue Dec 05 09:39:47 2017
 \begin{table}[ht]
 \centering
-\begin{tabular}{rrrrrrrr}
+\begin{tabular}{rrrrrrr}
   \hline
- & year & TX & KY & MI & LA & CT & NJ \\ 
+ & TX & KY & MI & LA & CT & NJ \\ 
   \hline
-1 &   1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
-  2 &   2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
+Year 1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
+  Year 2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
    \hline
@@
 
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Nov  6 13:58:09 2017
+% latex table generated in R 3.4.2 by xtable 1.8-2 package
+% Tue Dec 05 09:40:25 2017
 \begin{table}[ht]
@@
 &   \multicolumn{6}{c}{Year 1}\\%&\multicolumn{6}{c}{Year 2}\\
-&TX&KY&MI&LA&CT&Overall\\Final&699&625&242&129&91&1796\\Reassigned&600&140&220&5&92&1082\\Promoted&1619&9628&4582&1459&261&17561\\Mastered&15815&56199&31200&8648&1303&113374\\&\multicolumn{6}{c}{Year 2}\\
-Final&1252&660&236&137&114&2412\\Reassigned&1215&253&275&9&144&2042\\Promoted&1000&6296&4717&1998&450&14477\\Mastered&4929&36676&26509&10108&1513&79874\\
+&TX&KY&MI&LA&CT&Overall\\Final&699&625&242&129&91&1796\\Reassigned&600&140&220&5&92&1082\\Promoted&1619&9628&4582&1459&261&17561\\Mastered&15815&56199&31200&8648&1303&113374\\Total&18733&66592&36244&10241&1747&133813\\&\multicolumn{6}{c}{Year 2}\\
+Final&1252&660&236&137&114&2412\\Reassigned&1215&253&275&9&144&2042\\Promoted&1000&6296&4717&1998&450&14477\\Mastered&4929&36676&26509&10108&1513&79874\\Total&8396&43885&31737&12252&2221&98805\\
 
@@
     &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Nov  6 14:19:52 2017
+% latex table generated in R 3.4.2 by xtable 1.8-2 package
+% Tue Dec 05 10:03:13 2017
  TX &  18 & 582 & 281 & 932 \\ 

### 2017-12-08 paper.Rnw
Johns comments

@@
 \title{Mastery Learning in Practice:\\
-A (Mostly) Descriptive Analysis Based on the Cognitive Tutor Algebra I Effectiveness Trial}
+A (Mostly) Descriptive Analysis of Log Data from the Cognitive Tutor Algebra I Effectiveness Trial}
 
@@
 patterns of usage.
-First, we will discuss standard and customized CTAI curricula used in the study (\ref{sec:curricula}).
-Next, in Section\ref{sec:usage}, we will describe patterns in the amount CTAI usage, showing that
+First, we will discuss standard and customized CTAI curricula used in
+the study (Section \ref{sec:curricula}).
+Next, in Section \ref{sec:usage}, we will describe patterns in the amount CTAI usage, showing that
 it varied widely between states, between years, and between students.
@@
 though the effects vary widely across classrooms.
-Section \ref{sec:discussion} will conclude.
+Section \ref{sec:discussion} will conclude the article with a summary
+of findings and discussion.
 
@@
 algebra curriculum, and schools in the treatment group used Carnegie
-Learning's curriculum which includes textbook materials as well as
-CTAI. Each school participated for two years, with a different cohort
+Learning's curriculum which includes CTAI textbook materials and sofware. Each school participated for two years, with a different cohort
 of students participating the second year (with a small fraction of
@@
 of an imperfect ability to link log data to other study data files.
-For this reason, this study uses only data from on the
-\Sexpr{n_distinct(data[['schoolid2']])} schools for which at least 80\%
-students in both study years appear in the log data file.
+For this reason, this study uses only data from the
+\Sexpr{n_distinct(data[['schoolid2']])} treatment schools for which at least
+80\%
+of students in both study years appear in the log data file.
 This sample includes \Sexpr{n_distinct(data[['field_id']])} students, around
-\Sexpr{round(n_distinct(data[['field_id']])/n_distinct(stud[['field_id']]))*100}\%
-of the total high-school sample.
-Table \ref{tab:nByState} gives the number of students in each state of
-the sample in each year.
+\Sexpr{round(n_distinct(data[['field_id']])/n_distinct(stud[['field_id']])*100)}\%
+of the treated high-school sample.
+Table \ref{tab:nByState} gives the number of students in the sample by
+state and year.
 The states in the table are ordered by the total number of students
@@
 
-In this smaller sample, \Sexpr{sum(data[['obsUsage']]==FALSE)}
-students do not appear in the log data; they may have not used the CTAI
+In this sample from \Sexpr{n_distinct(data[['schoolid2']])} schools, \Sexpr{sum(data[['obsUsage']]==FALSE)}
+students who participated in the RAND study do not appear in the log data; they may have not used the CTAI
 software at all, or may have been excluded from the log data for other
@@
 @
-\caption{Percentage of worked problems coming from various levels
+\caption{Percentage of worked problems coming from various courses
   (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
@@
 there were teachers who exclusively used customized curricula the
-whole year.
+whole second year.
 
@@
 of hours worked decreased by
-\Sexpr{round(tab['Year 1','Hours']-tab['Year 2','Hours'])}
-, the median number of problems decreased by
-\Sexpr{round(tab['Year 1','Problems']-tab['Year 2','Problems'])}, and
+\Sexpr{round(tab['Year 1','Hours']-tab['Year 2','Hours'])}, the median number of problems decreased by
+\Sexpr{tab['Year 1','Problems']-tab['Year 2','Problems']}, and
 the median number of sections decreased by
-\Sexpr{round(tab['Year 1','Sections']-tab['Year 2','Sections'])} from
+\Sexpr{tab['Year 1','Sections']-tab['Year 2','Sections']} from
 years 1 to 2.
+Yet, as discussed below, the median number of units worked increased
+by \Sexpr{tab['Year 2','Units']-tab['Year 1','Units']}.
 
@@
 
-In contrast to the numbers of hours, problems, and sections students
-worked, Table \ref{tab:medUsage} shows that the median number of units
+In contrast to the decreasing numbers of hours, problems, and sections students
+worked in year 2, Table \ref{tab:medUsage} shows that the median number of units
 students worked increased by
 \Sexpr{round(tab['Year 2','Units']-tab['Year 1','Units'])}
-from year 1 to year 2.
+in year 2.
 This suggests students in year 2 were exposed, on average, to a
@@
 the proportions of worked sections in each state and study year that
-ended with mastery, promotion, or reassignment, or which were the
+ended with mastery, promotion, or reassignment, or as the
 student's final section.
@@
 mastered
-\Sexpr{round(statusCurr[['value']][statusCurr[['Curriculum']]=='Bridge-to-Algebra'
-  & statusCurr[['variable']]=='graduated']*100)}\% of the time;
+\Sexpr{round(
+  mean(statusOverall$status[statusOverall$Curriculum=='Bridge-to-Algebra']=='graduated',na.rm=T)*100)}\%
+of the time;
 those from Algebra I were mastered
@@
 However, it may suggest that some students studying advanced topics
-would fair better in more standard curricula.
+would fare better in more standard curricula.
 
@@
   variable=='Reassigned')[['value']])}\% vs.
-\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Reassigned')[['value']])}\%).
-This may an overall skepticism towards the Carnegie Learning standards
+\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' &
+  variable=='Reassigned' & Yr=='Yr 2')[['value']],1)}\%, in year 2).
+This may indicate an overall skepticism towards the Carnegie Learning standards
 among certain schools and teachers, manifested in both adoption of
@@
 state's reassignments the occurred in each month of the school year.
-(Note that while both Figure \ref{fig:byMonth} and \ref{fig:byMonth}
+(Note that while both Figure \ref{fig:byMonth} and \ref{fig:byMonthState}
 show proportions for each month, the denominators are not the same:
@@
 reassignments, at least 75\% of the rest of the class had exited the
-section on the same date or earlier.
+section through graduation or promotion on the same date or earlier.
 The figure also reveals that the proportions of students who had graduated or been
@@
   sum(trans1o['4-Quadrant Linear Graphs',])*100)}\%
-of the reassignments begun in ``4-Quadrant Linear Graphs'' end in the
+of the reassignments from ``4-Quadrant Linear Graphs'' end in the
 same unit.
@@
 year 2.
-On the other hand, it is rare for students to be reassigned within the same unit
+On the other hand, it is relatively rare for students to be reassigned within the same unit
 (these reassignments account for \Sexpr{round(sum(diag(trans1o))/sum(trans1o)*100)}\% and \Sexpr{round(sum(diag(trans2o))/sum(trans2o)*100)}\% in the two years).
@@
 Specifically, student reassignment was not randomized.
-Therefore, claims regarding the causal effects of reassignment on
+Therefore, precise estimates of causal effects of reassignment on
 learning require strong untestable assumptions that are unlikely to be
@@
 All that said, observational estimates of reassignment's average
-effects can be valuable, if interpreted cautiously.
+causal effects can be valuable, if interpreted cautiously, for
+instance by assessing their sensitivity to unmeasured confounding, as
+we do below.
 In the absence of evidence from randomized trials, observational
@@
 
+stopifnot(names(Rhos)[which.max(abs(Ts[names(Rhos)]))]=="raceHispanic")
+stopifnot(names(Rhos)[which.max(abs(Rhos))]=="spec_esl")
+
 @
@@
 %\subsection{Average Effects}
-Figure \ref{fig:cpYyear} shows students gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
-(The number of reassignments was jittered---random noise was added---to avoid overplotting.)
+Figure \ref{fig:cpYyear} shows students' gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
+(The number of reassignments was jittered---random noise was added on
+the horizontal dimension---to avoid overplotting.)
 Overall, the relationship between the two variables is positive.
@@
 pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
-\Sexpr{round(mean(is.na(cpDat[['xirt']]))*100)}\%).
-To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2014effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligibility.
+\Sexpr{round(mean(is.na(cpDat[['xirt']]))*100)}\%). To account for
+this, we used regression calibration based on the 20 ``multiple
+imputations'' used in the original CTAI study,
+\citet{pane2014effectiveness}.} race, sex, grade, special education
+and gifted status, English as a second language (ESL), and free and reduced-price lunch eligibility.
 
@@
 In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
-Roughly speaking, the idea is to widen the confidence interval from a causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
+Roughly speaking, the idea is to widen the confidence interval from an
+ostensibly causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
 These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.
@@
 omission of a less important hypothetical confounder---one that
-predicts posttests and reassignment as well as the next best observed
-covariates---is
+predicts posttests as well as ESL status and reassignment as well as
+Hispanic ethnicity, the next best observed
+predictors---is
 \Sexpr{round(adjustedEsts1[['mod3.1']][1,'ests'],2)} $\pm$
@@
 The wide variety in the use of reassignment that we have documented
-here might suggest that reassignment's treatment effect might vary as
+here might suggest that reassignment's treatment effect varies as
 well.
@@
 This paper exploited available log data from the high-school arm of the CTAI
-effectiveness trial---in the second year, an impressive result for the
-software---to describe variation and patterns in the software's
+effectiveness trial---which yielded an impressive result for the
+software in year 2---to describe variation and patterns in the software's
 usage, paying particular attention to issues of mastery learning.
@@
    curricula.
- \item Section mastery:
+ \item Examination of section mastery found that:
  \begin{itemize}
@@
  \end{itemize}
- \item Reassignment Patterns
+ \item Examination of reassignment patterns found that:
   \begin{itemize}
@@
 quite differently from the vision of CT's designers.
-In some cases, the departures were a matter of a student's inability
-to achieve mastery quickly enough, and in other cases departures were
-driven by educators' preferences that ran counter to CTAI's design.
+In some cases, the departures were apparently a matter of a student's inability
+to achieve mastery quickly enough, and in other cases of educators' preferences that ran counter to CTAI's design.
 

### 2017-12-08 paper.tex
Johns comments

@@
 of an imperfect ability to link log data to other study data files.
-For this reason, this study uses only data from on the
-18 schools for which at least 80\%
-students in both study years appear in the log data file.
+For this reason, this study uses only data from the
+18 treatment schools for which at least
+80\%
+of students in both study years appear in the log data file.
 This sample includes 4460 students, around
-100\%
-of the total high-school sample.
-Table \ref{tab:nByState} gives the number of students in each state of
-the sample in each year.
+75\%
+of the treated high-school sample.
+Table \ref{tab:nByState} gives the number of students in the sample by
+state and year.
 The states in the table are ordered by the total number of students
@@
 
-In this smaller sample, 164
-students do not appear in the log data; they may have not used the CTAI
+In this sample from 18 schools, 164
+students who participated in the RAND study do not appear in the log data; they may have not used the CTAI
 software at all, or may have been excluded from the log data for other
@@
 
-\caption{Percentage of worked problems coming from various levels
+\caption{Percentage of worked problems coming from various courses
   (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
@@
 of hours worked decreased by
-10
-, the median number of problems decreased by
+10, the median number of problems decreased by
 81, and
 the median number of sections decreased by
-6 from
+6.5 from
 years 1 to 2.
+Yet, as discussed below, the median number of units worked increased
+by 1.
 
@@
 
-In contrast to the numbers of hours, problems, and sections students
-worked, Table \ref{tab:medUsage} shows that the median number of units
+In contrast to the decreasing numbers of hours, problems, and sections students
+worked in year 2, Table \ref{tab:medUsage} shows that the median number of units
 students worked increased by
 1
-from year 1 to year 2.
+in year 2.
 This suggests students in year 2 were exposed, on average, to a
@@
 mastered
-\% of the time;
+92\%
+of the time;
 those from Algebra I were mastered
@@
 curriculum (3\% vs.
-1, 1\%).
-This may an overall skepticism towards the Carnegie Learning standards
+1.4\%, in year 2).
+This may indicate an overall skepticism towards the Carnegie Learning standards
 among certain schools and teachers, manifested in both adoption of
@@
 44\%
-of the reassignments begun in ``4-Quadrant Linear Graphs'' end in the
+of the reassignments from ``4-Quadrant Linear Graphs'' end in the
 same unit.
@@
 year 2.
-On the other hand, it is rare for students to be reassigned within the same unit
+On the other hand, it is relatively rare for students to be reassigned within the same unit
 (these reassignments account for 15\% and 18\% in the two years).
@@
 pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
-15\%).
-To account for this, we used regression calibration based on the 20 ``multiple imputations'' used in the original CTAI study, \citet{pane2014effectiveness}.} race, sex, grade, special education and gifted status, English as a second language, and free and reduced-price lunch eligibility.
+15\%). To account for
+this, we used regression calibration based on the 20 ``multiple
+imputations'' used in the original CTAI study,
+\citet{pane2014effectiveness}.} race, sex, grade, special education
+and gifted status, English as a second language (ESL), and free and reduced-price lunch eligibility.
 
@@
 omission of a less important hypothetical confounder---one that
-predicts posttests and reassignment as well as the next best observed
-covariates---is
+predicts posttests as well as ESL status and reassignment as well as
+Hispanic ethnicity, the next best observed
+predictors---is
 -0.15 $\pm$

### 2017-12-15 paper.Rnw
including anita's comments

@@
 library(knitr)
+library(lubridate)
+library(scales)
+library(reshape2)
+library(dplyr)
+library(ggplot2)
+library(xtable)
+
 #library(tikzDevice)
@@
 data from the RAND study, a seven-state randomized controlled trial of
-CTAI in high schools and middle schools. The study found a significant
+CTAI in high schools and middle schools. That study found a significant
 positive effect of CTAI in high schools, during their second year of
@@
 Next, in Section \ref{sec:usage}, we will describe patterns in the amount CTAI usage, showing that
-it varied widely between states, between years, and between students.
+it varied widely between states, between years, between schools, and between students.
 In particular, the amount of usage decreased from years 1 to
@@
 Algebra I sequence.
-A small but notable number of more advanced problems were worked in
-Kentucky in Year 2, and some less advanced problems were worked in
+A small but notable number of less advanced problems were worked in
+Kentucky in year 2, and some more advanced problems were worked in
 Michigan and Louisiana.
@@
 enrolled in three different curricula throughout the entire year
-(Bridge-to-Algebra, Algebra I, and a specialized Geometry curriculum),
+(Bridge-to-Algebra, Algebra I, and a customized Geometry curriculum),
 while a year 2 teacher in Michigan enrolled students in
 three curricula sequentially throughout the year: Algebra I
-until November, followed by a customized curriculum util February, and
+until November, followed by a customized curriculum until February, and
 ending with a different customized curriculum until June. While there are numerous instances of these
-uses of multiple curricula in Year 2, there are also many occurrences
+uses of multiple curricula in year 2, there are also many occurrences
 of teachers who had their students enrolled in the standard Algebra I
-throughout the entire year.  In TX mostly, with some instances in Michigan,
-there were teachers who exclusively used customized curricula the
-whole second year.
+throughout the entire year, including all Connecticut teachers.
+There were also teachers, mostly in Texas, who used customized
+curricula exclusively throughout the
+second year.
 
@@
 Analogous figures for the numbers of problems and sections students
-worked, available in an online supplement, showed similar patterns for
-the numbers of sections and problems students worked.
+worked, %available in an online supplement,
+showed similar patterns.% for
+%the numbers of sections and problems students worked.
 Usage time varied substantially between students and across states and
@@
 between-state variation, and a between-student increase in Louisiana.
-One intriguing possibility is that CT usage may have been better tailored to
-teachers and students in the second year than in the first---perhaps
-students who stood to gain more from the software increased their
-usage, and the rest of the students used it less.
+One intriguing possibility is that the amount of CT usage may have been better tailored to
+teachers and students in the second year than in the first. Perhaps
+usage increased for
+students who stood to gain more from the software and decreased for
+students who stood to gain less.
 
@@
 Louisiana but less in Texas, Connecticut, and New Jersey.
-However, in every state the median Year 2 student worked at least as
-many different units as the median Year 1 student.
+However, in every state the median year 2 student worked at least as
+many different units as the median year 1 student.
 Variation in the number of units worked also increased slightly
@@
 Thus we investigated whether, on average, there were more graduated sections per students the second year, and how this related to whether they were enrolled in the standard versus customized curricula.
-Overall, the number of graduated sections per student increased in Year 2.
-The number of graduated sections per student was 17.94 in Year 1 and 25.66 in Year 2 in standard curricula.
-However, the customized curricula had fewer average graduated sections per student enrolled compared to the standard curricula; these numbers were 5.64 in Year 1 and 11.47 in Year 2.
+Overall, the number of graduated sections per student increased in year 2.
+The number of graduated sections per student was 17.94 in year 1 and 25.66 in year 2 in standard curricula.
+However, the customized curricula had fewer average graduated sections per student enrolled compared to the standard curricula; these numbers were 5.64 in year 1 and 11.47 in year 2.
 The percentages of average graduated sections to the average total worked sections were: 33.33% in Year 1 and 48.92% in Year 2 in the standard curricula, and 57.55% in Year 2 for the customized curricula.
@@
 Some records were omitted from this analysis: sections missing information on date of completion, as well as each student's last section attempted because there is no completion date recorded for those sections.
-As presented in Figures 2 and 3 for Year 1 and 2, respectively, students in NJ increased the time spent per section from Year 1 to Year 2; however, there were relatively few students in the study from NJ.
+As presented in Figures 2 and 3 for year 1 and 2, respectively, students in NJ increased the time spent per section from year 1 to year 2; however, there were relatively few students in the study from NJ.
 From Figure 1, it was shown that CT and TX completed, on average, fewer sections per student.
 From the information presented in Figures 2 and 3, these states spent the highest average number of days per section.
-TX, with the largest enrollment of students and the greatest use of customized curricula in Year 2, had students spending slightly more days completing sections in customized curricula relative to the standard curricula.
+TX, with the largest enrollment of students and the greatest use of customized curricula in year 2, had students spending slightly more days completing sections in customized curricula relative to the standard curricula.
 
@@
 Results are similar to the data presented in Figures 2 and 3.
-However, in Year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time Texas students spent mastering sections in the standard curricula.
+However, in year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time Texas students spent mastering sections in the standard curricula.
 This difference is also seen in Kentucky and Louisiana.
@@
 expect if students adhered to the curriculum.
-Students varied in how many units they worked---with the variation due
+Students varied in the number of units they worked---with the variation due
 to both student ability and the amount of time allocated to CTAI
@@
 Most strikingly, ``Linear Equations with Variables on Both Sides'' was worked by a
-greater proportion of students in Year 2 than in Year 1, and by a
+greater proportion of students in year 2 than in year 1, and by a
 greater proportion of students than any of the previous six sections.
@@
 to about \Sexpr{round(tab2b['Reassigned','TX']*100)}\%.
-The proportion of Texas sections labeled ``Final'' increased as well,
-most likely due to the decrease in usage, and thus the total number of
-sections.
+The proportion of Texas sections labeled ``Final'' increased as
+well---the expected result of decreasing the overall number of worked
+sections and holding fixed the likelihood of ending
+usage while in the middle of a section.
 
@@
 After school, state was the most important, accounting for
-\Sexpr{vcs[['Overall_1']]['state',1]} and
+\Sexpr{vcs[['Overall_1']]['state',1]}\% and
 \Sexpr{vcs[['Overall_2']]['state',1]}\% in the two years, and unit,
-accounting for \Sexpr{vcs[['Overall_1']]['unit',1]} and \Sexpr{vcs[['Overall_2']]['unit',1]}\%.
+accounting for \Sexpr{vcs[['Overall_1']]['unit',1]}\% and \Sexpr{vcs[['Overall_2']]['unit',1]}\%.
 Surprisingly, classroom and student-level factors only accounted for
@@
     with the student having mastered all of the included skills.
-  \item There were three ways students worked a section with
+  \item There were three ways students worked a section without
     mastering its contents: exhausting its problems and being

### 2017-12-15 paper.tex
including anita's comments

@@
 state-by-state comparisons (but included analyses that pool across states).
-% latex table generated in R 3.4.2 by xtable 1.8-2 package
-% Tue Dec 05 09:39:47 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Fri Dec 15 13:54:41 2017
 \begin{table}[ht]
@@
 
-% latex table generated in R 3.4.2 by xtable 1.8-2 package
-% Tue Dec 05 09:40:25 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Nov  6 13:58:09 2017
 \begin{table}[ht]
@@
 to about 14\%.
-The proportion of Texas sections labeled ``Final'' increased as well,
-most likely due to the decrease in usage, and thus the total number of
-sections.
+The proportion of Texas sections labeled ``Final'' increased as
+well---the expected result of decreasing the overall number of worked
+sections and holding fixed the likelihood of ending
+usage while in the middle of a section.
 
@@
 After school, state was the most important, accounting for
-20 and
+20\% and
 21\% in the two years, and unit,
-accounting for 16 and 14\%.
+accounting for 16\% and 14\%.
 Surprisingly, classroom and student-level factors only accounted for
@@
     &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
-% latex table generated in R 3.4.2 by xtable 1.8-2 package
-% Tue Dec 05 10:03:13 2017
+% latex table generated in R 3.3.1 by xtable 1.8-2 package
+% Mon Nov  6 14:19:52 2017
  TX &  18 & 582 & 281 & 932 \\ 

### 2017-12-15 paper.Rnw
took out stuff commented out

@@
 middle school arm. The current study focuses on high school students
-only. %\marginpar{I think it reads better to use the spelled-out names,
-%  but I'm happy to switch if you feel strongly about this.}%, due to the large number of students in this cohort and the
-% significant positive treatment effect previously found for this
-% cohort. All discussion that follows refers exclusively to the high
-% school data.
+only.
 
@@
 
-## ggplot(bbb,aes(Yr,n,color=bbb))+geom_col(show.legend=FALSE)+geom_col(aes(fill=Curriculum))+geom_col(aes(fill=overall),alpha=0.5)+facet_grid(~state)
-##     scale_fill_manual(values=c('#8dd3c7','#8dd3c7','#ffffb3','#ffffb3','#bebada','white','black'))
-##     geom_col(aes(Yr,n,fill=overall),alpha=0.5)+
-##     #geom_col(aes(Yr,n,fill=overall))+#scale_fill_manual(values=c('white','black'))+
-##     facet_grid(~state)
-
-## omar <- par()$oma
-
-## par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
-## par(mfrow=c(1,9))
-## for(st in levels(data$State)){
-## #    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
-##     cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
-##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-##     for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-##     cwy <- cwy[nrow(cwy):1,]
-##     barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main=st,yaxt='n')
-##     if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
-##     #box()
-## }
-## cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
-##     summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
-## for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
-## cwy <- cwy[nrow(cwy):1,]
-## barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',50,NA),main='Overall',yaxt='n')
-
-## plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
-## legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
-##        fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
-##        angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,50),xpd = TRUE, bty = "n",inset=c(0,0))
-
-## par(oma=omar)
 @
@@
 from years 1 to 2.
-
-% All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year.
-% Figure
-% In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools.
-The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus.
-In Year 2, Carnegie Learning also developed many more customized curricula with input from participating schools, for a total of 33 curricula used in 13 schools.
-This led to fewer students enrolled in the standard curriculum and more enrolled in customized curricula overall in Year 2.
-
-% In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204.
-The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2.
-It should be noted that students could be enrolled in multiple curricula.
-For instance, teachers could choose to enroll students in any of the standard curricula provided, such as Algebra I or Algebra II, along with any of the customized curricula that were used by the school.
-
-% The use of customized curricula varied substantially by state.
-For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2.
-All other states contained schools that used customized curricula in addition to the standard curricula.
-CT schools maintained similar number of students in Algebra I across both years with 111 in Year 1 and 135 in Year 2.
-NJ substantially decreased the number of students in the CTAI program from 89 to 32 from Year 1 to Year 2.
-However, while all other states (KY, LA, MI, and TX) did decrease enrollment in the standard Algebra I curriculum the second year, those same states increased enrollment in the customized curricula.
-KY added 3 customized curricula with 146 students, LA added 15 customized curricula with 286 students, MI added 5 customized curricula with 370 students, and TX added 10 additional customized curricula and had the largest enrollment in these curricula with 1,531 students.
-Students enrolled in the standard Algebra I curriculum dropped from 974 in Year 1 to 61 in Year 2 in TX.
-As noted above, students could be enrolled in more than one curriculum.
-In Year 2, the average number of curricula per a student ranged from 1.00 (NJ and CT) to 2.53 (TX).
-The next highest value was LA at 2.10.
 
@@
 
-% \begin{figure}
-% <<mastSec,dependson=c('usageTime','data','setup'),fig.width=6,fig.height=3>>=
-% print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
-% @
-% \caption{Boxplots of the number of sections each of Cognitive Tutor
-%   software each student mastered, by year and state. Students
-%   mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
-%   with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
-%   were excluded.}
-% \label{fig:mastSec}
-% \end{figure}
-
 Figure \ref{fig:timeByStud} shows that the number of hours students
@@
 Analogous figures for the numbers of problems and sections students
-worked, %available in an online supplement,
-showed similar patterns.% for
-%the numbers of sections and problems students worked.
+worked,
+showed similar patterns.
 Usage time varied substantially between students and across states and
@@
 slightly wider range of topics, and varied somewhat more in their usage.
-
-\begin{comment}
-Of course, more worked sections does not necessarily mean better worked sections.
-For instance, students might attempt fewer sections in year two, but graduate a higher percentage.
-Thus we investigated whether, on average, there were more graduated sections per students the second year, and how this related to whether they were enrolled in the standard versus customized curricula.
-Overall, the number of graduated sections per student increased in year 2.
-The number of graduated sections per student was 17.94 in year 1 and 25.66 in year 2 in standard curricula.
-However, the customized curricula had fewer average graduated sections per student enrolled compared to the standard curricula; these numbers were 5.64 in year 1 and 11.47 in year 2.
-The percentages of average graduated sections to the average total worked sections were: 33.33% in Year 1 and 48.92% in Year 2 in the standard curricula, and 57.55% in Year 2 for the customized curricula.
-The average number of promoted sections per student also increased from Year 1 to Year 2, though only slightly.
-In Year 1, the average number of promotions were 1.20 for the customized curricula versus 3.64 for the standard curricula; in Year 2, those numbers were 2.20 and 4.95, respectively.
-
-We also evaluated the time spent per section across different curricula and states.
-Some records were omitted from this analysis: sections missing information on date of completion, as well as each student's last section attempted because there is no completion date recorded for those sections.
-As presented in Figures 2 and 3 for year 1 and 2, respectively, students in NJ increased the time spent per section from year 1 to year 2; however, there were relatively few students in the study from NJ.
-From Figure 1, it was shown that CT and TX completed, on average, fewer sections per student.
-From the information presented in Figures 2 and 3, these states spent the highest average number of days per section.
-TX, with the largest enrollment of students and the greatest use of customized curricula in year 2, had students spending slightly more days completing sections in customized curricula relative to the standard curricula.
-
-Figures 4 and 5 present information for only sections that were mastered by students, that is, cases in which a student was judged by the software to have mastered the material in the section.
-Results are similar to the data presented in Figures 2 and 3.
-However, in year 2, TX students in customized curricula spent fewer days on a section they mastered, compared to the time Texas students spent mastering sections in the standard curricula.
-This difference is also seen in Kentucky and Louisiana.
-If customized curriculum are more aligned with teachers' instruction, graduating from a section may require fewer days than sections in the standard curricula, possibly because students are working on the same topics in class as they are in the software.
-Figures 4 and 5 also show that, on average, Connecticut and Texas spent more days graduating from sections, regardless whether the section was part of a standard curricula or customized curricula.
-One explanation for the longer time periods in these states may be due to a teacher's use of the software simultaneously with the textbook instruction.
-In this case, the teachers could have switched back-and-forth as they instructed, hence extending the time worked by their students.
-For the states with shorter time periods, teachers may have followed a more sequential process where they first completed one component, e.g.
-textbook material, and then allowed students to focus on and complete the second component, e.g.
-software.
-\end{comment}
 
@@
 alternative curricula and reassignment.%\marginpar{does that make more sense?}
-
-% A somewhat lower proportion of Algebra I sections from customized
-% curricula were mastered
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])}\%)
-% than from the standard Algebra I curriculum
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])}\%).
-% This difference must be interpreted with caution, since a greater
-% proportion of sections of the customized curriculum ended in
-% reassignment
-
-% or were final
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])}\% vs.
-% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])}\%)
-% , as well.
-% The proportion of sections ending in promotion was almost identical:
-% (\Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])}\% vs.
-% \Sexpr{round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])}\%).
 
@@
 \end{figure}
-
-% \section{Notes}
-% [[don't forget to include these]]
-
-% \begin{itemize}
-%   \item \# of students per state?
-%   \item NJ excluded from state-by-state comparisons, cuz too
-%     small. Included in ``overall,'' etc.
-%   \item What schools included?
-%   \item Students with no mastery data: maybe they didn't work at all,
-%     maybe they worked but we don't have their data. Treating them as unknown.
-% \end{itemize}
 

### 2017-12-15 paper.tex
took out stuff commented out

@@
 
-% All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year.
-% Figure
-% In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools.
-The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus.
-In Year 2, Carnegie Learning also developed many more customized curricula with input from participating schools, for a total of 33 curricula used in 13 schools.
-This led to fewer students enrolled in the standard curriculum and more enrolled in customized curricula overall in Year 2.
-
-% In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204.
-The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2.
-It should be noted that students could be enrolled in multiple curricula.
-For instance, teachers could choose to enroll students in any of the standard curricula provided, such as Algebra I or Algebra II, along with any of the customized curricula that were used by the school.
-
-% The use of customized curricula varied substantially by state.
-For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2.
-All other states contained schools that used customized curricula in addition to the standard curricula.
-CT schools maintained similar number of students in Algebra I across both years with 111 in Year 1 and 135 in Year 2.
-NJ substantially decreased the number of students in the CTAI program from 89 to 32 from Year 1 to Year 2.
-However, while all other states (KY, LA, MI, and TX) did decrease enrollment in the standard Algebra I curriculum the second year, those same states increased enrollment in the customized curricula.
-KY added 3 customized curricula with 146 students, LA added 15 customized curricula with 286 students, MI added 5 customized curricula with 370 students, and TX added 10 additional customized curricula and had the largest enrollment in these curricula with 1,531 students.
-Students enrolled in the standard Algebra I curriculum dropped from 974 in Year 1 to 61 in Year 2 in TX.
-As noted above, students could be enrolled in more than one curriculum.
-In Year 2, the average number of curricula per a student ranged from 1.00 (NJ and CT) to 2.53 (TX).
-The next highest value was LA at 2.10.
-
 Throughout the school year, teachers could have a class of students
@@
 
-% \begin{figure}
-% <<mastSec,dependson=c('usageTime','data','setup'),fig.width=6,fig.height=3>>=
-% print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
-% @
-% \caption{Boxplots of the number of sections each of Cognitive Tutor
-%   software each student mastered, by year and state. Students
-%   mastering more than 250 sections (sum(secByStud[['mastered']]>250,na.rm=TRUE)/2) of nrow(secByStud)/2) and students
-%   with no mastery data (sum(is.na(secByStud[['mastered']]))/2)
-%   were excluded.}
-% \label{fig:mastSec}
-% \end{figure}
-
 Figure \ref{fig:timeByStud} shows that the number of hours students
@@
 alternative curricula and reassignment.%\marginpar{does that make more sense?}
-
-% A somewhat lower proportion of Algebra I sections from customized
-% curricula were mastered
-% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' & variable=='Graduated')[['value']])\%)
-% than from the standard Algebra I curriculum
-% (round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Graduated')[['value']])\%).
-% This difference must be interpreted with caution, since a greater
-% proportion of sections of the customized curriculum ended in
-% reassignment
-
-% or were final
-% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &   variable=='Final')[['value']])\% vs.
-% round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Final')[['value']])\%)
-% , as well.
-% The proportion of sections ending in promotion was almost identical:
-% (round(100*subset(statusCurr,Curr2=='Algebra I (Cust.)' &variable=='Promoted')[['value']])\% vs.
-% round(100*subset(statusCurr,Curr2=='Algebra I' & variable=='Promoted')[['value']])\%).
 

### 2018-01-02 jedm-latex-2017/instructions.tex
jedm format

\documentclass[notitlepage,12pt]{jedm}
%\usepackage[sc,sf,small]{titlesec}
\usepackage[table]{xcolor}
\usepackage{url}

% FINAL COPYEDITTED SUBMISSION - UNCOMMENT THIS TO SUPPRESS PAGE NUMBERS
%\pagenumbering{gobble}

\begin{document}

\title{Submission Instructions}
\date{} %do not delete this, it suppresses insertion of the date

\author{{\large Author One}\\Institution of Author one\\email.of@author.dom \and {\large Second Author}\\Affiliation\\email@xx.xx  \and {\large Third Author}\\Affiliation\\email@xx.xx }

\maketitle

\begin{abstract}
This document serves both as JEDM submission instruction and as a template file.  This is the abstract. It should contain from 100 to 250 words.
\end{abstract}

\section{Paper format}

Manuscripts should be formatted for letter sized paper, one side only, leaving 2.75cm margins on the right and left sides and 2.25cm on the top and bottom. Body font should be Times Roman 12pt and sections, title, and authors font should be Helvetica.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Style files}

For LaTeX, a style file named \texttt{jedm.cls} is provided on the journal's web site.  An MS Word file containing this example text is also provided, \texttt{jedm.doc}, which can be used as a template.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Figures and tables}

Each figure and table must be mentioned in the text, and must be numbered consecutively in order of appearance (with captions in lower case). For the review process, the figures should be integrated into the text rather than being inserted at the end of the document.

As JEDM is an electronic journal, authors are encouraged to use colours when it enhances visibility and understanding of figures.

Examples of a figure and a table are given in figure~\ref{tab:fig1} and table~\ref{tab:1}.

\begin{figure}
  \centerline{\fbox{\parbox{.66\textwidth}{\centering \raisebox{0pt}[2cm][2cm]{%
        \begin{tabular}{c}
          \cellcolor{blue!35}{\sf This is a figure}\\
          \cellcolor{blue!25}{\sf And yes, you can use colors!}\\
          \cellcolor{blue!15}{\sf As long as it enhances visibility and understanding}\\
        \end{tabular}
  }}}}
  \caption{This is the figure's caption.  It should be a centered paragraph of width 193mm (7.6in). Font size should be 11pt.}
  \label{tab:fig1}
\end{figure}

\begin{table}
  \caption{This is an example of a table that lists the margins of this template.  Captions should follow the same rules as a figure, except that they are put on top of the table.}\vspace*{1ex}
  \label{tab:1}
  \centering
  \begin{tabular}{| l | l |}
    \hline
    \multicolumn{1}{|c|}{\textbf{Margin}} & \multicolumn{1}{c|}{\textbf{Size}} \\
    \hline
    left&2.75cm\\
    right&2.75cm\\
    top&2.25cm\\
    bottom&2.25cm\\
    \hline
  \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Appendices}

Supplementary technical material (e.g. mathematical proofs or descriptions of experimental procedures) should be collected in an appendix at the end of the paper (before the acknowledgements and the references sections).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Footnotes and acknowledgements}

Footnotes should be used sparingly and indicated by consecutive superscript numbers in the text. Material to be footnoted should appear at the bottom of the page on which it is referenced. Acknowledgements and grant numbers should be put into a separate `Acknowledgement' section right before the list of references.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{References}

References should follow the ACM standard.  The example provided here uses the \texttt{jedm.cls} class file and \texttt{acmtrans.bst} bib style file.  For example, we could write that \citeN{JEDM:baker2009} published a review on EDM in this journal; other reviews were published later \cite[for eg.]{romero2010educational}. The provided ref.bib file contains examples of virtually every possible citation type.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Supporting materials}

Authors are encouraged to submit the data they use and the analysis code in order to replicate and perform rigorous comparisons across studies.  The data and code can be stored on the \texttt{educationaldatamining.org} site for public reference, or stored privately for reviewing purpose only if required. See the online submission instructions for guidance on using and citing code repositories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Page numbering and sectioning}

For the manuscript to be reviewed, page number should appear at the bottom of each page.  \textbf{For the final version, they must be taken out as the standard JEDM footer is be added.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sections and subsections}

Section style should follow the example in this document.

%%%%
\subsubsection{Subsection levels}

There should be no more than three levels of sections.

\paragraph{Fourth level.}

The fourth level should simply have their title in the same font style as those of of subsections, without numbering and at the beginning of a paragaph.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Submission}

The journal prefers PDF format, but can also accommodate most other popular formats.  Care should be taken to embed fonts in the rendered PDF.

You can submit your manuscript by email to \texttt{jedm.editor@gmail.com}, or directly on the JEDM review management website.

Instructions for submitting the papers on the website are at \url{http://jedm.educationaldatamining.org/index.php/JEDM/about/submissions#onlineSubmissions}.  You first need to register and log in to the site.  When registering, you must activate your role as ``author''.

% REMOVE NOCITE OR IT WILL LIST EVERYTHING IN YOUR DATABASE AS A REFERENCE
\nocite{*}

\bibliographystyle{acmtrans}
\bibliography{./ref}

\end{document}

### 2018-01-02 paper.Rnw
jedm format

@@
-\documentclass[12pt]{article}
+ \documentclass[notitlepage,12pt]{jedm}
 
@@
 \usepackage{verbatim}
-\usepackage{natbib}
+%\usepackage{natbib}
 \usepackage{bm}
@@
 
-\author{Anita Israni \and Adam C Sales \and John F Pane}
+\author{{\large Anita Israni}\\University of Texas College of
+  Education\\aisrani@gmail.com \and {\large Adam C Sales}\\University of Texas College of
+  Education\\asales@utexas.edu  \and {\large John F Pane}\\RAND Corporation\\jpane@rand.org}
+
+\date{}
 
@@
 
-<<data,include=FALSE>>=
+<<data,include=FALSE,cache=TRUE>>=
 if('cpPaper.RData'%in%list.files() & file.info('cpPaper.RData')$mtime>file.info('dataMerge.r')$mtime){
@@
 
+\begin{abstract}
+Mastery learning, the notion that students learn best if they move on
+from studying a topic only after having demonstrated mastery, sits at
+the foundation of the theory of intelligent tutoring. This paper is an
+exploration of how mastery learning plays out in practice, based on
+log data from a large randomized effectiveness trial of the Cognitive
+Tutor Algebra I (CTAI) curriculum. We find that students frequently
+progressed from CTAI sections they were working on without
+demonstrating mastery and worked units out of order. Moreover, these
+behaviors were substantially more common in the second year of the
+study, in which the CTAI effect was significantly larger. We explore
+the various ways students departed from the official CTAI curriculum,
+focusing on heterogeneity between years, states, schools, and
+students. The paper concludes with an observational study of the
+effect on post-test scores of teachers reassigning students out of
+their current sections before they mastered the requisite skills,
+finding that reassignment appears to lowers posttest scores---a finding
+that is fairly reseliant to confouding from omitted
+covariates---but that the effect varies substantially between
+classrooms.
+\end{abstract}
 \section{Introduction}
 Mastery learning sits at the foundation of intelligent tutoring
-systems \citep[e.g.][]{corbett2001cognitive,wenger2014artificial}.
+systems \cite[e.g.][]{corbett2001cognitive,wenger2014artificial}.
 The philosophy of mastery learning assumes a well-structured
@@
 they master its skills
-\citep{bloom1968learning,kulik1990effectiveness}. The Cognitive Tutor
+\cite{bloom1968learning,kulik1990effectiveness}. The Cognitive Tutor
 Algebra I (CTAI) system, developed by Carnegie Learning, Inc., is one of the best studied and best regarded examples of modern educational software.
@@
 The software component of the curriculum allows students to progress at their own pace and receive individualized feedback on their performance.
-A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students' scores on an Algebra-I posttest by about one fifth of a standard deviation \citep{pane2014effectiveness}.
+A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students' scores on an Algebra-I posttest by about one fifth of a standard deviation \cite{pane2014effectiveness}.
 CTAI's success in this experiment would seem to validate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
@@
 software data from the study reflect typical usage.
-Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects \citep{sales2015exploring,sales2016student}.
-\citet{sales2017role} found that students more likely to master worked
+Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects \cite{sales2015exploring,sales2016student}.
+\citeN{sales2017role} found that students more likely to master worked
 sections of the CTAI software may experience \emph{smaller} effects
@@
 imputations'' used in the original CTAI study,
-\citet{pane2014effectiveness}.} race, sex, grade, special education
+\citeN{pane2014effectiveness}.} race, sex, grade, special education
 and gifted status, English as a second language (ESL), and free and reduced-price lunch eligibility.
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(abs(adjustedEsts1[['mod3.1']][1,'ests'])/0.2*100)}\% of that.
+The magnitudes of the effects are rather large: \citeN{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is \Sexpr{round(abs(adjustedEsts1[['mod3.1']][1,'ests'])/0.2*100)}\% of that.
 
@@
 in ability, beyond what is captured in pretest scores.
-\citet{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
+\citeN{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
 In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
@@
 Unlike fixed effects models, multilevel models ``partially pool'' data
-across classrooms to estimate classroom specific effects more precisely \citep{gelmanHill}.
+across classrooms to estimate classroom specific effects more precisely \cite{gelmanHill}.
 This is especially important given the small sample sizes within
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{acmtrans}
 \bibliography{ct,cp}

### 2018-01-02 paper.tex
jedm format

@@
-\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+ \documentclass[notitlepage,12pt]{jedm}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citet{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is 73\% of that.
+The magnitudes of the effects are rather large: \citeN{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is 73\% of that.
 

### 2018-01-02 intro.tex
finish jedm format

--- 2018-01-02 paper.tex
@@
- \documentclass[notitlepage,12pt]{jedm}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-
-%\usepackage{endfloat}
-%\usepackage{soul}
-
-%\usepackage{type1ec}
-\usepackage{subcaption}
-\usepackage{fullpage}
-\usepackage{enumerate}
-%\usepackage{graphicx}
-%\usepackage{graphics}
-\usepackage{multirow}
-\usepackage{comment}
-\usepackage{amsmath,amssymb,amsfonts,amsthm}
-\usepackage{bbm}
-\usepackage{setspace}
-\usepackage{verbatim}
-%\usepackage{natbib}
-\usepackage{bm}
-\usepackage{pdflscape}
-%\usepackage[dvipsnames]{xcolor}
-\usepackage{tikz}
-\usepackage{xr}
-%\usepackage{mathabx}
-%\usepackage{filecontents}
-%\usepackage{bibentry}
-%\usepackage{hanging}
-%\usepackage{apacite}
-%\usepackage{hyperref}
-\usepackage{makecell}
-
-%\usepackage{endnotes}
-
-%\let\footnote=\endnote
-
-\title{Mastery Learning in Practice:\\
-A (Mostly) Descriptive Analysis of Log Data from the Cognitive Tutor Algebra I Effectiveness Trial}
-
-\author{{\large Anita Israni}\\University of Texas College of
-  Education\\aisrani@gmail.com \and {\large Adam C Sales}\\University of Texas College of
-  Education\\asales@utexas.edu  \and {\large John F Pane}\\RAND Corporation\\jpane@rand.org}
-
-\date{}
-
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\maketitle
-
 \begin{abstract}
@@
 Mastery learning sits at the foundation of intelligent tutoring
-systems \cite[e.g.][]{corbett2001cognitive,wenger2014artificial}.
+systems \cite{corbett2001cognitive,wenger2014artificial}.
 The philosophy of mastery learning assumes a well-structured
@@
 of findings and discussion.
-
-\section{The RAND Effectiveness Trial}\label{sec:RANDtrial}
-The study to measure the effectiveness of CTAI included 7 states, 73
-high schools, and 74 middle schools with nearly 18,700 high school
-students and 6,800 middle school students participating. Schools were enrolled in a
-total of 52 school districts that were distributed among urban,
-suburban, and rural areas. Schools were matched on a set of
-covariates, and then randomly assigned to the treatment or control
-group. Schools in the control group continued with their current
-algebra curriculum, and schools in the treatment group used Carnegie
-Learning's curriculum which includes CTAI textbook materials and sofware. Each school participated for two years, with a different cohort
-of students participating the second year (with a small fraction of
-students present in the study both years because they repeated
-algebra). It should be noted that this study did not include statewide
-implementations; the study results cannot be generalized to all
-schools within the state. In some states, one large school district
-participated, while in other states, a set of smaller school districts
-participated. The states included Alabama (AL), Connecticut (CT),
-Kentucky (KY), Louisiana (LA), Michigan (MI), New Jersey (NJ), and
-Texas (TX). Each state participated in both the middle school and high
-school arms of the study, except AL, which participated only in the
-middle school arm. The current study focuses on high school students
-only.
-
-There are some limitations to the available data for this study.
-Log data from some schools, and some students within schools, were
-missing either because the log files were not retrievable, or because
-of an imperfect ability to link log data to other study data files.
-For this reason, this study uses only data from the
-18 treatment schools for which at least
-80\%
-of students in both study years appear in the log data file.
-This sample includes 4460 students, around
-75\%
-of the treated high-school sample.
-Table \ref{tab:nByState} gives the number of students in the sample by
-state and year.
-The states in the table are ordered by the total number of students
-they represent in the sample; they will appear in this order in
-all of the forthcoming tables and figures.
-Some figures will only show data from a subset of states; since so few
-students were in New Jersey, it will be excluded from almost all
-state-by-state comparisons (but included analyses that pool across states).
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Fri Dec 15 13:54:41 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrrrrr}
-  \hline
- & TX & KY & MI & LA & CT & NJ \\ 
-  \hline
-Year 1 & 947 & 890 & 325 & 164 & 112 &  17 \\ 
-  Year 2 & 719 & 646 & 285 & 180 & 139 &  36 \\ 
-   \hline
-\end{tabular}
-\caption{Numbers of students in the sample by state and study year} 
-\label{tab:nByState}
-\end{table}
-
-In this sample from 18 schools, 164
-students who participated in the RAND study do not appear in the log data; they may have not used the CTAI
-software at all, or may have been excluded from the log data for other
-reasons.
-Since we don't know which is true, we exclude these students from most
-analyses.
-
-It is likely that some usage data were missing, even for students who
-appear in the usage dataset.
-However, it is impossible to know in which cases these data were
-missing or why; for the most part, we ignore this problem, but it
-should be kept in mind nonetheless.
-
-\section{Standard and Customized Curricula}\label{sec:curricula}
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/curricula-1}
-
-\caption{Percentage of worked problems coming from various courses
-  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
-  standard and customized variants, denoted by shading.}
-\label{fig:curricula}
-\end{figure}
-
-Students' automatic progress through the Cognitive Tutor (CT) software
-is normally governed by
-the sequences of sections and units embedded in the software.
-Without external meddling, the curriculum a student works on
-determines the sequence, and thus what section he or she will be directed towards next after
-mastering (or exhausting the problems) from a previous section.
-In the CTAI effectiveness trial, the most common curriculum was,
-naturally, Algebra I.
-This came with three closely related variants, due to new software releases.
-Students requiring more remediation were able to work on a less
-advanced curriculum, called ``Bridge to Algebra,'' and more advanced
-students could work on Algebra II or Geometry.
-
-In the second year of the study, some high schools, primarily in
-Texas, Michigan, and Kentucky, requested customized variants of the
-curricula.
-This was typically due to state standards, testing schedules, or local
-scope and sequence guidelines.
-These ``customized curricula'' altered the order of some sections and
-units, and were usually particular to schools.
-
-Figure \ref{fig:curricula} shows the percentage of worked problems
-from each curriculum, from standard and customized varieties, by state
-and year.
-First, note that the vast majority of worked problems were from the
-Algebra I sequence.
-A small but notable number of less advanced problems were worked in
-Kentucky in year 2, and some more advanced problems were worked in
-Michigan and Louisiana.
-Secondly, note the rise in ``customized curricula'' in year 2 in
-Texas, Kentucky, and Michigan, the three states with the most students
-in our dataset.
-In particular, Texas shifted almost entirely to customized curricula
-from years 1 to 2.
-
-Throughout the school year, teachers could have a class of students
-working on multiple curricula either sequentially, where the students
-changed curricula in lock step, or simultaneously, where students
-worked on different curricula at the same time. As an example, two
-teachers located in Kentucky had their students working on Algebra I
-throughout most of the year and then reassigned them to Algebra II in
-the last month of school. In contrast, a different teacher in Kentucky had students variously
-enrolled in three different curricula throughout the entire year
-(Bridge-to-Algebra, Algebra I, and a customized Geometry curriculum),
-while a year 2 teacher in Michigan enrolled students in
-three curricula sequentially throughout the year: Algebra I
-until November, followed by a customized curriculum until February, and
-ending with a different customized curriculum until June. While there are numerous instances of these
-uses of multiple curricula in year 2, there are also many occurrences
-of teachers who had their students enrolled in the standard Algebra I
-throughout the entire year, including all Connecticut teachers.
-There were also teachers, mostly in Texas, who used customized
-curricula exclusively throughout the
-second year.
-
-\section{Student Usage Across States and Years}\label{sec:usage}
-
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Nov  6 13:58:09 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrrrr}
-  \hline
- & Hours & Problems & Sections & Units \\ 
-  \hline
-Year 1 & 33.47 & 309 & 43.0 & 9 \\ 
-  Year 2 & 23.83 & 228 & 36.5 & 10 \\ 
-   \hline
-\end{tabular}
-\caption{Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study. Students with no usage data were excluded.} 
-\label{tab:medUsage}
-\end{table}
-
-Table \ref{tab:medUsage} shows the median numbers of hours,
-problems, sections, and units worked on by each student in the dataset
-in the two years of the study.
-Apparently, usage decreased markedly in the second year: the median
-of hours worked decreased by
-10, the median number of problems decreased by
-81, and
-the median number of sections decreased by
-6.5 from
-years 1 to 2.
-Yet, as discussed below, the median number of units worked increased
-by 1.
-
-\begin{figure}
-\centering
-
-\includegraphics[width=\maxwidth]{figure/usageTime-1}
-
-\caption{Boxplots of hours each student spent on Cognitive Tutor
-  software over by year and state. Students  with no timestamp data
-  ($n=$193), with anomalous negative time
-  ($n=$3) or with more than 110 hours ($n=$91)
-  were excluded.}
-\label{fig:timeByStud}
-\end{figure}
-
-Figure \ref{fig:timeByStud} shows that the number of hours students
-spent working on the CT software in some more detail, via
-state-by-year boxplots.
-Analogous figures for the numbers of problems and sections students
-worked,
-showed similar patterns.
-Usage time varied substantially between students and across states and
-years.
-Students in Texas,
-Connecticut, and New Jersey worked far fewer hours than students in
-Kentucky, Louisiana, and Michigan.
-Not every state reduced its usage from years 1 to 2---while students
-in Texas, Kentucky and New Jersey used the software less in the second
-year than in the first, students in Michigan, Louisiana, and
-Connecticut increased their usage.
-
-Overall, usage varied a bit more in year 2 than in year
-1---the median absolute deviation of time spent was 15.9 hours in the first year, compared to 16.9 in  the second year.
-The increase in variation seems to be driven both by increasing
-between-state variation, and a between-student increase in Louisiana.
-One intriguing possibility is that the amount of CT usage may have been better tailored to
-teachers and students in the second year than in the first. Perhaps
-usage increased for
-students who stood to gain more from the software and decreased for
-students who stood to gain less.
-
-In contrast to the decreasing numbers of hours, problems, and sections students
-worked in year 2, Table \ref{tab:medUsage} shows that the median number of units
-students worked increased by
-1
-in year 2.
-This suggests students in year 2 were exposed, on average, to a
-slightly wider range of topics.
-Figure \ref{fig:unitsByStud} shows boxplots of the numbers of units
-worked by state and year.
-The geographic variation in units worked mirrors the pattern in Figure
-\ref{fig:timeByStud}, with more usage in Kentucky, Michigan, and
-Louisiana but less in Texas, Connecticut, and New Jersey.
-However, in every state the median year 2 student worked at least as
-many different units as the median year 1 student.
-Variation in the number of units worked also increased slightly
-from years 1 to 2---the interquartile range (IQR) increased in every state
-except for Kentucky, where a decrease in IQR was accompanied by an
-increase in the number of outliers.
-
-\begin{figure}
-\centering
-
-\includegraphics[width=\maxwidth]{figure/unitsWorked-1}
-
-\caption{Boxplots of the number of units of Cognitive Tutor
-  software each student worked, by year and state. Students
-  working more than 55 units (
-  7 of
-  4460) and students
-  with no usage data (189)
-  were excluded.}
-\label{fig:unitsByStud}
-\end{figure}
-
-All in all, students used CT software less in year 2 than in year
-1.
-On the other hand, students in the second year tended to see a
-slightly wider range of topics, and varied somewhat more in their usage.
-
-\section{Working Units in Order---Or Not}\label{sec:order}
-
-Overall, students used CT less in the second year than in the
-first.
-How was this difference distributed across CTAI units?
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/whichUnits-1}
-
-\caption{The percentages of students with usage data who worked at
-  least one problem from each unit in the Algebra I curriculum. The
-  units are arranged in order for the standard curriculum. In the 2008
-version of the software, the ``Unit Conversions'' unit was broken up
-into two smaller units; for the sake of between-year comparisons, we
-re-combined them.}
-\label{fig:unitsWorked}
-\end{figure}
-
-Figure \ref{fig:unitsWorked} shows the units of algebra along the
-horizontal axis, according to their order in the standard CTAI
-curriculum.
-The vertical axis shows the percentage of students with usage data who
-worked each unit.
-
-In year 1, the curve is almost monotonically decreasing, as one would
-expect if students adhered to the curriculum.
-Students varied in the number of units they worked---with the variation due
-to both student ability and the amount of time allocated to CTAI
-within a classroom---but they mostly followed the standard curriculum.
-Students who worked fewer units stopped earlier in the sequence, and those who worked more units progressed farther.
-Hence, earlier sections were worked by higher proportions of students
-than later units.
-
-In contrast, in year 2 students were much more likely to depart from the
-standard unit order.
-For instance, Figure \ref{fig:unitsWorked} suggests that some students
-skipped ``Unit Conversions'' to work on  ``1st
-Quadrant Linear Graphs'' or skipped ``1 step Linear Equations'' to work on ``Independent Variables in Linear Models''
-In both these cases, the subsequent unit was worked on by a greater
-proportion of students than the immediately prior unit.
-
-Most strikingly, ``Linear Equations with Variables on Both Sides'' was worked by a
-greater proportion of students in year 2 than in year 1, and by a
-greater proportion of students than any of the previous six sections.
-Presumably teachers and administrators wanted students to focus on
-that unit, perhaps because they found it to be particularly effective,
-because students tend to struggle with its main topic, or because its
-topic may figure prominently in an upcoming standardized test.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/unitsWorkedCust-1}
-
-\caption{The percentages of year-2 students with usage data who worked at
-  least one problem from each unit in the Algebra I curriculum. The
-  units are arranged in order for the standard curriculum. Students
-  are divided between those attending schools using primarily a customized
-  curriculum and those using primarily the standard Algebra I curriculum.}
-\label{fig:unitsWorkedCust}
-\end{figure}
-
-Most of the variation in unit order was driven by the rise, in year 2,
-of customized curricula.
-Figure \ref{fig:unitsWorkedCust} divides year-2 students into those
-attending schools using primarily a customized curriculum, and those
-attending schools using primarily a standardized
-curriculum.\footnote{At least
-  90\% of
-  problems worked by students at ``Customized'' schools were from a customized curriculum,
-  and at most
-  1\% of
-  problems at ``Standard'' schools were from a customized curriculum.}
-Students using a standardized curriculum followed the standard
-sequence---more or less---while students using customized curricula
-did not.
-That said, there were some order violations in the standard group:
-specifically, more students worked problems from units ``2-step Linear
-Equations'' and ``Exponents'' than worked the preceding sections;
-this suggests that some teachers used the reassignment tool to
-prioritize particular topics.
-Of course, teacher reassignment may have occurred in schools with customized
-curricula as well---a possibility we will discuss in the next
-section.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/unitsBySchool-1}
-
-\caption{The percentages of year-2 students with usage data in each
-  school who worked at least one problem from each unit in the Algebra I curriculum. The
-  units are arranged in order for the standard curriculum. Schools are
-  classified as either using primarily customized curricula (solid
-  line) or using primarily the standard Algebra I curriculum (dotted).}
-\label{fig:unitsBySchool}
-\end{figure}
-
-Figure \ref{fig:unitsBySchool} further decomposes the year-2 results
-by school and state, showing a large amount of variation between
-states, as well as variation between schools within states.
-In Texas, every school used customized curricula, most of which seem to
-prioritize some of the same units, for instance, ``Linear Patterns,''
-``Independent Variables in Linear Models,'' and
-``Linear Equations with Variables on Both Sides''
-On the other hand, there was also variance between schools.
-For instance, one school prioritized units ``2 Step Linear Equations''
-and ``4-Quadrant Linear Graphs'' while nearly eliminating ``Linear Patterns.''
-
-Between-school variation is evident in the other states, as well.
-In four of the five Kentucky schools, nearly every student worked on
-the first nine units; in the one Kentucky school that used a
-customized curriculum, nearly every student worked on the first 13
-units, omitted the 15th (``Lin. Mod. in General Form''), and worked on
-the 16th and 17th (``Literal Equations'' and ``Linear Equations with Variables on Both Sides'').
-In the remaining school, nearly every student worked on the first
-section, but usage decreased rapidly from there.
-In one Michigan school which used the standard curriculum,
-no students seem to have worked on the ``Linear Models \& Ratios'' section.
-
-If unit order and topic scaffolding are important to CT's mastery
-learning mechanism, the wide variation in students' realized curricula
-would seem to pose a problem.
-The fact that the prescribed order was followed less in the second
-year of the study, when CTAI was effective, than in the first year,
-when it wasn't, suggests that the standard curriculum may play a
-smaller role than one might otherwise imagine.
-
-\section{Mastering the Material---Or Not}\label{sec:mastery}
-The central idea behind mastery learning is that students progress
-through the curriculum as they master skills.
-In the context of CT, skills are clustered within sections, which are
-in turn clustered within units.
-Students progress from the current section to the next section after
-mastering all of the current section's skills.
-Ideally, students would master all of the skills in all of the
-sections they work.
-
-By default, the software operates by automatically moving students
-from section to section based on the sequence of topics defined by the
-curriculum they were currently enrolled in. In this
-software-controlled sequencing, students ideally spend the time
-necessary to learn the material of a section, are judged by the
-software to have mastered the material, and then ``graduate'' to the
-next section.
-However, the software will also ``promote'' a student to the next
-section if the student exhausts a section's material without mastering
-its skills.
-Additionally, teachers are able to modify a student's path within the curriculum.
-They can ``reassign'' students from their current sections to other
-sections earlier or later in the intended sequence, including sections
-they worked on previously.
-Finally, if the semester ends, or a student stops using CT for some
-other reason, while in the middle of working through a
-section, the section is designated ``final.''
-All in all, each CT section a student encounters ends in one of four possible ways:
-mastery, promotion, reassignment, or as the student's final section.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/overallStatus-1}
-
-\caption{The distributions of outcomes of worked sections, by state and across the
-  entire sample, in the two study years.}
-\label{fig:overallStatus}
-\end{figure}
-
-\begin{table}
-  \centering
- \begin{tabular}{rllllll}%|llllll}
-
-&   \multicolumn{6}{c}{Year 1}\\%&\multicolumn{6}{c}{Year 2}\\
-&TX&KY&MI&LA&CT&Overall\\Final&699&625&242&129&91&1796\\Reassigned&600&140&220&5&92&1082\\Promoted&1619&9628&4582&1459&261&17561\\Mastered&15815&56199&31200&8648&1303&113374\\Total&18733&66592&36244&10241&1747&133813\\&\multicolumn{6}{c}{Year 2}\\
-Final&1252&660&236&137&114&2412\\Reassigned&1215&253&275&9&144&2042\\Promoted&1000&6296&4717&1998&450&14477\\Mastered&4929&36676&26509&10108&1513&79874\\Total&8396&43885&31737&12252&2221&98805\\
-
-\end{tabular}
-\caption{Numbers of worked sections that ended in each of the
-  four possible outcomes, across states and study years.}
-\label{tab:overallStatus}
-\end{table}
-
-Figure \ref{fig:overallStatus} and Table \ref{tab:overallStatus} show
-the proportions of worked sections in each state and study year that
-ended with mastery, promotion, or reassignment, or as the
-student's final section.
-In the first year, about
-85\% of
-worked sections are mastered, except in Connecticut.
-Other than in Texas, about
-13--15\%
-of sections end in promotion.
-About 3\% of sections in Texas
-and 5\%
-in Connecticut end in reassignment, which is even rarer in the other states.
-
-With the exception of Texas, sections tended to be completed similarly
-in both years.
-In Texas, however, the percentage of sections ending in reassignment increased by a factor
-of about about
-5,
-to about 14\%.
-The proportion of Texas sections labeled ``Final'' increased as
-well---the expected result of decreasing the overall number of worked
-sections and holding fixed the likelihood of ending
-usage while in the middle of a section.
-
-Across states, sections ended in reassignment at a rate of about
-1\% in year 1 and
-2\% in year 2.
-
-\subsection{Section Mastery and Curriculum}
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/statusCur-1}
-
-\caption{The distributions of outcomes of worked sections, by
-  curriculum, in the two study years. (There were no Bridge-to-Algebra
-sections in customized curricula in our dataset.)}
-\label{fig:statusCur}
-\end{figure}
-
-A well-designed curriculum, can, in theory, play an important role in
-students' attainment of mastery.
-Students who work on appropriate problems that build on their current
-set of skills should be more likely to master new skills than students
-working on problems above their level.
-What role did variations in the CT curriculum play in mastery during the effectiveness trial?
-
-Figure \ref{fig:statusCur} shows the proportions of worked sections
-that were mastered or ended in promotion, reassignment, or finality, in
-standard and customized versions of each CT curriculum.
-Mastery proportions do, indeed, depend on curriculum.
-Specifically, students mastered sections from more advanced curricula
-less frequently.
-Sections from the most basic curriculum, Bridge to Algebra, were
-mastered
-92\%
-of the time;
-those from Algebra I were mastered
-83\%
-of the time, and those from more advanced curricula were mastered
-at a rate of
-75\%.
-This is unsurprising, since more advanced curricula may be expected
-to be more challenging.
-However, it may suggest that some students studying advanced topics
-would fare better in more standard curricula.
-
-Algebra I sections from customized curricula tended to end in
-reassignment more often than sections from the standard Algebra I
-curriculum (3\% vs.
-1.4\%, in year 2).
-This may indicate an overall skepticism towards the Carnegie Learning standards
-among certain schools and teachers, manifested in both adoption of
-alternative curricula and reassignment.%\marginpar{does that make more sense?}
-
-\section{Digging Deeper into Section Reassignment}\label{sec:cp}
-The proportion of worked sections in our dataset ending in
-reassignment was small.
-Nevertheless, since reassignment represents the only mechanism by
-which individual teachers can affect their students' progress through
-the Cognitive Tutor, exploring patterns of reassignment can provide
-insight into how CT was used.
-
-\subsection{How Do Reassignment Patterns Vary?}
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/vcs-1}
-
-\caption{Results from a set of eight multilevel logistic regressions
-  predicting section reassignment. For each year, in the entire sample (``Overall'')
-  and in the three states with the largest numbers of reassignments (Texas,
-  Kentucky, and Michigan), we regressed a binary variable indicating
-  whether a section ended in reassignment on random intercepts for
-  school, class, student, and unit, and in the overall case, for state
-  as well, and recorded their variance. The residual variance was set
-  as the variance of the standard logistic distribution,
-  $\pi/3$. These bar charts give the proportion of the total variance
-  attributable to each random effect.}
-\label{fig:vc}
-\end{figure}
-Teachers alone control reassignment.
-Nevertheless, the factors influencing student reassignment vary at a
-number of levels.
-For instance, state and district standards may prod teachers into
-reassigning students to particular units.
-Some principals may encourage teachers to adhere to the official
-curriculum and avoid reassignment.
-Some students may be more prone to reassignment than others.
-Certain units in the CTAI curriculum may be harder than others,
-causing students to tarry and teachers to reassign.
-Finally, a host of other factors, at these levels and others, may spur reassignment.
-
-To better understand the source of the variation in
-reassignment---what drives some, but not other, sections worked by
-students to end in
-reassignment---we fit a set of multilevel models.
-We fit separate models to data from each the three states with the highest numbers of reassignments, Texas,
-Kentucky, and Michigan, and in the sample as a whole, in each of the
-two study years, yielding a total of eight models.
-Each model was a logistic regression: a binary indicator for section
-reassignment was regressed on a random intercept for unit, as well as
-nested random intercepts for student, classroom, and school.
-Models fit to data from all six states included an additional random
-intercept for state.
-
-Logistic regression can be represented in terms of an underlying
-latent variable $Z^*$: student $i$ working section $sec$ is reassigned
-when $Z_{sec,i}^*>0$.
-The model for $Z^*$ is:
-\begin{equation*}
- Z^*_{sec,i}=\alpha_0+\beta_{u[sec]}+\gamma_i+\delta_{c[i]}+\epsilon_{s[i]}+e_{sec,i}
-\end{equation*}
-Where $\alpha_0$ is an overall intercept,
-and $\beta_{u[sec]}$, $\gamma_i$, $\delta_{c[i]}$, and $\epsilon_{s[i]}$
-are random intercepts for the unit in which $sec$ appears, for student
-$i$, for $i$'s classroom, and for $i$'s school, respectively.
-Again, the model fit to all six states includes an additional random
-intercept for state.
-The random intercepts are modeled as independent and normally
-distributed, each with its own variance.
-The regression error $e_{sec,i}$ is given the standard logistic
-distribution, with ``residual'' variance $\pi/3$.
-It is convenient to represent variance in reassignment probabilities
-in terms of the variance of $Z^*$.
-
-Figure \ref{fig:vc} gives the variance components estimated from these
-logistic regressions: variances of the random intercept terms, as a
-percentage of the total variance of $Z^*$.
-Overall, in both years of the study, the largest determinant of
-reassignment was school, accounting for
-30\% of the variation in year 1,
-and 43\% in year 2.
-After school, state was the most important, accounting for
-20\% and
-21\% in the two years, and unit,
-accounting for 16\% and 14\%.
-Surprisingly, classroom and student-level factors only accounted for
-7\% and
-3\% in year 1, respectively,
-and 6\% and
-1\% in year 2.
-The pattern was similar in Texas---where school accounted for over half
-the variation in reassignment in both years---and in Michigan to a
-lesser extent.\footnote{Percentages in state specific models, in which
-  there is no between-state variance, cannot be
-  directly compared to those from the overall model.}
-In Kentucky, unit played the largest role
-(52\%) in year 1, and classroom played
-the largest role in year 2 (50\%).
-Across states and years, student level factors never accounted for
-more than 5\% of
-the variation in reassignment.
-Other than in Kentucky in year 2, classroom never accounted for more
-than 12\%  of
-the variation.
-
-\textbf{Summary.} Although teachers control reassignment, their decisions appear to be largely
-determined by broader policies, occurring at the state or school level.
-
-\subsection{When are Students Reassigned?}
-
-The timing of reassignments can also provide a window into what drives
-teachers' decisions to reassign students.
-Figure \ref{fig:byMonth} shows the proportion of worked sections in
-each month that end in reassignment.
-In both years, reassignments were much more common in the second half
-of the school year than in the first.
-This may be the result of teachers learning how to use the software as
-the year progresses, or responding the pressure of upcoming
-standardized tests by accelerating students' progress and reassigning
-students to relevant sections.
-
-As we've seen, reassignment was more common in year 2 than in year 1.
-In fact, reassignment increases fairly steadily over the entire length
-of the study.
-Through December of the first year, reassignment was rare. From
-January through May of year 1, between one and two percent of sections
-ended in reassignment.
-Year 2 begin where year 1 left off, with one to two percent of
-sections reassigned.
-Finally, from February through May of the second year, the rate of
-reassignment increased again.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/byMonth-1}
-
-\caption{The proportion of worked sections ending in reassignment, by
-  month and year.}
-\label{fig:byMonth}
-\end{figure}
-
-\begin{table}
-  \centering
-  \begin{tabular}{rll|ll}
-    &\multicolumn{2}{c}{Year 1}&\multicolumn{2}{c}{Year 2}\\
-    &Aug--Dec&Jan--Jun&Aug--Dec&Jan--Jun\\
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Mon Nov  6 14:19:52 2017
- TX &  18 & 582 & 281 & 932 \\ 
-  KY &   9 & 131 &  52 & 201 \\ 
-  MI &  47 & 173 & 245 &  30 \\ 
-  LA &   0 &   5 &   9 &   0 \\ 
-  CT &  14 &  78 &  84 &  60 \\ 
-  NJ &   3 &  22 &  17 & 129 \\ 
-   \hline
-Overall &  91 & 991 & 688 & 1352 \\
-
-\hline
-\end{tabular}
-\caption{The number of reassignments in each state and overall in the
-  first and second halves of the year}
-\label{tab:byMonth}
-\end{table}
-
-Figure \ref{fig:byMonthState} and Table \ref{tab:byMonth} decompose
-these trends by state, Figure \ref{fig:byMonthState} includes just
-the three states with the highest number of reassignments, Texas,
-Kentucky, and Michigan. Table \ref{tab:byMonth} shows data for each of
-the six states and overall.
-For the two study years, Figure \ref{fig:byMonthState} shows the proportion of all of each
-state's reassignments the occurred in each month of the school year.
-(Note that while both Figure \ref{fig:byMonth} and \ref{fig:byMonthState}
-show proportions for each month, the denominators are not the same:
-Figure \ref{fig:byMonth} gives the proportion of each month's worked
-sections that ended in reassignment, and Figure \ref{fig:byMonthState}
-gives the proportion of all of the state's reassignments over the
-course of the year, that occurred in each month.)
-
-Figure \ref{fig:byMonthState} reveals patterns that are not apparent
-in Figure 10.
-For example, the bulk of Kentucky's reassignments took place in May.
-A fifth of Michigan's reassignments in year 1 occurred in October;
-for the rest of the year, Michigan roughly followed the same pattern as Texas.
-In year 2, the vast majority of Michigan's reassignments also occurred near
-the beginning of the year, in September and October.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/byMonthState-1}
-
-\caption{The proportion of a state's reassignments that occurred in
-  each month of the year.}
-\label{fig:byMonthState}
-\end{figure}
-
-\textbf{Summary.} In Texas and overall, reassignments were more common
-in the second half of the school year than in the first half. In
-Kentucky and Michigan, they were clustered in a few specific months.
-
-\subsection{Does Reassignment Depend on Classmates?}
-
-Student individuality and independence might be the most important
-motivating factors behind mastery learning---each student learns at
-his or her own pace, and struggles on a unique set of skills.
-Students are supposed to move through the CT curriculum independently
-of each other.
-However, reassignments give teachers the ability to override this
-feature, and coordinate students' progress.
-Teachers might identify students who are behind their classmates and
-reassign them to later sections.
-They may also move an entire class together to a particular section or
-unit of interest.
-To what extent did these and similar considerations drive reassignment
-in the CTAI study?
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/classmates-1}
-
-\caption{Distributions of classmates' statuses at each reassignment in
-  Texas, Kentucky, and Michigan. Each reassignment that took place in
-  these three states is represented by a bar showing the proportions of
-  the reassigned student's classmates who had exited the section on
-  the same day or earlier via promotion or mastery, who had been
-  reassigned from that section on an earlier date, on the same date,
-  or on a later date, who mastered the section or were promoted on a
-  later date, or who never worked the section at all. The bars are
-  rank ordered according to those proportions, in the order listed.}
-\label{fig:classmates}
-\end{figure}
-
-Figure \ref{fig:classmates} addresses this question by plotting a
-students' classmates' statuses at the time he or she is reassigned.
-For each reassignment in Texas, Kentucky, and Michigan, Figure \ref{fig:classmates}
-plots a vertical bar colored to show the proportions of the reassigned
-students' classmates (represented in the usage data) who  had exited the section on
-the same day or earlier via promotion or mastery, who had been
-reassigned from that section on an earlier date, on the same date,
-or on a later date, who mastered the section or were promoted on a
-later date, or who never worked the section at all.
-The bars are rank ordered according to those same proportions: first by the
-proportion of classmates who were promoted or mastered the section on
-the same day as the reassignment in question, or earlier, next by the
-proportion who were reassigned from the same section on an earlier
-date, next by the proportion who were reassigned from the same section
-on the same date, and finally by the proportion who were reassigned
-from the same section on a later date.
-
-Do teachers reassign students in order to help them catch up with classmates?
-According to Figure \ref{fig:classmates}, that might be part of the
-story, but isn't all of it.
-Across years and states, in about 40\% of
-reassignments, at least 75\% of the rest of the class had exited the
-section through graduation or promotion on the same date or earlier.
-The figure also reveals that the proportions of students who had graduated or been
-promoted from the same section on the same day or earlier was smaller in year 2
-than in year 1, especially in Texas and Michigan.
-
-In Texas, year 2 saw a dramatic increase in the proportions of
-students reassigned from the same section on the same day, suggesting
-that some teachers may have been moving the class together through the
-curriculum.
-In Michigan in year 2, teachers reassigned almost all students who
-worked certain sections.
-That is, in 65\% of the instances in which year-two Michigan
-students were reassigned from a section, at least 75\% of their
-classmates were, at some point, reassigned from the same section, or
-never worked it.
-Across all three states and both years, it is exceedingly rare for
-students to master or be promoted from a section after someone in
-their class has been reassigned from the same section.
-
-\textbf{Summary.} Three different patterns emerge from Figure
-\ref{fig:classmates}: teachers reassigning students who have fallen
-behind their classmates, teachers reassigning an entire class from the
-same section on the same day, and teachers reassigning almost all
-students who begin to work on particular sections.
-Each of these patterns mostly takes place in different states and
-years.
-
-\subsection{Where To?}
-
-Teachers who reassign students may simply move them to the next
-section within the same unit.
-Say that a teacher believes that a particular student had already mastered
-the skills in one of the CTAI sections or is wheel-spinning---working
-problems without learning---or a teacher dislikes one of the sections in a CT unit.
-Still, the teacher wants the student to learn as much as possible from
-the current unit.
-Then moving the student to the next section within the same unit might
-make sense.
-Alternatively, teachers who believe that some of their students are
-not progressing quickly enough may reassign them out of their current
-units entirely, and into the next unit in the sequence---whatever that
-may be.
-Finally, a teacher who wanted his or her students to focus on a
-particular topic might reassign them all to the appropriate unit
-when the time is appropriate.
-
-Which of these patterns is most prevalent?
-More generally, when teachers reassign their students, where in the
-curriculum do they send them?
-
-\begin{figure}
-  \centering
-  \begin{subfigure}{5in}
-
-\includegraphics[width=\maxwidth]{figure/transition1-1}
-
-\end{subfigure}
-\begin{subfigure}{1in}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/trans1legend-1}
-
-\end{subfigure}
-\caption{Reassignment transition plot for study year 1. The units on
-  the left of the plot are the Algebra I units that ended in reassignment at
-  least 30 times in year 1. The units on the right are those that were
-  the destination of at least 30 reassignments. Also included on the
-  left are the top two sending units for each of the units on the
-  right, and also included on the right are all the units on the left,
-  along with those units' top two receiving units. The units are numbered
-  according their order in the standard Algebra I curriculum. There is an arrow from a
-  unit on the left to a unit on the right if a student was assigned
-  from the unit on the left to the one on the right. The thickness of
-  the arrows is proportional to the percentage of all reassignments from
-  the sending unit that ended in that receiving unit. The darkness of
-  the arrows is proportional to the number of reassignments from the
-  sending unit to the receiving unit.}
-\label{fig:trans1}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \begin{subfigure}{5in}
-
-\includegraphics[width=\maxwidth]{figure/transition2-1}
-
-\end{subfigure}
-\begin{subfigure}{1in}
-
-\includegraphics[width=\maxwidth]{figure/trans2legend-1}
-
-\end{subfigure}
-\caption{Reassignment transition plot for study year 2. See caption of
-Figure \ref{fig:trans1} for details.}
-\label{fig:trans2}
-\end{figure}
-
-To address these questions, we focus on the units in the standard
-Algebra I sequence.
-Figures \ref{fig:trans1} and \ref{fig:trans2} give transition plots
-for reassignment in the two study years.
-On the left of each figure, marked ``From,'' are the top ``sending''
-units: the units that students were reassigned \emph{from} at least 30
-times.
-They are numbered according to the order they appear in the standard Algebra I curriculum.
-On the right, under ``To,'' are the top ``receiving'' units: the units
-that students were reassigned \emph{to} at least 30 times.
-For completeness, each of the top two receivers for each sending unit,
-and each of the top two senders for each receiving unit were also included.
-Finally, all of the sending units listed on the left were also included in the
-receiving column.
-All in all, 75\% of the first-year reassignments and
-75\% of the second-year reassignments are
-captured in the figures.
-
-The arrows in the plot represent reassignments.
-An arrow from a sending unit to a receiving unit indicates
-reassignment from the former to the latter.
-The thickness of the arrows represents the proportion of reassignments
-originating in the sending unit whose destination was the receiving unit.
-The darkness of the arrows represents the number of such reassignments.
-For instance, in Figure \ref{fig:trans1}, the arrow from
-``4-Quadrant Linear Graphs'' on the left to ``4-Quadrant Linear
-Graphs'' on the right is fairly thick, since
-44\%
-of the reassignments from ``4-Quadrant Linear Graphs'' end in the
-same unit.
-Yet it is also fairly faint, since it only represents
-12
-reassignments.
-
-Inspecting the figures shows that the most common pattern is for
-students to be reassigned to the next unit in the curriculum---this pattern comprises
-52\% of the reassignments in year 1 and
-24\% of those in
-year 2.
-On the other hand, it is relatively rare for students to be reassigned within the same unit
-(these reassignments account for 15\% and 18\% in the two years).
-It is also rare for students to be reassigned to units earlier in the curriculum (comprising
-2\% and
-8\%).
-
-The transition plots also reveal some interesting cases worth highlighting.
-In year 1 (Figure \ref{fig:trans1}), students reassigned from the first unit, ``Linear Patterns'', were primarily placed two units ahead, in ``1st Quadrant Linear Graphs,'' skipping ``Unit Conversions,'' perhaps suggesting a disinterest in ``Unit Conversions'' on the part of the reassigning teachers.
-Similarly,
-9
-of the students
-(27\%)
-reassigned from the section ``Independent Variables in Linear Models'' were placed 13 units later in ``Systems of Linear Equations''
-and all 62 of the students reassigned from ``Quadratic Models \& Area'' were placed in ``Exponents.''
-This suggests that some teachers may have considered the units ``Systems of Linear Equations'' and ``Exponents'' to contain particularly important material.
-
-Since the total number of reassignments was higher in year 2, the corresponding plot (Figure \ref{fig:trans2}) is larger and more complex.
-It is also more common in year 2 for students to be reassigned to units other than the next unit in the sequences.
-This may be partly due to the proliferation of customized curricula.
-Two units, ``4-Quadrant Linear Graphs'' and ``Lin. Equations with
-Variables on Both Sides,'' were common destinations from a wide
-variety of earlier units, suggesting strong teacher interest in those units.
-The majority (76\%)
-of the students reassigned from ``Probability'' were placed in another section of the same unit, perhaps suggesting problems with some of that unit's sections; in fact, all of the students reassigned within the ``Probability'' unit were reassigned from one its first three sections (of seven).
-Finally, 44
-(81\%) of the reassignments from ``Pythagorean Theorem'' ended in the earlier ``Product Rule for Exponents'' unit.
-
-\textbf{Summary.} The most prevalent pattern was for students to be reassigned
-to the following unit, suggesting that teachers may be mostly
-interested in advancing students who are behind.
-On the other hand, a number of examples of other patterns---students moving within the same unit, or to units out of sequence---appear as well, suggesting that some teachers may be finely manipulating their students' curricula.
-
-\section{Effects of Reassignment}\label{sec:effects}
-The goal of CTAI is to help students learn Algebra, so
-the most important questions about reassignment are about its effect on learning.
-Although the data from this study came from a randomized trial, it was
-CTAI as a whole that was randomized, not individual behaviors within CTAI.
-Specifically, student reassignment was not randomized.
-Therefore, precise estimates of causal effects of reassignment on
-learning require strong untestable assumptions that are unlikely to be
-true.
-As in all observational studies, this includes the assumption that all
-confounding variables---variables that predict both reassignment and
-learning---have been measured well and modeled correctly.
-Further complicating matters, although reassignment itself is a
-well-defined process, in practice it can take many forms, as we have
-endeavored to show.
-There is no reason to expect the effects of reassignment to be
-the same regardless of whether the teacher used it to help lagging
-students catch up, to allocate time to important topics, or for some
-other reason.
-
-All that said, observational estimates of reassignment's average
-causal effects can be valuable, if interpreted cautiously, for
-instance by assessing their sensitivity to unmeasured confounding, as
-we do below.
-In the absence of evidence from randomized trials, observational
-studies can help guide intuition, future research, and even---when
-combined with other relevant information and theory---practice.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/cpYyear-1}
-
-\caption{Students' Gain scores (posttest minus pre-test) versus the number of times the were reassigned over the course of the year (jittered), with a Loess smoother added.}
-\label{fig:cpYyear}
-\end{figure}
-
-%\subsection{Average Effects}
-Figure \ref{fig:cpYyear} shows students' gain scores---the difference between their posttest and pre-test scores---as a function of the number of times they were reassigned.
-(The number of reassignments was jittered---random noise was added on
-the horizontal dimension---to avoid overplotting.)
-Overall, the relationship between the two variables is positive.
-Nevertheless, some non-linearity seems to be present, especially in year 2.
-Further, the distribution of the number of reassignments is right-skewed---again, especially in year 2.
-Care in modeling the number of reassignments, then, is especially important---observations from students reassigned an unusually large number of times can exert undo influence on a regression model and generate misleading results, particularly in the presence of non-monotonic relationships.
-We settled on three different strategies: first, the variable $R^{bin}$ dichotomizes reassignment---$R^{bin}=0$ for students who were never reassigned, and $R^{bin}=1$ for students who were.
-Next, $R^{cat}$ defines a categorical variable taking the values $R^{cat}=0,1,2,3,4+$ for students who were reassigned 0, 1, 2, 3 or four or more times, respectively.
-Finally, $R^{num}$ is the raw number of reassignments, which we include for completeness.
-
-We used linear models to estimate the effect of reassignment, regressing posttest scores on $R^{bin}$, $R^{cat}$, or $R^{num}$ along with fixed effects for classroom, essentially modeling reassignment as randomly assigned within classroom.
-Since this is unlikely to be the case (even approximately), we ran a
-second set of models including student level covariates as well:
-pretest scores,\footnote{These are measured with error, and missing at a relatively high rate (
-15\%). To account for
-this, we used regression calibration based on the 20 ``multiple
-imputations'' used in the original CTAI study,
-\citeN{pane2014effectiveness}.} race, sex, grade, special education
-and gifted status, English as a second language (ESL), and free and reduced-price lunch eligibility.
-
-\begin{table}
-  \centering
-\begin{tabular}{r|c|c|c|}
-&\multicolumn{3}{c}{Parametrization}\\
-&$R^{bin}$&$R^{cat}$&$R^{num}$\\
-&\makecell[c]{Effect of\\ $\ge 1$ Reassignment}&\makecell[l]{Effect of\\  \# Reassignments:}& \makecell[c]{Effect per\\ Reassignment:}\\
-\hline
-\makecell[r]{No\\Covariates}&-0.14 $\pm$ 0.06 &\makecell[l]{1 :   -0.15 $\pm$ 0.07 \\2 :   -0.15 $\pm$ 0.1 \\3 :   -0.07 $\pm$ 0.15 \\4 +:  -0.12 $\pm$ 0.17 }&-0.04 $\pm$ 0.03 \\
-\hline\makecell[r]{Covariate\\ Adjusted}&-0.15 $\pm$ 0.06 &\makecell[l]{1 :  -0.15 $\pm$ 0.07 \\2 :  -0.16 $\pm$ 0.1 \\3 :  -0.08 $\pm$ 0.15 \\4 +:  -0.14 $\pm$ 0.17 }&-0.04 $\pm$ 0.03 \\
-\hline
-\end{tabular}
-\caption{Estimates of effects of reassignment on posttests, with 95\% margins of error.}
-\label{effectResults}
-\end{table}
-
-The results are reported in Table \ref{effectResults}.
-All the effect estimates are negative, indicating that reassigning students may hurt their algebra learning.
-The estimated effects decrease in magnitude for three or four reassignments, but these estimates are very noisy---very few students were reassigned more than two times.
-The magnitudes of the effects are rather large: \citeN{pane2014effectiveness} reported an effect, in year 2, of about 0.2; the estimated effect of at least one reassignment is 73\% of that.
-
-But what of unmeasured confounding?
-For instance, the negative effect may be due to baseline differences
-in ability, beyond what is captured in pretest scores.
-\citeN{hhh} suggest a method of estimating the sensitivity of a regression to an omitted confounder based on bench-marking from observed confounders.
-In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
-Roughly speaking, the idea is to widen the confidence interval from an
-ostensibly causal linear model to account for the possibility of a hypothetical unmeasured confounder that predicts reassignment and posttests to the same extent as one of the observed covariates.
-These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.
-As is typical, the pretest is our most important measured covariate, both in terms of its prediction of reassignment and of posttest scores.
-The sensitivity interval for the effect of being reassigned at least
-once on posttest scores, allowing for the possible omission of a
-hypothetical confounder at most as important as pretest, is
--0.15 $\pm$
-0.14.
-This interval is quite wide, implying that such a covariate could
-explain much of the observed relationship between reassignment and
-posttest scores (or that the relationship may be much stronger).
-On the other hand, the sensitivity interval allowing for the possible
-omission of a less important hypothetical confounder---one that
-predicts posttests as well as ESL status and reassignment as well as
-Hispanic ethnicity, the next best observed
-predictors---is
--0.15 $\pm$
-0.07.
-This interval is substantially tighter.
-All in all, unmeasured confounding may play an important role here,
-but there is good reason to believe that it does not explain all of
-the observed relationship.
-
-The wide variety in the use of reassignment that we have documented
-here might suggest that reassignment's treatment effect varies as
-well.
-Figure \ref{fig:trtHet} shows estimated classroom-specific treatment
-effects of $R^{bin}$, being reassigned at least once.
-The estimates came from a multilevel model in which posttest scores
-were regressed on $R^{bin}$ and student-level covariates, along with
-random effects for classroom and random slopes for $R^{bin}$ varying
-by classroom.
-Unlike fixed effects models, multilevel models ``partially pool'' data
-across classrooms to estimate classroom specific effects more precisely \cite{gelmanHill}.
-This is especially important given the small sample sizes within
-classrooms.
-Figure \ref{fig:trtHet} shows a wide variation in the effect of
-reassignment across classrooms---the estimated standard deviation of
-these effects was
-0.18,
-larger than the average effect itself.
-While the effect was negative in most classrooms, it was positive in some.
-This variation could be due to a number of factors, including
-differences in the composition of classrooms, but supports the
-hypothesis that differences in when and how reassignment is used
-lead to differences in its effect.
-
-\begin{figure}
-  \centering
-
-\includegraphics[width=\maxwidth]{figure/hetroPlot-1}
-
-\caption{The effect of being reassigned at least once, in each
-  classroom, as estimated by a multilevel model. The classrooms are
-  sorted by these estimated effects. The dotted line indicated an
-  effect of zero.}
-\label{fig:trtHet}
-\end{figure}
-
-\section{Summary and Discussion}\label{sec:discussion}
-
-The effectiveness of the Cognitive Tutor software presumably depends
-on how much, and how, it is used.
-This paper exploited available log data from the high-school arm of the CTAI
-effectiveness trial---which yielded an impressive result for the
-software in year 2---to describe variation and patterns in the software's
-usage, paying particular attention to issues of mastery learning.
-We found that:
-\begin{itemize}
- \item The amount the software was used varied widely between states
-   and decreased, overall, from years 1 to 2.
- \item Year 2 saw the proliferation of ``customized'' curricula in
-   three states, altering which units students worked, and in what
-   order.
- \item Year 2 saw frequent departures from the standard CTAI unit
-   sequence, driven mainly, but not entirely, by customized
-   curricula.
- \item Examination of section mastery found that:
- \begin{itemize}
-  \item About 85\% of
-    worked sections in year 1 and
-    81\% in  year 2 ended
-    with the student having mastered all of the included skills.
-  \item There were three ways students worked a section without
-    mastering its contents: exhausting its problems and being
-    promoted, being reassigned to a new section by the teacher, and
-    ending CT use altogether.
-  \item Reassignment was rare, though it was more prevalent in Texas
-    and Connecticut than in other states, and more common in year 2
-    than 1.
-  \item Mastery rates were lower for more advanced curricula.
- \end{itemize}
- \item Examination of reassignment patterns found that:
-  \begin{itemize}
-   \item Reassignment rates were determined more by factors varying by state and school
-     than by classroom.
-   \item Reassignment was more common in the second half of the year
-     than in the first---particularly in Texas.
-   \item Depending on state and year, reassignment typically takes
-     place in one of three scenarios: teachers reassigning students
-     who have fallen behind, teachers reassigning (almost) the entire
-     class together, and teachers assigning (almost) all students
-     \emph{out} of a particular section.
-   \item Students are most commonly reassigned to the next unit in the
-     curriculum---suggesting that teachers may be advancing lagging
-     students---but in some cases they might be finely manipulating
-     their students' curricula.
-   \end{itemize}
- \item Reassignment appears to lower students' performance on the
-   post-test relative to that of their peers; however, the effect
-   appears to vary widely between classrooms (and, presumably, how
-   reassignment is used).
- \end{itemize}
-
-In practice, mastery learning and topic scaffolding often unfolded
-quite differently from the vision of CT's designers.
-In some cases, the departures were apparently a matter of a student's inability
-to achieve mastery quickly enough, and in other cases of educators' preferences that ran counter to CTAI's design.
-
-Notably, both the amount of usage and fidelity to CTAI's design
-decreased from years 1 to 2---just as the estimated \emph{effect} of
-CTAI increased.
-In year 2, when the effect was substantial, students spent less time,
-and followed the CTAI curriculum and guidelines less closely than in
-year 1, when the estimated effect was negative but statistically
-insignificant.
-This raises questions as to the roles of structure and mastery in
-CTAI's effectiveness.
-Does flexibility lead to higher effects? Is mastery learning an
-important mechanism for CTAI?
-Answers to these questions could prove crucial for optimizing the
-realized effectiveness of CTAI and other intelligent tutors.
-
-On the other hand, reassignment appears to hurt students' performance
-on the posttest, relative to their classmates (though based on data
-from a randomized experiment, this analysis was observational, so
-confounding can't be ruled out).
-Does this, in contrast, suggest that achieving mastery is an important
-component of effective intelligent tutoring?
-
-As educational technology spreads, attention to the details of
-implementation may yield important insights---or important
-questions---about effectiveness, and the science of when
-intelligent tutors work, and when they don't.
-
-\bibliographystyle{acmtrans}
-\bibliography{ct,cp}
-
-\end{document}

### 2018-01-02 paper.Rnw
finish jedm format

@@
 
-%\usepackage{endfloat}
-%\usepackage{soul}
-
-%\usepackage{type1ec}
-\usepackage{subcaption}
-\usepackage{fullpage}
-\usepackage{enumerate}
-%\usepackage{graphicx}
-%\usepackage{graphics}
-\usepackage{multirow}
-\usepackage{comment}
-\usepackage{amsmath,amssymb,amsfonts,amsthm}
-\usepackage{bbm}
-\usepackage{setspace}
-\usepackage{verbatim}
-%\usepackage{natbib}
-\usepackage{bm}
-\usepackage{pdflscape}
-%\usepackage[dvipsnames]{xcolor}
-\usepackage{tikz}
-\usepackage{xr}
-%\usepackage{mathabx}
-%\usepackage{filecontents}
-%\usepackage{bibentry}
-%\usepackage{hanging}
-%\usepackage{apacite}
-%\usepackage{hyperref}
-\usepackage{makecell}
-
-%\usepackage{endnotes}
-
-%\let\footnote=\endnote
-
-\title{Mastery Learning in Practice:\\
-A (Mostly) Descriptive Analysis of Log Data from the Cognitive Tutor Algebra I Effectiveness Trial}
-
-\author{{\large Anita Israni}\\University of Texas College of
-  Education\\aisrani@gmail.com \and {\large Adam C Sales}\\University of Texas College of
-  Education\\asales@utexas.edu  \and {\large John F Pane}\\RAND Corporation\\jpane@rand.org}
-
-\date{}
+\input{preamble.tex}
 
@@
 
-\begin{abstract}
-Mastery learning, the notion that students learn best if they move on
-from studying a topic only after having demonstrated mastery, sits at
-the foundation of the theory of intelligent tutoring. This paper is an
-exploration of how mastery learning plays out in practice, based on
-log data from a large randomized effectiveness trial of the Cognitive
-Tutor Algebra I (CTAI) curriculum. We find that students frequently
-progressed from CTAI sections they were working on without
-demonstrating mastery and worked units out of order. Moreover, these
-behaviors were substantially more common in the second year of the
-study, in which the CTAI effect was significantly larger. We explore
-the various ways students departed from the official CTAI curriculum,
-focusing on heterogeneity between years, states, schools, and
-students. The paper concludes with an observational study of the
-effect on post-test scores of teachers reassigning students out of
-their current sections before they mastered the requisite skills,
-finding that reassignment appears to lowers posttest scores---a finding
-that is fairly reseliant to confouding from omitted
-covariates---but that the effect varies substantially between
-classrooms.
-\end{abstract}
-\section{Introduction}
-Mastery learning sits at the foundation of intelligent tutoring
-systems \cite[e.g.][]{corbett2001cognitive,wenger2014artificial}.
-The philosophy of mastery learning assumes a well-structured
-curriculum, and posits that students progress within the curriculum as
-they master its skills
-\cite{bloom1968learning,kulik1990effectiveness}. The Cognitive Tutor
-Algebra I (CTAI) system, developed by Carnegie Learning, Inc., is one of the best studied and best regarded examples of modern educational software.
-It is a blended learning system for teaching algebraic concepts and principles, to middle and high school students, including both textbook materials and software.
-The software component of the curriculum allows students to progress at their own pace and receive individualized feedback on their performance.
-A large-scale randomized effectiveness trial conducted by the RAND corporation showed that, in some circumstances, CTAI boosts students' scores on an Algebra-I posttest by about one fifth of a standard deviation \cite{pane2014effectiveness}.
-CTAI's success in this experiment would seem to validate its pedagogy: mastery learning, and the algebra I curriculum on which it is based.
-
-However, the theory underlying CTAI does not always determine its
-use. To be sure, the software has a standard set of algebra topics,
-divided into units and further into sections; and a standard sequence
-for presenting them. But this precise curriculum is not mandatory. At
-the request of educators, it can be customized by altering what units
-or sections are included (including, possibly, material from a
-different standard curriculum such as geometry), as well as their
-sequence, to conform to local or state standards or scope and sequence
-guides. Further, teachers have the option of moving students within
-the curriculum, regardless of the software's estimate of their skill
-mastery.
-
-This article examines teachers' and schools' adherence and
-non-adherence to the standard, mastery-based CTAI curriculum, using
-data from the RAND study, a seven-state randomized controlled trial of
-CTAI in high schools and middle schools. That study found a significant
-positive effect of CTAI in high schools, during their second year of
-implementation but not the first. Students in the treatment group of
-the study were enrolled in one or more of the standard or customized
-curricula during their participation in the study, and the software
-logged aspects of their usage, including time spent, sections
-encountered, and whether the software judged the students to have
-mastered the sections. Adopting standard procedures of an
-effectiveness trial, both Carnegie Learning and the researchers
-running the study restricted their support and oversight to what is
-typically provided outside of an experimental context. Thus, the
-software data from the study reflect typical usage.
-Secondary data analyses used principal stratification to show that students who attempted more sections experienced larger treatment effects, and students who had high or low assistance levels, as opposed to an average level, experienced smaller treatment effects \cite{sales2015exploring,sales2016student}.
-\citeN{sales2017role} found that students more likely to master worked
-sections of the CTAI software may experience \emph{smaller} effects
-than those less likely to achieve mastery, casting doubt on the role
-of mastery learning as a mechanism for the treatment effect.
-
-Here we contextualize the previous findings to describe, in detail, the ways
-in which schools, teachers, and students violate mastery learning.
-We will begin with short discussions of the RAND effectiveness trial and the usage
-data it produced.
-Sections \ref{sec:curricula}---\ref{sec:order} will describe overall
-patterns of usage.
-First, we will discuss standard and customized CTAI curricula used in
-the study (Section \ref{sec:curricula}).
-Next, in Section \ref{sec:usage}, we will describe patterns in the amount CTAI usage, showing that
-it varied widely between states, between years, between schools, and between students.
-In particular, the amount of usage decreased from years 1 to
-2---though more in some states than others.
-Then in section \ref{sec:order} we will describe how the amount of usage
-changed---which units of the CTAI curriculum were worked more and less
-from years 1 to 2---and find that order in which students worked
-CTAI's units varied across years. We show that this change is due
-mostly, but not completely, to the presence of customized curricula in
-year 2.
-
-In Section \ref{sec:mastery} we will describe patterns of mastery in general,
-and in Section \ref{sec:cp} we will delve deeply into ``reassignment'': the process in which a
-teacher moves a student out of a section he or she has not (yet)
-mastered into a new section.
-In particular, we will attempt to elucidate teachers' goals in
-reassigning students. One hypothesis is the need for teachers to push
-ahead students who were falling behind, i.e. to reassign them from
-sections on which they were struggling, to allow them catch up with
-the rest of the class. Another hypothesis is that teachers sought to
-push students past easier sections to begin working on more relevant
-or challenging topics for them.
-A third hypothesis is that teachers needed to cover certain topics in preparation for an upcoming state exam, and might have reassigned groups of students all at the same time to cover topics that might otherwise not have been covered.
-This hypothesis may lead to an increase in reassignments as the exam approaches.
-We find evidence of all three motivations for reassignment, varying
-across classrooms, schools, states, and study  years.
-
-Finally, in Section \ref{sec:effects} will will give
-quasi-experimental estimates of the effects of reassignment on
-students' posttest scores.
-We find that reassignment probably decreases student learning,
-though the effects vary widely across classrooms.
-Section \ref{sec:discussion} will conclude the article with a summary
-of findings and discussion.
+\input{intro.tex}
 

### 2018-01-02 paper.tex
finish jedm format

@@
 state-by-state comparisons (but included analyses that pool across states).
-% latex table generated in R 3.3.1 by xtable 1.8-2 package
-% Fri Dec 15 13:54:41 2017
+% latex table generated in R 3.4.2 by xtable 1.8-2 package
+% Tue Dec 05 09:39:47 2017
 \begin{table}[ht]

### 2018-01-02 preamble.tex
finish jedm format

\usepackage{xcolor}
\usepackage{url}

%\usepackage{endfloat}
%\usepackage{soul}

%% \usepackage{type1ec}
 \usepackage{subcaption}
% \usepackage{fullpage}
% \usepackage{enumerate}
% %\usepackage{graphicx}
% %\usepackage{graphics}
% \usepackage{multirow}
% \usepackage{comment}
% \usepackage{amsmath,amssymb,amsfonts,amsthm}
% \usepackage{bbm}
% \usepackage{setspace}
% \usepackage{verbatim}
% %\usepackage{natbib}
% \usepackage{bm}
% \usepackage{pdflscape}
% %\usepackage[dvipsnames]{xcolor}
% \usepackage{tikz}
% \usepackage{xr}
% %\usepackage{mathabx}
% %\usepackage{filecontents}
% %\usepackage{bibentry}
% %\usepackage{hanging}
% %\usepackage{apacite}
%\usepackage{hyperref}
\usepackage{makecell}
\usepackage{amsmath}
%\usepackage{endnotes}

%\let\footnote=\endnote

\title{Mastery Learning in Practice:\\
A (Mostly) Descriptive Analysis of Log Data from the Cognitive Tutor Algebra I Effectiveness Trial}

\author{{\large Anita Israni}\\University of Texas College of
  Education\\aisrani@gmail.com \and {\large Adam C Sales}\\University of Texas College of
  Education\\asales@utexas.edu  \and {\large John F Pane}\\RAND Corporation\\jpane@rand.org}

\date{}
